### 2017-04-03 paperIEEE.tex
Initial Structure
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@2 db7c848b-7755-4df2-b763-8796797afa52

%\documentclass{vgtc}
\documentclass[journal]{vgtc}                % final (journal style)
%\documentclass[review]{vgtc}                 % review
%\documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint]{vgtc}               % preprint
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
%\documentclass[electronic]{vgtc}             % electronic version
%\documentclass[electronic,journal]{vgtc}     % electronic version, journal

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% These three lines bring in essential packages: ``mathptmx'' for Type 1
%% typefaces, ``graphicx'' for inclusion of EPS figures. and ``times''
%% for proper handling of the times font family.
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
%\usepackage{tabularx}
\usepackage{amsmath}
%\usepackage{url}
\usepackage{subfigure}
\usepackage{multirow}
%\usepackage{teaser}

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{0}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}

%% allow for this line if you want the electronic option to work properly
\vgtcinsertpkg

%% In preprint mode you may define your own headline.
%\preprinttext{To appear in an IEEE VGTC sponsored conference.}

\usepackage{color}
\definecolor{RED}{rgb}{1,0,0}
\definecolor{BLUE}{rgb}{0,0,1}
\newcommand{\FIXME}[1]{\textbf{\color{BLUE}{FIXME: #1}}}
%\newcommand{\FIXME}[1]{}

% suppress  single floating lines on top (widow) and bottom (club)
%  10000 is infinity
%  tradeoff: possible underful vboxes
\clubpenalty=10000
\widowpenalty=10000

%% Paper title.
%----------------------------------------------------------------------
\title{Equalizer: A Mature Parallel Rendering Framework}
%----------------------------------------------------------------------

%% Author and Affiliation (multiple authors with multiple affiliations)
\author{Stefan Eilemann\thanks{email: eilemann@gmail.com} \\ %
\and Renato Pajarola\thanks{email: pajarola@acm.org}}

\affiliation{\scriptsize Visualization and MultiMedia Lab \\ Department of Informatics \\ University of Z\"urich}

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Parallel Rendering, Scalable Visualization, Cluster Graphics, Immersive Environments, Display Walls}

%% ACM Computing Review (CR) categories.
%% See <http://www.acm.org/class/1998/> for details.
%% The ``\CRcat'' command takes four arguments.

\CRcatlist{ % not used in journal version
  \CRcat{I.3.2}{Graphics Systems}{Distributed Graphics}{Parallel Rendering};
  \CRcat{I.3.m}{Miscellaneous}{Rendering Clusters}{Scalable Rendering};
  \CRcat{I.3.7}{Three-Dimensional Graphics and Realism}{Virtual Reality}{Immersive Environments}
}

\vgtccategory{Research}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

%%%%%% START OF THE PAPER %%%%%%
\begin{document}

%----------------------------------------------------------------------
%% Abstract section.
\abstract {
  We present the features, algorithms and system integration necessary to
  implement a parallel rendering framework usable in a wide range of real-world
  applications, based on the basic architecture and implementation of the Equalizer
  parallel rendering framework presented in \cite{EP:07}.
} % end of abstract

\maketitle

\begin{figure*}[ht]\center
  \includegraphics[width=2\columnwidth]{images/teaser} \\
  (a) \hfil \hfil (b) \hfil \hfil (c)
  \vspace{-2mm}
  \caption{Various Equalizer use cases: (a) immersive CAVE, (b) display wall and
    (c) scalable volume rendering.}
  \label{FIG_teaser}
\end{figure*}

%----------------------------------------------------------------------
\section{Introduction}
%----------------------------------------------------------------------

The continuing improvements in hardware integration lead to ever faster CPUs and GPUs, as well as higher resolution sensor and display devices. Moreover, increased hardware parallelism is applied in form of multi-core CPU workstations, massive parallel super computers, or cluster systems. Hand in hand goes the rapid growth in complexity of data sets from numerical simulations, high-resolution 3D scanning systems or bio-medical imaging, which causes interactive exploration and visualization of such large data sets to become a serious challenge. It is thus crucial for a visualization solution to take advantage of hardware accelerated scalable parallel rendering. In this systems paper we describe a new scalable parallel rendering framework called {\em Equalizer} that is aimed primarily at cluster-parallel rendering, but works as well in a shared-memory system. Cluster systems are the main focus  because workstation graphics hardware is developing faster than high-end graphics (super-) computers can absorb new developments, and also because clusters offer a better cost-performance balance.

Previous parallel rendering approaches typically failed in one of the following system requirements:
%
\begin{enumerate}\renewcommand{\labelenumi}{\alph{enumi})}
\addtolength{\itemsep}{-0.5\baselineskip}
\item generic application support, instead of special domain solution
\item scalable abstraction of the graphics layer
\item exploit existing code infrastructure, such as proprietary scene graphs, molecular data structures, level-of-detail and geometry databases
\end{enumerate}

To date, generic and scalable parallel rendering frameworks  that can be adopted to a wide range of scientific visualization domains are not yet readily available. Furthermore, flexible
configurability to arbitrary cluster and display-wall configurations has also not been addressed in the past, but is of immense practical importance to scientists depending high-performance interactive visualization as a scientific tool. In this paper we present Equalizer, which is a novel flexible framework for parallel rendering that supports scalable performance, configuration flexibility, is {\em minimally invasive} with respect to adapting existing visualization applications, and is applicable to virtually any scientific visualization application domain.

The main contributions that Equalizer introduces in a single parallel rendering system, and which are presented in this paper are:
%
\begin{enumerate}\renewcommand{\labelenumi}{\roman{enumi})}
\addtolength{\itemsep}{-0.5\baselineskip}
\item novel concept of compound trees for flexible configuration of graphics system resources,
\item easy specification of parallel task decomposition and image compositing choice through compound tree layouts,
\item automatic decomposition and distributed execution of rendering tasks according to compound tree,
\item support for parallel surface as well as transparent (volume) rendering through $z$-visibility as well as $\alpha$-blending compositing,
\item fully decentralized architecture providing network swap barrier (synchronization) and distributed objects functionality,
\item support for low-latency distributed frame synchronization and image compositing,
\item minimally invasive programming model.
\end{enumerate}

Equalizer is open source, available under the LGPL license from http://www.equalizergraphcis.com/, which allows it to be used both for open source and commercial applications. It is source-code portable, and has been tested on Linux, Microsoft Windows, and Mac OS X in 32 and 64 bit mode using both little endian and big endian processors.

%----------------------------------------------------------------------
\section{Related Work}
%----------------------------------------------------------------------
\label{SEC_related}

The early fundamental concepts of parallel rendering have been laid down in \cite{MCEF:94} and \cite{Crockett:97}. A number of domain specific parallel rendering algorithms and special-purpose hardware solutions have been proposed in the past, however, only few generic parallel rendering frameworks have been developed.

\subsubsection*{Domain specific solutions}

Cluster-based parallel rendering has been commercialized for off-line rendering (i.e. distributed ray-tracing) for computer generated animated movies or special effects, since the ray-tracing technique is inherently amenable to parallelization for off-line processing. Other special-purpose solutions exist for parallel rendering in specific application domains such as volume rendering \cite{LWMT:97,Wittenbrink:98,HSCSM:00,SL:02,GS:02,NSJLYZ:05} or geo-visualization \cite{VR:91,AG:95,LDC:96,JLMV:06}. However, such specific solutions are typically not applicable as a generic parallel rendering paradigm and do not translate to arbitrary scientific visualization and distributed graphics problems.

Recently in \cite{NC:07}, parallel rendering of hierarchical level-of-detail (LOD) data has been addressed and a solution specific to sort-first tile-based parallel rendering has been presented. While the presented approach is not a generic parallel rendering system, basic concepts presented in \cite{NC:07} such as load management and adaptive LOD data traversal can be carried over to other sort-first parallel rendering solutions.

\subsubsection*{Special-purpose architectures}

Traditionally,  high-performance real-time rendering systems have relied on an integrated proprietary system architecture, such as the SGI graphics super computers. These special-purpose solutions have become a niche product as their graphics performance does not keep up with off-the-shelf workstation graphics hardware and scalability of clusters. However, cluster systems need more sophisticated parallel graphics rendering libraries, such as the one proposed in this paper.

Due to its conceptual simplicity, a number of special-purpose image compositing hardware solutions for sort-last parallel rendering have been developed. The proposed hardware architectures include Sepia \cite {MHS:99a,sepia}, Sepia~2 \cite{LMSBHa:01,LMSBH:01}, Lightning~2 \cite{Stoll01}, Metabuffer \cite{Blanke00,Zhang01}, MPC Compositor \cite{Muraki01} and PixelFlow \cite{Molnar92, Eyles97}, of which only a few have reached the commercial product stage (i.e. Sepia~2 and MPC Compositor). However, the inherent inflexibility and setup overhead have limited their distribution and application support. Moreover, with the recent advances in the speed of CPU-GPU interfaces, such as PCI Express and other modern interconnects, combinations of software and GPU-based solutions offer more flexibility at comparable performance.

% FlowVR \cite{AGLLMRR:04}, not really related that much here

\subsubsection*{Generic approaches}

A number of algorithms and systems for parallel rendering have been developed in the past. On one hand, some general concepts applicable to cluster parallel rendering have been presented in \cite{Mueller:95,Mueller:97} (sort-first architecture), \cite{SZFLS:99,SFLS:00} (load balancing), \cite{SFL:01} (data replication), or \cite{CMF:05,CM:06} (scalability). On the other hand, specific algorithms have been developed for cluster based rendering and compositing such as \cite{AP:98}, \cite{CKS:02} and \cite{YYC:01,SMLAP:03}. However, these approaches do not constitute APIs and libraries that can readily be integrated into existing visualization applications, although the issue of the design of a parallel graphics interface has been addressed in \cite{Igehy98}.
%
Only few generic APIs and (cluster-) parallel rendering systems exist which include VR Juggler \cite{BJHMBC:01} (and its derivatives), Chromium \cite{HHNFAKK:02} (an evolution of \cite{Humphreys99,Humphreys00,HEBSEH:01}) and OpenGL Multipipe SDK \cite{JDBJBCER:04,BRE:05,MPK}.

VR Juggler \cite{BJHMBC:01,JBBC:98} is a graphics framework for virtual
reality applications which shields the application developer from the
underlying hardware architecture, devices and operating system. Its main
aim is to make virtual reality configurations easy to set up and use
without the need to know details about the devices and hardware
configuration, but not specifically to provide scalable parallel
rendering. Extensions of VR Juggler, such as for example ClusterJuggler
\cite{BC:03} and NetJuggler \cite{AGLMR:02}, are typically based on the
replication of application and data on each cluster node and basically
take care of synchronization issues, but fail to provide a flexible and
powerful configuration mechanism that efficiently supports scalable
rendering as also noted in \cite{SWNH:03}. The presented system is different from VR Juggler in that it
fully supports scalable parallel rendering such as sort-first and
sort-last task decomposition and image compositing, it provides more
flexible node configurations which for example allow specifying
arbitrary task decomposition and image compositing combinations as
simple compound layouts. Furthermore, it it is fully distributed which includes
support for network swap barriers (synchronization), distributed objects
as well as image compression and transmission. In contrast to VR
Juggler, Equalizer supports multiple rendering threads per process,
which is important for multi-GPU systems.

While Chromium \cite{HHNFAKK:02} provides a powerful and transparent
abstraction of the OpenGL API, that allows a flexible configuration of
display resources, its main limitation with respect to scalable
rendering is that it is focused on streaming OpenGL commands through a
network of nodes, often initiated from a single source. This has also been
observed in \cite{SWNH:03}. The problem
comes in when the OpenGL stream is large in size, due to not only
containing OpenGL calls but also the rendered data such as geometry
and image data. Only if the geometry and textures are mostly static
and can be kept in GPU memory on the graphics card, no significant
bottleneck can be expected as then the OpenGL stream is composed of a
relatively small number of rendering instructions. However, as it is
typical in real-world visualization applications, display and object
settings are interactively manipulated, data and parameters may change
dynamically, and large data sets do not fit statically in GPU memory but
are often dynamically loaded from out-of-core and/or multiresolution
data structures. This can lead to frequent updates not only of commands and
parameters wich have to be distributed but also of
the rendered data itself (geometry and texture), thus causing the OpenGL stream
to expand dramatically. Furthermore, this stream of function calls and
data must be packaged and broadcast in real-time over the network to
multiple nodes for each rendered frame. This makes CPU performance and
network bandwidth a more likely limiting factor. While preserving a minimally
invasive API, the novel proposed system is better aimed at scalability
as the actual data access is decentralized in the distributed rendering
clients.

The performance experiments in \cite{HHNFAKK:02} indicate that Chromium is working
quite well when the rendering problem is fill-rate limited.
This is due to the fact that the OpenGL commands and a non-critical
amount of rendering data can be distributed to multiple nodes without significant problems
and since the critical fill-rate work is then performed locally on the graphics hardware.

Chromium also provides some facilities for parallel application
development, namely a sort-last, binary-swap compositing SPU and an
OpenGL extension providing synchronization primitives, such as a barrier
and semaphore. It leaves other problems, such as configuration, task
decomposition as well as process and thread management unaddressed, thus
making the development of parallel OpenGL applications harder than with
Equalizer. Parallel Chromium applications tend to be written for one
specific parallel rendering use case, such as for example the sort-first distributed
memory volume renderer \cite{BHPB:03} or the sort-last parallel
volume renderer raptor \cite{Raptor}. We are not aware of a generic Chromium-based
application using many-to-one sort-first or stereo decompositions. This is
another difference to Equalizer which provides a much more flexible
task decomposition configuration. Applications written once for Equalizer can
easily be run in any different task decomposition mode and for any physical
display configuration without any changes to the application itself.
While Equalizer provides an abstraction of all entities of the rendering pipeline
(see Sections~\ref{SEC_architecture} and~\ref{SEC_development}), Chromium's
infrastructure is primarily the compositing stage.

OpenGL Multipipe SDK (MPK) \cite{BRE:05} implements an effective
parallel rendering API for a shared memory multi-CPU/GPU system. It is
similar to IRIS Performer \cite{RH:94} in that it handles multi-pipe
rendering by a lean abstraction layer via a conceptual callback
mechanism, and that it runs different application tasks in
parallel. However, MPK is not designed nor meant for rendering nodes
separated by a network. MPK focuses on providing a parallel rendering
framework for a single application, parts of which are run in parallel
on multiple rendering channels, such as the culling, rendering and final
image compositing processes. Compared to MPK, Equalizer supports a
fully distributed parallel rendering paradigm and features a more flexible
task decomposition approach.

%------------------------------------------------------------------------------
\section{Usability}
%------------------------------------------------------------------------------

\subsection{Physical and Logical Visualization Setup}

Canvas, Layout, Observer

multi-view with runtime switching,

2D overlays

Swap Barriers

\subsection{Auto-Configuration}

hwsd with VirtualGL support

\subsection{Sequel}

application, renderer, view data

\subsection{Qt Windowing}

Challenges in threading model, architecture

\subsection{Tide Integration}

Tiled interactive display environment, parallel pixel streaming, events

%------------------------------------------------------------------------------
\section{The Collage Network Library}
%------------------------------------------------------------------------------

\subsection{Distributed, Versioned Objects}

types (instance, delta, static), versioning, multicast, compression,
serializable with dirty bits, mapping, blocking commits

\subsection{Reliable Stream Protocol}

UDP-based reliability protocol

\subsection{Infiniband RDMA}

reverse-engr impl

%------------------------------------------------------------------------------
\section{Virtual Reality}
%------------------------------------------------------------------------------

\subsection{Dynamic Focus Distance}

Focus what user is looking at

\subsection{Asymmetric Eye Position}

Better HMD by measuring user geometry

\subsection{Application-specific Scaling}

Gullivers world

\subsection{Runtime Stereo Switch}

\subsection{Swap Synchronization and GPU affinity}

%------------------------------------------------------------------------------
\section{Performance}
%------------------------------------------------------------------------------

\subsection{New Decomposition Modes}

The initial version of Equalizer implemented sort-first (2D), sort-last (DB) and
stereo (EYE) decomposition. In the following we present new decomposition modes
and motivate their use case.

\subsubsection{Time-Multiplex}

Interaction with threading

\subsubsection{Tiles and Chunks}

Equalizers do what?

\subsubsection{Pixel}

Equal load for fill-limited apps

\subsubsection{Subpixel}

MSAA and DOF, plus idle refinement algo

\subsection{Equalizers}

Equalizer are one addition to compound trees. They modify parameters of their
respective subtree at runtime to optimize one aspect of the decomposition.

\subsubsection{Sort-First and Sort-Last Load Equalizer}

reactive load equalization, describe tunables (damping, resistance, delta,
granularity), sort-first respects input channel size

\subsubsection{Cross-Segment Load Equalizer}

summarize \cite{EEP:11}

\subsubsection{Dynamic Frame Resolution}

Constant frame rate for fill-limited applications

\subsubsection{Frame Rate Equalizer}

Predominantly used by DPlex to smooth output framerate

\subsubsection{Monitoring}

Control workstation in VR setups

\subsection{Optimizations}

\subsubsection{Region of Interest}

for compositing and load equalizer.

\subsubsection{Asynchronous Readback}

pipelines rendering with compositing

\subsubsection{Download and Compression Plugins}

GPU-CPU transfer plugins with optional compression (eg YUV) linked to CPU
compression for network transfer

\subsubsection{Thread Synchronization Modes}

Per-node sync, draw sync, async

%----------------------------------------------------------------------
\section{Applications}
%----------------------------------------------------------------------

\subsection{Livre}
\subsection{RTT Deltagen}
\subsection{RTNeuron}

%----------------------------------------------------------------------
\section{Experimental Results}
%----------------------------------------------------------------------
\label{SEC_results}

%----------------------------------------------------------------------
\section{Discussion and Conclusion}
%----------------------------------------------------------------------
\label{SEC_conclusions}

%----------------------------------------------------------------------
%\section*{Acknowledgements}
%----------------------------------------------------------------------
%% if specified like this the section will be ommitted in review mode
\acknowledgements{
We would like to thank and acknowledge the following institutions and projects for providing the 3D geometry and volume test data sets:
the Digital Michelangelo Project, Stanford 3D Scanning Repository, Cyberware Inc., volvis.org and the Visual Human Project.
This work was partially supported by the Swiss National Science Foundation Grant 200021-116329/1.
}

\vspace{-2mm}
\footnotesize
\bibliographystyle{abbrv}
\bibliography{references}

\end{document}

### 2017-04-03 paperIEEE.tex
Filling in
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@3 db7c848b-7755-4df2-b763-8796797afa52

@@
 \usepackage{multirow}
-%\usepackage{teaser}
 
@@
 %% Abstract section.
-\abstract {
-  We present the features, algorithms and system integration necessary to
-  implement a parallel rendering framework usable in a wide range of real-world
-  applications, based on the basic architecture and implementation of the Equalizer
-  parallel rendering framework presented in \cite{EP:07}.
-} % end of abstract
+\abstract{ We present the features, algorithms and system integration necessary
+  to implement a parallel rendering framework usable in a wide range of
+  real-world research and industry applications, based on the basic architecture
+  and implementation of the Equalizer parallel rendering framework presented in
+  \cite{EMP:09}.  } % end of abstract
 
@@
   \vspace{-2mm}
-  \caption{Various Equalizer use cases: (a) immersive CAVE, (b) display wall and
-    (c) scalable volume rendering.}
+  \caption{Example Equalizer application use cases: (a) 192 Megapixel CAVE at
+    KAUST running RTNeuron, (b) Immersive HMD with external tracked and
+    untracked views running RTT DeltaGen for virtual car usability studies (c)
+    Cave2 running a molecular visualization build using Omegalib.}
   \label{FIG_teaser}
@@
 
-The early fundamental concepts of parallel rendering have been laid down in \cite{MCEF:94} and \cite{Crockett:97}.
-A number of domain specific parallel rendering algorithms and special-purpose hardware solutions have been proposed in the past, however, only few generic parallel rendering frameworks have been developed.
+\cite{EMP:09} summarized the work in parallel rendering at the time of
+publication. In the following we will present the related work published since
+this publication.
 
-\subsubsection*{Domain specific solutions}
+\cite{EEP:11} cross-segment load balancing
 
-Cluster-based parallel rendering has been commercialized for off-line rendering (i.e.
-distributed ray-tracing) for computer generated animated movies or special effects, since the ray-tracing technique is inherently amenable to parallelization for off-line processing.
-Other special-purpose solutions exist for parallel rendering in specific application domains such as volume rendering \cite{LWMT:97,Wittenbrink:98,HSCSM:00,SL:02,GS:02,NSJLYZ:05} or geo-visualization \cite{VR:91,AG:95,LDC:96,JLMV:06}.
-However, such specific solutions are typically not applicable as a generic parallel rendering paradigm and do not translate to arbitrary scientific visualization and distributed graphics problems.
+\cite{DK:11}: Framework to develop apps for multitile, thus targeted
+specifically for tiled-wall visualization systems: grid configuration,
+scalability (?), multiple work sessions, multiuser events. Master-slave approach
+like Eq. with grid nodes running instances of the application. Is maybe more
+like a GLUT and window replacement, or enhancement to deal with the OpenGL
+contexts and events
 
-Recently in \cite{NC:07}, parallel rendering of hierarchical level-of-detail (LOD) data has been addressed and a solution specific to sort-first tile-based parallel rendering has been presented.
-While the presented approach is not a generic parallel rendering system, basic concepts presented in \cite{NC:07} such as load management and adaptive LOD data traversal can be carried over to other sort-first parallel rendering solutions.
+\cite{CKP:12} tiled display wall virtual environment
 
-\subsubsection*{Special-purpose architectures}
+\cite{TBD} DisplayCluster
 
-Traditionally,  high-performance real-time rendering systems have relied on an integrated proprietary system architecture, such as the SGI graphics super computers.
-These special-purpose solutions have become a niche product as their graphics performance does not keep up with off-the-shelf workstation graphics hardware and scalability of clusters.
-However, cluster systems need more sophisticated parallel graphics rendering libraries, such as the one proposed in this paper.
+\cite{NHM:11} ClusterGL?
 
-Due to its conceptual simplicity, a number of special-purpose image compositing hardware solutions for sort-last parallel rendering have been developed.
-The proposed hardware architectures include Sepia \cite {MHS:99a,sepia}, Sepia~2 \cite{LMSBHa:01,LMSBH:01}, Lightning~2 \cite{Stoll01}, Metabuffer \cite{Blanke00,Zhang01}, MPC Compositor \cite{Muraki01} and PixelFlow \cite{Molnar92, Eyles97}, of which only a few have reached the commercial product stage (i.e.
-Sepia~2 and MPC Compositor).
-However, the inherent inflexibility and setup overhead have limited their distribution and application support.
-Moreover, with the recent advances in the speed of CPU-GPU interfaces, such as PCI Express and other modern interconnects, combinations of software and GPU-based solutions offer more flexibility at comparable performance.
-
-% FlowVR \cite{AGLLMRR:04}, not really related that much here
-
-\subsubsection*{Generic approaches}
-
-A number of algorithms and systems for parallel rendering have been developed in the past.
-On one hand, some general concepts applicable to cluster parallel rendering have been presented in \cite{Mueller:95,Mueller:97} (sort-first architecture), \cite{SZFLS:99,SFLS:00} (load balancing), \cite{SFL:01} (data replication), or \cite{CMF:05,CM:06} (scalability).
-On the other hand, specific algorithms have been developed for cluster based rendering and compositing such as \cite{AP:98}, \cite{CKS:02} and \cite{YYC:01,SMLAP:03}.
-However, these approaches do not constitute APIs and libraries that can readily be integrated into existing visualization applications, although the issue of the design of a parallel graphics interface has been addressed in \cite{Igehy98}.
-%
-Only few generic APIs and (cluster-) parallel rendering systems exist which include VR Juggler \cite{BJHMBC:01} (and its derivatives), Chromium \cite{HHNFAKK:02} (an evolution of \cite{Humphreys99,Humphreys00,HEBSEH:01}) and OpenGL Multipipe SDK \cite{JDBJBCER:04,BRE:05,MPK}.
-
-VR Juggler \cite{BJHMBC:01,JBBC:98} is a graphics framework for virtual
-reality applications which shields the application developer from the
-underlying hardware architecture, devices and operating system. Its main
-aim is to make virtual reality configurations easy to set up and use
-without the need to know details about the devices and hardware
-configuration, but not specifically to provide scalable parallel
-rendering. Extensions of VR Juggler, such as for example ClusterJuggler
-\cite{BC:03} and NetJuggler \cite{AGLMR:02}, are typically based on the
-replication of application and data on each cluster node and basically
-take care of synchronization issues, but fail to provide a flexible and
-powerful configuration mechanism that efficiently supports scalable
-rendering as also noted in \cite{SWNH:03}. The presented system is different from VR Juggler in that it
-fully supports scalable parallel rendering such as sort-first and
-sort-last task decomposition and image compositing, it provides more
-flexible node configurations which for example allow specifying
-arbitrary task decomposition and image compositing combinations as
-simple compound layouts. Furthermore, it it is fully distributed which includes
-support for network swap barriers (synchronization), distributed objects
-as well as image compression and transmission. In contrast to VR
-Juggler, Equalizer supports multiple rendering threads per process,
-which is important for multi-GPU systems.
-
-While Chromium \cite{HHNFAKK:02} provides a powerful and transparent
-abstraction of the OpenGL API, that allows a flexible configuration of
-display resources, its main limitation with respect to scalable
-rendering is that it is focused on streaming OpenGL commands through a
-network of nodes, often initiated from a single source. This has also been
-observed in \cite{SWNH:03}. The problem
-comes in when the OpenGL stream is large in size, due to not only
-containing OpenGL calls but also the rendered data such as geometry
-and image data. Only if the geometry and textures are mostly static
-and can be kept in GPU memory on the graphics card, no significant
-bottleneck can be expected as then the OpenGL stream is composed of a
-relatively small number of rendering instructions. However, as it is
-typical in real-world visualization applications, display and object
-settings are interactively manipulated, data and parameters may change
-dynamically, and large data sets do not fit statically in GPU memory but
-are often dynamically loaded from out-of-core and/or multiresolution
-data structures. This can lead to frequent updates not only of commands and
-parameters wich have to be distributed but also of
-the rendered data itself (geometry and texture), thus causing the OpenGL stream
-to expand dramatically. Furthermore, this stream of function calls and
-data must be packaged and broadcast in real-time over the network to
-multiple nodes for each rendered frame. This makes CPU performance and
-network bandwidth a more likely limiting factor. While preserving a minimally
-invasive API, the novel proposed system is better aimed at scalability
-as the actual data access is decentralized in the distributed rendering
-clients.
-
-The performance experiments in \cite{HHNFAKK:02} indicate that Chromium is working
-quite well when the rendering problem is fill-rate limited.
-This is due to the fact that the OpenGL commands and a non-critical
-amount of rendering data can be distributed to multiple nodes without significant problems
-and since the critical fill-rate work is then performed locally on the graphics hardware.
-
-Chromium also provides some facilities for parallel application
-development, namely a sort-last, binary-swap compositing SPU and an
-OpenGL extension providing synchronization primitives, such as a barrier
-and semaphore. It leaves other problems, such as configuration, task
-decomposition as well as process and thread management unaddressed, thus
-making the development of parallel OpenGL applications harder than with
-Equalizer. Parallel Chromium applications tend to be written for one
-specific parallel rendering use case, such as for example the sort-first distributed
-memory volume renderer \cite{BHPB:03} or the sort-last parallel
-volume renderer raptor \cite{Raptor}. We are not aware of a generic Chromium-based
-application using many-to-one sort-first or stereo decompositions. This is
-another difference to Equalizer which provides a much more flexible
-task decomposition configuration. Applications written once for Equalizer can
-easily be run in any different task decomposition mode and for any physical
-display configuration without any changes to the application itself.
-While Equalizer provides an abstraction of all entities of the rendering pipeline
-(see Sections~\ref{SEC_architecture} and~\ref{SEC_development}), Chromium's
-infrastructure is primarily the compositing stage.
-
-OpenGL Multipipe SDK (MPK) \cite{BRE:05} implements an effective
-parallel rendering API for a shared memory multi-CPU/GPU system. It is
-similar to IRIS Performer \cite{RH:94} in that it handles multi-pipe
-rendering by a lean abstraction layer via a conceptual callback
-mechanism, and that it runs different application tasks in
-parallel. However, MPK is not designed nor meant for rendering nodes
-separated by a network. MPK focuses on providing a parallel rendering
-framework for a single application, parts of which are run in parallel
-on multiple rendering channels, such as the culling, rendering and final
-image compositing processes. Compared to MPK, Equalizer supports a
-fully distributed parallel rendering paradigm and features a more flexible
-task decomposition approach.
+Omegalib
 
@@
 
+In this section we present features motivated by real-world application use
+cases, i.e., new functionalities rather then performance improvements. We
+motivate the use case, explain the architecture and integration into our
+parallel rendering framework, and, where applicable, show the steps needed to
+use this functionality in applications.
+
 \subsection{Physical and Logical Visualization Setup}
 
-Canvas, Layout, Observer
+Real-world visualization setups are often complex, and having an abstract
+representation of the display system can simplify the configuration
+process. Real-world applications often have the need to be aware of spatial
+relationship of the display setup, for example to render 2D overlays or to
+configure multiple views on a tiled display wall.
 
-multi-view with runtime switching,
+We addressed this need through a new configuration section interspersed between
+the node/pipe/window/channel hardware resources and the compound trees
+configurating the resource usage for parallel rendering.
 
-2D overlays
+A typical installation consists of one projection canvas, which is one
+aggregated projection surface, e.g., a tiled display wall or a CAVE. Desktop
+windows are considered a canvas. Each canvas is made of one or more segments,
+which are the individual outputs connected to a display or projector. Segments
+can be planar or non-planar to each other, and can overlap or have gaps between
+each other. A segment is referencing a channel, which defines the output area of
+this segment, e.g., on a DVI connector connected to a projector.
 
-Swap Barriers
+A canvas can define a frustum, which will create default, planar sub-frusta for
+all of its segments. A segment can also define a frustum, which overrides the
+canvas frustum, e.g., for non-planar setups such as CAVEs or curved
+screens. These frusta describe a physically-correct display setup for a Virtual
+Reality installation. A canvas may have a software or hardware swap barrier,
+which will synchronize the rendering of all contributing GPUs.
 
-\subsection{Auto-Configuration}
+On each canvas, the application can display one or more views. A view is a view
+on a model, in the sense used by the MVC pattern. The view class is used by
+Equalizer applications to define view-specific data for rendering, e.g., a
+scene, viewing mode or camera. The application process manages this data, and
+the render clients receive it for rendering.
 
-hwsd with VirtualGL support
+A layout groups one or more views which logically belong together. A layout is
+applied to a canvas. The layout assignment can be changed at run-time by the
+application. The intersection between views and segments defines which output
+channels are available, and which frustum they should use for rendering. These
+output channels are then used as destination channels in a compound. They are
+automatically created during configuration.
 
-\subsection{Sequel}
+A view may have a frustum description. The view's frustum overrides frusta
+specified at the canvas or segment level. This is used for non-physically
+correct rendering, e.g., to compare two models side-by-side on a tiled display
+wall. If the view does not specify a frustum, the corresponding destination
+channels will use the physically correct sub-frustum resulting from the
+view/segment intersection.
 
-application, renderer, view data
+An observer looks at one or more views. It is described by the observer position
+in the world and its eye separation. Each observer will have its own stereo
+mode, focus distance and frame loop (framerate). This allows to have untracked
+views and multiple tracked views, e.g., two HMDs, in the same application.
+
+\subsection{Automatic Configuration}
+
+Automatic configuration implements the discovery of local and remote resources
+as well as the creation of typical configurations using the discovered resources
+at application launch time.
+
+The discovery is implemented in a separate library, hwsd (HardWare Service
+Discovery), which uses a plugin-based approach to discover GPUs for GLX, AGL or
+WGL windowing systems, as well as network interfaces on Linux, Mac OS X and
+Windows. Furthermore, it detects the presence of VirtualGL to allow optimal
+configuration of remote visualization clusters. The resources can be discovered
+on the local workstation, and through the help of a simple daemon using the
+zeroconf protocol, on a set of remote nodes within a visualization cluster. A
+session identifier may be used to support multiple users on a single cluster.
+
+The Equalizer server uses the hwsd library to discover local and remote
+resources when an hwsd session name instead of a \textsf{.eqc} configuration
+file is provided. A set of standard decomposition modes is configured, which can
+be selected through activating the corresponding layout.
+
+This versatile mechanism allows non-experts to configure asnd profit from
+multi-GPU workstations and visualization clusters, as well as to provide system
+administrators with the tools to implement easy to use integration with cluster
+schedulers. This feature is transparent to Equalizer application developers.
 
@@
 Tiled interactive display environment, parallel pixel streaming, events
+
+\subsection{Sequel}
+
+application, renderer, view data
 
@@
 
-Equalizer are one addition to compound trees. They modify parameters of their
+Equalizer are an addition to compound trees. They modify parameters of their
 respective subtree at runtime to optimize one aspect of the decomposition.
@@
 \subsection{RTNeuron}
+\subsection{Omegalib}
 
@@
 \label{SEC_results}
+
+\subsection{Object Distribution}
+
+Distribute large ply model to 1..16 nodes using TCP, IB and RSP
+
+\subsection{Decomposition Modes}
+
+Two graphs: Time to render a 4k, 256-step MSAA image of largest ply model and
+Livre using 1..16 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps),
+tiles, chunks, pixel, subpixel)
 

### 2017-04-03 paperIEEE.tex
ROI & async compositing
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@4 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-for compositing and load equalizer.
+The region of interest is the screen-space 2D bounding box enclosing the
+geometry rendered by a single resource. We have extended the core parallel
+rendering framework to use an application-provided ROI to optimize the load
+equalizer as well as image compositing performance. The load equalizer uses the
+ROI to refine its load grid to the regions containing data. The compositing code
+uses the ROI to minimize image readback and network transmission. In
+\cite{EBAHMP:12}, we provide the details of the algorithm, and show that this
+improves rendering performance of up to 25\%.
 
-\subsubsection{Asynchronous Readback}
+\subsubsection{Asynchronous Compositing}
 
-pipelines rendering with compositing
+Asynchronous compositing pipelines rendering with compositing operations, by
+executing the image readback, network transfer and image assembly from threads
+running in parallel to the rendering threads. In \cite{EBAHMP:12}, we provide
+the details of the implementation and experimental data showing an improvement
+of the rendering performance of over 25\% for large node counts.
 
@@
 \subsection{RTT Deltagen}
+
 \subsection{RTNeuron}
+\cite{HBBES:13}
+
+\subsection{Raster}
 \subsection{Omegalib}

### 2017-04-03 paperIEEE.tex
Draft Decompostion section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@5 db7c848b-7755-4df2-b763-8796797afa52

@@
 \newcommand{\FIXME}[1]{\textbf{\color{BLUE}{FIXME: #1}}}
+\newcommand{\sref}[1]{Section~\ref{#1}}
 %\newcommand{\FIXME}[1]{}
@@
 
-Interaction with threading
+Time-multiplexing, or DPlex, was already implemented in \cite{BRE:05}. While it
+increases the framerate linearly, it does not decrease the latency between user
+input and the corresponding output. Consequently, this decomposition mode is
+mostly useful for non-interactive movie generation. It is transparent to
+Equalizer applications, but does require the configuration latency to be equal
+or greater than the number of source channels. Furthermore, to work in a
+multi-threaded, multi-GPU configuration, the application needs to support
+running the rendering threads asynchronously, as outlined in
+\sref{SEC_threading}.
 
@@
 
-Equalizers do what?
+Tile and chunk decompositions are a variant of sort-first and sort-last
+rendering, respectively. They decompose the scene into a predefined set of
+fixed-size image tiles or database ranges. These tasks are queued and processed
+by all source channels by polling a server-central queue. Prefetching ensures
+that the task communication overlaps with rendering. As shown in \cite{SPEP:16},
+these modes provide better performance due to being inherently load-balanced, as
+long as there is an insignificant overhead for the render task setup. This mode
+is transparent to Equalizer applications.
 
@@
 
-Equal load for fill-limited apps
+Pixel compounds decompose the destination channel by interleaving rows or
+columns in image space. They are a variant of sort-first decomposition which
+works well for fill-limited applications which are not geometry bound. Source
+channels cannot reduce geometry load through view frustum culling, since each
+source channel has almost the same frustum (only shifted by some pixels),
+but applied to a reduced 2D viewport. However, the fragment load on all source
+channels is very similar due to the interleaved distribution of pixels. This
+functionality is transparent to Equalizer applications, and the default
+compositing implementation uses the OpenGL stencil buffer to blit pixels onto
+the destination channel.
 
@@
 
-MSAA and DOF, plus idle refinement algo
+Subpixel compounds are similar to pixel compounds, but they decompose the work
+for a single pixel, for example when using multisampling or depth of
+field. Composition typically uses accumulation and averaging of all computed
+fragments for a pixel. This feature is not fully transparent to the application,
+since it needs to adapt (jitter or tilt) the frustum based on the iteration
+executed. Furthermore, subpixel compounds interact with idle image refinements,
+e.g., they can accelerate idle anti-aliasing of a scene when the camera and
+scene are not changed.
 
@@
 
-\subsubsection{Thread Synchronization Modes}
+\subsubsection{Thread Synchronization Modes}\label{SEC_threading}
 

### 2017-04-03 paperIEEE.tex
Fill in some equalizers
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@6 db7c848b-7755-4df2-b763-8796797afa52

@@
 \usepackage{multirow}
+\usepackage{paralist}
 
@@
 
+Virtual Reality
+
+\subsection{Head Tracking}
+
 \subsection{Dynamic Focus Distance}
@@
 running the rendering threads asynchronously, as outlined in
-\sref{SEC_threading}.
+\sref{SEC_threading}. The output frame rate of the destination channel is
+smoothened using a frame rate equalizer (\sref{SEC_framerateEq}).
 
@@
 
-reactive load equalization, describe tunables (damping, resistance, delta,
-granularity), sort-first respects input channel size
+Sort-first and sort-last load balancing is the most obvious optimization for
+these parallel rendering modes. Our load equalizer is fully transparent for
+application developers, that is, it uses a reactive approach based on past
+rendering times. This assumes a reasonable frame-to-frame coherency. Our
+implementation stores a 2D or 1D grid of the load, mapping the load of each
+channel. The load is stored in normalized 2D/1D coordinates using
+$\frac{time}{area}$ as the load, the contributing source channels are organized
+in a binary tree, and then the algorithm balances the two branches of each level
+by equalizing the integral over the area on each side.
 
-\subsubsection{Cross-Segment Load Equalizer}
+We have implemented various tunable parameters allowing application developers
+to optimize the load balancing based on the characteristics of their rendering
+algorithm:
+\begin{compactdesc}
+\item[Damping] reduces frame-to-frame oscillations. It is a normalized scalar
+  defining how much of the computed delta from the previous position is
+  applied. The equal load distribution within the region of interest assumed by
+  the load equalizer is in reality not equal, causing the load balancing to
+  overshoot.
+\item[Resistance] eliminates small deltas in the load balancing step. This might
+  help the application to cache some computations since the frustum does not
+  change each frame.
+\item[Boundaries] define the modulo factor in pixels onto which a load split may
+  fall. Some rendering algorithms produce artefacts related to the OpenGL raster
+  position, e.g., screen door transparency, which can be eliminated by aligning
+  the boundary to the pixel repetition. Furthermore, some rendering algorithms
+  are sensitive to cache alignments, which can again be exploited by chosing the
+  corresponding boundary.
+\end{compactdesc}
 
-summarize \cite{EEP:11}
+\subsubsection{Cross-Segment Load Balancing}
+
+Cross-segment load balancing addresses the optimal resource allocation of $n$
+rendering resources to $m$ output channels (with $n\geq m$). The view equalizer
+works in conjunction with load equalizer balancing the individual output
+channels. It monitors the usage of shared source channels (across outputs) and
+activates them to balance the rendering time of all outputs. In \cite{EEP:11},
+we provide a detailed description and evaluation of our algorithm.
 
@@
 
-\subsubsection{Frame Rate Equalizer}
+\subsubsection{Frame Rate Equalizer}\label{SEC_framerateEq}
 

### 2017-04-03 paperIEEE.tex
Frame rate eq
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@7 db7c848b-7755-4df2-b763-8796797afa52

@@
 Equalizer are an addition to compound trees. They modify parameters of their
-respective subtree at runtime to optimize one aspect of the decomposition.
+respective subtree at runtime to optimize one aspect of the decomposition. Due
+to their nature, they are transparent to application developers, but might have
+application-accessible parameters to tune their behaviour.
 
@@
 
-Predominantly used by DPlex to smooth output framerate
+The framerate equalizer smoothens the output frame rate of a destination
+channel by instructing the corresponding window to delay its buffer swap to a
+minimum time between swaps. This is regularly used for time-multiplexed
+decompostions, where source channels tend to drift and finish their rendering
+not evenly distributed over time. This equalizer is however fully independent of
+DPlex compounds, and may be used to smoothen irregular application rendering
+algorithms.
 

### 2017-04-03 paperIEEE.tex
VR and Equalizer content
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@8 db7c848b-7755-4df2-b763-8796797afa52

@@
 \newcommand{\sref}[1]{Section~\ref{#1}}
+\newcommand{\fig}[1]{Figure~\ref{#1}}
 %\newcommand{\FIXME}[1]{}
@@
 
-\cite{EMP:09} summarized the work in parallel rendering at the time of
-publication. In the following we will present the related work published since
-this publication.
+In \cite{EMP:09}, we presented the Equalizer parallel rendering framework. This
+paper also summarized the work in parallel rendering up to 2009. In the
+following we will present the related work published since then.
 
@@
 
+\label{SEC_observer}
 An observer looks at one or more views. It is described by the observer position
@@
 
-Virtual Reality
+Virtual Reality is an important field for parallel rendering. It does however
+require special attention to support it as a first-class citizen in a generic
+parallel rendering framework. Equalizer has been used in many virtual reality
+installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
+at the KAUST visualization laboratory and head-mounted displays
+(\fig{FIG_teaser}). In the following we lay out the features needed support
+these installations. All features presented were motivated by application use
+cases and have been validated with their respective users.
 
@@
 
+Head tracking is the minimal feature needed to support immersive
+installations. Equalizer does support multiple, independent tracked views
+through the observer abstraction introduced in \sref{SEC_observer}. Built-in
+VRPN support enables the direct, application-transparent configuration of a VRPN
+tracker device. Alternatively, an application can provide a $4\times 4$ tracking
+matrix defining the transformation from the canvas coordinate system to the
+observer. Both CAVE-like tracking with fixed projection surfaces and HMD
+tracking modes are implemented.
+
 \subsection{Dynamic Focus Distance}
 
-Focus what user is looking at
+To our knowledge, all parallel rendering systems have the focal plane coincide
+with the physical display surface. For better viewing comfort, we introduce a
+new dynamic focus mode, where the application defines the distance of the focal
+plane from the observer, based on the current view direction af the
+user. Initial experiments show that this is particularly effective for objects
+placed within the immersive space, that is, in front of all display segments.
 
@@
 
-Better HMD by measuring user geometry
+Traditional head tracking computes the left and right eye positions by using a
+configurable interocular distance and the tracking matrix. However, human heads
+are not symmetric, and by measuring individual users a more precise frustum can
+be computed. Equalizer supports this through the optional configuration (file or
+programmatically) of individual 3D positions for the left and right eye.
 
@@
 
-Constant frame rate for fill-limited applications
+The DFR equalizer provides a functionality similar to dynamic video resizing
+\cite{MBDM:97}, that is, it maintains a constant framerate by adapting the
+rendering resolution of a fill-limited application. In Equalizer, this works by
+rendering into a source channel (typically on a FBO) separate to the destination
+channel, and then scaling the rendering during the transfer (typically through
+an on-GPU texture) to the destination channel. The DFR equalizer monitors the
+rendering performance and accordingly adapts the resolution of the source
+channel and zoom factor for the source to destination transfer. If the
+performance and source channel resolutions allows, this will not only subsample,
+but also supersample the destination channel to reduce aliasing artefacts.
 
@@
 Livre using 1..16 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps),
-tiles, chunks, pixel, subpixel)
+tiles, chunks, pixel, subpixel) (+DFR for Livre)
 

### 2017-04-03 paperIEEE.tex
Add 2010 compositing paper
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@9 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+Qt is a popular window system with many application developers. Unfortunately,
+it imposes a different threading model for window creation and event handling
+then Equalizer. In Equalizer, each GPU rendering thread is independently
+responsible for creating its windows, receiving the events and eventually
+dispatching them to the application process main thread. This design is
+motivated by the natural threading model of X11 and WGL, and allows simple
+sequential semantics between OpenGL rendering and event handling. In contrast,
+Qt requires all windows and QOpenGLContext to be created from the Qt main
+thread. An existing Qt window or context may be moved to a different thread, and
+events are signalled from the main thread.
+
 Challenges in threading model, architecture
@@
 uses the ROI to minimize image readback and network transmission. In
-\cite{EBAHMP:12}, we provide the details of the algorithm, and show that this
-improves rendering performance of up to 25\%.
+\cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of the algorithm, and
+show that using ROI can quadruple the rendering performance, in particular for
+the costly compositing step in sort-last rendering.
 
@@
 GPU-CPU transfer plugins with optional compression (eg YUV) linked to CPU
-compression for network transfer
+compression for network transfer \cite{MEP:10}
 

### 2017-04-03 paperIEEE.tex
Update TVCG style
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@10 db7c848b-7755-4df2-b763-8796797afa52

@@
-%\documentclass{vgtc}
-\documentclass[journal]{vgtc}                % final (journal style)
-%\documentclass[review]{vgtc}                 % review
-%\documentclass[review,journal]{vgtc}         % review (journal style)
-%\documentclass[widereview]{vgtc}             % wide-spaced review
-%\documentclass[preprint]{vgtc}               % preprint
-%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)
-%\documentclass[electronic]{vgtc}             % electronic version
-%\documentclass[electronic,journal]{vgtc}     % electronic version, journal
+%% bare_jrnl_compsoc.tex
+%% V1.4b
+%% 2015/08/26
+%% by Michael Shell
+%% See:
+%% http://www.michaelshell.org/
+%% for current contact information.
+%%
+%% This is a skeleton file demonstrating the use of IEEEtran.cls
+%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
+%% Computer Society journal paper.
+%%
+%% Support sites:
+%% http://www.michaelshell.org/tex/ieeetran/
+%% http://www.ctan.org/pkg/ieeetran
+%% and
+%% http://www.ieee.org/
 
-%% Uncomment one of the lines above depending on where your paper is
-%% in the conference process. ``review'' and ``widereview'' are for review
-%% submission, ``preprint'' is for pre-publication, and the final version
-%% doesn't use a specific qualifier. Further, ``electronic'' includes
-%% hyperreferences for more convenient online viewing.
+%%*************************************************************************
+%% Legal Notice:
+%% This code is offered as-is without any warranty either expressed or
+%% implied; without even the implied warranty of MERCHANTABILITY or
+%% FITNESS FOR A PARTICULAR PURPOSE!
+%% User assumes all risk.
+%% In no event shall the IEEE or any contributor to this code be liable for
+%% any damages or losses, including, but not limited to, incidental,
+%% consequential, or any other damages, resulting from the use or misuse
+%% of any information contained here.
+%%
+%% All comments are the opinions of their respective authors and are not
+%% necessarily endorsed by the IEEE.
+%%
+%% This work is distributed under the LaTeX Project Public License (LPPL)
+%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
+%% distributed and modified. A copy of the LPPL, version 1.3, is included
+%% in the base LaTeX documentation of all distributions of LaTeX released
+%% 2003/12/01 or later.
+%% Retain all contribution notices and credits.
+%% ** Modified files should be clearly indicated as such, including  **
+%% ** renaming them and changing author support contact information. **
+%%*************************************************************************
 
-%% Please use one of the ``review'' options in combination with the
-%% assigned online id (see below) ONLY if your paper uses a double blind
-%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
-%% in the past.
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. The IEEE's font choices and paper sizes can   ***
+% *** trigger bugs that do not appear when using other class files.       ***                          ***
+% The testflow support page is at:
+% http://www.michaelshell.org/tex/testflow/
 
-%% These three lines bring in essential packages: ``mathptmx'' for Type 1
-%% typefaces, ``graphicx'' for inclusion of EPS figures. and ``times''
-%% for proper handling of the times font family.
-%% We encourage the use of mathptmx for consistent usage of times font
-%% throughout the proceedings. However, if you encounter conflicts
-%% with other math-related packages, you may want to disable it.
+\documentclass[10pt,journal,compsoc]{IEEEtran}
+%
+% If IEEEtran.cls has not been installed into the LaTeX system files,
+% manually specify the path to it like:
+% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}
+
+% Some very useful LaTeX packages include:
+% (uncomment the ones you want to load)
+
+% *** MISC UTILITY PACKAGES ***
+%
+%\usepackage{ifpdf}
+% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
+% compilation based on whether the output is pdf or dvi.
+% usage:
+% \ifpdf
+%   % pdf code
+% \else
+%   % dvi code
+% \fi
+% The latest version of ifpdf.sty can be obtained from:
+% http://www.ctan.org/pkg/ifpdf
+% Also, note that IEEEtran.cls V1.7 and later provides a builtin
+% \ifCLASSINFOpdf conditional that works the same way.
+% When switching from latex to pdflatex and vice-versa, the compiler may
+% have to be run twice to clear warning/error messages.
+
+% *** CITATION PACKAGES ***
+%
+\ifCLASSOPTIONcompsoc
+  % IEEE Computer Society needs nocompress option
+  % requires cite.sty v4.0 or later (November 2003)
+  \usepackage[nocompress]{cite}
+\else
+  % normal IEEE
+  \usepackage{cite}
+\fi
+% cite.sty was written by Donald Arseneau
+% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
+% \cite{} output to follow that of the IEEE. Loading the cite package will
+% result in citation numbers being automatically sorted and properly
+% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
+% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
+% \cite will automatically add leading space, if needed. Use cite.sty's
+% noadjust option (cite.sty V3.8 and later) if you want to turn this off
+% such as if a citation ever needs to be enclosed in parenthesis.
+% cite.sty is already installed on most LaTeX systems. Be sure and use
+% version 5.0 (2009-03-20) and later if using hyperref.sty.
+% The latest version can be obtained at:
+% http://www.ctan.org/pkg/cite
+% The documentation is contained in the cite.sty file itself.
+%
+% Note that some packages require special options to format as the Computer
+% Society requires. In particular, Computer Society  papers do not use
+% compressed citation ranges as is done in typical IEEE papers
+% (e.g., [1]-[4]). Instead, they list every citation separately in order
+% (e.g., [1], [2], [3], [4]). To get the latter we need to load the cite
+% package with the nocompress option which is supported by cite.sty v4.0
+% and later. Note also the use of a CLASSOPTION conditional provided by
+% IEEEtran.cls V1.7 and later.
+
+% *** GRAPHICS RELATED PACKAGES ***
+%
+\ifCLASSINFOpdf
+  % \usepackage[pdftex]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../pdf/}{../jpeg/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+\else
+  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
+  % will default to the driver specified in the system graphics.cfg if no
+  % driver is specified.
+  % \usepackage[dvips]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../eps/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.eps}
+\fi
+% graphicx was written by David Carlisle and Sebastian Rahtz. It is
+% required if you want graphics, photos, etc. graphicx.sty is already
+% installed on most LaTeX systems. The latest version and documentation
+% can be obtained at:
+% http://www.ctan.org/pkg/graphicx
+% Another good source of documentation is "Using Imported Graphics in
+% LaTeX2e" by Keith Reckdahl which can be found at:
+% http://www.ctan.org/pkg/epslatex
+%
+% latex, and pdflatex in dvi mode, support graphics in encapsulated
+% postscript (.eps) format. pdflatex in pdf mode supports graphics
+% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
+% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
+% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
+% which can result in "jaggedy"/blurry rendering of lines and letters as
+% well as large increases in file sizes.
+%
+% You can find documentation about the pdfTeX application at:
+% http://www.tug.org/applications/pdftex
+
+% *** MATH PACKAGES ***
+%
+%\usepackage{amsmath}
+% A popular package from the American Mathematical Society that provides
+% many useful and powerful commands for dealing with mathematics.
+%
+% Note that the amsmath package sets \interdisplaylinepenalty to 10000
+% thus preventing page breaks from occurring within multiline equations. Use:
+%\interdisplaylinepenalty=2500
+% after loading amsmath to restore such page breaks as IEEEtran.cls normally
+% does. amsmath.sty is already installed on most LaTeX systems. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/pkg/amsmath
+
+% *** SPECIALIZED LIST PACKAGES ***
+%
+%\usepackage{algorithmic}
+% algorithmic.sty was written by Peter Williams and Rogerio Brito.
+% This package provides an algorithmic environment fo describing algorithms.
+% You can use the algorithmic environment in-text or within a figure
+% environment to provide for a floating algorithm. Do NOT use the algorithm
+% floating environment provided by algorithm.sty (by the same authors) or
+% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
+% algorithm float types and packages that provide these will not provide
+% correct IEEE style captions. The latest version and documentation of
+% algorithmic.sty can be obtained at:
+% http://www.ctan.org/pkg/algorithms
+% Also of interest may be the (relatively newer and more customizable)
+% algorithmicx.sty package by Szasz Janos:
+% http://www.ctan.org/pkg/algorithmicx
+
+% *** ALIGNMENT PACKAGES ***
+%
+%\usepackage{array}
+% Frank Mittelbach's and David Carlisle's array.sty patches and improves
+% the standard LaTeX2e array and tabular environments to provide better
+% appearance and additional user controls. As the default LaTeX2e table
+% generation code is lacking to the point of almost being broken with
+% respect to the quality of the end results, all users are strongly
+% advised to use an enhanced (at the very least that provided by array.sty)
+% set of table tools. array.sty is already installed on most systems. The
+% latest version and documentation can be obtained at:
+% http://www.ctan.org/pkg/array
+
+% IEEEtran contains the IEEEeqnarray family of commands that can be used to
+% generate multiline equations as well as matrices, tables, etc., of high
+% quality.
+
+% *** SUBFIGURE PACKAGES ***
+%\ifCLASSOPTIONcompsoc
+%  \usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
+%\else
+%  \usepackage[caption=false,font=footnotesize]{subfig}
+%\fi
+% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
+% for subfigure.sty, the latter of which is no longer maintained and is
+% incompatible with some LaTeX packages including fixltx2e. However,
+% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
+% which will override IEEEtran.cls' handling of captions and this will result
+% in non-IEEE style figure/table captions. To prevent this problem, be sure
+% and invoke subfig.sty's "caption=false" package option (available since
+% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
+% handling of captions.
+% Note that the Computer Society format requires a sans serif font rather
+% than the serif font used in traditional IEEE formatting and thus the need
+% to invoke different subfig.sty package options depending on whether
+% compsoc mode has been enabled.
+%
+% The latest version and documentation of subfig.sty can be obtained at:
+% http://www.ctan.org/pkg/subfig
+
+% *** FLOAT PACKAGES ***
+%
+%\usepackage{fixltx2e}
+% fixltx2e, the successor to the earlier fix2col.sty, was written by
+% Frank Mittelbach and David Carlisle. This package corrects a few problems
+% in the LaTeX2e kernel, the most notable of which is that in current
+% LaTeX2e releases, the ordering of single and double column floats is not
+% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
+% single column figure to be placed prior to an earlier double column
+% figure.
+% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
+% corrections already built into the system in which case a warning will
+% be issued if an attempt is made to load fixltx2e.sty as it is no longer
+% needed.
+% The latest version and documentation can be found at:
+% http://www.ctan.org/pkg/fixltx2e
+
+%\usepackage{stfloats}
+% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
+% the ability to do double column floats at the bottom of the page as well
+% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
+% LaTeX2e). It also provides a command:
+%\fnbelowfloat
+% to enable the placement of footnotes below bottom floats (the standard
+% LaTeX2e kernel puts them above bottom floats). This is an invasive package
+% which rewrites many portions of the LaTeX2e float routines. It may not work
+% with other packages that modify the LaTeX2e float routines. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/pkg/stfloats
+% Do not use the stfloats baselinefloat ability as the IEEE does not allow
+% \baselineskip to stretch. Authors submitting work to the IEEE should note
+% that the IEEE rarely uses double column equations and that authors should try
+% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
+% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
+% such ways.
+% Do not attempt to use stfloats with fixltx2e as they are incompatible.
+% Instead, use Morten Hogholm'a dblfloatfix which combines the features
+% of both fixltx2e and stfloats:
+%
+% \usepackage{dblfloatfix}
+% The latest version can be found at:
+% http://www.ctan.org/pkg/dblfloatfix
+
+%\ifCLASSOPTIONcaptionsoff
+%  \usepackage[nomarkers]{endfloat}
+% \let\MYoriglatexcaption\caption
+% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
+%\fi
+% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and
+% Axel Sommerfeldt. This package may be useful when used in conjunction with
+% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
+% submissions have lists of figures/tables at the end of the paper and that
+% figures/tables without any captions are placed on a page by themselves at
+% the end of the document. If needed, the draftcls IEEEtran class option or
+% \CLASSINPUTbaselinestretch interface can be used to increase the line
+% spacing as well. Be sure and use the nomarkers option of endfloat to
+% prevent endfloat from "marking" where the figures would have been placed
+% in the text. The two hack lines of code above are a slight modification of
+% that suggested by in the endfloat docs (section 8.4.1) to ensure that
+% the full captions always appear in the list of figures/tables - even if
+% the user used the short optional argument of \caption[]{}.
+% IEEE papers do not typically make use of \caption[]'s optional argument,
+% so this should not be an issue. A similar trick can be used to disable
+% captions of packages such as subfig.sty that lack options to turn off
+% the subcaptions:
+% For subfig.sty:
+% \let\MYorigsubfloat\subfloat
+% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
+% However, the above trick will not work if both optional arguments of
+% the \subfloat command are used. Furthermore, there needs to be a
+% description of each subfigure *somewhere* and endfloat does not add
+% subfigure captions to its list of figures. Thus, the best approach is to
+% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
+% and instead reference/explain all the subfigures within the main caption.
+% The latest version of endfloat.sty and its documentation can obtained at:
+% http://www.ctan.org/pkg/endfloat
+%
+% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
+% later in the document, say, to conditionally put the References on a
+% page by themselves.
+
+% *** PDF, URL AND HYPERLINK PACKAGES ***
+%
+%\usepackage{url}
+% url.sty was written by Donald Arseneau. It provides better support for
+% handling and breaking URLs. url.sty is already installed on most LaTeX
+% systems. The latest version and documentation can be obtained at:
+% http://www.ctan.org/pkg/url
+% Basically, \url{my_url_here}.
+
+% *** Do not adjust lengths that control margins, column widths, etc. ***
+% *** Do not use packages that alter fonts (such as pslatex).         ***
+% There should be no need to do such things with IEEEtran.cls V1.6 and later.
+% (Unless specifically asked to do so by the journal or conference you plan
+% to submit to, of course. )
 
@@
 
-%% If you are submitting a paper to a conference for review with a double
-%% blind reviewing process, please replace the value ``0'' below with your
-%% OnlineID. Otherwise, you may safely leave it at ``0''.
-\onlineid{0}
-
-%% declare the category of your paper, only shown in review mode
-\vgtccategory{Research}
-
-%% allow for this line if you want the electronic option to work properly
-\vgtcinsertpkg
-
-%% In preprint mode you may define your own headline.
-%\preprinttext{To appear in an IEEE VGTC sponsored conference.}
-
 \usepackage{color}
@@
 \newcommand{\fig}[1]{Figure~\ref{#1}}
-%\newcommand{\FIXME}[1]{}
 
@@
 
-%% Paper title.
-%----------------------------------------------------------------------
+% correct bad hyphenation here
+\hyphenation{op-tical net-works semi-conduc-tor}
+
+\begin{document}
+%
+% paper title
+% Titles are generally capitalized except for words such as a, an, and, as,
+% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
+% not capitalized unless they are the first or last word of the title.
+% Linebreaks \\ can be used within to get better formatting as desired.
+% Do not put math or special symbols in the title.
 \title{Equalizer: A Mature Parallel Rendering Framework}
-%----------------------------------------------------------------------
+%
+%
+% author names and IEEE memberships
+% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
+% a structure at a ~ so this keeps an author's name from being broken across
+% two lines.
+% use \thanks{} to gain access to the first footnote area
+% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
+% was not built to handle multiple paragraphs
+%
+%
+%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
+% lists the Computer Society journals use for "first footnote" author
+% affiliations. Use \IEEEcompsocthanksitem which works much like \item
+% for each affiliation group. When not in compsoc mode,
+% \IEEEcompsocitemizethanks becomes like \thanks and
+% \IEEEcompsocthanksitem becomes a line break with idention. This
+% facilitates dual compilation, although admittedly the differences in the
+% desired content of \author between the different types of papers makes a
+% one-size-fits-all approach a daunting prospect. For instance, compsoc
+% journal papers have the author affiliations above the "Manuscript
+% received ..."  text while in non-compsoc journals this is reversed. Sigh.
 
@@
 
-\affiliation{\scriptsize Visualization and MultiMedia Lab \\ Department of Informatics \\ University of Z\"urich}
+\author{Stefan~Eilemann, David~Steiner and Renato Pajarola% <-this % stops a space
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Visualization and MultiMedia Lab, Department of Informatics, University of Z\"urich.
+\IEEEcompsocthanksitem S. Eilemann is also with the Blue Brain Project, Ecole
+Polytechnique Federale de Lausanne.}}
 
-%% Keywords that describe your work. Will show as 'Index Terms' in journal
-%% please capitalize first letter and insert punctuation after last keyword
-\keywords{Parallel Rendering, Scalable Visualization, Cluster Graphics, Immersive Environments, Display Walls}
+% note the % following the last \IEEEmembership and also \thanks -
+% these prevent an unwanted space from occurring between the last author name
+% and the end of the author line. i.e., if you had this:
+%
+% \author{....lastname \thanks{...} \thanks{...} }
+%                     ^------------^------------^----Do not want these spaces!
+%
+% a space would be appended to the last name and could cause every name on that
+% line to be shifted left slightly. This is one of those "LaTeX things". For
+% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
+% "AB" then you have to do: "\textbf{A}\textbf{B}"
+% \thanks is no different in this regard, so shield the last } of each \thanks
+% that ends a line with a % and do not let a space in before the next \thanks.
+% Spaces after \IEEEmembership other than the last one are OK (and needed) as
+% you are supposed to have spaces between the names. For what it is worth,
+% this is a minor point as most people would not even notice if the said evil
+% space somehow managed to creep in.
 
-%% ACM Computing Review (CR) categories.
-%% See <http://www.acm.org/class/1998/> for details.
-%% The ``\CRcat'' command takes four arguments.
+% The paper headers
+\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
+{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Computer Society Journals}
+% The only time the second header will appear is for the odd numbered pages
+% after the title page when using the twoside option.
+%
+% *** Note that you probably will NOT want to include the author's ***
+% *** name in the headers of peer review papers.                   ***
+% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
+% you desire.
 
-\CRcatlist{ % not used in journal version
-  \CRcat{I.3.2}{Graphics Systems}{Distributed Graphics}{Parallel Rendering};
-  \CRcat{I.3.m}{Miscellaneous}{Rendering Clusters}{Scalable Rendering};
-  \CRcat{I.3.7}{Three-Dimensional Graphics and Realism}{Virtual Reality}{Immersive Environments}
-}
+% The publisher's ID mark at the bottom of the page is less important with
+% Computer Society journal papers as those publications place the marks
+% outside of the main text columns and, therefore, unlike regular IEEE
+% journals, the available text space is not reduced by their presence.
+% If you want to put a publisher's ID mark on the page you can do it like
+% this:
+%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
+% or like this to get the Computer Society new two part style.
+%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2015 IEEE}%
+%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
+% Remember, if you use this you must call \IEEEpubidadjcol in the second
+% column for its text to clear the IEEEpubid mark (Computer Society jorunal
+% papers don't need this extra clearance.)
 
-\vgtccategory{Research}
+% use for special paper notices
+%\IEEEspecialpapernotice{(Invited Paper)}
 
-%% Copyright space is enabled by default as required by guidelines.
-%% It is disabled by the 'review' option or via the following command:
-% \nocopyrightspace
+% for Computer Society papers, we must declare the abstract and index terms
+% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
+% command as these need to go into the title area created by \maketitle.
+% As a general rule, do not put math, special symbols or citations
+% in the abstract or keywords.
+\IEEEtitleabstractindextext{%
+\begin{abstract}
+  We present the features, algorithms and system integration necessary to
+  implement a parallel rendering framework usable in a wide range of real-world
+  research and industry applications, based on the basic architecture and
+  implementation of the Equalizer parallel rendering framework presented in
+  \cite{EMP:09}.
+\end{abstract}
 
-%%%%%% START OF THE PAPER %%%%%%
-\begin{document}
+% Note that keywords are not normally used for peerreview papers.
+\begin{IEEEkeywords}
+Parallel Rendering, Scalable Visualization, Cluster Graphics, Immersive Environments, Display Walls
+\end{IEEEkeywords}}
+
+% make the title area
+\maketitle
+
+% To allow for easy dual compilation without having to reenter the
+% abstract/keywords data, the \IEEEtitleabstractindextext text will
+% not be used in maketitle, but will appear (i.e., to be "transported")
+% here as \IEEEdisplaynontitleabstractindextext when the compsoc
+% or transmag modes are not selected <OR> if conference mode is selected
+% - because all conference papers position the abstract like regular
+% papers do.
+\IEEEdisplaynontitleabstractindextext
+% \IEEEdisplaynontitleabstractindextext has no effect when using
+% compsoc or transmag under a non-conference mode.
+
+% For peer review papers, you can put extra information on the cover
+% page as needed:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+%
+% For peerreview papers, this IEEEtran command inserts a page break and
+% creates the second title. It will be ignored for other modes.
+\IEEEpeerreviewmaketitle
 
 %----------------------------------------------------------------------
-%% Abstract section.
-\abstract{ We present the features, algorithms and system integration necessary
-  to implement a parallel rendering framework usable in a wide range of
-  real-world research and industry applications, based on the basic architecture
-  and implementation of the Equalizer parallel rendering framework presented in
-  \cite{EMP:09}.  } % end of abstract
-
-\maketitle
-
 \begin{figure*}[ht]\center
@@
 %----------------------------------------------------------------------
-\section{Introduction}
+% Computer Society journal (but not conference!) papers do something unusual
+% with the very first section heading (almost always called "Introduction").
+% They place it ABOVE the main text! IEEEtran.cls does not automatically do
+% this for you, but you can achieve this effect with the provided
+% \IEEEraisesectionheading{} command. Note the need to keep any \label that
+% is to refer to the section immediately after \section in the above as
+% \IEEEraisesectionheading puts \section within a raised box.
+\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
 %----------------------------------------------------------------------
@@
 %
-\begin{enumerate}\renewcommand{\labelenumi}{\alph{enumi})}
-\addtolength{\itemsep}{-0.5\baselineskip}
+\begin{compactenum}\renewcommand{\labelenumi}{\alph{enumi})}
 \item generic application support, instead of special domain solution
@@
 \item exploit existing code infrastructure, such as proprietary scene graphs, molecular data structures, level-of-detail and geometry databases
-\end{enumerate}
+\end{compactenum}
 
@@
 %
-\begin{enumerate}\renewcommand{\labelenumi}{\roman{enumi})}
-\addtolength{\itemsep}{-0.5\baselineskip}
+\begin{compactenum}\renewcommand{\labelenumi}{\roman{enumi})}
 \item novel concept of compound trees for flexible configuration of graphics system resources,
@@
 \item minimally invasive programming model.
-\end{enumerate}
+\end{compactenum}
 
@@
 
-GPU-CPU transfer plugins with optional compression (eg YUV) linked to CPU
+GPU-CPU transfer plugins with optional compression (YUV, RLE) linked to CPU
 compression for network transfer \cite{MEP:10}
@@
 %----------------------------------------------------------------------
-%\section*{Acknowledgements}
+\appendices
+% use section* for acknowledgment
+\ifCLASSOPTIONcompsoc
+  % The Computer Society usually uses the plural form
+  \section*{Acknowledgments}
+\else
+  % regular IEEE prefers the singular form
+  \section*{Acknowledgment}
+\fi
 %----------------------------------------------------------------------
-%% if specified like this the section will be ommitted in review mode
-\acknowledgements{
-We would like to thank and acknowledge the following institutions and projects for providing the 3D geometry and volume test data sets:
-the Digital Michelangelo Project, Stanford 3D Scanning Repository, Cyberware Inc., volvis.org and the Visual Human Project.
-This work was partially supported by the Swiss National Science Foundation Grant 200021-116329/1.
-}
+We would like to thank and acknowledge the following institutions and projects
+for providing the 3D geometry and volume test data sets: the Digital
+Michelangelo Project, Stanford 3D Scanning Repository, Cyberware Inc.,
+volvis.org and the Visual Human Project.  This work was partially supported by
+the Swiss National Science Foundation Grant 200021-116329/1.
 
-\vspace{-2mm}
-\footnotesize
+% trigger a \newpage just before the given reference
+% number - used to balance the columns on the last page
+% adjust value as needed - may need to be readjusted if
+% the document is modified later
+%\IEEEtriggeratref{8}
+% The "triggered" command can be changed if desired:
+%\IEEEtriggercmd{\enlargethispage{-5in}}
+
+% biography section
+%
+% If you have an EPS/PDF photo (graphicx package needed) extra braces are
+% needed around the contents of the optional argument to biography to prevent
+% the LaTeX parser from getting confused when it sees the complicated
+% \includegraphics command within an optional argument. (You could create
+% your own custom macro containing the \includegraphics command to make things
+% simpler here.)
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
+% or if you just want to reserve a space for a photo:
+
+\begin{IEEEbiography}{Michael Shell}
+Biography text here.
+\end{IEEEbiography}
+
+% if you will not have a photo at all:
+\begin{IEEEbiographynophoto}{John Doe}
+Biography text here.
+\end{IEEEbiographynophoto}
+
+% insert where needed to balance the two columns on the last page with
+% biographies
+%\newpage
+
+\begin{IEEEbiographynophoto}{Jane Doe}
+Biography text here.
+\end{IEEEbiographynophoto}
+
 \bibliographystyle{abbrv}
@@
 
+% that's all folks
 \end{document}

### 2017-04-03 paperIEEE.tex
co::Object section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@11 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+An important part of writing a parallel rendering application is the
+communication layer between the individual processes. Equalizer relies on the
+Collage network library for its internal operation. Collage furthermore provides
+powerful abstractions for writing Equalizer applications, which are introduced
+in this section.
+
+\subsection{Architecture}
+
+Collage provides networking functionality of different abstraction layers,
+gradually providing higher level functionality for the programmer. The main
+primitives in Collage are:
+
+\begin{compactdesc}
+\item[Connection] A stream-oriented point-to-point communication
+  line. The connections
+  transmit raw data reliably between two endpoints for unicast connections, and
+  between a set of endpoints for multicast connections. For unicast,
+  process-local pipes, TCP and Infiniband RDMA are implemented. For multicast,
+  a reliable, UDP-based protocol is discussed in \sref{SEC_RSP}.
+\item[DataI/OStream] Abstracts the input and output of C++ data types from or to
+  a set of connections by implementing output stream operators. Uses buffering
+  to aggregate data for network transmission. Performs endian swapping during
+  input if the endianness differs between the originating and local node.
+\item[Node and LocalNode] The abstraction of a process in the cluster. Nodes
+  communicate with each other using connections. A LocalNode listens on various
+  connections and processes requests for a given process. Received data is
+  wrapped in ICommands and dispatched to command handler methods. A Node is a
+  proxy for a remote LocalNode. The Equalizer Client object is a LocalNode.
+\item[Object] Provides object-oriented, versioned data distribution of C++
+  objects between nodes. Objects are registered or mapped on a Local\-Node.
+\end{compactdesc}
+
+\begin{figure}[ht]\center
+  \includegraphics[width=\columnwidth]{EqDocs/Developer/ProgrammingGuide/images/netObject}
+  \caption{\label{fNetObject}Communication between two Collage Objects}
+\end{figure}
+\fig{fNetObject} illustrates this architecture.
+
+\subsection{Infiniband RDMA}
+
+needed? reverse-engr impl
+
+\subsection{Reliable Stream Protocol}\label{SEC_RSP}
+
+UDP-based reliability protocol
+
 \subsection{Distributed, Versioned Objects}
 
-types (instance, delta, static), versioning, multicast, compression,
-serializable with dirty bits, mapping, blocking commits
-
-\subsection{Reliable Stream Protocol}
-
-UDP-based reliability protocol
-
-\subsection{Infiniband RDMA}
-
-reverse-engr impl
+Adapting an existing application for parallel rendering requires the
+synchronization of application data across the processes in the parallel
+rendering setup. Existing parallel rendering frameworks address this often
+poorly, at best they rely on MPI to distribute data. Real-world, interactive
+visualization applications are typically written in C++ and have complex data
+models and class hierarchies to represent their application state. As outlined
+in \cite{EMP:09}, the parallel rendering code in an Equalizer application only
+needs access to the data needed for rendering, as all application logic is
+centralized in the application main thread. We have encountered two main
+approaches to address this distribution: Using a shared filesystem for static
+data and using data distribution for static and dynamic data.
+
+Distributed objects in Collage provide powerful, object-oriented data
+distribution for C++ objects. They facilitate the implementation of data
+distribution in a cluster environment. Distributed objects are created by
+subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
+application programmer implements serialization and deserialization of the
+distributed data. Distributed objects can be static (immutable) or
+dynamic. Objects have a universally unique identifier (UUID) to address them
+cluster-wide. A master-slave model is used to establish mapping and data
+synchronization across processes. Typically, the application mail loop registers
+a master instance and communicates the UUID to the render clients, which map
+their instance to the given identifier. The following object types are
+available:
+
+\begin{compactdesc}
+\item[Static] The object is not versioned nor buffered. The instance data is
+  serialized whenever a new slave instance is mapped. No additional data is
+  stored.
+\item[Instance] The object is versioned and buffered. The instance and delta
+  data are identical, that is, only instance data is serialized. Previous
+  instance data is saved to be able to map old versions.
+\item[Delta] The object is versioned and buffered. The delta data is typically
+  smaller than the instance data. The delta data is transmitted to slave
+  instances for synchronization. Previous instance and delta data is saved to be
+  able to map and sync old versions.
+\item[Unbuffered] The object is versioned and unbuffered. No data is stored, and
+  no previous versions can be mapped. No old versions can be mapped and no
+  additional data is stored.
+\end{compactdesc}
+
+All distributed objects have to implement \textsf{getInstanceData} and
+\textsf{applyInstanceData} to serialize and deserialize the object's distributed
+data. These methods provide an output or input stream as a parameter, which
+abstracts the data transmission and can be used like a \textsf{std::stream}. The
+data streams implement efficient buffering and compression, and automatically
+select the best connection for data transport. Custom data type serializers can
+be implemented by providing the appropriate serialization functions. No pointers
+should be directly transmitted through the data streams. For pointers, the
+corresponding object is typically a distributed object as well, and its
+identifier and potentially version is transmitted in place of the memory
+address.
+
+Dynamic objects are versioned, and have to override \textsf{getChangeType} to
+indicate how they want to have changes to be handled. Upon \textsf{commit} the
+delta data from the previous version is sent, if available using multicast, to
+all mapped slave instances. The data is queued on the remote node, and is
+applied when the application calls \textsf{sync} to synchronize the object to a
+new version. The \textsf{sync} method might block if a version has not yet been
+committed or is still in transmission. All versioned objects the following
+characteristics:
+
+\begin{compactitem}
+\item The master instance of the object generates new versions for all
+  slaves. These versions are continuous. It is possible to commit on slave
+  instances, but special care has to be taken to handle possible
+  conflicts.
+\item Slave instance versions can only be advanced, that is, \textsf{sync(
+  version )} with a version smaller than the current version will fail.
+\item Newly mapped slave instances are mapped to the oldest available
+  version by default, or to the version specified when calling
+  \textsf{mapObject}.
+\end{compactitem}
+
+Besides the instance data serialization methods used to map an object, versioned
+objects may implement \textsf{pack} and \textsf{unpack} to serialize or
+deserialize the changes since the last version.
+
+The Collage Serializable implements one convenient usage pattern for object data
+distribution. The \textsf{co::Serializable} data distribution is based on the
+concept of dirty bits, allowing inheritance with data distribution. Dirty bits
+form a 64-bit mask which marks the parts of the object to be distributed during
+the next commit. For serialization, the application developer implements
+\textsf{serialize} or \textsf{deserialize}, which are called with the bit mask
+specifying which data has to be transmitted or received. During a commit or
+sync, the current dirty bits are given, whereas during object mapping all dirty
+bits are passed to the serialization methods.
+
+Blocking commits allow to limit the number of outstanding, queued versions on
+the slave nodes. A token-based protocol will block the commit on the master
+instance if to many unsynchronized versions exist.
 

### 2017-04-03 paperIEEE.tex
More Collage text
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@12 db7c848b-7755-4df2-b763-8796797afa52

@@
 it imposes a different threading model for window creation and event handling
-then Equalizer. In Equalizer, each GPU rendering thread is independently
+compared to Equalizer. In Equalizer, each GPU rendering thread is independently
 responsible for creating its windows, receiving the events and eventually
@@
 
-\subsection{Sequel}
+\subsection{Sequel}\label{sec:sequel}
 
@@
 
+\subsection{Barriers, Queues and Object Maps}
+
+Collage implements a few generic distributed objects which are used by Equalizer
+and applicatons. A barrier is a distributed barrier primitive used for software
+swap barriers in Equalizer (\sref{sec:swap}). Its implementation follows a
+simple master-slave approach, which has shown to by sufficient for this use
+case. Queues are single producer, multiple consumer FIFOs. To hide network
+latencies, the consumers use prefetching of queue items into a local queue. They
+are used for tile and chunk compounds (\sref{sec:tile}).
+
+The object map facilitates distribution and synchronization of a collection of
+distributed objects. Master versions can be registered on a central node, e.g.,
+the application node in Equalizer. Consumers, e.g., Equalizer render clients,
+can selectively map the objects they are interested in. Committing the object
+map will commit all registered objects and sync their new version to the
+slaves. Syncing the map on the slaves will synchronize all mapped instances to
+the new version recorded in the object map. This effective design allows data
+distribution with minimal application logic. It is used by Sequel
+(\sref{sec:sequel}) and other Collage applications.
+
 %------------------------------------------------------------------------------
@@
 
-\subsection{Swap Synchronization and GPU affinity}
+\subsection{Swap Synchronization and GPU affinity}\label{sec:swap}
 
@@
 
-\subsubsection{Tiles and Chunks}
+\subsubsection{Tiles and Chunks}\label{sec:tile}
 

### 2017-04-03 paperIEEE.tex
Minor tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@13 db7c848b-7755-4df2-b763-8796797afa52
Filling in more gaps
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@14 db7c848b-7755-4df2-b763-8796797afa52

@@
 paper also summarized the work in parallel rendering up to 2009. In the
-following we will present the related work published since then.
+following we will present the related work published since then. An extensive
+Programming and User Guide provides in-depth documentation on using and
+programming Equalizer\cite{Eilemann:13}.
 
@@
 Reality installation. A canvas may have a software or hardware swap barrier,
-which will synchronize the rendering of all contributing GPUs.
+which will synchronize the rendering of all contributing GPUs. The software
+barrier executes a \textsf{glFinish} to ensure the GPU is ready to swap, a
+Collage barrier (\sref{SEC_barrier}) to synchronize all segments, the swap
+buffers call followed by a \textsf{glFlush} to ensure timely execution of the
+swap command. The hardware swap barrier is implemented using the
+\textsf{WGL\_NV\_swap\_group} or \textsf{GLX\_NV\_swap\_group} extensions.
 
@@
 views and multiple tracked views, e.g., two HMDs, in the same application.
+
+\subsection{Runtime Reconfiguration}\label{SEC_reconfig}
+
+Switching a layout, as described above, or switching the stereo rendering mode,
+may involve a different set of resources after the change, including the launch
+and exit of render client processes. Equalizer solves this through a
+reconfiguration step at the beginning of each rendering frame. Each resource
+(channel, window, pipe, node) has an activation count, which is updated when the
+layout or any other relevant rendering parameter is changed. When a resource is
+found whose activation count does not match its current start/stopped state, the
+resource is created or destroyed and \textsf{configInit} or \textsf{configExit}
+are called accordingly. In the current implementation, a normal configuration
+initialization or exit, as described in \cite{EMP:09}, uses the same code path
+with all used resources transitioning to a running or stopped state,
+accordingly. Since starting new resources typically requires object mapping and
+associated data distribution, it is a costly operation. To improve performance
+and robustness, Equalizer will flush the rendering pipeline whenever a runtime
+reconfiguration is detected, before executing the change. This implementation
+allows the late join or early exit, transparent to application developers.
 
@@
 
-\subsection{Barriers, Queues and Object Maps}
+\subsection{Barriers, Queues and Object Maps}\label{SEC_barrier}
 
@@
 
-\subsection{Application-specific Scaling}
-
-Gullivers world
+\subsection{Model Unit}
+
+This feature allows applications to specify a scaling factor between the model
+and the real world, to allow exploration of macroscopic or microscopic worlds in
+virtual reality. The unit is per view, allowing different scale factors within
+the same configuration. It scales both the specified projection surface as well
+as the eye position (and therefore separation) to achieve the necessary effect.
 
@@
 
-\subsection{Swap Synchronization and GPU affinity}\label{sec:swap}
+Applications can switch each view between mono and stereo rendering at runtime,
+and can run both monoscopic and stereoscopic views concurrently
+(\fig{FIG_teaser} (b)). While this is trivial for active stereo rendering, it
+does potentially involve the start and stop of resources and processes for
+passive stereo or stereo-dependent task decompositions, as described in
+\sref{SEC_reconfig}.
 
@@
 
-Control workstation in VR setups
+The monitor allows to reuse the rendering on another channel, typically for
+monitoring a larger setup on a control workstation. Output frames on the display
+channels are connected to input frames on a single monitoring channel. The
+monitor equalizer changes the scaling factor and offset between the output and
+input, so that the monitor channel has the same, but typically downscaled view,
+as the originating segments.
 

### 2017-04-03 paperIEEE.tex
Filling in more gaps
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@15 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-application, renderer, view data
+Sequal is a simplification layer for Equalizer. It is based on the realization
+that while fully expressive, the verbatim abstraction layer of nodes, pipes,
+windows and channels in Equalizer requires significant learning to fully
+understand and exploit. In reality, a higher abstraction of \textsf{Application}
+and \textsf{Renderer} is sufficient for most use cases. In Sequel, the
+application class drives the configuration, and one renderer instance is created
+for each (pipe) render thread. They also provide the natural place to store and
+distribute data. Last, but not least, view data provides a convenient way to
+manage multiple views by storing the camera, model or any other view-specific
+information.
 
@@
 The initial version of Equalizer implemented sort-first (2D), sort-last (DB) and
-stereo (EYE) decomposition. In the following we present new decomposition modes
-and motivate their use case.
+stereo (EYE) decomposition \cite{EMP:09}. In the following we present new
+decomposition modes and motivate their use case.
 
@@
 
-GPU-CPU transfer plugins with optional compression (YUV, RLE) linked to CPU
-compression for network transfer \cite{MEP:10}
+Compression for the compositing step in parallel rendering is critical for
+performance. This not only applies to the well-researched network transfer step,
+but also for the transfer between GPU and CPU. Equalizer supports a variety of
+compression algorithms, from very fast RLE encoding, JPEG compression to YUV
+subsampling on the GPU. These algorithms are implemented as runtime-loaded
+plugins, allowing easy extension and customization to application-specific
+compression. In \cite{MEP:10}, we show this to be a critical step for
+interactive performance at scale.
 
@@
 
-Per-node sync, draw sync, async
+Different applications have different degrees on how decoupled and thread-safe
+the rendering code is from the application logic. For full decoupling all
+mutable data has to have a copy in each render thread, which is not feasible in
+large data scenarios. To easily customize the synchronization of all threads on
+a single process, Equalizer implements three threading modes:  Full
+synchronization, draw synchronization and asynchronous. Note that the execution
+between nodes is always asynchronous, for up to \textsf{latency} frames.
+
+In full synchronization, all threads always execute the same frame, that is, the
+render threads are unlocked after \textsf{Node::frameStart}, and the node is
+blocked for all render threads to finish the frame before executing
+\textsf{Node::frameFinish}. This allows the render threads to read shared data
+from all their operations, but provides the slowest performance.
+
+In draw synchronization, the node thread and all render threads are synchronized
+for all \textsf{frameDraw} operations, that is, \textsf{Node::frameFinish} is
+executed after the last channel is done drawing. This allows the render threads
+to read shared data during their draw operation, but not during
+compositing. Since compositing is typically independent of the rendered data,
+this is the default mode. This mode allows to overlap compositing with data
+synchronization on multi-GPU machines.
+
+In asynchronous execution, all threads run asynchronously. Render threads may
+work on different frames at any given time. This mode is the fastest, and
+requires the application to have one instance of each mutable object in each
+render thread. It is required for using time-multiplex compounds in
+multithreaded rendering.
 

### 2017-04-03 paperIEEE.tex
Compound figure
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@16 db7c848b-7755-4df2-b763-8796797afa52

@@
 %------------------------------------------------------------------------------
+\section{Performance}
+%------------------------------------------------------------------------------
+
+This section introduces new decomposition modes, runtime adaptations and other
+optimizations to increase the rendering performance for a wide variety of
+algorithms.
+
+\subsection{New Decomposition Modes}
+
+The initial version of Equalizer implemented sort-first (2D), sort-last (DB),
+stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we present new
+decomposition modes and motivate their use case. \fig{fig:compounds} provides an
+overview of the new modes.
+
+\begin{figure*}[ht]\center
+  \subfigure[Time-Multiplex]{\label{fig:dplex}
+    \includegraphics[width=0.24\textwidth]{images/dplex}}\hfil
+  \subfigure[Tiles]{\label{fig:tile}
+    \includegraphics[width=0.24\textwidth]{images/tile}}\hfil
+  \subfigure[Pixel]{\label{fig:pixel}
+    \includegraphics[width=0.24\textwidth]{images/pixel}}\hfil
+  \subfigure[Subpixel]{\label{fig:subpixel}
+    \includegraphics[width=0.24\textwidth]{images/subpixel}}\hfil
+  \caption{New Equalizer task decompositions for parallel rendering}
+  \label{fig:compounds}
+\end{figure*}
+
+\subsubsection{Time-Multiplex}
+
+Time-multiplexing (\fig{fig:dplex}), or DPlex, was implemented in \cite{BRE:05}
+for shared memory machines. It is however a better fit for distributed memory
+systems, since the separate memory space makes concurrent rendering of different
+frames easier to implement. While it increases the framerate linearly, it does
+not decrease the latency between user input and the corresponding
+output. Consequently, this decomposition mode is mostly useful for
+non-interactive movie generation. It is transparent to Equalizer applications,
+but does require the configuration latency to be equal or greater than the
+number of source channels. Furthermore, to work in a multi-threaded, multi-GPU
+configuration, the application needs to support running the rendering threads
+asynchronously, as outlined in \sref{SEC_threading}. The output frame rate of
+the destination channel is smoothened using a frame rate equalizer
+(\sref{SEC_framerateEq}).
+
+\subsubsection{Tiles and Chunks}\label{sec:tile}
+
+Tile (\fig{fig:tile}) and chunk decompositions are a variant of sort-first and
+sort-last rendering, respectively. They decompose the scene into a predefined
+set of fixed-size image tiles or database ranges. These tasks are queued and
+processed by all source channels by polling a server-central queue. Prefetching
+ensures that the task communication overlaps with rendering. As shown in
+\cite{SPEP:16}, these modes provide better performance due to being inherently
+load-balanced, as long as there is an insignificant overhead for the render task
+setup. This mode is transparent to Equalizer applications.
+
+\subsubsection{Pixel}
+
+Pixel compounds (\fig{fig:pixel}) decompose the destination channel by
+interleaving rows or columns in image space. They are a variant of sort-first
+decomposition which works well for fill-limited applications which are not
+geometry bound. Source channels cannot reduce geometry load through view frustum
+culling, since each source channel has almost the same frustum (only shifted by
+some pixels), but applied to a reduced 2D viewport. However, the fragment load
+on all source channels is very similar due to the interleaved distribution of
+pixels. This functionality is transparent to Equalizer applications, and the
+default compositing implementation uses the OpenGL stencil buffer to blit pixels
+onto the destination channel.
+
+\subsubsection{Subpixel}
+
+Subpixel compounds (\fig{fig:subpixel}) are similar to pixel compounds, but they
+decompose the work for a single pixel, for example when using multisampling or
+depth of field. Composition typically uses accumulation and averaging of all
+computed fragments for a pixel. This feature is not fully transparent to the
+application, since it needs to adapt (jitter or tilt) the frustum based on the
+iteration executed. Furthermore, subpixel compounds interact with idle image
+refinements, e.g., they can accelerate idle anti-aliasing of a scene when the
+camera and scene are not changed.
+
+\subsection{Equalizers}
+
+Equalizer are an addition to compound trees. They modify parameters of their
+respective subtree at runtime to optimize one aspect of the decomposition. Due
+to their nature, they are transparent to application developers, but might have
+application-accessible parameters to tune their behaviour.
+\subsubsection{Sort-First and Sort-Last Load Equalizer}
+
+Sort-first and sort-last load balancing is the most obvious optimization for
+these parallel rendering modes. Our load equalizer is fully transparent for
+application developers, that is, it uses a reactive approach based on past
+rendering times. This assumes a reasonable frame-to-frame coherency. Our
+implementation stores a 2D or 1D grid of the load, mapping the load of each
+channel. The load is stored in normalized 2D/1D coordinates using
+$\frac{time}{area}$ as the load, the contributing source channels are organized
+in a binary tree, and then the algorithm balances the two branches of each level
+by equalizing the integral over the area on each side.
+
+We have implemented various tunable parameters allowing application developers
+to optimize the load balancing based on the characteristics of their rendering
+algorithm:
+\begin{compactdesc}
+\item[Damping] reduces frame-to-frame oscillations. It is a normalized scalar
+  defining how much of the computed delta from the previous position is
+  applied. The equal load distribution within the region of interest assumed by
+  the load equalizer is in reality not equal, causing the load balancing to
+  overshoot.
+\item[Resistance] eliminates small deltas in the load balancing step. This might
+  help the application to cache some computations since the frustum does not
+  change each frame.
+\item[Boundaries] define the modulo factor in pixels onto which a load split may
+  fall. Some rendering algorithms produce artefacts related to the OpenGL raster
+  position, e.g., screen door transparency, which can be eliminated by aligning
+  the boundary to the pixel repetition. Furthermore, some rendering algorithms
+  are sensitive to cache alignments, which can again be exploited by chosing the
+  corresponding boundary.
+\end{compactdesc}
+
+\subsubsection{Cross-Segment Load Balancing}
+
+Cross-segment load balancing addresses the optimal resource allocation of $n$
+rendering resources to $m$ output channels (with $n\geq m$). The view equalizer
+works in conjunction with load equalizer balancing the individual output
+channels. It monitors the usage of shared source channels (across outputs) and
+activates them to balance the rendering time of all outputs. In \cite{EEP:11},
+we provide a detailed description and evaluation of our algorithm.
+
+\subsubsection{Dynamic Frame Resolution}
+
+The DFR equalizer provides a functionality similar to dynamic video resizing
+\cite{MBDM:97}, that is, it maintains a constant framerate by adapting the
+rendering resolution of a fill-limited application. In Equalizer, this works by
+rendering into a source channel (typically on a FBO) separate to the destination
+channel, and then scaling the rendering during the transfer (typically through
+an on-GPU texture) to the destination channel. The DFR equalizer monitors the
+rendering performance and accordingly adapts the resolution of the source
+channel and zoom factor for the source to destination transfer. If the
+performance and source channel resolutions allows, this will not only subsample,
+but also supersample the destination channel to reduce aliasing artefacts.
+
+\subsubsection{Frame Rate Equalizer}\label{SEC_framerateEq}
+
+The framerate equalizer smoothens the output frame rate of a destination
+channel by instructing the corresponding window to delay its buffer swap to a
+minimum time between swaps. This is regularly used for time-multiplexed
+decompostions, where source channels tend to drift and finish their rendering
+not evenly distributed over time. This equalizer is however fully independent of
+DPlex compounds, and may be used to smoothen irregular application rendering
+algorithms.
+
+\subsubsection{Monitoring}
+
+The monitor allows to reuse the rendering on another channel, typically for
+monitoring a larger setup on a control workstation. Output frames on the display
+channels are connected to input frames on a single monitoring channel. The
+monitor equalizer changes the scaling factor and offset between the output and
+input, so that the monitor channel has the same, but typically downscaled view,
+as the originating segments.
+
+\subsection{Optimizations}
+
+\subsubsection{Region of Interest}
+
+The region of interest is the screen-space 2D bounding box enclosing the
+geometry rendered by a single resource. We have extended the core parallel
+rendering framework to use an application-provided ROI to optimize the load
+equalizer as well as image compositing performance. The load equalizer uses the
+ROI to refine its load grid to the regions containing data. The compositing code
+uses the ROI to minimize image readback and network transmission. In
+\cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of the algorithm, and
+show that using ROI can quadruple the rendering performance, in particular for
+the costly compositing step in sort-last rendering.
+
+\subsubsection{Asynchronous Compositing}
+
+Asynchronous compositing pipelines rendering with compositing operations, by
+executing the image readback, network transfer and image assembly from threads
+running in parallel to the rendering threads. In \cite{EBAHMP:12}, we provide
+the details of the implementation and experimental data showing an improvement
+of the rendering performance of over 25\% for large node counts.
+
+\subsubsection{Download and Compression Plugins}
+
+Compression for the compositing step in parallel rendering is critical for
+performance. This not only applies to the well-researched network transfer step,
+but also for the transfer between GPU and CPU. Equalizer supports a variety of
+compression algorithms, from very fast RLE encoding, JPEG compression to YUV
+subsampling on the GPU. These algorithms are implemented as runtime-loaded
+plugins, allowing easy extension and customization to application-specific
+compression. In \cite{MEP:10}, we show this to be a critical step for
+interactive performance at scale.
+
+\subsubsection{Thread Synchronization Modes}\label{SEC_threading}
+
+Different applications have different degrees on how decoupled and thread-safe
+the rendering code is from the application logic. For full decoupling all
+mutable data has to have a copy in each render thread, which is not feasible in
+large data scenarios. To easily customize the synchronization of all threads on
+a single process, Equalizer implements three threading modes:  Full
+synchronization, draw synchronization and asynchronous. Note that the execution
+between nodes is always asynchronous, for up to \textsf{latency} frames.
+
+In full synchronization, all threads always execute the same frame, that is, the
+render threads are unlocked after \textsf{Node::frameStart}, and the node is
+blocked for all render threads to finish the frame before executing
+\textsf{Node::frameFinish}. This allows the render threads to read shared data
+from all their operations, but provides the slowest performance.
+
+In draw synchronization, the node thread and all render threads are synchronized
+for all \textsf{frameDraw} operations, that is, \textsf{Node::frameFinish} is
+executed after the last channel is done drawing. This allows the render threads
+to read shared data during their draw operation, but not during
+compositing. Since compositing is typically independent of the rendered data,
+this is the default mode. This mode allows to overlap compositing with data
+synchronization on multi-GPU machines.
+
+In asynchronous execution, all threads run asynchronously. Render threads may
+work on different frames at any given time. This mode is the fastest, and
+requires the application to have one instance of each mutable object in each
+render thread. It is required for using time-multiplex compounds in
+multithreaded rendering.
+
+%------------------------------------------------------------------------------
+\section{Virtual Reality}
+%------------------------------------------------------------------------------
+
+Virtual Reality is an important field for parallel rendering. It does however
+require special attention to support it as a first-class citizen in a generic
+parallel rendering framework. Equalizer has been used in many virtual reality
+installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
+at the KAUST visualization laboratory and head-mounted displays
+(\fig{FIG_teaser}). In the following we lay out the features needed support
+these installations. All features presented were motivated by application use
+cases and have been validated with their respective users.
+
+\subsection{Head Tracking}
+
+Head tracking is the minimal feature needed to support immersive
+installations. Equalizer does support multiple, independent tracked views
+through the observer abstraction introduced in \sref{SEC_observer}. Built-in
+VRPN support enables the direct, application-transparent configuration of a VRPN
+tracker device. Alternatively, an application can provide a $4\times 4$ tracking
+matrix defining the transformation from the canvas coordinate system to the
+observer. Both CAVE-like tracking with fixed projection surfaces and HMD
+tracking modes are implemented.
+
+\subsection{Dynamic Focus Distance}
+
+To our knowledge, all parallel rendering systems have the focal plane coincide
+with the physical display surface. For better viewing comfort, we introduce a
+new dynamic focus mode, where the application defines the distance of the focal
+plane from the observer, based on the current view direction af the
+user. Initial experiments show that this is particularly effective for objects
+placed within the immersive space, that is, in front of all display segments.
+
+\subsection{Asymmetric Eye Position}
+
+Traditional head tracking computes the left and right eye positions by using a
+configurable interocular distance and the tracking matrix. However, human heads
+are not symmetric, and by measuring individual users a more precise frustum can
+be computed. Equalizer supports this through the optional configuration (file or
+programmatically) of individual 3D positions for the left and right eye.
+
+\subsection{Model Unit}
+
+This feature allows applications to specify a scaling factor between the model
+and the real world, to allow exploration of macroscopic or microscopic worlds in
+virtual reality. The unit is per view, allowing different scale factors within
+the same configuration. It scales both the specified projection surface as well
+as the eye position (and therefore separation) to achieve the necessary effect.
+
+\subsection{Runtime Stereo Switch}
+
+Applications can switch each view between mono and stereo rendering at runtime,
+and can run both monoscopic and stereoscopic views concurrently
+(\fig{FIG_teaser} (b)). While this is trivial for active stereo rendering, it
+does potentially involve the start and stop of resources and processes for
+passive stereo or stereo-dependent task decompositions, as described in
+\sref{SEC_reconfig}.
+
+%------------------------------------------------------------------------------
 \section{Usability}
@@
 for each (pipe) render thread. They also provide the natural place to store and
-distribute data. Last, but not least, view data provides a convenient way to
-manage multiple views by storing the camera, model or any other view-specific
-information.
+distribute data. Last, but not least, \textsf{ViewData} provides a convenient
+way to manage multiple views by storing the camera, model or any other
+view-specific information.
 
@@
 
-%------------------------------------------------------------------------------
-\section{Virtual Reality}
-%------------------------------------------------------------------------------
-
-Virtual Reality is an important field for parallel rendering. It does however
-require special attention to support it as a first-class citizen in a generic
-parallel rendering framework. Equalizer has been used in many virtual reality
-installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
-at the KAUST visualization laboratory and head-mounted displays
-(\fig{FIG_teaser}). In the following we lay out the features needed support
-these installations. All features presented were motivated by application use
-cases and have been validated with their respective users.
-
-\subsection{Head Tracking}
-
-Head tracking is the minimal feature needed to support immersive
-installations. Equalizer does support multiple, independent tracked views
-through the observer abstraction introduced in \sref{SEC_observer}. Built-in
-VRPN support enables the direct, application-transparent configuration of a VRPN
-tracker device. Alternatively, an application can provide a $4\times 4$ tracking
-matrix defining the transformation from the canvas coordinate system to the
-observer. Both CAVE-like tracking with fixed projection surfaces and HMD
-tracking modes are implemented.
-
-\subsection{Dynamic Focus Distance}
-
-To our knowledge, all parallel rendering systems have the focal plane coincide
-with the physical display surface. For better viewing comfort, we introduce a
-new dynamic focus mode, where the application defines the distance of the focal
-plane from the observer, based on the current view direction af the
-user. Initial experiments show that this is particularly effective for objects
-placed within the immersive space, that is, in front of all display segments.
-
-\subsection{Asymmetric Eye Position}
-
-Traditional head tracking computes the left and right eye positions by using a
-configurable interocular distance and the tracking matrix. However, human heads
-are not symmetric, and by measuring individual users a more precise frustum can
-be computed. Equalizer supports this through the optional configuration (file or
-programmatically) of individual 3D positions for the left and right eye.
-
-\subsection{Model Unit}
-
-This feature allows applications to specify a scaling factor between the model
-and the real world, to allow exploration of macroscopic or microscopic worlds in
-virtual reality. The unit is per view, allowing different scale factors within
-the same configuration. It scales both the specified projection surface as well
-as the eye position (and therefore separation) to achieve the necessary effect.
-
-\subsection{Runtime Stereo Switch}
-
-Applications can switch each view between mono and stereo rendering at runtime,
-and can run both monoscopic and stereoscopic views concurrently
-(\fig{FIG_teaser} (b)). While this is trivial for active stereo rendering, it
-does potentially involve the start and stop of resources and processes for
-passive stereo or stereo-dependent task decompositions, as described in
-\sref{SEC_reconfig}.
-
-%------------------------------------------------------------------------------
-\section{Performance}
-%------------------------------------------------------------------------------
-
-\subsection{New Decomposition Modes}
-
-The initial version of Equalizer implemented sort-first (2D), sort-last (DB) and
-stereo (EYE) decomposition \cite{EMP:09}. In the following we present new
-decomposition modes and motivate their use case.
-
-\subsubsection{Time-Multiplex}
-
-Time-multiplexing, or DPlex, was already implemented in \cite{BRE:05}. While it
-increases the framerate linearly, it does not decrease the latency between user
-input and the corresponding output. Consequently, this decomposition mode is
-mostly useful for non-interactive movie generation. It is transparent to
-Equalizer applications, but does require the configuration latency to be equal
-or greater than the number of source channels. Furthermore, to work in a
-multi-threaded, multi-GPU configuration, the application needs to support
-running the rendering threads asynchronously, as outlined in
-\sref{SEC_threading}. The output frame rate of the destination channel is
-smoothened using a frame rate equalizer (\sref{SEC_framerateEq}).
-
-\subsubsection{Tiles and Chunks}\label{sec:tile}
-
-Tile and chunk decompositions are a variant of sort-first and sort-last
-rendering, respectively. They decompose the scene into a predefined set of
-fixed-size image tiles or database ranges. These tasks are queued and processed
-by all source channels by polling a server-central queue. Prefetching ensures
-that the task communication overlaps with rendering. As shown in \cite{SPEP:16},
-these modes provide better performance due to being inherently load-balanced, as
-long as there is an insignificant overhead for the render task setup. This mode
-is transparent to Equalizer applications.
-
-\subsubsection{Pixel}
-
-Pixel compounds decompose the destination channel by interleaving rows or
-columns in image space. They are a variant of sort-first decomposition which
-works well for fill-limited applications which are not geometry bound. Source
-channels cannot reduce geometry load through view frustum culling, since each
-source channel has almost the same frustum (only shifted by some pixels),
-but applied to a reduced 2D viewport. However, the fragment load on all source
-channels is very similar due to the interleaved distribution of pixels. This
-functionality is transparent to Equalizer applications, and the default
-compositing implementation uses the OpenGL stencil buffer to blit pixels onto
-the destination channel.
-
-\subsubsection{Subpixel}
-
-Subpixel compounds are similar to pixel compounds, but they decompose the work
-for a single pixel, for example when using multisampling or depth of
-field. Composition typically uses accumulation and averaging of all computed
-fragments for a pixel. This feature is not fully transparent to the application,
-since it needs to adapt (jitter or tilt) the frustum based on the iteration
-executed. Furthermore, subpixel compounds interact with idle image refinements,
-e.g., they can accelerate idle anti-aliasing of a scene when the camera and
-scene are not changed.
-
-\subsection{Equalizers}
-
-Equalizer are an addition to compound trees. They modify parameters of their
-respective subtree at runtime to optimize one aspect of the decomposition. Due
-to their nature, they are transparent to application developers, but might have
-application-accessible parameters to tune their behaviour.
-
-\subsubsection{Sort-First and Sort-Last Load Equalizer}
-
-Sort-first and sort-last load balancing is the most obvious optimization for
-these parallel rendering modes. Our load equalizer is fully transparent for
-application developers, that is, it uses a reactive approach based on past
-rendering times. This assumes a reasonable frame-to-frame coherency. Our
-implementation stores a 2D or 1D grid of the load, mapping the load of each
-channel. The load is stored in normalized 2D/1D coordinates using
-$\frac{time}{area}$ as the load, the contributing source channels are organized
-in a binary tree, and then the algorithm balances the two branches of each level
-by equalizing the integral over the area on each side.
-
-We have implemented various tunable parameters allowing application developers
-to optimize the load balancing based on the characteristics of their rendering
-algorithm:
-\begin{compactdesc}
-\item[Damping] reduces frame-to-frame oscillations. It is a normalized scalar
-  defining how much of the computed delta from the previous position is
-  applied. The equal load distribution within the region of interest assumed by
-  the load equalizer is in reality not equal, causing the load balancing to
-  overshoot.
-\item[Resistance] eliminates small deltas in the load balancing step. This might
-  help the application to cache some computations since the frustum does not
-  change each frame.
-\item[Boundaries] define the modulo factor in pixels onto which a load split may
-  fall. Some rendering algorithms produce artefacts related to the OpenGL raster
-  position, e.g., screen door transparency, which can be eliminated by aligning
-  the boundary to the pixel repetition. Furthermore, some rendering algorithms
-  are sensitive to cache alignments, which can again be exploited by chosing the
-  corresponding boundary.
-\end{compactdesc}
-
-\subsubsection{Cross-Segment Load Balancing}
-
-Cross-segment load balancing addresses the optimal resource allocation of $n$
-rendering resources to $m$ output channels (with $n\geq m$). The view equalizer
-works in conjunction with load equalizer balancing the individual output
-channels. It monitors the usage of shared source channels (across outputs) and
-activates them to balance the rendering time of all outputs. In \cite{EEP:11},
-we provide a detailed description and evaluation of our algorithm.
-
-\subsubsection{Dynamic Frame Resolution}
-
-The DFR equalizer provides a functionality similar to dynamic video resizing
-\cite{MBDM:97}, that is, it maintains a constant framerate by adapting the
-rendering resolution of a fill-limited application. In Equalizer, this works by
-rendering into a source channel (typically on a FBO) separate to the destination
-channel, and then scaling the rendering during the transfer (typically through
-an on-GPU texture) to the destination channel. The DFR equalizer monitors the
-rendering performance and accordingly adapts the resolution of the source
-channel and zoom factor for the source to destination transfer. If the
-performance and source channel resolutions allows, this will not only subsample,
-but also supersample the destination channel to reduce aliasing artefacts.
-
-\subsubsection{Frame Rate Equalizer}\label{SEC_framerateEq}
-
-The framerate equalizer smoothens the output frame rate of a destination
-channel by instructing the corresponding window to delay its buffer swap to a
-minimum time between swaps. This is regularly used for time-multiplexed
-decompostions, where source channels tend to drift and finish their rendering
-not evenly distributed over time. This equalizer is however fully independent of
-DPlex compounds, and may be used to smoothen irregular application rendering
-algorithms.
-
-\subsubsection{Monitoring}
-
-The monitor allows to reuse the rendering on another channel, typically for
-monitoring a larger setup on a control workstation. Output frames on the display
-channels are connected to input frames on a single monitoring channel. The
-monitor equalizer changes the scaling factor and offset between the output and
-input, so that the monitor channel has the same, but typically downscaled view,
-as the originating segments.
-
-\subsection{Optimizations}
-
-\subsubsection{Region of Interest}
-
-The region of interest is the screen-space 2D bounding box enclosing the
-geometry rendered by a single resource. We have extended the core parallel
-rendering framework to use an application-provided ROI to optimize the load
-equalizer as well as image compositing performance. The load equalizer uses the
-ROI to refine its load grid to the regions containing data. The compositing code
-uses the ROI to minimize image readback and network transmission. In
-\cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of the algorithm, and
-show that using ROI can quadruple the rendering performance, in particular for
-the costly compositing step in sort-last rendering.
-
-\subsubsection{Asynchronous Compositing}
-
-Asynchronous compositing pipelines rendering with compositing operations, by
-executing the image readback, network transfer and image assembly from threads
-running in parallel to the rendering threads. In \cite{EBAHMP:12}, we provide
-the details of the implementation and experimental data showing an improvement
-of the rendering performance of over 25\% for large node counts.
-
-\subsubsection{Download and Compression Plugins}
-
-Compression for the compositing step in parallel rendering is critical for
-performance. This not only applies to the well-researched network transfer step,
-but also for the transfer between GPU and CPU. Equalizer supports a variety of
-compression algorithms, from very fast RLE encoding, JPEG compression to YUV
-subsampling on the GPU. These algorithms are implemented as runtime-loaded
-plugins, allowing easy extension and customization to application-specific
-compression. In \cite{MEP:10}, we show this to be a critical step for
-interactive performance at scale.
-
-\subsubsection{Thread Synchronization Modes}\label{SEC_threading}
-
-Different applications have different degrees on how decoupled and thread-safe
-the rendering code is from the application logic. For full decoupling all
-mutable data has to have a copy in each render thread, which is not feasible in
-large data scenarios. To easily customize the synchronization of all threads on
-a single process, Equalizer implements three threading modes:  Full
-synchronization, draw synchronization and asynchronous. Note that the execution
-between nodes is always asynchronous, for up to \textsf{latency} frames.
-
-In full synchronization, all threads always execute the same frame, that is, the
-render threads are unlocked after \textsf{Node::frameStart}, and the node is
-blocked for all render threads to finish the frame before executing
-\textsf{Node::frameFinish}. This allows the render threads to read shared data
-from all their operations, but provides the slowest performance.
-
-In draw synchronization, the node thread and all render threads are synchronized
-for all \textsf{frameDraw} operations, that is, \textsf{Node::frameFinish} is
-executed after the last channel is done drawing. This allows the render threads
-to read shared data during their draw operation, but not during
-compositing. Since compositing is typically independent of the rendered data,
-this is the default mode. This mode allows to overlap compositing with data
-synchronization on multi-GPU machines.
-
-In asynchronous execution, all threads run asynchronously. Render threads may
-work on different frames at any given time. This mode is the fastest, and
-requires the application to have one instance of each mutable object in each
-render thread. It is required for using time-multiplex compounds in
-multithreaded rendering.
-
 %----------------------------------------------------------------------

### 2017-04-03 paperIEEE.tex
Equalizers figure
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@17 db7c848b-7755-4df2-b763-8796797afa52

@@
 application-accessible parameters to tune their behaviour.
+
+\begin{figure*}[ht]\center
+  \subfigure[Load-Balancing]{\label{fig:loadeq}
+    \includegraphics[width=0.24\textwidth]{images/loadeq}}\hfil
+  \subfigure[Cross-Segment Load-Balancing]{\label{fig:vieweq}
+    \includegraphics[width=0.24\textwidth]{images/vieweq}}\hfil
+  \subfigure[Dynamic Frame Resolution]{\label{fig:dfr}
+    \includegraphics[width=0.24\textwidth]{images/dfr}}\hfil
+  \subfigure[Monitoring]{\label{fig:monitor}
+    \includegraphics[width=0.24\textwidth]{images/monitoreq}}\hfil
+  \caption{Runtime modifications}
+  \label{fig:equalizers}
+\end{figure*}
+
 \subsubsection{Sort-First and Sort-Last Load Equalizer}
 
-Sort-first and sort-last load balancing is the most obvious optimization for
-these parallel rendering modes. Our load equalizer is fully transparent for
-application developers, that is, it uses a reactive approach based on past
-rendering times. This assumes a reasonable frame-to-frame coherency. Our
-implementation stores a 2D or 1D grid of the load, mapping the load of each
-channel. The load is stored in normalized 2D/1D coordinates using
+Sort-first (\fig{fig:loadeq}) and sort-last load balancing is the most obvious
+optimization for these parallel rendering modes. Our load equalizer is fully
+transparent for application developers, that is, it uses a reactive approach
+based on past rendering times. This assumes a reasonable frame-to-frame
+coherency. Our implementation stores a 2D or 1D grid of the load, mapping the
+load of each channel. The load is stored in normalized 2D/1D coordinates using
 $\frac{time}{area}$ as the load, the contributing source channels are organized
@@
 
-Cross-segment load balancing addresses the optimal resource allocation of $n$
-rendering resources to $m$ output channels (with $n\geq m$). The view equalizer
-works in conjunction with load equalizer balancing the individual output
-channels. It monitors the usage of shared source channels (across outputs) and
-activates them to balance the rendering time of all outputs. In \cite{EEP:11},
-we provide a detailed description and evaluation of our algorithm.
+Cross-segment load balancing (\fig{fig:vieweq}) addresses the optimal resource
+allocation of $n$ rendering resources to $m$ output channels (with $n\geq
+m$). The view equalizer works in conjunction with load equalizer balancing the
+individual output channels. It monitors the usage of shared source channels
+(across outputs) and activates them to balance the rendering time of all
+outputs. In \cite{EEP:11}, we provide a detailed description and evaluation of
+our algorithm.
 
@@
 
-The DFR equalizer provides a functionality similar to dynamic video resizing
-\cite{MBDM:97}, that is, it maintains a constant framerate by adapting the
-rendering resolution of a fill-limited application. In Equalizer, this works by
-rendering into a source channel (typically on a FBO) separate to the destination
-channel, and then scaling the rendering during the transfer (typically through
-an on-GPU texture) to the destination channel. The DFR equalizer monitors the
-rendering performance and accordingly adapts the resolution of the source
-channel and zoom factor for the source to destination transfer. If the
-performance and source channel resolutions allows, this will not only subsample,
-but also supersample the destination channel to reduce aliasing artefacts.
+The DFR equalizer (\fig{fig:dfr}) provides a functionality similar to dynamic
+video resizing \cite{MBDM:97}, that is, it maintains a constant framerate by
+adapting the rendering resolution of a fill-limited application. In Equalizer,
+this works by rendering into a source channel (typically on a FBO) separate to
+the destination channel, and then scaling the rendering during the transfer
+(typically through an on-GPU texture) to the destination channel. The DFR
+equalizer monitors the rendering performance and accordingly adapts the
+resolution of the source channel and zoom factor for the source to destination
+transfer. If the performance and source channel resolutions allows, this will
+not only subsample, but also supersample the destination channel to reduce
+aliasing artefacts.
 
@@
 
-The monitor allows to reuse the rendering on another channel, typically for
-monitoring a larger setup on a control workstation. Output frames on the display
-channels are connected to input frames on a single monitoring channel. The
-monitor equalizer changes the scaling factor and offset between the output and
-input, so that the monitor channel has the same, but typically downscaled view,
-as the originating segments.
+The monitor equalizer (\fig{fig:monitor}) allows to reuse the rendering on
+another channel, typically for monitoring a larger setup on a control
+workstation. Output frames on the display channels are connected to input frames
+on a single monitoring channel. The monitor equalizer changes the scaling factor
+and offset between the output and input, so that the monitor channel has the
+same, but typically downscaled view, as the originating segments.
 

### 2017-04-03 paperIEEE.tex
Some related work
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@18 db7c848b-7755-4df2-b763-8796797afa52

@@
 %----------------------------------------------------------------------
-\section{Related Work}
+\section{Related Work}\label{SEC_related}
 %----------------------------------------------------------------------
-\label{SEC_related}
-
-In \cite{EMP:09}, we presented the Equalizer parallel rendering framework. This
-paper also summarized the work in parallel rendering up to 2009. In the
-following we will present the related work published since then. An extensive
-Programming and User Guide provides in-depth documentation on using and
-programming Equalizer\cite{Eilemann:13}.
-
-\cite{EEP:11} cross-segment load balancing
+
+In~\cite{EMP:09}, we presented the Equalizer parallel rendering framework. This
+paper summarized the work in parallel rendering up to 2009. In the following we
+will present the related work published since then. An extensive Programming and
+User Guide provides in-depth documentation on using and programming
+Equalizer~\cite{Eilemann:13}.
+
+The concept of transparent OpenGL interception popularized by WireGL and
+Chromium~\cite{HHNFAKK:02} has received little attention. While some commercial
+implementations such as TechViz and MechDyne Conduit continue to exists, only
+ClusterGL?~\cite{NHM:11} has been published. ClusterGL employs the same approach
+as Chromium, but delivers a significantly faster implementation of transparent
+OpenGL interception and distribution for parallel rendering.
+
+On the other hand, software for driving and interacting with tiled display walls
+has received significant attention. Sage~\cite{TBD} and Sage2~\cite{TBD}
+
+Equalizer itself has received significant attention to research and validate
+parallel rendering performance, by using compression and region of interest
+during compositing~\cite{MEP:10}, load-balancing resources for multi-display
+installations~\cite{EEP:11}, asynchronous compositing and NUMA
+optimizations~\cite{EBAHMP:12} as well as work queueing~\cite{SPEP:16}.
+
+Various applications and frameworks have used Equalizer for new research in
+visualization, such as the RTNeuron~\cite{HBBES:13}, an OpenGL renderer for
+visualizing simulations of neucortical brain regions, Omegalib~\cite{TBD},
+extending Equalizer for the hybrid 2D/3D environments such as the CAVE2,
+Siegen~\cite{TBD}
 
@@
 \cite{TBD} DisplayCluster
-
-\cite{NHM:11} ClusterGL?
 
@@
 screens. These frusta describe a physically-correct display setup for a Virtual
-Reality installation. A canvas may have a software or hardware swap barrier,
-which will synchronize the rendering of all contributing GPUs. The software
-barrier executes a \textsf{glFinish} to ensure the GPU is ready to swap, a
-Collage barrier (\sref{SEC_barrier}) to synchronize all segments, the swap
-buffers call followed by a \textsf{glFlush} to ensure timely execution of the
-swap command. The hardware swap barrier is implemented using the
+Reality installation. \label{sec:swap} A canvas may have a software or hardware
+swap barrier, which will synchronize the rendering of all contributing GPUs. The
+software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
+swap, a Collage barrier (\sref{SEC_barrier}) to synchronize all segments, the
+swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
+the swap command. The hardware swap barrier is implemented using the
 \textsf{WGL\_NV\_swap\_group} or \textsf{GLX\_NV\_swap\_group} extensions.

### 2017-04-03 paperIEEE.tex
References and related work
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@19 db7c848b-7755-4df2-b763-8796797afa52

@@
 The concept of transparent OpenGL interception popularized by WireGL and
-Chromium~\cite{HHNFAKK:02} has received little attention. While some commercial
-implementations such as TechViz and MechDyne Conduit continue to exists, only
-ClusterGL?~\cite{NHM:11} has been published. ClusterGL employs the same approach
-as Chromium, but delivers a significantly faster implementation of transparent
-OpenGL interception and distribution for parallel rendering.
+Chromium~\cite{HHNFAKK:02} has received little attention since 2009. While some
+commercial implementations such as TechViz and MechDyne Conduit continue to
+exist, on the research side only ClusterGL~\cite{NHM:11} has been
+presented. ClusterGL employs the same approach as Chromium, but delivers a
+significantly faster implementation of transparent OpenGL interception and
+distribution for parallel rendering. CGLX~\cite{CGLX:11} tries to bring parallel
+execution transparently to OpenGL applications, by emulating the GLUT API and
+intercepting certain OpenGL calls. This approach works transparently for trivial
+applications, but quickly requires the application developer to address the
+complexities of a distributed application when mutable application state needs
+to be synchronized across processes.
 
 On the other hand, software for driving and interacting with tiled display walls
-has received significant attention. Sage~\cite{TBD} and Sage2~\cite{TBD}
+has received significant attention. Sage~\cite{Sage} and Sage2~\cite{Sage2} have
+received significant attention. Sage is build entirely around the concept of a
+shared framebuffer where all content windows are separate applications using
+pixel streaming. It is no longer actively supported. Sage 2 is a complete,
+browser-centric reimplementation where each application is a web application
+distributed across browser instances. DisplayCluster~\cite{DC}, and its
+continuation Tide, also implement the shared framebuffer concept of Sage, but
+provide a few native content applications integrated into the display servers.
 
@@
 visualization, such as the RTNeuron~\cite{HBBES:13}, an OpenGL renderer for
-visualizing simulations of neucortical brain regions, Omegalib~\cite{TBD},
+visualizing simulations of neucortical brain regions, Omegalib~\cite{Omegalib},
 extending Equalizer for the hybrid 2D/3D environments such as the CAVE2,
-Siegen~\cite{TBD}
+visualization for remote sensing date~\cite{LK:09}, .
 
@@
 \cite{CKP:12} tiled display wall virtual environment
-
-\cite{TBD} DisplayCluster
-
-Omegalib
 
@@
 
-Tiled interactive display environment, parallel pixel streaming, events
+Tide (Tiled interactive display environment) is an improved version of
+DisplayCluster~\cite{DC}, providing touch-based, multi-window user interface for
+high-resolution tiled display walls. Remote applications receive input events
+and send pixel streams using the Deflect client library. Equalizer includes full
+support, enabling application-transparent integration with Tide. When a Tide
+server is configured, all output channels of a view stream in parallel to one
+window on the wall. In \cite{deflect} we have shown interactive framerates for
+an 24 megapixel resolution over a WAN link. Events received are translated into
+Equalizer events and injected into the normal event flow, allowing seamless
+application integration.
 
@@
 
+In this section, we present some major applications built using Equalizer, and
+show how they interact with the framework to solve complex parallel rendering
+problems.
+
 \subsection{Livre}
+
+Livre (Large-scale interactive volume rendering engine) is an out-of-core
+parallel 4D volume renderer.
+
 \subsection{RTT Deltagen}

### 2017-04-03 paperIEEE.tex
Missing CGLX ref
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@20 db7c848b-7755-4df2-b763-8796797afa52

@@
 significantly faster implementation of transparent OpenGL interception and
-distribution for parallel rendering. CGLX~\cite{CGLX:11} tries to bring parallel
+distribution for parallel rendering. CGLX~\cite{DK:11} tries to bring parallel
 execution transparently to OpenGL applications, by emulating the GLUT API and
@@
 visualization for remote sensing date~\cite{LK:09}, .
-
-\cite{DK:11}: Framework to develop apps for multitile, thus targeted
-specifically for tiled-wall visualization systems: grid configuration,
-scalability (?), multiple work sessions, multiuser events. Master-slave approach
-like Eq. with grid nodes running instances of the application. Is maybe more
-like a GLUT and window replacement, or enhancement to deal with the OpenGL
-contexts and events
 

### 2017-04-03 paperIEEE.tex
Compound configs
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@21 db7c848b-7755-4df2-b763-8796797afa52

@@
 %\usepackage{url}
-\usepackage{subfigure}
+\usepackage{caption,subcaption}
 \usepackage{multirow}
 \usepackage{paralist}
+\usepackage{listings}
 
@@
 The initial version of Equalizer implemented sort-first (2D), sort-last (DB),
-stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we present new
-decomposition modes and motivate their use case. \fig{fig:compounds} provides an
-overview of the new modes.
+stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
+present new decomposition modes and motivate their use case. \fig{fig:compounds}
+provides an overview of the new modes. We describe their configuration in
+same \textit{compound} framework introduced in \cite{EMP:09}
 
 \begin{figure*}[ht]\center
-  \subfigure[Time-Multiplex]{\label{fig:dplex}
-    \includegraphics[width=0.24\textwidth]{images/dplex}}\hfil
-  \subfigure[Tiles]{\label{fig:tile}
-    \includegraphics[width=0.24\textwidth]{images/tile}}\hfil
-  \subfigure[Pixel]{\label{fig:pixel}
-    \includegraphics[width=0.24\textwidth]{images/pixel}}\hfil
-  \subfigure[Subpixel]{\label{fig:subpixel}
-    \includegraphics[width=0.24\textwidth]{images/subpixel}}\hfil
+  \begin{subfigure}[b]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{images/dplex}
+      \caption{\label{fig:dplex}Time-Multiplex}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{images/tile}
+      \caption{\label{fig:tile}Tiles}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{images/pixel}
+      \caption{\label{fig:pixel}Pixel}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{images/subpixel}
+      \caption{\label{fig:subpixel}Subpixel}
+  \end{subfigure}
   \caption{New Equalizer task decompositions for parallel rendering}
@@
 
-Time-multiplexing (\fig{fig:dplex}), or DPlex, was implemented in \cite{BRE:05}
-for shared memory machines. It is however a better fit for distributed memory
-systems, since the separate memory space makes concurrent rendering of different
-frames easier to implement. While it increases the framerate linearly, it does
-not decrease the latency between user input and the corresponding
-output. Consequently, this decomposition mode is mostly useful for
-non-interactive movie generation. It is transparent to Equalizer applications,
-but does require the configuration latency to be equal or greater than the
-number of source channels. Furthermore, to work in a multi-threaded, multi-GPU
-configuration, the application needs to support running the rendering threads
-asynchronously, as outlined in \sref{SEC_threading}. The output frame rate of
-the destination channel is smoothened using a frame rate equalizer
+Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
+implemented in \cite{BRE:05} for shared memory machines. It is however a better
+fit for distributed memory systems, since the separate memory space makes
+concurrent rendering of different frames easier to implement. While it increases
+the framerate linearly, it does not decrease the latency between user input and
+the corresponding output. Consequently, this decomposition mode is mostly useful
+for non-interactive movie generation. It is transparent to Equalizer
+applications, but does require the configuration latency to be equal or greater
+than the number of source channels. Furthermore, to work in a multi-threaded,
+multi-GPU configuration, the application needs to support running the rendering
+threads asynchronously, as outlined in \sref{SEC_threading}. The output frame
+rate of the destination channel is smoothened using a frame rate equalizer
 (\sref{SEC_framerateEq}).
+
+\begin{figure*}[ht]\center
+  \begin{subfigure}[b]{0.24\textwidth}
+    {\tiny\begin{lstlisting}
+compound {
+  channel "destination"
+  framerate_equalizer {}
+  compound {
+    channel "source1"
+    phase 0 period 2
+    outputframe  { name "frame" }
+  }
+  compound {
+    channel "source2"
+    phase 1 period 2
+    outputframe  { name "frame" }
+  }
+  inputframe { name "frame" }
+}
+    \end{lstlisting}}
+    \caption{\label{eqc:dplex}Time-Multiplex}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    {\tiny\begin{lstlisting}
+compound {
+  channel "destination"
+  outputtiles {
+    name "queue"
+    size [ 64 64 ]
+  }
+  compound {
+    channel "destination"
+    inputtiles { name "queue" }
+  }
+  compound {
+    channel "source"
+    inputtiles { name "queue" }
+    outputframe {}
+  }
+  inputframe { name "frame.source" }
+}
+    \end{lstlisting}}
+    \caption{\label{eqc:tiles}Tiles}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    {\tiny\begin{lstlisting}
+compound {
+  channel "dest"
+  compound {
+    channel "dest"
+    pixel [ 0 0 2 1 ]
+    outputframe { type texture }
+  }
+  compound {
+    channel "source"
+    pixel [ 1 0 2 1 ]
+    outputframe{}
+  }
+  inputframe { name "frame.dest" }
+  inputframe { name "frame.source" }
+}
+    \end{lstlisting}}
+    \caption{\label{eqc:pixel}Pixel}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    {\tiny\begin{lstlisting}
+compound {
+  channel "dest"
+  compound {
+    channel "dest"
+    subpixel [ 0 2 ]
+    outputframe { type texture }
+  }
+  compound {
+    channel "source"
+    subpixel [ 1 2 ]
+    outputframe{}
+  }
+  inputframe { name "frame.dest" }
+  inputframe { name "frame.source" }
+}
+    \end{lstlisting}}
+    \caption{\label{eqc:subpixel}Subpixel}
+  \end{subfigure}
+  \caption{Compound configurations for \fig{fig:compounds}}
+  \label{eqc:compounds}
+\end{figure*}
 
@@
 \begin{figure*}[ht]\center
-  \subfigure[Load-Balancing]{\label{fig:loadeq}
-    \includegraphics[width=0.24\textwidth]{images/loadeq}}\hfil
-  \subfigure[Cross-Segment Load-Balancing]{\label{fig:vieweq}
-    \includegraphics[width=0.24\textwidth]{images/vieweq}}\hfil
-  \subfigure[Dynamic Frame Resolution]{\label{fig:dfr}
-    \includegraphics[width=0.24\textwidth]{images/dfr}}\hfil
-  \subfigure[Monitoring]{\label{fig:monitor}
-    \includegraphics[width=0.24\textwidth]{images/monitoreq}}\hfil
+  \begin{subfigure}[b]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{images/loadeq}
+      \caption{\label{fig:loadeq}Load-Balancing}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{images/vieweq}
+      \caption{\label{fig:vieweq}Cross-Segment Load-Balancing}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{images/dfr}
+      \caption{\label{fig:dfr}Dynamic Frame Resolution}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.24\textwidth}
+    \includegraphics[width=\textwidth]{images/monitoreq}
+      \caption{\label{fig:monitor}Monitoring}
+  \end{subfigure}
   \caption{Runtime modifications}

### 2017-04-03 paperIEEE.tex
Refine compound configs
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@22 db7c848b-7755-4df2-b763-8796797afa52
General review
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@23 db7c848b-7755-4df2-b763-8796797afa52

@@
 % Do not put math or special symbols in the title.
-\title{Equalizer: A Mature Parallel Rendering Framework}
+\title{Equalizer 2.0\\Convergence of a Parallel Rendering Framework}
 %
@@
 
-The continuing improvements in hardware integration lead to ever faster CPUs and GPUs, as well as higher resolution sensor and display devices.
-Moreover, increased hardware parallelism is applied in form of multi-core CPU workstations, massive parallel super computers, or cluster systems.
-Hand in hand goes the rapid growth in complexity of data sets from numerical simulations, high-resolution 3D scanning systems or bio-medical imaging, which causes interactive exploration and visualization of such large data sets to become a serious challenge.
-It is thus crucial for a visualization solution to take advantage of hardware accelerated scalable parallel rendering.
-In this systems paper we describe a new scalable parallel rendering framework called {\em Equalizer} that is aimed primarily at cluster-parallel rendering, but works as well in a shared-memory system.
-Cluster systems are the main focus  because workstation graphics hardware is developing faster than high-end graphics (super-) computers can absorb new developments, and also because clusters offer a better cost-performance balance.
-
-Previous parallel rendering approaches typically failed in one of the following system requirements:
-%
-\begin{compactenum}\renewcommand{\labelenumi}{\alph{enumi})}
-\item generic application support, instead of special domain solution
-\item scalable abstraction of the graphics layer
-\item exploit existing code infrastructure, such as proprietary scene graphs, molecular data structures, level-of-detail and geometry databases
-\end{compactenum}
-
-To date, generic and scalable parallel rendering frameworks  that can be adopted to a wide range of scientific visualization domains are not yet readily available. Furthermore, flexible
-configurability to arbitrary cluster and display-wall configurations has also not been addressed in the past, but is of immense practical importance to scientists depending high-performance interactive visualization as a scientific tool.
-In this paper we present Equalizer, which is a novel flexible framework for parallel rendering that supports scalable performance, configuration flexibility, is {\em minimally invasive} with respect to adapting existing visualization applications, and is applicable to virtually any scientific visualization application domain.
-
-The main contributions that Equalizer introduces in a single parallel rendering system, and which are presented in this paper are:
-%
-\begin{compactenum}\renewcommand{\labelenumi}{\roman{enumi})}
-\item novel concept of compound trees for flexible configuration of graphics system resources,
-\item easy specification of parallel task decomposition and image compositing choice through compound tree layouts,
-\item automatic decomposition and distributed execution of rendering tasks according to compound tree,
-\item support for parallel surface as well as transparent (volume) rendering through $z$-visibility as well as $\alpha$-blending compositing,
-\item fully decentralized architecture providing network swap barrier (synchronization) and distributed objects functionality,
-\item support for low-latency distributed frame synchronization and image compositing,
-\item minimally invasive programming model.
-\end{compactenum}
-
-Equalizer is open source, available under the LGPL license from http://www.equalizergraphcis.com/, which allows it to be used both for open source and commercial applications.
-It is source-code portable, and has been tested on Linux, Microsoft Windows, and Mac OS X in 32 and 64 bit mode using both little endian and big endian processors.
+Highlight integration of many features into a general framework vs one-off
+research implemenation of one algorithm.
 
 %----------------------------------------------------------------------
-\section{Related Work}\label{SEC_related}
+\section{Related Work}\label{sec:related}
 %----------------------------------------------------------------------
@@
     \includegraphics[width=\textwidth]{images/dplex}
-      \caption{\label{fig:dplex}Time-Multiplex}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{images/tile}
-      \caption{\label{fig:tile}Tiles}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{images/pixel}
-      \caption{\label{fig:pixel}Pixel}
   \end{subfigure}
@@
     \includegraphics[width=\textwidth]{images/subpixel}
+  \end{subfigure}\\[\medskipamount]
+  \begin{subfigure}[t]{0.24\textwidth}
+    {\tiny\begin{lstlisting}
+compound {
+  channel "destination"
+  framerate_equalizer {}
+  compound {
+    channel "source1"
+    phase 0 period 3
+    outputframe  { name "frame" }
+  }
+  compound {
+    channel "source2"
+    phase 1 period 3
+    outputframe  { name "frame" }
+  }
+  compound {
+    channel "source3"
+    phase 2 period 3
+    outputframe  { name "frame" }
+  }
+  inputframe { name "frame" }
+}
+    \end{lstlisting}\vspace{24ex}}
+    \caption{\label{fig:dplex}Time-Multiplex}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.24\textwidth}
+    {\tiny\begin{lstlisting}
+compound {
+  channel "destination"
+  outputtiles {
+    name "queue"
+    size [ 64 64 ]
+  }
+  compound {
+    channel "destination"
+    inputtiles { name "queue" }
+  }
+  compound {
+    channel "source1"
+    inputtiles { name "queue" }
+    outputframe {}
+  }
+  compound {
+    channel "source2"
+    inputtiles { name "queue" }
+    outputframe {}
+  }
+  compound {
+    channel "source3"
+    inputtiles { name "queue" }
+    outputframe {}
+  }
+  inputframe { name "frame.source1" }
+  inputframe { name "frame.source2" }
+  inputframe { name "frame.source3" }
+}
+    \end{lstlisting}}
+    \caption{\label{fig:tiles}Tiles}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.24\textwidth}
+    {\tiny\begin{lstlisting}
+compound {
+  channel "dest"
+  compound {
+    channel "dest"
+    pixel [ 0 0 3 1 ]
+    outputframe { type texture }
+  }
+  compound {
+    channel "source1"
+    pixel [ 1 0 3 1 ]
+    outputframe {}
+  }
+  compound {
+    channel "source2"
+    pixel [ 2 0 3 1 ]
+    outputframe {}
+  }
+  inputframe { name "frame.dest" }
+  inputframe { name "frame.source1" }
+  inputframe { name "frame.source2" }
+}
+    \end{lstlisting}\vspace{21ex}}
+    \caption{\label{fig:pixel}Pixel}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.24\textwidth}
+    {\tiny\begin{lstlisting}
+compound {
+  channel "dest"
+  compound {
+    channel "dest"
+    subpixel [ 0 3 ]
+    outputframe { type texture }
+  }
+  compound {
+    channel "source1"
+    subpixel [ 1 3 ]
+    outputframe {}
+  }
+  compound {
+    channel "source2"
+    subpixel [ 1 3 ]
+    outputframe {}
+  }
+  inputframe { name "frame.dest" }
+  inputframe { name "frame.source1" }
+  inputframe { name "frame.source2" }
+}
+    \end{lstlisting}\vspace{21ex}}
     \caption{\label{fig:subpixel}Subpixel}
   \end{subfigure}
-  \caption{New Equalizer task decompositions for parallel rendering}
+  \caption{New Equalizer task decompositions and their compounds for parallel
+    rendering}
   \label{fig:compounds}
@@
 multi-GPU configuration, the application needs to support running the rendering
-threads asynchronously, as outlined in \sref{SEC_threading}. The output frame
+threads asynchronously, as outlined in \sref{sec:threading}. The output frame
 rate of the destination channel is smoothened using a frame rate equalizer
-(\sref{SEC_framerateEq}).
-
-\begin{figure*}[ht]\center
-  \begin{subfigure}[b]{0.24\textwidth}
-    {\tiny\begin{lstlisting}
-compound {
-  channel "destination"
-  framerate_equalizer {}
-  compound {
-    channel "source1"
-    phase 0 period 2
-    outputframe  { name "frame" }
-  }
-  compound {
-    channel "source2"
-    phase 1 period 2
-    outputframe  { name "frame" }
-  }
-  inputframe { name "frame" }
-}
-    \end{lstlisting}}
-    \caption{\label{eqc:dplex}Time-Multiplex}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.24\textwidth}
-    {\tiny\begin{lstlisting}
-compound {
-  channel "destination"
-  outputtiles {
-    name "queue"
-    size [ 64 64 ]
-  }
-  compound {
-    channel "destination"
-    inputtiles { name "queue" }
-  }
-  compound {
-    channel "source"
-    inputtiles { name "queue" }
-    outputframe {}
-  }
-  inputframe { name "frame.source" }
-}
-    \end{lstlisting}}
-    \caption{\label{eqc:tiles}Tiles}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.24\textwidth}
-    {\tiny\begin{lstlisting}
-compound {
-  channel "dest"
-  compound {
-    channel "dest"
-    pixel [ 0 0 2 1 ]
-    outputframe { type texture }
-  }
-  compound {
-    channel "source"
-    pixel [ 1 0 2 1 ]
-    outputframe{}
-  }
-  inputframe { name "frame.dest" }
-  inputframe { name "frame.source" }
-}
-    \end{lstlisting}}
-    \caption{\label{eqc:pixel}Pixel}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.24\textwidth}
-    {\tiny\begin{lstlisting}
-compound {
-  channel "dest"
-  compound {
-    channel "dest"
-    subpixel [ 0 2 ]
-    outputframe { type texture }
-  }
-  compound {
-    channel "source"
-    subpixel [ 1 2 ]
-    outputframe{}
-  }
-  inputframe { name "frame.dest" }
-  inputframe { name "frame.source" }
-}
-    \end{lstlisting}}
-    \caption{\label{eqc:subpixel}Subpixel}
-  \end{subfigure}
-  \caption{Compound configurations for \fig{fig:compounds}}
-  \label{eqc:compounds}
-\end{figure*}
+(\sref{sec:framerateEq}).
 
@@
 
-\subsubsection{Frame Rate Equalizer}\label{SEC_framerateEq}
+\subsubsection{Frame Rate Equalizer}\label{sec:framerateEq}
 
@@
 
-\subsubsection{Thread Synchronization Modes}\label{SEC_threading}
+\subsubsection{Thread Synchronization Modes}\label{sec:threading}
 
@@
 installations. Equalizer does support multiple, independent tracked views
-through the observer abstraction introduced in \sref{SEC_observer}. Built-in
+through the observer abstraction introduced in \sref{sec:observer}. Built-in
 VRPN support enables the direct, application-transparent configuration of a VRPN
@@
 passive stereo or stereo-dependent task decompositions, as described in
-\sref{SEC_reconfig}.
+\sref{sec:reconfig}.
 
@@
 software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
-swap, a Collage barrier (\sref{SEC_barrier}) to synchronize all segments, the
+swap, a Collage barrier (\sref{sec:barrier}) to synchronize all segments, the
 swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
@@
 
-\label{SEC_observer}
+\label{sec:observer}
 An observer looks at one or more views. It is described by the observer position
@@
 
-\subsection{Runtime Reconfiguration}\label{SEC_reconfig}
+\subsection{Runtime Reconfiguration}\label{sec:reconfig}
 
@@
   process-local pipes, TCP and Infiniband RDMA are implemented. For multicast,
-  a reliable, UDP-based protocol is discussed in \sref{SEC_RSP}.
+  a reliable, UDP-based protocol is discussed in \sref{sec:RSP}.
 \item[DataI/OStream] Abstracts the input and output of C++ data types from or to
@@
 
-\subsection{Reliable Stream Protocol}\label{SEC_RSP}
+\subsection{Reliable Stream Protocol}\label{sec:RSP}
 
@@
 
-The Collage Serializable implements one convenient usage pattern for object data
-distribution. The \textsf{co::Serializable} data distribution is based on the
-concept of dirty bits, allowing inheritance with data distribution. Dirty bits
-form a 64-bit mask which marks the parts of the object to be distributed during
-the next commit. For serialization, the application developer implements
-\textsf{serialize} or \textsf{deserialize}, which are called with the bit mask
-specifying which data has to be transmitted or received. During a commit or
-sync, the current dirty bits are given, whereas during object mapping all dirty
-bits are passed to the serialization methods.
+\label{sec:Serializable}The Collage Serializable implements one convenient usage
+pattern for object data distribution. The \textsf{co::Serializable} data
+distribution is based on the concept of dirty bits, allowing inheritance with
+data distribution. Dirty bits form a 64-bit mask which marks the parts of the
+object to be distributed during the next commit. For serialization, the
+application developer implements \textsf{serialize} or \textsf{deserialize},
+which are called with the bit mask specifying which data has to be transmitted
+or received. During a commit or sync, the current dirty bits are given, whereas
+during object mapping all dirty bits are passed to the serialization methods.
 
@@
 
-\subsection{Barriers, Queues and Object Maps}\label{SEC_barrier}
+\subsection{Barriers, Queues and Object Maps}\label{sec:barrier}
 
@@
 
-Livre (Large-scale interactive volume rendering engine) is an out-of-core
-parallel 4D volume renderer.
+Livre (Large-scale Interactive Volume Rendering Engine) is an out-of-core
+parallel 4D volume renderer. It supports sort-first decomposition, both for
+scalability as well as for driving large-scale tiled display walls. The
+out-of-core loading of data relies on a shared file system, only the high-level
+state, e.g., camera position and rendering settings, are shared through Collage
+objects.
 
 \subsection{RTT Deltagen}
+
+RTT Deltagen (now Dassault 3D Excite) is a commercial application for
+interactive, high quality rendering of CAD data. The RTT Scale module,
+delivering multi-GPU and distributed execution, is based on Equalizer and
+Collage, and has driven many of the aforementioned features.
+
+RTT Scale uses a master-slave execution mode, were a single running Deltagen
+instance can go into "Scale mode" at any time by launching an Equalizer
+configuration. Consequently, the whole internal representation needed for
+rendering is based on a Collage-based data distribution. The rendering clients
+are separate, smaller applications which will map their scenes during
+startup. At runtime, any change performed in the main application is committed
+as a delta at the beginning of the next frame, following a design pattern
+similar to the Collage \textsf{Serializable}
+(\sref{sec:Serializable}). Multicast (\sref{sec:RSP}) is used to keep data
+distribution times during session launch reasonable for larger cluster sizes
+(tens to hundreds of nodes).
+
+RTT Scale is used for a wide variety of use cases. In virtual reality, the
+application is used for virtual prototyping and design reviews in front of
+high-resolution display walls or in a CAVE and for virtual prototyping of
+human-machine interactions in a CAVE or using HMDs. For scalability, sort-first
+and tile compounds are used to achieve fast, high-quality rendering, primarily
+for interactive raytracing, both based on CPUs and GPUs. For CPU-based
+raytracing, often Linux-based rendering clients are used with a Windows-based
+application node.
 
@@
 %----------------------------------------------------------------------
-\label{SEC_results}
+\label{sec:results}
 
@@
 Distribute large ply model to 1..16 nodes using TCP, IB and RSP
+
+Distribute a video texture (per frame)
 
@@
 %----------------------------------------------------------------------
-\label{SEC_conclusions}
+\label{sec:conclusions}
 
@@
 
+We would also like to thank all supporters and contributors of Equalizer, most
+notably RTT, the Blue Brain Project, the University of Siegen, EVL, Dardo, TBD.
+
 % trigger a \newpage just before the given reference

### 2017-04-03 paperIEEE.tex
Abstract
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@24 db7c848b-7755-4df2-b763-8796797afa52

@@
   \begin{abstract}
-  We present the features, algorithms and system integration necessary to
-  implement a parallel rendering framework usable in a wide range of real-world
-  research and industry applications, based on the basic architecture and
+    Developing complex, real world applications which utilize multiple GPUs and
+    computers for 3D interactive rendering tasks can be a daunting task, in
+    particular since it requires expertise in distributed systems and parallel
+    rendering in addition to the application domain. We present a mature
+    parallel rendering framework which provides a large set of features,
+    algorithms and system integration for a wide range of real-world research
+    and industry applications. Based on the basic architecture and
     implementation of the Equalizer parallel rendering framework presented in
-  \cite{EMP:09}.
+    \cite{EMP:09}, we show how the generic algorithms are integrated in the
+    framework to help application development in many different domains.
   \end{abstract}
@@
     Parallel Rendering, Scalable Visualization, Cluster Graphics, Immersive Environments, Display Walls
-\end{IEEEkeywords}}
+  \end{IEEEkeywords}
+}
 
@@
 
-Tile (\fig{fig:tile}) and chunk decompositions are a variant of sort-first and
+Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
 sort-last rendering, respectively. They decompose the scene into a predefined
@@
 
-\subsection{Object Distribution}
-
-Distribute large ply model to 1..16 nodes using TCP, IB and RSP
-
-Distribute a video texture (per frame)
-
 \subsection{Decomposition Modes}
@@
 tiles, chunks, pixel, subpixel) (+DFR for Livre)
+
+\subsection{Object Distribution}
+
+Distribute large ply model to 1..16 nodes using TCP, IB and RSP
+
+Distribute a video texture (per frame)
 

### 2017-04-03 paperIEEE.tex
Improve object section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@25 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-UDP-based reliability protocol
+RSP is an implementation of a reliable multicast protocol over unreliable UDP
+transport. RSP behaves similarly to TCP, it provides full reliability and
+ordering of the data, and slow receivers will eventually throttle the
+sender. This behaviour is needed to guarantee delivery of data in all
+situations. Pragmatic multicast (PGM~\cite{pgm}) provides full ordering, but
+slow clients will disconnect from the multicast session instead of throttling
+the send rate.
+
+RSP is optimized for LAN performance. In the following we will outline the RSP
+protocol and implementation and motivate the design decisions. Any defaults
+given below are for Linux or OS X, the Windows UDP stack requires different
+default values which can be found in the implementation.
+
+Our RSP implementation uses a separate protocol thread for each RSP connection,
+which handles all reads and writes on the multicast group. This thread
+implements the protocol handling and communicates with the application threads
+through thread-safe queues, as shown in \fig{fRSPProto}. These queues contain
+datagrams, which are prefixed by a type-specific header. Each connection has a
+configurable number of buffers, (1024 by default) of a configurable MTU (1470
+bytes default) which are either free for use or in-use for reading or writing.
+
+%%     1,     // RSP_ERROR_DOWNSCALE
+%%     5,     // RSP_ERROR_UPSCALE
+%%     20,    // RSP_ERROR_MAXSCALE
+%%     3,     // RSP_MIN_SENDRATE_SHIFT
+
+%%     17,    // RSP_ACK_FREQUENCY
+
+\begin{figure}[ht]\center
+  \includegraphics[width=\columnwidth]{images/rspPackets}
+  \caption{\label{fRSPProto}Reliable Stream Protocol}
+\end{figure}
+
+A client opens a listening multicast RSP connection. Similar to TCP, this
+listener will create one connected RSP connection for each remote member in the
+multicast group, that is, on each node $n$ RSP connection instances exist, $1$
+for the local listener and $n-1$ for all other nodes. A multicast group
+therefore looks like a fully connected set of TCP connections between $n$ nodes
+for the user.
+
+Each member of a multicast group has a group-unique 16 bit identifier, which is
+randomly selected, announced to the group, and if no collisions are notified by
+any other member, confirmed for this connection. This mechanism allows
+addressing of connections with a low protocol overhead. The datagrams used in
+this exchange contain no payload, and they contain two bytes type (HELLO,
+HELLO\_REPLAY, HELLO\_DENY, HELLO\_CONFIRM), two bytes protocol version and two
+bytes connection identifier.
+
+The \textsf{\_appBuffers} queue, which is blocking, contains received UDP
+datagrams on incoming connections, or empty buffers for writing on outgoing
+connections. The \textsf{\_threadBuffers} queue is lock-free and used by the
+protocol thread to store empty buffers on incoming connections and pre-packaged
+UDP datagrams on outgoing connections. The application thread pushes buffers
+into this queue, consumed by the protocol thread. A datagram is up to 64
+kilobytes (UDP limit, configurable) and contains a header (two bytes type, two
+bytes size, two bytes source identifier, two bytes sequence number) and its
+payload. The sequence number is used for positive (ack) and negative (nack)
+acknowledgments and reordering of packets.
+
+The \textsf{\_writeBuffers std::deque} is only used by the protocol
+thread. It contains already sent, but not yet acknowledged datagrams. Its
+datagrams are resent upon reception of a nack or recycled into
+\textsf{\_appBuffers} when acknowledged by all receivers. Similarly, the
+\textsf{\_recvBuffers std::deque} contains received out-of-order packets already
+received by an incoming connection. Upon each reception it is checked, and
+in-order datagrams are placed into \textsf{\_appBuffers} for consumption by the
+application thread.
+
+Handling a smooth packet flow is critical for performance. RSP actively flow
+control to advance the stream. Each incoming connection actively acknowledges
+every $n$ (17 by default) packets fully received. The incoming connections shift
+this acknowledgment by their connection identifier to avoid bursts of acks. Any
+missed datagram is actively nack'ed as soon as it is detected. Write connections
+continuously retransmit packets from nack datagrams, and advance their
+\textsf{\_writeBuffers} upon reception of acks from all readers. Furthermore,
+the writer will explicitly request an ack or nack when it runs out of empty
+buffers or finishes its write queue.
+
+Congestion control is necessary to optimize bandwidth usage. While TCP uses the
+well-known AIMD (additive increase, multiplicative decrease) algorithm, we have
+chosen a more aggressive congestion control algorithm of additive increase and
+decrease. This has proven experimentally to be more optimal: UDP is often
+rate-limited by switches, that is, packets are discarded regularly and not
+exceptionally. Only slowly backing of the current send rate helps to stay close
+to this limit. Furthermore, our RSP traffic is limited to the local subnet,
+making cooparation between multiple data stream less of an issue. Send rate
+limiting uses a bucket algorithm, where over time the bucket fills with send
+credits, from which sends are substracted. If there are no available credits,
+the sender sleeps until sufficient credits are available.
 
@@
 approaches to address this distribution: Using a shared filesystem for static
-data and using data distribution for static and dynamic data.
+data, or using data distribution for static and dynamic data.
+
+Distributed objects are not required to build Equalizer applications. While most
+developers choose to use this abstraction for convenience, we have seen
+applications using other means for data distribution, e.g., MPI.
+
+\subsubsection{Programming Interface}
 
@@
 cluster-wide. A master-slave model is used to establish mapping and data
-synchronization across processes. Typically, the application mail loop registers
+synchronization across processes. Typically, the application main loop registers
 a master instance and communicates the UUID to the render clients, which map
@@
 data. These methods provide an output or input stream as a parameter, which
-abstracts the data transmission and can be used like a \textsf{std::stream}. The
-data streams implement efficient buffering and compression, and automatically
-select the best connection for data transport. Custom data type serializers can
-be implemented by providing the appropriate serialization functions. No pointers
-should be directly transmitted through the data streams. For pointers, the
-corresponding object is typically a distributed object as well, and its
-identifier and potentially version is transmitted in place of the memory
-address.
+abstracts the data transmission and can be used like a \textsf{std::stream}.
+The data streams implement efficient buffering and compression, and
+automatically select the best connection for data transport. Custom data type
+serializers can be implemented by providing the appropriate serialization
+functions. No pointers should be directly transmitted through the data
+streams. For pointers, the corresponding object is typically a distributed
+object as well, and its identifier and potentially version is transmitted in
+place of the memory address.
 
@@
 
-Besides the instance data serialization methods used to map an object, versioned
-objects may implement \textsf{pack} and \textsf{unpack} to serialize or
-deserialize the changes since the last version.
+Besides the instance data serialization methods used to map an object, delta and
+unbuffered objects may implement \textsf{pack} and \textsf{unpack} to serialize
+or deserialize the changes since the last version.
 
@@
 instance if to many unsynchronized versions exist.
+
+\subsubsection{Optimizations}
+
+The API presented in the previous section provides sufficient abstraction to
+implement various optimizations for faster mapping and synchronization of data:
+compression, chunking, caching, preloading and multicast. The results section
+evaluates these optimizations.
+
+The most obvious one is compression. Recently, many new compression algorithms
+have been developed which exploit modern CPU architectures and deliver
+compression rates well above one gigabyte per second. Collage uses the Pression
+library~\cite{pression}, which provides an unified interface for a number of
+compression libraries, such as FastLZ~\cite{fastlz}, Snappy~\cite{snappy} and
+ZStandard~\cite{zstd}. It also contains a custom, virtually zero-cost RLE
+compressor. Pression also parallelizes the compression and decompression using
+data decomposition.
+
+The data streaming interface also implements chunking, which aims to pipeline
+the serialization code with the network transmission. After a configurable
+number of bytes has been serialized to the internal buffer, it is transmitted
+and serialization continues. This is used both for the initial mapping data and
+for commit data.
+
+Caching retains instance data of objects in a cache, and reuses this data to
+accelarate mapping of objects. The instance cache is either filled by "snooping"
+on multicast transmissions or by an explicit preloading when the master objects
+are registered. The preloading ("send on register") sends instance data of
+recently registered master objects while the correspinding node is idle to all
+connected nodes. These nodes simply enter the received data to their
+cache. Preloading uses multicast when available.
+
+Due to the master-slave model of data distribution, multicast can be used to
+optimize the transmission time of data. If the contributing nodes share a
+multicast session, and more than one slave instance is mapped, Collage
+automatically uses the multicast connection to send the new version
+information.
 
@@
 \subsection{Raster}
+
+\subsection{Bino}
+
 \subsection{Omegalib}
@@
 
-Distribute large ply model to 1..16 nodes using TCP, IB and RSP
+Pression data compressor benchmark
+
+Distribute large ply and raw model to 1..16 nodes using TCP, IB and RSP
 

### 2017-04-03 paperIEEE.tex
Results for data compression engines
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@26 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-RSP is optimized for LAN performance. In the following we will outline the RSP
-protocol and implementation and motivate the design decisions. Any defaults
-given below are for Linux or OS X, the Windows UDP stack requires different
-default values which can be found in the implementation.
+RSP combines various established algorithms~\cite{nack, flowcontrol} for
+multicast in an open source implementation capable of delivering wire speed
+transmission rates on high-speed LAN interfaces. In the following we will
+outline the RSP protocol and implementation and motivate the design
+decisions. Any defaults given below are for Linux or OS X, the Windows UDP stack
+requires different default values which can be found in the implementation.
 
@@
 connections. The \textsf{\_threadBuffers} queue is lock-free and used by the
-protocol thread to store empty buffers on incoming connections and pre-packaged
+protocol thread to store empty buffers on incoming connections, and pre-packaged
 UDP datagrams on outgoing connections. The application thread pushes buffers
@@
 \subsection{Object Distribution}
+
+\begin{figure}[ht]\center
+  \includegraphics[width=\columnwidth]{images/compressor}
+  \caption{\label{fCompressor}Data Compression Engines}
+\end{figure}
 

### 2017-04-03 paperIEEE.tex
More data compression results
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@27 db7c848b-7755-4df2-b763-8796797afa52

@@
-%% bare_jrnl_compsoc.tex
-%% V1.4b
-%% 2015/08/26
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
-%% Computer Society journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/pkg/ieeetran
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE!
-%% User assumes all risk.
-%% In no event shall the IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. The IEEE's font choices and paper sizes can   ***
-% *** trigger bugs that do not appear when using other class files.       ***                          ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
 \documentclass[10pt,journal,compsoc]{IEEEtran}
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/pkg/ifpdf
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
 
@@
 \fi
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of the IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off
-% such as if a citation ever needs to be enclosed in parenthesis.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 5.0 (2009-03-20) and later if using hyperref.sty.
-% The latest version can be obtained at:
-% http://www.ctan.org/pkg/cite
-% The documentation is contained in the cite.sty file itself.
-%
-% Note that some packages require special options to format as the Computer
-% Society requires. In particular, Computer Society  papers do not use
-% compressed citation ranges as is done in typical IEEE papers
-% (e.g., [1]-[4]). Instead, they list every citation separately in order
-% (e.g., [1], [2], [3], [4]). To get the latter we need to load the cite
-% package with the nocompress option which is supported by cite.sty v4.0
-% and later. Note also the use of a CLASSOPTION conditional provided by
-% IEEEtran.cls V1.7 and later.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation
-% can be obtained at:
-% http://www.ctan.org/pkg/graphicx
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found at:
-% http://www.ctan.org/pkg/epslatex
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics.
-%
-% Note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/amsmath
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/pkg/algorithms
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/pkg/algorithmicx
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/array
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a sans serif font rather
-% than the serif font used in traditional IEEE formatting and thus the need
-% to invoke different subfig.sty package options depending on whether
-% compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/pkg/subfig
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure.
-% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
-% corrections already built into the system in which case a warning will
-% be issued if an attempt is made to load fixltx2e.sty as it is no longer
-% needed.
-% The latest version and documentation can be found at:
-% http://www.ctan.org/pkg/fixltx2e
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/stfloats
-% Do not use the stfloats baselinefloat ability as the IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that the IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-% \usepackage{dblfloatfix}
-% The latest version can be found at:
-% http://www.ctan.org/pkg/dblfloatfix
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and
-% Axel Sommerfeldt. This package may be useful when used in conjunction with
-% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
-% submissions have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.4.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/pkg/endfloat
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/url
-% Basically, \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
 
@@
 sender. This behaviour is needed to guarantee delivery of data in all
-situations. Pragmatic multicast (PGM~\cite{pgm}) provides full ordering, but
-slow clients will disconnect from the multicast session instead of throttling
-the send rate.
-
-RSP combines various established algorithms~\cite{nack, flowcontrol} for
-multicast in an open source implementation capable of delivering wire speed
-transmission rates on high-speed LAN interfaces. In the following we will
-outline the RSP protocol and implementation and motivate the design
-decisions. Any defaults given below are for Linux or OS X, the Windows UDP stack
-requires different default values which can be found in the implementation.
+situations. Pragmatic generic multicast (PGM~\cite{pgm}) provides full ordering,
+but slow clients will disconnect from the multicast session instead of
+throttling the send rate.
+
+RSP combines various established algorithms~\cite{adamson2004negative,
+  Gau:2002:MFC:506824.506832} for multicast in an open source implementation
+capable of delivering wire speed transmission rates on high-speed LAN
+interfaces. In the following we will outline the RSP protocol and implementation
+and motivate the design decisions. Any defaults given below are for Linux or OS
+X, the Windows UDP stack requires different default values which can be found in
+the implementation.
 
@@
 library~\cite{pression}, which provides an unified interface for a number of
-compression libraries, such as FastLZ~\cite{fastlz}, Snappy~\cite{snappy} and
-ZStandard~\cite{zstd}. It also contains a custom, virtually zero-cost RLE
+compression libraries, such as FastLZ~\cite{jesperfast}, Snappy~\cite{snappy}
+and ZStandard~\cite{zstd}. It also contains a custom, virtually zero-cost RLE
 compressor. Pression also parallelizes the compression and decompression using
@@
 %----------------------------------------------------------------------
-\section{Experimental Results}
+\section{Experimental Results}\label{sec:results}
 %----------------------------------------------------------------------
-\label{sec:results}
+
+hardware description
 
@@
 
+\subsubsection{Data Compression Engines}
+
+A critical piece for data distribution performance are the characteristics of
+the data compression algorithms. This microbenchmark compresses a set of binary
+files to precalculate the speed and compression ratio of the various
+engines. \fig{fCompressor} shows the compression and decompression speed in
+gigabyte per second as well as the size of the compressed data. The different
+ZSTD$x$ engines use the ZStandard compression library at level $x$.
+
 \begin{figure}[ht]\center
   \includegraphics[width=\columnwidth]{images/compressor}
-  \caption{\label{fCompressor}Data Compression Engines}
+  \caption{\label{fCompressor}Data Compression for Generic Binary Data}
 \end{figure}
 
-Pression data compressor benchmark
-
+RLE compression has a very low overhead, but does not compress the data will
+since it only removes "blank space" in the data. The Snappy compression, used as
+default in Collage, achieves the same compression ratio as the LZ variants at a
+much higher speed. The ZStandard compressor has roughly the same speed as the LZ
+variants at the lowest compression level, but provides significantly better
+compression. At higher compression levels it can improve the compression ratio
+slightly, but at a high cost for the compression speed.
+
+\subsubsection{Model Distribution and Update}
 Distribute large ply and raw model to 1..16 nodes using TCP, IB and RSP

### 2017-04-03 paperIEEE.tex
More data compression results
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@28 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+To benchmark the data distribution we used two datasets: The david statue at 2mm
+resolution (\fig{fDavid2mm}) and 3D volumes of a spike frequencies of a 3
+million neuron simulation (\fig{fVolumes}).
+
+\begin{figure}[ht]\center
+  \includegraphics[width=.49\columnwidth]{images/spikes512}\hfil
+  \includegraphics[width=.49\columnwidth]{images/spikes1280}
+  \caption{\label{fVolumes}Spike frequencies of a 3 million neuron simulation at
+    437x240x512 (51MB, left) and 1092x600x1280 (800MB, right) resolution}
+\end{figure}
+
+The spike frequency volumes aggregate the number of spikes which happened within
+a given voxel over a given time. The absolute spike count is renormalized to an
+unsigned byte value (0..255) during creation. Higher densities in the volume
+represent higher spiking activity in the voxel. Consequently, the volume is more
+sparsely populated and has a more even distribution of values at higher
+resolution.
+
 \subsubsection{Data Compression Engines}
@@
 slightly, but at a high cost for the compression speed.
+
+The compression ratio for the models used in the following benchmark deviate
+from this averaged distribution. \fig{fCompressorDetail} shows the compression
+ratios for the triangle and volume data.
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\textwidth]{images/compressorDetail}
+  \caption{\label{fCompressorDetail}Data Compression for ply data (left, David
+    statue 2mm) and raw volume data shown in \fig{fVolumes} (middle and right)}
+\end{figure*}
+
+The ply data is little compressible, the default compressor achieves a 10\%
+reduction. This is due to a high entropy in the data and the dominant use of
+floating point representations. Overall the profile is similar to the generic
+benchmark, at a smaller compression rate.
+
+The volume data on the other hand is sparsely populated and using integer (byte)
+representation, which is easier to compress. The naive RLE implementation
+already achieves a good compression, showing that the smaller volume contains at
+most 28\% empty space and the bigger volume at most 90\%. Snappy and ZStandard
+can reduce the data much further, reducing the data to a few megabytes.
 

### 2017-04-03 paperIEEE.tex
Resolve references
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@29 db7c848b-7755-4df2-b763-8796797afa52

@@
 To benchmark the data distribution we used two datasets: The david statue at 2mm
-resolution (\fig{fDavid2mm}) and 3D volumes of a spike frequencies of a 3
-million neuron simulation (\fig{fVolumes}).
+resolution (\fig{fDavid2mm}) and 3D volumes of a spike frequencies of an
+electrical simulation of three million neurons (\fig{fVolumes}).
 
@@
   \includegraphics[width=.49\columnwidth]{images/spikes1280}
-  \caption{\label{fVolumes}Spike frequencies of a 3 million neuron simulation at
-    437x240x512 (51MB, left) and 1092x600x1280 (800MB, right) resolution}
+  \caption{\label{fVolumes}Spike frequencies of a three million neuron
+    electrical simulation at 437x240x512 (51MB, left) and 1092x600x1280 (800MB,
+    right) resolution}
 \end{figure}

### 2017-04-03 paperIEEE.tex
Flow and my CV
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@30 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+The rest of this paper is structured as follows: First we introduce new related
+work since the introduction of Equalizer. The main body of this paper presents
+new performance features, virtual reality algorithms, usability features to
+build complex applications, main novel features of the underlying Collage
+network library and a quick overview of the main Equalizer-based applications. A
+result section presents new experiments not previously published, followed by
+the discussion and conclusion.
+
 %----------------------------------------------------------------------
@@
 In~\cite{EMP:09}, we presented the Equalizer parallel rendering framework. This
-paper summarized the work in parallel rendering up to 2009. In the following we
-will present the related work published since then. An extensive Programming and
-User Guide provides in-depth documentation on using and programming
-Equalizer~\cite{Eilemann:13}.
+paper summarized the work in parallel rendering up to 2009. An extensive
+Programming and User Guide provides in-depth documentation on using and
+programming Equalizer~\cite{Eilemann:13}. In the following we assume these two
+publications and their references as the basis, and will present the related
+work published since then.
 
@@
 execution transparently to OpenGL applications, by emulating the GLUT API and
-intercepting certain OpenGL calls. This approach works transparently for trivial
-applications, but quickly requires the application developer to address the
-complexities of a distributed application when mutable application state needs
-to be synchronized across processes.
+intercepting certain OpenGL calls. In contrast to frameworks like Chromium, CGLX
+follows the distribution application approach, whichworks transparently for
+trivial applications, but quickly requires the application developer to address
+the complexities of a distributed application when mutable application state
+needs to be synchronized across processes.
 
 On the other hand, software for driving and interacting with tiled display walls
-has received significant attention. Sage~\cite{Sage} and Sage2~\cite{Sage2} have
-received significant attention. Sage is build entirely around the concept of a
-shared framebuffer where all content windows are separate applications using
-pixel streaming. It is no longer actively supported. Sage 2 is a complete,
+has received significant attention. Sage~\cite{Sage} and Sage~2~\cite{Sage2}
+have received significant attention. Sage is build entirely around the concept
+of a shared framebuffer where all content windows are separate applications
+using pixel streaming. It is no longer actively supported. Sage 2 is a complete,
 browser-centric reimplementation where each application is a web application
 distributed across browser instances. DisplayCluster~\cite{DC}, and its
-continuation Tide, also implement the shared framebuffer concept of Sage, but
-provide a few native content applications integrated into the display servers.
-
-Equalizer itself has received significant attention to research and validate
-parallel rendering performance, by using compression and region of interest
-during compositing~\cite{MEP:10}, load-balancing resources for multi-display
-installations~\cite{EEP:11}, asynchronous compositing and NUMA
-optimizations~\cite{EBAHMP:12} as well as work queueing~\cite{SPEP:16}.
+continuation Tide~\cite{tide}, also implement the shared framebuffer concept of
+Sage, but provide a few native content applications integrated into the display
+servers.
+
+Equalizer itself has received significant attention within the research
+community. Various algorithms to improve the parallel rendering performance have
+been proposed: compression and region of interest during
+compositing~\cite{MEP:10}; load-balancing resources for multi-display
+installations~\cite{EEP:11}; asynchronous compositing and NUMA
+optimizations~\cite{EBAHMP:12}; as well as work queueing~\cite{SPEP:16}.
 
@@
 visualization, such as the RTNeuron~\cite{HBBES:13}, an OpenGL renderer for
-visualizing simulations of neucortical brain regions, Omegalib~\cite{Omegalib},
-extending Equalizer for the hybrid 2D/3D environments such as the CAVE2,
-visualization for remote sensing date~\cite{LK:09}, .
+visualizing simulations of neucortical brain regions; Omegalib~\cite{Omegalib},
+extending Equalizer for the hybrid 2D/3D environments such as the CAVE2;
+visualization for remote sensing data~\cite{LK:09}, bino.
 
@@
 
-hardware description
+This section presents new experiments, complementing the results of previous
+publications~\cite{EP:07, EMP:09, MEP:10, EEP:11, EBAHMP:12, HBBES:13, deflect,
+  SPEP:16}. The first part analyses data distribution performance, in particular
+how the various optimizations in Collage effect complex operations in realistic
+scenarios. The second part summarizes rendering performance over all
+decomposition modes with a few representative workloads.
 
@@
   \includegraphics[width=.49\columnwidth]{images/spikes1280}
-  \caption{\label{fVolumes}Spike frequencies of a three million neuron
-    electrical simulation at 437x240x512 (51MB, left) and 1092x600x1280 (800MB,
-    right) resolution}
+  \caption{\label{fVolumes}Volumes used for object distribution benchmarks:
+    spike frequencies of a three million neuron electrical simulation at
+    437x240x512 (51MB, left) and 1092x600x1280 (800MB, right) resolution}
 \end{figure}
@@
 sparsely populated and has a more even distribution of values at higher
-resolution.
+resolutions.
 
@@
 
-\begin{IEEEbiography}{Michael Shell}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{Stefan Eilemann}
+Stefan is the technical manager of the Visualization Team in the Blue Brain
+Project. His work involves working towards large-scale visualization for
+Exascale simulations; interactive integration of simulation, analysis and
+visualization as well as flexible frameworks for data sharing and dynamic
+allocation of heterogenous resources. He is the CEO and founder of Eyescale and
+also the lead developer of the Equalizer parallel rendering framework. He has
+worked in SGI's Advanced Graphics Division as the technical lead of OpenGL
+Multipipe SDK - a toolkit for scalable, parallel visualization software on
+shared memory supercomputers. He received his masters degree in Computer Science
+from EPFL in 2015, and his Engineering Diploma in Computer Science in 1998.
 \end{IEEEbiography}

### 2017-04-03 paperIEEE.tex
Polishing
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@31 db7c848b-7755-4df2-b763-8796797afa52

@@
 publications~\cite{EP:07, EMP:09, MEP:10, EEP:11, EBAHMP:12, HBBES:13, deflect,
-  SPEP:16}. The first part analyses data distribution performance, in particular
-how the various optimizations in Collage effect complex operations in realistic
-scenarios. The second part summarizes rendering performance over all
-decomposition modes with a few representative workloads.
+  SPEP:16}. The first part summarizes rendering performance over all
+decomposition modes with a few representative workloads. The second part
+analyses data distribution performance, in particular how the various
+optimizations in Collage effect complex operations in realistic scenarios.
 
@@
 
+The data distribution benchmarks have been performed on a cluster using
+dual processor Intel Xeon X5690 CPUs (3.47GHz), 24 GB main memory, 10GBit/s
+Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake 3.2 release
+mode settings to compile the software stack.
+
 To benchmark the data distribution we used two datasets: The david statue at 2mm
@@
 electrical simulation of three million neurons (\fig{fVolumes}).
+
+\begin{figure}[ht]\center
+  \includegraphics[width=.49\columnwidth]{images/david2mm}
+  \caption{\label{fDavid2mm}David statue at 2mm resolution (145MB)}
+\end{figure}
+
+The ply file is converted into a kd-Tree for fast view frustum culling and
+rendering, and the resulting data structure is serialized in binary form for
+data transmission.
 
@@
 gigabyte per second as well as the size of the compressed data. The different
-ZSTD$x$ engines use the ZStandard compression library at level $x$.
+ZSTD$x$ engines use the ZStandard compression library at level $x$. The
+measurements were performed on a single, isolated node.
 
@@
 shared memory supercomputers. He received his masters degree in Computer Science
-from EPFL in 2015, and his Engineering Diploma in Computer Science in 1998.
+from EPFL in 2015, and his Engineering Diploma in Computer Science in 1998. He
+is currently working towards a PhD in Computer Science at the Visualization and
+Multimedia Lab at the University of Zurich.
 \end{IEEEbiography}

### 2017-04-03 paperIEEE.tex
More benchmarks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@32 db7c848b-7755-4df2-b763-8796797afa52

@@
     untracked views running RTT DeltaGen for virtual car usability studies (c)
-    Cave2 running a molecular visualization build using Omegalib.}
+    Cave2 running a molecular visualization build based on Omegalib.}
   \label{FIG_teaser}
@@
 present new decomposition modes and motivate their use case. \fig{fig:compounds}
-provides an overview of the new modes. We describe their configuration in
-same \textit{compound} framework introduced in \cite{EMP:09}
+provides an overview of the new modes. We describe their configuration in the
+\textit{compound} framework introduced in \cite{EMP:09}.
 
@@
   \includegraphics[width=\columnwidth]{EqDocs/Developer/ProgrammingGuide/images/netObject}
-  \caption{\label{fNetObject}Communication between two Collage Objects}
+  \caption{\label{fNetObject}Communication between two Collage objects}
 \end{figure}
@@
 and ZStandard~\cite{zstd}. It also contains a custom, virtually zero-cost RLE
-compressor. Pression also parallelizes the compression and decompression using
-data decomposition.
+compressor. Pression parallelizes the compression and decompression using
+data decomposition. Note that this compression is generic, and implemented
+transparently for the application. Applications can still use data-specific
+compression and feed the compressed data to Collage, typically disabling the
+generic compression.
 
@@
 \begin{figure}[ht]\center
-  \includegraphics[width=.49\columnwidth]{images/spikes512}\hfil
-  \includegraphics[width=.49\columnwidth]{images/spikes1280}
+  \includegraphics[height=.37\columnwidth]{images/spikes512}\hfil
+  \includegraphics[height=.37\columnwidth]{images/beechnut}
   \caption{\label{fVolumes}Volumes used for object distribution benchmarks:
     spike frequencies of a three million neuron electrical simulation at
-    437x240x512 (51MB, left) and 1092x600x1280 (800MB, right) resolution}
+    437x240x512 unsigned byte (51MB, left) and an MRI scan of a beechnut at
+    1024x1024x1546 unsigned short (3GB, right) resolution}
 \end{figure}
@@
 unsigned byte value (0..255) during creation. Higher densities in the volume
-represent higher spiking activity in the voxel. Consequently, the volume is more
-sparsely populated and has a more even distribution of values at higher
-resolutions.
+represent higher spiking activity in the voxel.
 
@@
   \includegraphics[width=\columnwidth]{images/compressor}
-  \caption{\label{fCompressor}Data Compression for Generic Binary Data}
+  \caption{\label{fCompressor}Data compression for generic binary data}
 \end{figure}
@@
   \includegraphics[width=\textwidth]{images/compressorDetail}
-  \caption{\label{fCompressorDetail}Data Compression for ply data (left, David
-    statue 2mm) and raw volume data shown in \fig{fVolumes} (middle and right)}
+  \caption{\label{fCompressorDetail}Data compression for ply data (left, David
+    statue 2mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
 \end{figure*}
@@
 \subsubsection{Model Distribution and Update}
+
+In this section we present results on how data distribution and synchronization
+performs in real-world applications. We extracted the existing data distribution
+code in a PLY and a volume renderer into a benchmark application to measure the
+time to initially map all the objects on the render client nodes, and to perform
+a commit-sync of the full data set after mapping has been established. All
+figures observe a noticable measurement jitter due to other services running on
+the cluster. The details of the benchmark algorithm can be found in the implementation~\cite{eqObjectBench}.
+
+David 2mm
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\textwidth]{images/spikesdist}
+  \caption{\label{fSpikesDist}Object mapping (left )and data synchronization
+    times (middle, detail view right) for the spike volume data set in
+    \fig{fVolumes}, right}
+\end{figure*}
+
 Distribute large ply and raw model to 1..16 nodes using TCP, IB and RSP
-
-Distribute a video texture (per frame)
 

### 2017-04-03 paperIEEE.tex
Updated beechnut
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@33 db7c848b-7755-4df2-b763-8796797afa52

@@
 \begin{figure}[ht]\center
-  \includegraphics[height=.37\columnwidth]{images/spikes512}\hfil
-  \includegraphics[height=.37\columnwidth]{images/beechnut}
+  \includegraphics[height=.44\columnwidth]{images/spikes512}\hfil
+  \includegraphics[height=.44\columnwidth]{images/beechnut}
   \caption{\label{fVolumes}Volumes used for object distribution benchmarks:
     spike frequencies of a three million neuron electrical simulation at
-    437x240x512 unsigned byte (51MB, left) and an MRI scan of a beechnut at
+    437x240x512 unsigned byte (51MB, left) and an MicroCT scan of a beechnut at
     1024x1024x1546 unsigned short (3GB, right) resolution}

### 2017-04-03 paperIEEE.tex
More benchmarks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@34 db7c848b-7755-4df2-b763-8796797afa52

@@
 Caching retains instance data of objects in a cache, and reuses this data to
-accelarate mapping of objects. The instance cache is either filled by "snooping"
-on multicast transmissions or by an explicit preloading when the master objects
-are registered. The preloading ("send on register") sends instance data of
-recently registered master objects while the correspinding node is idle to all
-connected nodes. These nodes simply enter the received data to their
-cache. Preloading uses multicast when available.
+accelarate mapping of objects. The instance cache is either filled by
+``snooping'' on multicast transmissions or by an explicit preloading when the
+master objects are registered. The preloading (``send on register'') sends
+instance data of recently registered master objects while the correspinding node
+is idle to all connected nodes. These nodes simply enter the received data to
+their cache. Preloading uses multicast when available.
 
@@
 RTT Scale uses a master-slave execution mode, were a single running Deltagen
-instance can go into "Scale mode" at any time by launching an Equalizer
+instance can go into ``Scale mode'' at any time by launching an Equalizer
 configuration. Consequently, the whole internal representation needed for
@@
     spike frequencies of a three million neuron electrical simulation at
-    437x240x512 unsigned byte (51MB, left) and an MicroCT scan of a beechnut at
+    512x437x240 unsigned byte (51MB, left) and an MicroCT scan of a beechnut at
     1024x1024x1546 unsigned short (3GB, right) resolution}
@@
 RLE compression has a very low overhead, but does not compress the data will
-since it only removes "blank space" in the data. The Snappy compression, used as
-default in Collage, achieves the same compression ratio as the LZ variants at a
-much higher speed. The ZStandard compressor has roughly the same speed as the LZ
-variants at the lowest compression level, but provides significantly better
+since it only removes ``blank space'' in the data. The Snappy compression, used
+as default in Collage, achieves the same compression ratio as the LZ variants at
+a much higher speed. The ZStandard compressor has roughly the same speed as the
+LZ variants at the lowest compression level, but provides significantly better
 compression. At higher compression levels it can improve the compression ratio
@@
 
-In this section we present results on how data distribution and synchronization
-performs in real-world applications. We extracted the existing data distribution
-code in a PLY and a volume renderer into a benchmark application to measure the
-time to initially map all the objects on the render client nodes, and to perform
-a commit-sync of the full data set after mapping has been established. All
-figures observe a noticable measurement jitter due to other services running on
-the cluster. The details of the benchmark algorithm can be found in the implementation~\cite{eqObjectBench}.
-
-David 2mm
+In this section we present how data distribution and synchronization performs in
+real-world applications. We extracted the existing data distribution code in a
+PLY and a volume renderer into a benchmark application to measure the time to
+initially map all the objects on the render client nodes, and to perform a
+commit-sync of the full data set after mapping has been established. All figures
+observe a noticable measurement jitter due to other services running on the
+cluster. The details of the benchmark algorithm can be found in the
+implementation~\cite{eqObjectBench}. We used the same data sets as in the
+previous section. After $8$ processes, a nodes start to run two processes per
+node, which share the CPU and network interface.
+
+Object mapping is measured using the following settings: \textbf{none}
+distributes the raw, uncompressed, and unbuffered data, \textbf{compression}
+uses the Snappy compresser to distribute unbuffered data, \textbf{buffered}
+reuses uncompressed, serialized data for mappings of other nodes, and
+\textbf{compression buffered} compresses this buffer. Unbuffered operations need
+to reserialize, and potentially recompress, the master object data for each
+slave node. Each slave instance needs to deserialize and decompress the data.
+
+During data synchronization, the master instance commits the object data to all
+mapped slave instances simultaneously. This is a ``push'' operation, whereas the
+mapping is a slave-triggered ``pull'' operation with additional synchronization
+points. Slave nodes queue this data and consume it during synchronization. We
+test the time to commit and sync the data using different compression engines.
+
+The David statue at 2mm resolution is organized in a kD-tree for rendering. Each
+node is a separate distributable object, which has two child node objects. A
+total of $1023$ objects is distributed and synchronized.
+
+\fig{fSpikesDist} has a significant compression ratio, which reflects in the
+results. Futhermore, only a single distributed object is used to transmit the
+volume information in one chunk. Compression for this data is benefitial for
+transmitting data over a 10Gb/s link (880 MB/s vs 660MB/s raw data
+transfer). Buffering has little benefit since the serialization of volume data
+is trivial. Buffered compression makes a huge difference, since the compression
+cost can be amortized over $n$ nodes, reaching raw data transmission rates of
+3.7 GB/s.
 

### 2017-04-03 paperIEEE.tex
Remove EqDocs submodule
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@36 db7c848b-7755-4df2-b763-8796797afa52
add
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@37 db7c848b-7755-4df2-b763-8796797afa52
Add
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@38 db7c848b-7755-4df2-b763-8796797afa52

@@
 \author{Stefan Eilemann\thanks{email: eilemann@gmail.com} \\ %
-\and Renato Pajarola\thanks{email: pajarola@acm.org}}
+\and Renato Pajarola~\IEEEmembership{Senior Member,~IEEE}\thanks{email: pajarola@acm.org}}
 
@@
   \begin{abstract}
-    Developing complex, real world applications which utilize multiple GPUs and
-    computers for 3D interactive rendering tasks can be a daunting task, in
+    Developing complex, real world graphics applications which utilize multiple GPUs and
+    computers for interactive 3D rendering tasks can be a daunting task, in
     particular since it requires expertise in distributed systems and parallel
-    rendering in addition to the application domain. We present a mature
+    rendering in addition to the application domain itself. We present a mature
     parallel rendering framework which provides a large set of features,
@@
     implementation of the Equalizer parallel rendering framework presented in
-    \cite{EMP:09}, we show how the generic algorithms are integrated in the
-    framework to help application development in many different domains.
+    \cite{EMP:09}, we show how a variety of generic algorithms are integrated in the
+    framework to help application development in many different domains, and we show
+    a number of applications highlighting the diverse aspects and use cases of Equalizer.
   \end{abstract}
@@
 Highlight integration of many features into a general framework vs one-off
-research implemenation of one algorithm.
+research implementation of one algorithm. Also point out the various and very
+diverse examples in Section 7, highlighting the variety of application and implementation
+scenarios that can benefit from Equalizer.
 
@@
 
-In~\cite{EMP:09}, we presented the Equalizer parallel rendering framework. This
-paper summarized the work in parallel rendering up to 2009. An extensive
+In~\cite{EMP:09}, we presented the initial {\em Equalizer} parallel rendering framework,
+which summarized the work in parallel rendering up to 2009. An extensive
 Programming and User Guide provides in-depth documentation on using and
 programming Equalizer~\cite{Eilemann:13}. In the following we assume these two
-publications and their references as the basis, and will present the related
+publications and their references as a starting point, and focus on the related
 work published since then.
@@
 intercepting certain OpenGL calls. In contrast to frameworks like Chromium, CGLX
-follows the distribution application approach, whichworks transparently for
+follows the distributed application approach, which works transparently for
 trivial applications, but quickly requires the application developer to address
-the complexities of a distributed application when mutable application state
-needs to be synchronized across processes.
+the complexities of a distributed application when mutable application states
+need to be synchronized across processes.\FIXME{do we need to state how Equalizer
+is different from the new ClusterGL and CGLX?}
 
 On the other hand, software for driving and interacting with tiled display walls
-has received significant attention. Sage~\cite{Sage} and Sage~2~\cite{Sage2}
-have received significant attention. Sage is build entirely around the concept
+has received significant attention, including Sage~\cite{Sage} and Sage~2~\cite{Sage2}
+in particular. Sage was built entirely around the concept
 of a shared framebuffer where all content windows are separate applications
-using pixel streaming. It is no longer actively supported. Sage 2 is a complete,
+using pixel streaming but is no longer actively supported. Sage 2 is a complete,
 browser-centric reimplementation where each application is a web application
@@
 Sage, but provide a few native content applications integrated into the display
-servers.
+servers.\FIXME{also here, maybe a differentiation to Sage2 and DC is needed,
+showing the advantages of EQ}
 
@@
 optimizations~\cite{EBAHMP:12}; as well as work queueing~\cite{SPEP:16}.
-
-Various applications and frameworks have used Equalizer for new research in
+Additionally, complex large scale and out-of-core multiresolution
+rendering approaches have been parallelized and implemented with
+Equalizer~\cite{GMBP:10, GEMPG:13}, demonstrating the feasibility of
+the framework to be used with complex rendering algorithms and 3D model
+representations.
+
+Furthermore, various applications and frameworks have used Equalizer for new research in
 visualization, such as the RTNeuron~\cite{HBBES:13}, an OpenGL renderer for
-visualizing simulations of neucortical brain regions; Omegalib~\cite{Omegalib},
+visualizing simulations of neurocortical brain regions; Omegalib~\cite{Omegalib},
 extending Equalizer for the hybrid 2D/3D environments such as the CAVE2;
@@
 \begin{figure}[ht]\center
-  \includegraphics[width=\columnwidth]{EqDocs/Developer/ProgrammingGuide/images/netObject}
+  \includegraphics[width=\columnwidth]{images/netObject}
   \caption{\label{fNetObject}Communication between two Collage objects}

### 2017-04-03 paperIEEE.tex
Tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@39 db7c848b-7755-4df2-b763-8796797afa52

@@
     KAUST running RTNeuron, (b) Immersive HMD with external tracked and
-    untracked views running RTT DeltaGen for virtual car usability studies (c)
+    untracked views running RTT DeltaGen for virtual car usability studies, and (c)
     Cave2 running a molecular visualization build based on Omegalib.}
@@
 
+%------------------------------------------------------------------------------
 \subsection{New Decomposition Modes}
 
-The initial version of Equalizer implemented sort-first (2D), sort-last (DB),
+The initial version of Equalizer implemented the basic sort-first (2D), sort-last (DB),
 stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
-present new decomposition modes and motivate their use case. \fig{fig:compounds}
-provides an overview of the new modes. We describe their configuration in the
-\textit{compound} framework introduced in \cite{EMP:09}.
+present the newly added decomposition modes and motivate their use case. \fig{fig:compounds}
+provides an overview of the new modes. We describe how to setup compound
+descriptions in the framework in \cite{EMP:09}.
 
@@
   \end{subfigure}
-  \caption{New Equalizer task decompositions and their compounds for parallel
-    rendering}
+  \caption{New Equalizer task decomposition modes and their compound descriptions for parallel rendering}
   \label{fig:compounds}
@@
 multi-GPU configuration, the application needs to support running the rendering
-threads asynchronously, as outlined in \sref{sec:threading}. The output frame
-rate of the destination channel is smoothened using a frame rate equalizer
+threads asynchronously, as outlined below in \sref{sec:threading}. The output frame
+rate of the destination channel is smoothened using a {\em frame rate equalizer}
 (\sref{sec:framerateEq}).
@@
 processed by all source channels by polling a server-central queue. Prefetching
-ensures that the task communication overlaps with rendering. As shown in
+ensures that the task communication overlaps with rendering. As shown in~
 \cite{SPEP:16}, these modes provide better performance due to being inherently
@@
 
+%------------------------------------------------------------------------------
 \subsection{Equalizers}
 
-Equalizer are an addition to compound trees. They modify parameters of their
-respective subtree at runtime to optimize one aspect of the decomposition. Due
+{\em Equalizers} are an addition to compound trees. They modify parameters of their
+respective subtree at runtime to dynamically optimize one aspect of the decomposition. Due
 to their nature, they are transparent to application developers, but might have
-application-accessible parameters to tune their behaviour.
+application-accessible parameters to tune their behavior.
+\FIXME{ok, this is a bit tricky to Equalizer as a name for the framework as well as for this, can't we resolve this?}
 
@@
 algorithm:
+
 \begin{compactdesc}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Optimizations}
@@
 but also for the transfer between GPU and CPU. Equalizer supports a variety of
-compression algorithms, from very fast RLE encoding, JPEG compression to YUV
-subsampling on the GPU. These algorithms are implemented as runtime-loaded
+compression algorithms, from very fast RLE encoding, YUV
+subsampling on the GPU to JPEG compression. These algorithms are implemented as runtime-loaded
 plugins, allowing easy extension and customization to application-specific
@@
 at the KAUST visualization laboratory and head-mounted displays
-(\fig{FIG_teaser}). In the following we lay out the features needed support
+(\fig{FIG_teaser}). In the following we lay out the features needed to support
 these installations. All features presented were motivated by application use
@@
 new dynamic focus mode, where the application defines the distance of the focal
-plane from the observer, based on the current view direction af the
+plane from the observer, based on the current view direction of the
 user. Initial experiments show that this is particularly effective for objects
@@
 
+%------------------------------------------------------------------------------
 \subsection{Physical and Logical Visualization Setup}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Runtime Reconfiguration}\label{sec:reconfig}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Automatic Configuration}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Qt Windowing}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Tide Integration}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Sequel}\label{sec:sequel}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Architecture}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Infiniband RDMA}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Reliable Stream Protocol}\label{sec:RSP}
@@
 ordering of the data, and slow receivers will eventually throttle the
-sender. This behaviour is needed to guarantee delivery of data in all
+sender. This behavior is needed to guarantee delivery of data in all
 situations. Pragmatic generic multicast (PGM~\cite{pgm}) provides full ordering,
@@
 
+%------------------------------------------------------------------------------
 \subsection{Distributed, Versioned Objects}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Barriers, Queues and Object Maps}\label{sec:barrier}

### 2017-04-03 paperIEEE.tex
Tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@40 db7c848b-7755-4df2-b763-8796797afa52

@@
 application-accessible parameters to tune their behavior.
-\FIXME{ok, this is a bit tricky to Equalizer as a name for the framework as well as for this, can't we resolve this?}
+\FIXME{ok, this is a bit tricky to have Equalizer as a name for the framework as well as for this component, can't we change this?}
 
@@
 
-The discovery is implemented in a separate library, hwsd (HardWare Service
+The discovery is implemented in a separate library, hwsd\FIXME{need to somehow identify this term in the text,
+UPPERCASE or some special font type} (HardWare Service
 Discovery), which uses a plugin-based approach to discover GPUs for GLX, AGL or
@@
 
+\FIXME{Tide and Deflect and other 'names' should be clearly identifiable in the text.
+We have many of them and sometimes this isn't visually clear in the text.}
 Tide (Tiled interactive display environment) is an improved version of
-DisplayCluster~\cite{DC}, providing touch-based, multi-window user interface for
+DisplayCluster~\cite{DC}, providing a touch-based, multi-window user interface for
 high-resolution tiled display walls. Remote applications receive input events
@@
 window on the wall. In \cite{deflect} we have shown interactive framerates for
-an 24 megapixel resolution over a WAN link. Events received are translated into
+a 24 megapixel resolution over a WAN link. Events received are translated into
 Equalizer events and injected into the normal event flow, allowing seamless
@@
 interfaces. In the following we will outline the RSP protocol and implementation
-and motivate the design decisions. Any defaults given below are for Linux or OS
+as well as motivate the design decisions. Any defaults given below are for Linux or OS
 X, the Windows UDP stack requires different default values which can be found in
@@
 
-Handling a smooth packet flow is critical for performance. RSP actively flow
-control to advance the stream. Each incoming connection actively acknowledges
+Handling a smooth packet flow is critical for performance. \FIXME{RSP actively flow
+control to advance the stream}. Each incoming connection actively acknowledges
 every $n$ (17 by default) packets fully received. The incoming connections shift
@@
 volvis.org and the Visual Human Project.  This work was partially supported by
-the Swiss National Science Foundation Grant 200021-116329/1.
+the Swiss National Science Foundation Grant 200021-116329/1.\FIXME{SNF and Hasler as well as EU grants}
 
@@
 is currently working towards a PhD in Computer Science at the Visualization and
-Multimedia Lab at the University of Zurich.
+MultiMedia Lab at the University of Zurich.
 \end{IEEEbiography}

### 2017-04-03 paperIEEE.tex
Resolve some fixmes
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@41 db7c848b-7755-4df2-b763-8796797afa52

@@
 commercial implementations such as TechViz and MechDyne Conduit continue to
-exist, on the research side only ClusterGL~\cite{NHM:11} has been
-presented. ClusterGL employs the same approach as Chromium, but delivers a
-significantly faster implementation of transparent OpenGL interception and
-distribution for parallel rendering. CGLX~\cite{DK:11} tries to bring parallel
-execution transparently to OpenGL applications, by emulating the GLUT API and
-intercepting certain OpenGL calls. In contrast to frameworks like Chromium, CGLX
-follows the distributed application approach, which works transparently for
-trivial applications, but quickly requires the application developer to address
-the complexities of a distributed application when mutable application states
-need to be synchronized across processes.\FIXME{do we need to state how Equalizer
-is different from the new ClusterGL and CGLX?}
+exist, on the research side only ClusterGL~\cite{NHM:11} has been presented.
+ClusterGL employs the same approach as Chromium, but delivers a significantly
+faster implementation of transparent OpenGL interception and distribution for
+parallel rendering. CGLX~\cite{DK:11} tries to bring parallel execution
+transparently to OpenGL applications, by emulating the GLUT API and intercepting
+certain OpenGL calls. In contrast to frameworks like Chromium and ClusterGL
+which distribute OpenGL calls, CGLX follows the distributed application
+approach. This works transparently for trivial applications, but quickly
+requires the application developer to address the complexities of a distributed
+application, when mutable application state needs to be synchronized across
+processes. For realistic applications, writing parallel applications remains the
+only viable approach for scalable parallel rendering, as shown by the success of
+Paraview, Visit and Equalizer-based applications.
 
 On the other hand, software for driving and interacting with tiled display walls
-has received significant attention, including Sage~\cite{Sage} and Sage~2~\cite{Sage2}
-in particular. Sage was built entirely around the concept
-of a shared framebuffer where all content windows are separate applications
-using pixel streaming but is no longer actively supported. Sage 2 is a complete,
+has received significant attention, including Sage~\cite{Sage} and
+Sage~2~\cite{Sage2} in particular. Sage was built entirely around the concept of
+a shared framebuffer where all content windows are separate applications using
+pixel streaming but is no longer actively supported. Sage 2 is a complete,
 browser-centric reimplementation where each application is a web application
@@
 Sage, but provide a few native content applications integrated into the display
-servers.\FIXME{also here, maybe a differentiation to Sage2 and DC is needed,
-showing the advantages of EQ}
+servers. These solutions implement a scalable display environment and are a
+target display platform for scalable 3D graphics applications.
 
@@
 
-{\em Equalizers} are an addition to compound trees. They modify parameters of their
-respective subtree at runtime to dynamically optimize one aspect of the decomposition. Due
-to their nature, they are transparent to application developers, but might have
-application-accessible parameters to tune their behavior.
-\FIXME{ok, this is a bit tricky to have Equalizer as a name for the framework as well as for this component, can't we change this?}
+{\em Equalizers} are an addition to compound trees. They modify parameters of
+their respective subtree at runtime to dynamically optimize the resource usage,
+by each tuning one aspect of the decomposition. Due to their nature, they are
+transparent to application developers, but might have application-accessible
+parameters to tune their behavior. Resource equalization is the critical
+component for scalable parallel rendering, and therefore the eponym for the
+Equalizer project name. \FIXME{ok, this is a bit tricky to have Equalizer as a
+name for the framework as well as for this component, can't we change this?
+eile: We can't change the product name, and changing the component name will be
+hard. In fact, this is an artefact of the architecture: The runtime balancing of
+resources is the core contribution of Equalizer, and therefore the equalizers
+performing this task are the name-givers for the product. Modified paragraph
+above for clarity.}
 

### 2017-04-03 paperIEEE.tex
More FIXMEs address, typeset projects
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@42 db7c848b-7755-4df2-b763-8796797afa52

@@
     and industry applications. Based on the basic architecture and
-    implementation of the Equalizer parallel rendering framework presented in
+    implementation of the \textsf{Equalizer} parallel rendering framework presented in
     \cite{EMP:09}, we show how a variety of generic algorithms are integrated in the
     framework to help application development in many different domains, and we show
-    a number of applications highlighting the diverse aspects and use cases of Equalizer.
+    a number of applications highlighting the diverse aspects and use cases of \textsf{Equalizer}.
   \end{abstract}
@@
   \vspace{-2mm}
-  \caption{Example Equalizer application use cases: (a) 192 Megapixel CAVE at
+  \caption{Example \textsf{Equalizer} application use cases: (a) 192 Megapixel CAVE at
     KAUST running RTNeuron, (b) Immersive HMD with external tracked and
@@
 diverse examples in Section 7, highlighting the variety of application and implementation
-scenarios that can benefit from Equalizer.
+scenarios that can benefit from \textsf{Equalizer}.
 
 The rest of this paper is structured as follows: First we introduce new related
-work since the introduction of Equalizer. The main body of this paper presents
+work since the introduction of \textsf{Equalizer}. The main body of this paper presents
 new performance features, virtual reality algorithms, usability features to
-build complex applications, main novel features of the underlying Collage
-network library and a quick overview of the main Equalizer-based applications. A
+build complex applications, main novel features of the underlying \textsf{Collage}
+network library and a quick overview of the main \textsf{Equalizer}-based applications. A
 result section presents new experiments not previously published, followed by
@@
 
-In~\cite{EMP:09}, we presented the initial {\em Equalizer} parallel rendering framework,
+In~\cite{EMP:09}, we presented the initial \textsf{Equalizer} parallel rendering framework,
 which summarized the work in parallel rendering up to 2009. An extensive
 Programming and User Guide provides in-depth documentation on using and
-programming Equalizer~\cite{Eilemann:13}. In the following we assume these two
+programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following we assume these two
 publications and their references as a starting point, and focus on the related
@@
 
-The concept of transparent OpenGL interception popularized by WireGL and
-Chromium~\cite{HHNFAKK:02} has received little attention since 2009. While some
-commercial implementations such as TechViz and MechDyne Conduit continue to
-exist, on the research side only ClusterGL~\cite{NHM:11} has been presented.
-ClusterGL employs the same approach as Chromium, but delivers a significantly
-faster implementation of transparent OpenGL interception and distribution for
-parallel rendering. CGLX~\cite{DK:11} tries to bring parallel execution
+The concept of transparent OpenGL interception popularized by \textsf{WireGL}
+and \textsf{Chromium}~\cite{HHNFAKK:02} has received little attention since
+2009. While some commercial implementations such as \textsf{TechViz} and
+\textsf{MechDyne Conduit} continue to exist, on the research side only
+\textsf{ClusterGL}~\cite{NHM:11} has been presented. \textsf{ClusterGL} employs
+the same approach as \textsf{Chromium}, but delivers a significantly faster
+implementation of transparent OpenGL interception and distribution for parallel
+rendering. \textsf{CGLX}~\cite{DK:11} tries to bring parallel execution
 transparently to OpenGL applications, by emulating the GLUT API and intercepting
-certain OpenGL calls. In contrast to frameworks like Chromium and ClusterGL
-which distribute OpenGL calls, CGLX follows the distributed application
-approach. This works transparently for trivial applications, but quickly
-requires the application developer to address the complexities of a distributed
-application, when mutable application state needs to be synchronized across
-processes. For realistic applications, writing parallel applications remains the
-only viable approach for scalable parallel rendering, as shown by the success of
-Paraview, Visit and Equalizer-based applications.
+certain OpenGL calls. In contrast to frameworks like \textsf{Chromium} and
+\textsf{ClusterGL} which distribute OpenGL calls, \textsf{CGLX} follows the
+distributed application approach. This works transparently for trivial
+applications, but quickly requires the application developer to address the
+complexities of a distributed application, when mutable application state needs
+to be synchronized across processes. For realistic applications, writing
+parallel applications remains the only viable approach for scalable parallel
+rendering, as shown by the success of \textsf{Paraview}, \textsf{Visit} and
+\textsf{Equalizer}-based applications.
 
 On the other hand, software for driving and interacting with tiled display walls
-has received significant attention, including Sage~\cite{Sage} and
-Sage~2~\cite{Sage2} in particular. Sage was built entirely around the concept of
-a shared framebuffer where all content windows are separate applications using
-pixel streaming but is no longer actively supported. Sage 2 is a complete,
-browser-centric reimplementation where each application is a web application
-distributed across browser instances. DisplayCluster~\cite{DC}, and its
-continuation Tide~\cite{tide}, also implement the shared framebuffer concept of
-Sage, but provide a few native content applications integrated into the display
-servers. These solutions implement a scalable display environment and are a
-target display platform for scalable 3D graphics applications.
-
-Equalizer itself has received significant attention within the research
+has received significant attention, including \textsf{Sage}~\cite{Sage} and
+\textsf{Sage~2}~\cite{Sage2} in particular. \textsf{Sage} was built entirely
+around the concept of a shared framebuffer where all content windows are
+separate applications using pixel streaming but is no longer actively supported.
+\textsf{Sage 2} is a complete, browser-centric reimplementation where each
+application is a web application distributed across browser instances.
+\textsf{DisplayCluster}~\cite{DC}, and its continuation
+\textsf{Tide}~\cite{tide}, also implement the shared framebuffer concept of
+\textsf{Sage}, but provide a few native content applications integrated into the
+display servers. These solutions implement a scalable display environment and
+are a target display platform for scalable 3D graphics applications.
+
+\textsf{Equalizer} itself has received significant attention within the research
 community. Various algorithms to improve the parallel rendering performance have
@@
 
-Furthermore, various applications and frameworks have used Equalizer for new research in
-visualization, such as the RTNeuron~\cite{HBBES:13}, an OpenGL renderer for
-visualizing simulations of neurocortical brain regions; Omegalib~\cite{Omegalib},
-extending Equalizer for the hybrid 2D/3D environments such as the CAVE2;
+Furthermore, various applications and frameworks have used \textsf{Equalizer}
+for new research in visualization, such as \textsf{RTNeuron}~\cite{HBBES:13}, an
+OpenGL renderer for visualizing simulations of neurocortical brain regions;
+
+\textsf{Omegalib}~\cite{Omegalib}, extending \textsf{Equalizer} for the hybrid
+2D/3D environments such as the CAVE2;
+
 visualization for remote sensing data~\cite{LK:09}, bino.
@@
 
-The initial version of Equalizer implemented the basic sort-first (2D), sort-last (DB),
+The initial version of \textsf{Equalizer} implemented the basic sort-first (2D), sort-last (DB),
 stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
@@
   \end{subfigure}
-  \caption{New Equalizer task decomposition modes and their compound descriptions for parallel rendering}
+  \caption{New \textsf{Equalizer} task decomposition modes and their compound descriptions for parallel rendering}
   \label{fig:compounds}
@@
 the corresponding output. Consequently, this decomposition mode is mostly useful
-for non-interactive movie generation. It is transparent to Equalizer
+for non-interactive movie generation. It is transparent to \textsf{Equalizer}
 applications, but does require the configuration latency to be equal or greater
@@
 load-balanced, as long as there is an insignificant overhead for the render task
-setup. This mode is transparent to Equalizer applications.
+setup. This mode is transparent to \textsf{Equalizer} applications.
 
@@
 on all source channels is very similar due to the interleaved distribution of
-pixels. This functionality is transparent to Equalizer applications, and the
+pixels. This functionality is transparent to \textsf{Equalizer} applications, and the
 default compositing implementation uses the OpenGL stencil buffer to blit pixels
@@
 component for scalable parallel rendering, and therefore the eponym for the
-Equalizer project name. \FIXME{ok, this is a bit tricky to have Equalizer as a
+\textsf{Equalizer} project name. \FIXME{ok, this is a bit tricky to have Equalizer as a
 name for the framework as well as for this component, can't we change this?
 eile: We can't change the product name, and changing the component name will be
-hard. In fact, this is an artefact of the architecture: The runtime balancing of
-resources is the core contribution of Equalizer, and therefore the equalizers
+hard. In fact, this is an artefact of the sw architecture: The runtime balancing
+of resources is the core contribution of Equalizer, and therefore the equalizers
 performing this task are the name-givers for the product. Modified paragraph
@@
 video resizing \cite{MBDM:97}, that is, it maintains a constant framerate by
-adapting the rendering resolution of a fill-limited application. In Equalizer,
+adapting the rendering resolution of a fill-limited application. In \textsf{Equalizer},
 this works by rendering into a source channel (typically on a FBO) separate to
@@
 performance. This not only applies to the well-researched network transfer step,
-but also for the transfer between GPU and CPU. Equalizer supports a variety of
+but also for the transfer between GPU and CPU. \textsf{Equalizer} supports a variety of
 compression algorithms, from very fast RLE encoding, YUV
@@
 large data scenarios. To easily customize the synchronization of all threads on
-a single process, Equalizer implements three threading modes:  Full
+a single process, \textsf{Equalizer} implements three threading modes:  Full
 synchronization, draw synchronization and asynchronous. Note that the execution
@@
 require special attention to support it as a first-class citizen in a generic
-parallel rendering framework. Equalizer has been used in many virtual reality
+parallel rendering framework. \textsf{Equalizer} has been used in many virtual reality
 installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
@@
 Head tracking is the minimal feature needed to support immersive
-installations. Equalizer does support multiple, independent tracked views
+installations. \textsf{Equalizer} does support multiple, independent tracked views
 through the observer abstraction introduced in \sref{sec:observer}. Built-in
@@
 are not symmetric, and by measuring individual users a more precise frustum can
-be computed. Equalizer supports this through the optional configuration (file or
+be computed. \textsf{Equalizer} supports this through the optional configuration (file or
 programmatically) of individual 3D positions for the left and right eye.
@@
 software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
-swap, a Collage barrier (\sref{sec:barrier}) to synchronize all segments, the
+swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all segments, the
 swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
@@
 on a model, in the sense used by the MVC pattern. The view class is used by
-Equalizer applications to define view-specific data for rendering, e.g., a
+\textsf{Equalizer} applications to define view-specific data for rendering, e.g., a
 scene, viewing mode or camera. The application process manages this data, and
@@
 may involve a different set of resources after the change, including the launch
-and exit of render client processes. Equalizer solves this through a
+and exit of render client processes. \textsf{Equalizer} solves this through a
 reconfiguration step at the beginning of each rendering frame. Each resource
@@
 associated data distribution, it is a costly operation. To improve performance
-and robustness, Equalizer will flush the rendering pipeline whenever a runtime
+and robustness, \textsf{Equalizer} will flush the rendering pipeline whenever a runtime
 reconfiguration is detected, before executing the change. This implementation
@@
 
-The discovery is implemented in a separate library, hwsd\FIXME{need to somehow identify this term in the text,
-UPPERCASE or some special font type} (HardWare Service
+The discovery is implemented in a separate library, \textsf{hwsd} (HardWare Service
 Discovery), which uses a plugin-based approach to discover GPUs for GLX, AGL or
@@
 
-The Equalizer server uses the hwsd library to discover local and remote
-resources when an hwsd session name instead of a \textsf{.eqc} configuration
+The \textsf{Equalizer} server uses the hwsd library to discover local and remote
+resources when an hwsd session name instead of a \texttt{.eqc} configuration
 file is provided. A set of standard decomposition modes is configured, which can
@@
 
-This versatile mechanism allows non-experts to configure asnd profit from
+This versatile mechanism allows non-experts to configure and profit from
 multi-GPU workstations and visualization clusters, as well as to provide system
 administrators with the tools to implement easy to use integration with cluster
-schedulers. This feature is transparent to Equalizer application developers.
+schedulers. This feature is transparent to \textsf{Equalizer} application developers.
 
@@
 it imposes a different threading model for window creation and event handling
-compared to Equalizer. In Equalizer, each GPU rendering thread is independently
+compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread is independently
 responsible for creating its windows, receiving the events and eventually
@@
 
-\FIXME{Tide and Deflect and other 'names' should be clearly identifiable in the text.
-We have many of them and sometimes this isn't visually clear in the text.}
-Tide (Tiled interactive display environment) is an improved version of
-DisplayCluster~\cite{DC}, providing a touch-based, multi-window user interface for
-high-resolution tiled display walls. Remote applications receive input events
-and send pixel streams using the Deflect client library. Equalizer includes full
-support, enabling application-transparent integration with Tide. When a Tide
-server is configured, all output channels of a view stream in parallel to one
-window on the wall. In \cite{deflect} we have shown interactive framerates for
-a 24 megapixel resolution over a WAN link. Events received are translated into
-Equalizer events and injected into the normal event flow, allowing seamless
-application integration.
-
-%------------------------------------------------------------------------------
-\subsection{Sequel}\label{sec:sequel}
-
-Sequal is a simplification layer for Equalizer. It is based on the realization
+\textsf{Tide} (Tiled interactive display environment) is an improved version of
+\textsf{DisplayCluster}~\cite{DC}, providing a touch-based, multi-window user
+interface for high-resolution tiled display walls. Remote applications receive
+input events and send pixel streams using the \textsf{Deflect} client library.
+\textsf{Equalizer} includes full support, enabling application-transparent
+integration with \textsf{Tide}. When a \textsf{Tide} server is configured, all
+output channels of a view stream in parallel to one window on the wall. In
+\cite{deflect}, we have shown interactive framerates for a 24 megapixel
+resolution over a WAN link. Events received are translated into
+\textsf{Equalizer} events and injected into the normal event flow, allowing
+seamless application integration.
+
+%------------------------------------------------------------------------------
+\subsection{\textsf{Sequel}}\label{sec:sequel}
+
+\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based on the realization
 that while fully expressive, the verbatim abstraction layer of nodes, pipes,
-windows and channels in Equalizer requires significant learning to fully
+windows and channels in \textsf{Equalizer} requires significant learning to fully
 understand and exploit. In reality, a higher abstraction of \textsf{Application}
-and \textsf{Renderer} is sufficient for most use cases. In Sequel, the
+and \textsf{Renderer} is sufficient for most use cases. In \textsf{Sequel}, the
 application class drives the configuration, and one renderer instance is created
@@
 %------------------------------------------------------------------------------
-\section{The Collage Network Library}
+\section{The \textsf{Collage} Network Library}
 %------------------------------------------------------------------------------
@@
 An important part of writing a parallel rendering application is the
-communication layer between the individual processes. Equalizer relies on the
-Collage network library for its internal operation. Collage furthermore provides
-powerful abstractions for writing Equalizer applications, which are introduced
+communication layer between the individual processes. \textsf{Equalizer} relies on the
+\textsf{Collage} network library for its internal operation. \textsf{Collage} furthermore provides
+powerful abstractions for writing \textsf{Equalizer} applications, which are introduced
 in this section.
@@
 
-Collage provides networking functionality of different abstraction layers,
+\textsf{Collage} provides networking functionality of different abstraction layers,
 gradually providing higher level functionality for the programmer. The main
-primitives in Collage are:
+primitives in \textsf{Collage} are:
 
@@
   wrapped in ICommands and dispatched to command handler methods. A Node is a
-  proxy for a remote LocalNode. The Equalizer Client object is a LocalNode.
+  proxy for a remote LocalNode. The \textsf{Equalizer} Client object is a LocalNode.
 \item[Object] Provides object-oriented, versioned data distribution of C++
@@
   \includegraphics[width=\columnwidth]{images/netObject}
-  \caption{\label{fNetObject}Communication between two Collage objects}
+  \caption{\label{fNetObject}Communication between two \textsf{Collage} objects}
 \end{figure}
@@
 
-Handling a smooth packet flow is critical for performance. \FIXME{RSP actively flow
-control to advance the stream}. Each incoming connection actively acknowledges
-every $n$ (17 by default) packets fully received. The incoming connections shift
-this acknowledgment by their connection identifier to avoid bursts of acks. Any
-missed datagram is actively nack'ed as soon as it is detected. Write connections
-continuously retransmit packets from nack datagrams, and advance their
-\textsf{\_writeBuffers} upon reception of acks from all readers. Furthermore,
-the writer will explicitly request an ack or nack when it runs out of empty
-buffers or finishes its write queue.
+Handling a smooth packet flow is critical for performance. RSP uses an activ
+flow control to advance the byte stream buffered by the implementation. Each
+incoming connection actively acknowledges every $n$ (17 by default) packets
+fully received. The incoming connections shift this acknowledgment by their
+connection identifier to avoid bursts of acks. Any missed datagram is actively
+nack'ed as soon as it is detected. Write connections continuously retransmit
+packets from nack datagrams, and advance their \textsf{\_writeBuffers} upon
+reception of acks from all readers. Furthermore, the writer will explicitly
+request an ack or nack when it runs out of empty buffers or finishes its write
+queue.
 
@@
 models and class hierarchies to represent their application state. As outlined
-in \cite{EMP:09}, the parallel rendering code in an Equalizer application only
+in \cite{EMP:09}, the parallel rendering code in an \textsf{Equalizer} application only
 needs access to the data needed for rendering, as all application logic is
@@
 
-Distributed objects are not required to build Equalizer applications. While most
+Distributed objects are not required to build \textsf{Equalizer} applications. While most
 developers choose to use this abstraction for convenience, we have seen
@@
 
-Distributed objects in Collage provide powerful, object-oriented data
+Distributed objects in \textsf{Collage} provide powerful, object-oriented data
 distribution for C++ objects. They facilitate the implementation of data
@@
 
-\label{sec:Serializable}The Collage Serializable implements one convenient usage
+\label{sec:Serializable}The \textsf{Collage} Serializable implements one convenient usage
 pattern for object data distribution. The \textsf{co::Serializable} data
@@
 have been developed which exploit modern CPU architectures and deliver
-compression rates well above one gigabyte per second. Collage uses the Pression
+compression rates well above one gigabyte per second. \textsf{Collage} uses the Pression
 library~\cite{pression}, which provides an unified interface for a number of
@@
 transparently for the application. Applications can still use data-specific
-compression and feed the compressed data to Collage, typically disabling the
+compression and feed the compressed data to \textsf{Collage}, typically disabling the
 generic compression.
@@
 optimize the transmission time of data. If the contributing nodes share a
-multicast session, and more than one slave instance is mapped, Collage
+multicast session, and more than one slave instance is mapped, \textsf{Collage}
 automatically uses the multicast connection to send the new version
@@
 
-Collage implements a few generic distributed objects which are used by Equalizer
+\textsf{Collage} implements a few generic distributed objects which are used by \textsf{Equalizer}
 and applicatons. A barrier is a distributed barrier primitive used for software
-swap barriers in Equalizer (\sref{sec:swap}). Its implementation follows a
+swap barriers in \textsf{Equalizer} (\sref{sec:swap}). Its implementation follows a
 simple master-slave approach, which has shown to by sufficient for this use
@@
 distributed objects. Master versions can be registered on a central node, e.g.,
-the application node in Equalizer. Consumers, e.g., Equalizer render clients,
+the application node in \textsf{Equalizer}. Consumers, e.g., \textsf{Equalizer} render clients,
 can selectively map the objects they are interested in. Committing the object
@@
 the new version recorded in the object map. This effective design allows data
-distribution with minimal application logic. It is used by Sequel
-(\sref{sec:sequel}) and other Collage applications.
+distribution with minimal application logic. It is used by \textsf{Sequel}
+(\sref{sec:sequel}) and other \textsf{Collage} applications.
 
@@
 
-In this section, we present some major applications built using Equalizer, and
+In this section, we present some major applications built using \textsf{Equalizer}, and
 show how they interact with the framework to solve complex parallel rendering
@@
 out-of-core loading of data relies on a shared file system, only the high-level
-state, e.g., camera position and rendering settings, are shared through Collage
+state, e.g., camera position and rendering settings, are shared through \textsf{Collage}
 objects.
@@
 interactive, high quality rendering of CAD data. The RTT Scale module,
-delivering multi-GPU and distributed execution, is based on Equalizer and
-Collage, and has driven many of the aforementioned features.
+delivering multi-GPU and distributed execution, is based on \textsf{Equalizer} and
+\textsf{Collage}, and has driven many of the aforementioned features.
 
 RTT Scale uses a master-slave execution mode, were a single running Deltagen
-instance can go into ``Scale mode'' at any time by launching an Equalizer
+instance can go into ``Scale mode'' at any time by launching an \textsf{Equalizer}
 configuration. Consequently, the whole internal representation needed for
-rendering is based on a Collage-based data distribution. The rendering clients
+rendering is based on a \textsf{Collage}-based data distribution. The rendering clients
 are separate, smaller applications which will map their scenes during
@@
 as a delta at the beginning of the next frame, following a design pattern
-similar to the Collage \textsf{Serializable}
+similar to the \textsf{Collage} \textsf{Serializable}
 (\sref{sec:Serializable}). Multicast (\sref{sec:RSP}) is used to keep data
@@
 analyses data distribution performance, in particular how the various
-optimizations in Collage effect complex operations in realistic scenarios.
+optimizations in \textsf{Collage} effect complex operations in realistic scenarios.
 
@@
 since it only removes ``blank space'' in the data. The Snappy compression, used
-as default in Collage, achieves the same compression ratio as the LZ variants at
+as default in \textsf{Collage}, achieves the same compression ratio as the LZ variants at
 a much higher speed. The ZStandard compressor has roughly the same speed as the
@@
 
-We would also like to thank all supporters and contributors of Equalizer, most
+We would also like to thank all supporters and contributors of \textsf{Equalizer}, most
 notably RTT, the Blue Brain Project, the University of Siegen, EVL, Dardo, TBD.
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{Stefan Eilemann}
-Stefan is the technical manager of the Visualization Team in the Blue Brain
-Project. His work involves working towards large-scale visualization for
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{Stefan
+Eilemann} Stefan is the technical manager of the Visualization Team in the Blue
+Brain Project. His work involves working towards large-scale visualization for
 Exascale simulations; interactive integration of simulation, analysis and
@@
 allocation of heterogenous resources. He is the CEO and founder of Eyescale and
-also the lead developer of the Equalizer parallel rendering framework. He has
-worked in SGI's Advanced Graphics Division as the technical lead of OpenGL
-Multipipe SDK - a toolkit for scalable, parallel visualization software on
-shared memory supercomputers. He received his masters degree in Computer Science
-from EPFL in 2015, and his Engineering Diploma in Computer Science in 1998. He
-is currently working towards a PhD in Computer Science at the Visualization and
-MultiMedia Lab at the University of Zurich.
+also the lead developer of the \textsf{Equalizer} parallel rendering framework.
+He has worked in SGI's Advanced Graphics Division as the technical lead of
+OpenGL Multipipe SDK - a toolkit for scalable, parallel visualization software
+on shared memory supercomputers. He received his masters degree in Computer
+Science from EPFL in 2015, and his Engineering Diploma in Computer Science in
+1998. He is currently working towards a PhD in Computer Science at the
+Visualization and MultiMedia Lab at the University of Zurich.
 \end{IEEEbiography}

### 2017-04-03 paperIEEE.tex
Tweak
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@43 db7c848b-7755-4df2-b763-8796797afa52
Minor introducation tweak
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@44 db7c848b-7755-4df2-b763-8796797afa52

@@
 OpenGL renderer for visualizing simulations of neurocortical brain regions;
-
-\textsf{Omegalib}~\cite{Omegalib}, extending \textsf{Equalizer} for the hybrid
-2D/3D environments such as the CAVE2;
+\textsf{Omegalib}~\cite{Omegalib}, a framework used in the Cave2, extending
+\textsf{Equalizer} for hybrid collaborative 2D/3D environments;
 
@@
   \includegraphics[width=\textwidth]{images/spikesdist}
-  \caption{\label{fSpikesDist}Object mapping (left )and data synchronization
+  \caption{\label{fSpikesDist}Object mapping (left) and data synchronization
     times (middle, detail view right) for the spike volume data set in

### 2017-04-03 paperIEEE.tex
Add beechnut distribution graphs
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@45 db7c848b-7755-4df2-b763-8796797afa52

@@
 Furthermore, various applications and frameworks have used \textsf{Equalizer}
-for new research in visualization, such as \textsf{RTNeuron}~\cite{HBBES:13}, an
-OpenGL renderer for visualizing simulations of neurocortical brain regions;
-\textsf{Omegalib}~\cite{Omegalib}, a framework used in the Cave2, extending
-\textsf{Equalizer} for hybrid collaborative 2D/3D environments;
-
-visualization for remote sensing data~\cite{LK:09}, bino.
+for new research in visualization. On the application side, \textsf{RTT
+Deltagen}, \textsf{Bino}, \textsf{Livre} and \textsf{RTNeuron}~\cite{HBBES:13}
+are the most mature examples and are presented in \sref{sApplications}. On the
+framework side, \textsf{Omegalib}~\cite{Omegalib}, a framework used in the
+Cave2, made significant progress in integrating 2D collaborative workspaces
+like \textsf{Sage~2} with 3D immersive content.
+
+visualization for remote sensing data~\cite{LK:09}
 
@@
 %----------------------------------------------------------------------
-\section{Applications}
+\section{Applications\label{sApplications}}
 %----------------------------------------------------------------------
@@
 
-\fig{fSpikesDist} has a significant compression ratio, which reflects in the
-results. Futhermore, only a single distributed object is used to transmit the
-volume information in one chunk. Compression for this data is benefitial for
-transmitting data over a 10Gb/s link (880 MB/s vs 660MB/s raw data
-transfer). Buffering has little benefit since the serialization of volume data
-is trivial. Buffered compression makes a huge difference, since the compression
-cost can be amortized over $n$ nodes, reaching raw data transmission rates of
-3.7 GB/s.
+The spike volume data set shown has a significant compression ratio, which
+reflects in the results in \fig{fSpikesDist}. Futhermore, only a single
+distributed object is used to transmit the volume information in one chunk.
+Compression for this data is benefitial for transmitting data over a 10Gb/s link
+even for a single slave process only (970 MB/s vs 570 MB/s raw data transfer).
+Buffering has little benefit since the serialization of volume data is trivial.
+Buffered compression makes a huge difference, since the compression cost can be
+amortized over $n$ nodes, reaching raw data transmission rates of 3.7 GB/s with
+the default Snappy compressor and at best 4.4 GB/s with ZStandard at the lowest
+compression ratio.
 
@@
     times (middle, detail view right) for the spike volume data set in
+    \fig{fVolumes}, left}
+\end{figure*}
+
+\fig{fBeechnutDist}
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\textwidth]{images/beechnutdist}
+  \caption{\label{fBeechnutDist}Object mapping (left) and data synchronization
+    times (middle, detail view right) for the beechnut volume data set in
     \fig{fVolumes}, right}

### 2017-04-03 paperIEEE.tex
Improved results wording
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@47 db7c848b-7755-4df2-b763-8796797afa52

@@
 A critical piece for data distribution performance are the characteristics of
-the data compression algorithms. This microbenchmark compresses a set of binary
-files to precalculate the speed and compression ratio of the various
-engines. \fig{fCompressor} shows the compression and decompression speed in
-gigabyte per second as well as the size of the compressed data. The different
-ZSTD$x$ engines use the ZStandard compression library at level $x$. The
-measurements were performed on a single, isolated node.
+the data compression algorithm. Our microbenchmark compresses a set of binary
+files, precalculating the speed and compression ratio of the various engines.
+\fig{fCompressor} shows the compression and decompression speed in gigabyte per
+second as well as the size of the compressed data relative to the uncompressed
+data. The different ZSTD$x$ engines use the ZStandard compression library at
+compression level $x$. The measurements were performed on a single, isolated
+node.
 
@@
 
-RLE compression has a very low overhead, but does not compress the data will
-since it only removes ``blank space'' in the data. The Snappy compression, used
-as default in \textsf{Collage}, achieves the same compression ratio as the LZ variants at
-a much higher speed. The ZStandard compressor has roughly the same speed as the
-LZ variants at the lowest compression level, but provides significantly better
-compression. At higher compression levels it can improve the compression ratio
-slightly, but at a high cost for the compression speed.
-
-The compression ratio for the models used in the following benchmark deviate
-from this averaged distribution. \fig{fCompressorDetail} shows the compression
-ratios for the triangle and volume data.
+RLE compression has a very low overhead, but does not compress the data, since
+it only removes ``blank space'' in the data. The Snappy compression, used as
+default in \textsf{Collage}, achieves the same compression ratio as the LZ
+variants at a much higher speed. The ZStandard compressor has roughly the same
+speed as the LZ variants at the lowest compression level, but provides
+significantly better compression. At higher compression levels it can improve
+the compression ratio slightly, but at a high cost for the compression speed.
+
+The compression ratio for the models used in the following section deviate from
+this averaged distribution. \fig{fCompressorDetail} shows the compression ratios
+for the triangle and volume data.
 
@@
 reduction. This is due to a high entropy in the data and the dominant use of
-floating point representations. Overall the profile is similar to the generic
-benchmark, at a smaller compression rate.
-
-The volume data on the other hand is sparsely populated and using integer (byte)
-representation, which is easier to compress. The naive RLE implementation
+floating point values. Overall the profile is similar to the generic benchmark,
+at a smaller compression rate.
+
+The volume data on the other hand is sparsely populated and using integer (byte
+and short) values, which is easier to compress. The naive RLE implementation
 already achieves a good compression, showing that the smaller volume contains at
-most 28\% empty space and the bigger volume at most 90\%. Snappy and ZStandard
-can reduce the data much further, reducing the data to a few megabytes.
+most 28\% empty space and the bigger volume at most 43\%. Snappy and ZStandard
+can reduce the spike data much further, reducing the data to a few megabytes.
+Surprisingly, the beechnut data set does not yield significant higher
+compression with the modern Snappy and ZStandard libraries.
 
@@
 real-world applications. We extracted the existing data distribution code in a
-PLY and a volume renderer into a benchmark application to measure the time to
-initially map all the objects on the render client nodes, and to perform a
-commit-sync of the full data set after mapping has been established. All figures
-observe a noticable measurement jitter due to other services running on the
-cluster. The details of the benchmark algorithm can be found in the
-implementation~\cite{eqObjectBench}. We used the same data sets as in the
-previous section. After $8$ processes, a nodes start to run two processes per
-node, which share the CPU and network interface.
+PLY (eqPly) and a volume renderer (Livre) into a benchmark application to
+measure the time to initially map all the objects on the render client nodes,
+and to perform a commit-sync of the full data set after mapping has been
+established. All figures observe a noticable measurement jitter due to other
+services running on the cluster. The details of the benchmark algorithm can be
+found in the implementation~\cite{eqObjectBench}. We used the same data sets as
+in the previous section. After eight processes, a nodes start to run two
+processes per node, which share CPU, memory and network interface bandwidth.
 
@@
 distributes the raw, uncompressed, and unbuffered data, \textbf{compression}
-uses the Snappy compresser to distribute unbuffered data, \textbf{buffered}
+uses the Snappy compressor to distribute unbuffered data, \textbf{buffered}
 reuses uncompressed, serialized data for mappings of other nodes, and
@@
 
-The spike volume data set shown has a significant compression ratio, which
-reflects in the results in \fig{fSpikesDist}. Futhermore, only a single
-distributed object is used to transmit the volume information in one chunk.
-Compression for this data is benefitial for transmitting data over a 10Gb/s link
-even for a single slave process only (970 MB/s vs 570 MB/s raw data transfer).
-Buffering has little benefit since the serialization of volume data is trivial.
-Buffered compression makes a huge difference, since the compression cost can be
-amortized over $n$ nodes, reaching raw data transmission rates of 3.7 GB/s with
-the default Snappy compressor and at best 4.4 GB/s with ZStandard at the lowest
+The volume data sets are each distributed in a single object, serializing the
+raw volume buffer. The spike volume data set has a significant compression
+ratio, which reflects in the results in \fig{fSpikesDist}. Compression for this
+data is benefitial for transmitting data over a 10~Gb/s link even for a single
+slave process only (970 MB/s vs 570 MB/s raw data transfer). Buffering has
+little benefit since the serialization of volume data is trivial. Buffered
+compression makes a huge difference, since the compression cost can be amortized
+over $n$ nodes, reaching raw data transmission rates of 3.7 GB/s with the
+default Snappy compressor and at best 4.4 GB/s with ZStandard at the lowest
 compression ratio.
@@
 
-\fig{fBeechnutDist}
+The distribution of the beechnut data set also behaves as expected
+(\fig{fBeechnutDist}). Due to the larger object size, uncompressed transmission
+is slightly faster compared to the spike data set at 700 MB/s, and compressed
+transmission does not improve the mapping performance, likely due to increased
+memory pressure caused by the large data size. The comparison of the various
+compression engines is consisten with the benchmarks in
+\fig{fCompressorDetail}; RLE, Snappy and the LZ variants are very close to each
+other, and ZSTD1 can provide better performance after four nodes due to the
+better compression ratio.
 
@@
 
-Distribute large ply and raw model to 1..16 nodes using TCP, IB and RSP
+\FIXME{Missing: david on 10GE, spike comparisoon 10GE-IB-RSP}
 

### 2017-04-03 paperIEEE.tex
Network transport results
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@49 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-\author{Stefan~Eilemann, David~Steiner and Renato Pajarola% <-this % stops a space
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the Visualization and MultiMedia Lab, Department of Informatics, University of Z\"urich.
+\author{Stefan~Eilemann, David~Steiner and Renato Pajarola%
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the
+Visualization and MultiMedia Lab, Department of Informatics, University of
+Z\"urich.
 \IEEEcompsocthanksitem S. Eilemann is also with the Blue Brain Project, Ecole
@@
 %------------------------------------------------------------------------------
-\subsection{Infiniband RDMA}
-
-needed? reverse-engr impl
-
-%------------------------------------------------------------------------------
 \subsection{Reliable Stream Protocol}\label{sec:RSP}
@@
 interfaces. In the following we will outline the RSP protocol and implementation
-as well as motivate the design decisions. Any defaults given below are for Linux or OS
-X, the Windows UDP stack requires different default values which can be found in
-the implementation.
-
-Our RSP implementation uses a separate protocol thread for each RSP connection,
+as well as motivate the design decisions. Any defaults given below are for
+Linux or OS X, the Windows UDP stack requires different default values which can
+be found in the implementation.
+
+Our RSP implementation uses a separate protocol thread for each RSP group,
 which handles all reads and writes on the multicast group. This thread
@@
 configurable number of buffers, (1024 by default) of a configurable MTU (1470
-bytes default) which are either free for use or in-use for reading or writing.
-
-%%     1,     // RSP_ERROR_DOWNSCALE
-%%     5,     // RSP_ERROR_UPSCALE
-%%     20,    // RSP_ERROR_MAXSCALE
-%%     3,     // RSP_MIN_SENDRATE_SHIFT
-
-%%     17,    // RSP_ACK_FREQUENCY
+bytes default) which are either free for use or in use for reading or writing.
 
@@
 
-Handling a smooth packet flow is critical for performance. RSP uses an activ
-flow control to advance the byte stream buffered by the implementation. Each
-incoming connection actively acknowledges every $n$ (17 by default) packets
-fully received. The incoming connections shift this acknowledgment by their
-connection identifier to avoid bursts of acks. Any missed datagram is actively
-nack'ed as soon as it is detected. Write connections continuously retransmit
-packets from nack datagrams, and advance their \textsf{\_writeBuffers} upon
-reception of acks from all readers. Furthermore, the writer will explicitly
-request an ack or nack when it runs out of empty buffers or finishes its write
-queue.
+Handling a smooth packet flow is critical for performance. RSP uses active flow
+control to advance the byte stream buffered by the implementation. Each incoming
+connection actively acknowledges every $n$ (17 by default) packets fully
+received. The incoming connections temporarily offset this acknowledgment by
+their connection identifier to avoid bursts of acks. Any missed datagram is
+actively nack'ed as soon as it is detected. Write connections continuously
+retransmit packets from nack datagrams, and advance their
+\textsf{\_writeBuffers} upon reception of acks from all readers. Furthermore,
+the writer will explicitly request an ack or nack when it runs out of empty
+buffers or finishes its write queue. Nack datagrams may contain multiple ranges
+of missed datagrams, as UDP implementations often drop multiple packets at
+once.
 
@@
 
+Bino is a stereoscopic 3D video player capable of running on very large display
+systems. Originally written for the immersive semi-cylindrical projection
+system at the University of Siegen, it has been used in many installations
+thanks to its flexibility of configuration. Bino decodes the video on each
+Equalizer rendering process, and only synchronizes the time step globally,
+therefore providing a scalable solution to video playback.
+
 \subsection{Omegalib}
@@
 a given voxel over a given time. The absolute spike count is renormalized to an
-unsigned byte value (0..255) during creation. Higher densities in the volume
-represent higher spiking activity in the voxel.
+unsigned byte value during creation. Higher densities in the volume represent
+higher spiking activity in the voxel.
 
@@
 The David statue at 2mm resolution is organized in a kD-tree for rendering. Each
-node is a separate distributable object, which has two child node objects. A
-total of $1023$ objects is distributed and synchronized.
-
-The volume data sets are each distributed in a single object, serializing the
+node is a separate distributed object, which has two child node objects. A
+total of $1023$ objects is distributed and synchronized. \fig{fDavidDist} shows
+the distribution times for this data set. Due to limited compressibility of the
+data, the results are relative similar. Compressing the data repeatedly for
+each client leads to decreased performance, since the compression overhead
+cannot be amortized by the decreased transmission time. Buffering data slightly
+improves performance by reducing the CPU and copy overhead. Combining
+compression and buffering leads to the best performance, although only by about
+10\%.
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\textwidth]{images/daviddist}
+  \caption{\label{fDavidDist}Object mapping (left) and data synchronization
+    time (middle, detail view right) for the David 2mm data set in
+    \fig{fDavid2mm}}
+\end{figure*}
+
+During synchronization, data is pushed from the master process to all mapped
+slaves using a unicast connection to each slave. While the results in the
+\fig{fDavidDist} (middle) are relatively close to each other, we can still
+observe how the tradeoff between compression ratio and speed influences overall
+performance. Better, slower compression algorithms lead to better overall better
+performance when amortized over many send operations.
+
+The volume data sets are distributed in a single object, serializing the
 raw volume buffer. The spike volume data set has a significant compression
@@
 over $n$ nodes, reaching raw data transmission rates of 3.7 GB/s with the
-default Snappy compressor and at best 4.4 GB/s with ZStandard at the lowest
-compression ratio.
+default Snappy compressor and at best 4.4 GB/s with ZStandard at level 1.
 
@@
   \caption{\label{fSpikesDist}Object mapping (left) and data synchronization
-    times (middle, detail view right) for the spike volume data set in
+    time (middle, detail view right) for the spike data set in
     \fig{fVolumes}, left}
@@
 memory pressure caused by the large data size. The comparison of the various
-compression engines is consisten with the benchmarks in
+compression engines is consistent with the benchmarks in
 \fig{fCompressorDetail}; RLE, Snappy and the LZ variants are very close to each
@@
   \caption{\label{fBeechnutDist}Object mapping (left) and data synchronization
-    times (middle, detail view right) for the beechnut volume data set in
+    time (middle, detail view right) for the beechnut data set in
     \fig{fVolumes}, right}
@@
 
-\FIXME{Missing: david on 10GE, spike comparisoon 10GE-IB-RSP}
+Finally, we compare data distribution speed using different protocols. In this
+benchmark, data synchronization time of the spike volume data set is measured,
+as in \fig{fSpikesDist} (middle). Buffering is enabled, and compression is
+disabled to focus on the raw network performance. \fig{fIFDist} shows the
+performance using various protocols.
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\textwidth]{images/ifdist}
+  \caption{\label{fIFDist}Object synchronization using different network
+    transports}
+\end{figure*}
+
+Unsurprisingly, TCP over the faster InfiniBand link outperforms the cheaper ten
+gigabit ethernet link by more than a factor of two. Unexpectedly, the native
+RDMA connection performs worse, even though it outperforms IPoIB in a simple
+peer-to-peer connection benchmark. This needs further investigations, but we
+suspect the abstraction of a byte stream connection chosen by \textsf{Collage}
+for historical reasons is not will suited for remote DMA semantics, that is,
+one needs to design the network API around zero-copy semantics with managed
+memory for modern high-speed transports. Both Infiniband connections show
+significant measurement jitter.
+
+Our RSP multicast implementation performs as expected. \textsf{Collage} starts
+using multicast to commit new object versions when two or more clients are
+mapped, since the transmission to a single client is better done using unicast.
+RSP consistently outperforms unicast on the same physical interface and shows
+good scaling behavior (2.5 times slower on 16 vs 2 clients on ethernet, 1.8
+times slower on InfiniBand). The scaling is significantly better when only one
+process per node is used (\fig{fIFDist}, middle: 30\% slower on ethernet, flat
+on InfiniBand). The increased transmission time with multiple clients is caused
+by a higher probability of packet loss, which increases significantly when using
+more than one process per node, as shown in \fig{fIFDist}, right. Infiniband
+outperforms ethernet slightly, but is largely limited by the RSP implementation
+throughput of preparing and queueing the datagrams to and from the protocol
+thread, which we observed in a profiling analysis.
 
@@
 \label{sec:conclusions}
+
+RDMA needs further investigation.
 

### 2017-04-03 paperIEEE.tex
Minor tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@50 db7c848b-7755-4df2-b763-8796797afa52

@@
 % Do not put math or special symbols in the title.
-\title{Equalizer 2.0\\Convergence of a Parallel Rendering Framework}
+\title{Equalizer 2\\Convergence of a Parallel Rendering Framework}
 %
@@
   \begin{abstract}
-    Developing complex, real world graphics applications which utilize multiple GPUs and
-    computers for interactive 3D rendering tasks can be a daunting task, in
-    particular since it requires expertise in distributed systems and parallel
-    rendering in addition to the application domain itself. We present a mature
-    parallel rendering framework which provides a large set of features,
-    algorithms and system integration for a wide range of real-world research
-    and industry applications. Based on the basic architecture and
-    implementation of the \textsf{Equalizer} parallel rendering framework presented in
-    \cite{EMP:09}, we show how a variety of generic algorithms are integrated in the
-    framework to help application development in many different domains, and we show
-    a number of applications highlighting the diverse aspects and use cases of \textsf{Equalizer}.
+    Developing complex, real world graphics applications which utilize multiple
+    GPUs and computers for interactive 3D rendering tasks is a complex task, it
+    requires expertise in distributed systems and parallel rendering in addition
+    to the application domain itself. We present a mature parallel rendering
+    framework which provides a large set of features, algorithms and system
+    integration for a wide range of real-world research and industry
+    applications. Based on the basic architecture and implementation of the
+    \textsf{Equalizer} parallel rendering framework presented in \cite{EMP:09},
+    we show how a variety of generic algorithms are integrated in the framework
+    to help application scalability and development in many different domains,
+    highlight how concrete applications benefit from the diverse aspects and use
+    cases of \textsf{Equalizer}, and show new experimental results for parallel
+    rendering and data distribution.
   \end{abstract}
@@
   \vspace{-2mm}
-  \caption{Example \textsf{Equalizer} application use cases: (a) 192 Megapixel CAVE at
+  \caption{Example Equalizer applications: (a) 192 Megapixel CAVE at
     KAUST running RTNeuron, (b) Immersive HMD with external tracked and
-    untracked views running RTT DeltaGen for virtual car usability studies, and (c)
-    Cave2 running a molecular visualization build based on Omegalib.}
+    untracked views running RTT DeltaGen for virtual car usability studies,
+    and (c) Cave2 running a molecular visualization.}
   \label{FIG_teaser}

### 2017-04-03 paperIEEE.tex
Final Collage results
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@51 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-\begin{figure*}[ht]\center
-  \includegraphics[width=\textwidth]{images/daviddist}
-  \caption{\label{fDavidDist}Object mapping (left) and data synchronization
-    time (middle, detail view right) for the David 2mm data set in
-    \fig{fDavid2mm}}
-\end{figure*}
-
 During synchronization, data is pushed from the master process to all mapped
@@
 
-\begin{figure*}[ht]\center
-  \includegraphics[width=\textwidth]{images/spikesdist}
-  \caption{\label{fSpikesDist}Object mapping (left) and data synchronization
-    time (middle, detail view right) for the spike data set in
-    \fig{fVolumes}, left}
-\end{figure*}
-
 The distribution of the beechnut data set also behaves as expected
@@
 
-\begin{figure*}[ht]\center
-  \includegraphics[width=\textwidth]{images/beechnutdist}
-  \caption{\label{fBeechnutDist}Object mapping (left) and data synchronization
-    time (middle, detail view right) for the beechnut data set in
-    \fig{fVolumes}, right}
-\end{figure*}
-
 Finally, we compare data distribution speed using different protocols. In this
@@
 performance using various protocols.
-
-\begin{figure*}[ht]\center
-  \includegraphics[width=\textwidth]{images/ifdist}
-  \caption{\label{fIFDist}Object synchronization using different network
-    transports}
-\end{figure*}
 
@@
 times slower on InfiniBand). The scaling is significantly better when only one
-process per node is used (\fig{fIFDist}, middle: 30\% slower on ethernet, flat
-on InfiniBand). The increased transmission time with multiple clients is caused
-by a higher probability of packet loss, which increases significantly when using
-more than one process per node, as shown in \fig{fIFDist}, right. Infiniband
-outperforms ethernet slightly, but is largely limited by the RSP implementation
-throughput of preparing and queueing the datagrams to and from the protocol
-thread, which we observed in a profiling analysis.
+process per node is used (\fig{fIFDist}, middle: 30\% slower on ethernet,
+nearly flat on InfiniBand). The increased transmission time with multiple
+clients is caused by a higher probability of packet loss, which increases
+significantly when using more than one process per node, as shown in
+\fig{fIFDist}, right. This figure plots the number of retransmissions
+divided by the number of datagrams. Infiniband outperforms ethernet slightly,
+but is largely limited by the RSP implementation throughput of preparing and
+queueing the datagrams to and from the protocol thread, which we observed in a
+profiling analysis.
+
+\begin{figure*}[ht]\center
+  \includegraphics[height=.19\textheight]{images/daviddist}
+  \caption{\label{fDavidDist}Object mapping (left) and data synchronization
+    time (middle, detail view right) for the David 2mm data set in
+    \fig{fDavid2mm}}
+\end{figure*}
+\begin{figure*}[ht]\center
+  \includegraphics[height=.19\textheight]{images/spikesdist}
+  \caption{\label{fSpikesDist}Object mapping (left) and data synchronization
+    time (middle, detail view right) for the spike data set in
+    \fig{fVolumes}, left}
+\end{figure*}
+\begin{figure*}[ht]\center
+  \includegraphics[height=.19\textheight]{images/beechnutdist}
+  \caption{\label{fBeechnutDist}Object mapping (left) and data synchronization
+    time (middle, detail view right) for the beechnut data set in
+    \fig{fVolumes}, right}
+\end{figure*}
+\begin{figure*}[ht]\center
+  \includegraphics[height=.19\textheight]{images/ifdist}
+  \caption{\label{fIFDist}Object synchronization using different network
+    transports}
+\end{figure*}
+\clearpage
 

### 2017-04-03 paperIEEE.tex
grammar
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@52 db7c848b-7755-4df2-b763-8796797afa52

@@
 optimization for these parallel rendering modes. Our load equalizer is fully
-transparent for application developers, that is, it uses a reactive approach
+transparent for application developers; that is, it uses a reactive approach
 based on past rendering times. This assumes a reasonable frame-to-frame
@@
 The DFR equalizer (\fig{fig:dfr}) provides a functionality similar to dynamic
-video resizing \cite{MBDM:97}, that is, it maintains a constant framerate by
-adapting the rendering resolution of a fill-limited application. In \textsf{Equalizer},
+video resizing \cite{MBDM:97}; that is, it maintains a constant framerate by
+adapting the rendering resolution of a fill-limited application. In
+\textsf{Equalizer},
 this works by rendering into a source channel (typically on a FBO) separate to
@@
 
-In full synchronization, all threads always execute the same frame, that is, the
+In full synchronization, all threads always execute the same frame; that is, the
 render threads are unlocked after \textsf{Node::frameStart}, and the node is
@@
 In draw synchronization, the node thread and all render threads are synchronized
-for all \textsf{frameDraw} operations, that is, \textsf{Node::frameFinish} is
+for all \textsf{frameDraw} operations; that is, \textsf{Node::frameFinish} is
 executed after the last channel is done drawing. This allows the render threads
@@
 user. Initial experiments show that this is particularly effective for objects
-placed within the immersive space, that is, in front of all display segments.
+placed within the immersive space; that is, in front of all display segments.
 
@@
 listener will create one connected RSP connection for each remote member in the
-multicast group, that is, on each node $n$ RSP connection instances exist, $1$
+multicast group; that is, on each node $n$ RSP connection instances exist, $1$
 for the local listener and $n-1$ for all other nodes. A multicast group
@@
 decrease. This has proven experimentally to be more optimal: UDP is often
-rate-limited by switches, that is, packets are discarded regularly and not
+rate-limited by switches; that is, packets are discarded regularly and not
 exceptionally. Only slowly backing of the current send rate helps to stay close
@@
 \item[Instance] The object is versioned and buffered. The instance and delta
-  data are identical, that is, only instance data is serialized. Previous
+  data are identical; that is, only instance data is serialized. Previous
   instance data is saved to be able to map old versions.
@@
   conflicts.
-\item Slave instance versions can only be advanced, that is, \textsf{sync(
+\item Slave instance versions can only be advanced; that is, \textsf{sync(
   version )} with a version smaller than the current version will fail.
@@
 RDMA connection performs worse, even though it outperforms IPoIB in a simple
-peer-to-peer connection benchmark. This needs further investigations, but we
+peer-to-peer connection benchmark. This needs further investigation, but we
 suspect the abstraction of a byte stream connection chosen by \textsf{Collage}
-for historical reasons is not will suited for remote DMA semantics, that is,
+for historical reasons is not well suited for remote DMA semantics; that is,
 one needs to design the network API around zero-copy semantics with managed

### 2017-04-03 paperIEEE.tex
Fix result figure placement
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@53 db7c848b-7755-4df2-b763-8796797afa52
Minor edits.
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@54 db7c848b-7755-4df2-b763-8796797afa52

@@
 % Do not put math or special symbols in the title.
-\title{Equalizer 2\\Convergence of a Parallel Rendering Framework}
+\title{Equalizer 2.0 --\\Convergence of a Parallel Rendering Framework}
 %
@@
 component for scalable parallel rendering, and therefore the eponym for the
-\textsf{Equalizer} project name. \FIXME{ok, this is a bit tricky to have Equalizer as a
-name for the framework as well as for this component, can't we change this?
-eile: We can't change the product name, and changing the component name will be
-hard. In fact, this is an artefact of the sw architecture: The runtime balancing
-of resources is the core contribution of Equalizer, and therefore the equalizers
-performing this task are the name-givers for the product. Modified paragraph
-above for clarity.}
+\textsf{Equalizer} project name. \FIXME{modified paragraph is ok; but we
+still must make sure that the font differs between the Equalizer project and the
+Equalizer components consistently throughout the text}
 
@@
 
+% Was meinst du damit konkret?
+Die Implementierung verwendet einen impliziten Octtree (dh kein “baum” ist im Speicher, die Koordinaten werden anhand der Tiefe & position implizit berechnet), asynchrones laden von der DataSource, asychronen Upload auf die GPU.
+Die “source” Datenstruktur kennt Livre nicht, das hängt von der Implementierung der Datasource ab (zZ UVF, RAW (one octree node), Fivox (3D events-to-volume), Cubist (mesh-to-volume)).
+
+% A VisibleNodesVisitor traverses the Octtree to select the nodes based on the given screen space error (~max pixel/voxel) and their availability.
+How the nodes are selected, in particular when between different neighbour LOD levels is beyond my knowledge.
+Another (?) visitor is used to select desired nodes, which are submitted to the data loader threads, which use the data source to load them, then upload them (async) and add them to the available set.
+
 \subsection{RTT Deltagen}
@@
 
-\begin{figure*}[ht]\center
+\begin{figure*}[p!]\center
   \includegraphics[height=.19\textheight]{images/daviddist}
@@
 \end{figure*}
-\begin{figure*}[ht]\center
+\begin{figure*}[p!]\center
   \includegraphics[height=.19\textheight]{images/spikesdist}
@@
 \end{figure*}
-\begin{figure*}[ht]\center
+\begin{figure*}[p!]\center
   \includegraphics[height=.19\textheight]{images/beechnutdist}
@@
 \end{figure*}
-\begin{figure*}[ht]\center
+\begin{figure*}[p!]\center
   \includegraphics[height=.19\textheight]{images/ifdist}
@@
 \end{figure*}
-\clearpage
 

### 2017-04-03 paperIEEE.tex
Livre
Update Livre description and volume reference.

git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@55 db7c848b-7755-4df2-b763-8796797afa52

--- 2017-04-03 paperIEEE.tex
@@
 component for scalable parallel rendering, and therefore the eponym for the
-\textsf{Equalizer} project name. \FIXME{ok, this is a bit tricky to have Equalizer as a
-name for the framework as well as for this component, can't we change this?
-eile: We can't change the product name, and changing the component name will be
-hard. In fact, this is an artefact of the sw architecture: The runtime balancing
-of resources is the core contribution of Equalizer, and therefore the equalizers
-performing this task are the name-givers for the product. Modified paragraph
-above for clarity.}
+\textsf{Equalizer} project name. \FIXME{modified paragraph is ok; we
+have to ensure that the font differs between the Equalizer project and the
+Equalizer components consistently throughout the text}
 
@@
 
-Livre (Large-scale Interactive Volume Rendering Engine) is an out-of-core
-parallel 4D volume renderer. It supports sort-first decomposition, both for
-scalability as well as for driving large-scale tiled display walls. The
-out-of-core loading of data relies on a shared file system, only the high-level
-state, e.g., camera position and rendering settings, are shared through \textsf{Collage}
-objects.
+Livre (Large-scale Interactive Volume Rendering Engine) is a GPU ray-casting based parallel
+4D volume renderer, implementing state-of-the-art view-dependent level-of-detail rendering (LOD)
+and out-of-core data management~\cite{EHKRW:06}. Hierarchical and out-of-core LOD
+data management is supported by an implicit volume octree, accessed asynchronously
+by the renderer from a data source on a shared file system. Different data sources providing
+an octree conform access to RAW or compressed as well as to implicitly generated volume data
+(e.g. such as from event simulations or surface meshes) can be used.
+
+High-level state information, e.g., camera position and rendering settings, is shared in Livre
+through \textsf{Collage} objects between the parallel applications and rendering threads.
+Sort-first decomposition is efficiently supported through octree traversal and culling both
+for scalability as well as for driving large-scale tiled display walls.
 

### 2017-04-03 paperIEEE.tex
Raster added.
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@56 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+RASTeR~\cite{BGP:09} is an out-of-core and view-dependent real-time multiresolution
+terrain rendering approach using a patch-based restricted quadtree triangulation.
+For load-balanced parallel rendering~\cite{GMBP:10} it exploits fast hierarchical
+view-frustum culling of the level-of-detail (LOD) quadtree for sort-first decomposition, and
+uniform distribution of the visible LOD triangle patches for sort-last decomposition.
+The latter is enabled by a fast traversal of the patch-based restricted quadtree
+triangulation hierarchy which results in a list of selected LOD nodes,
+constituting a view-dependent cut or \emph{front of activated nodes} through the LOD hierarchy.
+Assigning and distributing equally sized segments of this active LOD front to the concurrent
+rendering threads results in a near-optimal sort-last decomposition for each frame.
+
 \subsection{Bino}

### 2017-04-04 paperIEEE.tex
Fixup svn merge fus
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@60 db7c848b-7755-4df2-b763-8796797afa52

@@
 \author{Stefan Eilemann\thanks{email: eilemann@gmail.com} \\ %
-\and Renato Pajarola~\IEEEmembership{Senior Member,~IEEE}\thanks{email: pajarola@acm.org}}
+\and Renato Pajarola~\IEEEmembership{Senior Member,~IEEE}\thanks{email:
+pajarola@acm.org}}
 
@@
 \markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Computer Society Journals}
+{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Computer
+Society Journals}
 % The only time the second header will appear is for the odd numbered pages
@@
 % you desire.
-
-% The publisher's ID mark at the bottom of the page is less important with
-% Computer Society journal papers as those publications place the marks
-% outside of the main text columns and, therefore, unlike regular IEEE
-% journals, the available text space is not reduced by their presence.
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
-% or like this to get the Computer Society new two part style.
-%\IEEEpubid{\makebox[\columnwidth]{\hfill 0000--0000/00/\$00.00~\copyright~2015 IEEE}%
-%\hspace{\columnsep}\makebox[\columnwidth]{Published by the IEEE Computer Society\hfill}}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark (Computer Society jorunal
-% papers don't need this extra clearance.)
 
@@
   \begin{IEEEkeywords}
-    Parallel Rendering, Scalable Visualization, Cluster Graphics, Immersive Environments, Display Walls
+    Parallel Rendering, Scalable Visualization, Cluster Graphics, Immersive
+Environments, Display Walls
   \end{IEEEkeywords}
@@
 research implementation of one algorithm. Also point out the various and very
-diverse examples in Section 7, highlighting the variety of application and implementation
+diverse examples in Section 7, highlighting the variety of application and
+implementation
 scenarios that can benefit from \textsf{Equalizer}.
@@
 The rest of this paper is structured as follows: First we introduce new related
-work since the introduction of \textsf{Equalizer}. The main body of this paper presents
+work since the introduction of \textsf{Equalizer}. The main body of this paper
+presents
 new performance features, virtual reality algorithms, usability features to
-build complex applications, main novel features of the underlying \textsf{Collage}
-network library and a quick overview of the main \textsf{Equalizer}-based applications. A
+build complex applications, main novel features of the underlying
+\textsf{Collage}
+network library and a quick overview of the main \textsf{Equalizer}-based
+applications. A
 result section presents new experiments not previously published, followed by
@@
 
-In~\cite{EMP:09}, we presented the initial \textsf{Equalizer} parallel rendering framework,
+In~\cite{EMP:09}, we presented the initial \textsf{Equalizer} parallel
+rendering framework,
 which summarized the work in parallel rendering up to 2009. An extensive
 Programming and User Guide provides in-depth documentation on using and
-programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following we assume these two
+programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following we assume
+these two
 publications and their references as a starting point, and focus on the related
@@
 
-The initial version of \textsf{Equalizer} implemented the basic sort-first (2D), sort-last (DB),
+The initial version of \textsf{Equalizer} implemented the basic sort-first
+(2D), sort-last (DB),
 stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
-present the newly added decomposition modes and motivate their use case. \fig{fig:compounds}
+present the newly added decomposition modes and motivate their use case.
+\fig{fig:compounds}
 provides an overview of the new modes. We describe how to setup compound
@@
   \end{subfigure}
-  \caption{New \textsf{Equalizer} task decomposition modes and their compound descriptions for parallel rendering}
+  \caption{New \textsf{Equalizer} task decomposition modes and their compound
+descriptions for parallel rendering}
   \label{fig:compounds}
@@
 multi-GPU configuration, the application needs to support running the rendering
-threads asynchronously, as outlined below in \sref{sec:threading}. The output frame
+threads asynchronously, as outlined below in \sref{sec:threading}. The output
+frame
 rate of the destination channel is smoothened using a {\em frame rate equalizer}
@@
 on all source channels is very similar due to the interleaved distribution of
-pixels. This functionality is transparent to \textsf{Equalizer} applications, and the
+pixels. This functionality is transparent to \textsf{Equalizer} applications,
+and the
 default compositing implementation uses the OpenGL stencil buffer to blit pixels
@@
 performance. This not only applies to the well-researched network transfer step,
-but also for the transfer between GPU and CPU. \textsf{Equalizer} supports a variety of
+but also for the transfer between GPU and CPU. \textsf{Equalizer} supports a
+variety of
 compression algorithms, from very fast RLE encoding, YUV
-subsampling on the GPU to JPEG compression. These algorithms are implemented as runtime-loaded
+subsampling on the GPU to JPEG compression. These algorithms are implemented as
+runtime-loaded
 plugins, allowing easy extension and customization to application-specific
@@
 require special attention to support it as a first-class citizen in a generic
-parallel rendering framework. \textsf{Equalizer} has been used in many virtual reality
+parallel rendering framework. \textsf{Equalizer} has been used in many virtual
+reality
 installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
@@
 Head tracking is the minimal feature needed to support immersive
-installations. \textsf{Equalizer} does support multiple, independent tracked views
+installations. \textsf{Equalizer} does support multiple, independent tracked
+views
 through the observer abstraction introduced in \sref{sec:observer}. Built-in
@@
 are not symmetric, and by measuring individual users a more precise frustum can
-be computed. \textsf{Equalizer} supports this through the optional configuration (file or
+be computed. \textsf{Equalizer} supports this through the optional
+configuration (file or
 programmatically) of individual 3D positions for the left and right eye.
@@
 software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
-swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all segments, the
+swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all
+segments, the
 swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
@@
 on a model, in the sense used by the MVC pattern. The view class is used by
-\textsf{Equalizer} applications to define view-specific data for rendering, e.g., a
+\textsf{Equalizer} applications to define view-specific data for rendering,
+e.g., a
 scene, viewing mode or camera. The application process manages this data, and
@@
 associated data distribution, it is a costly operation. To improve performance
-and robustness, \textsf{Equalizer} will flush the rendering pipeline whenever a runtime
+and robustness, \textsf{Equalizer} will flush the rendering pipeline whenever a
+runtime
 reconfiguration is detected, before executing the change. This implementation
@@
 
-The discovery is implemented in a separate library, \textsf{hwsd} (HardWare Service
+The discovery is implemented in a separate library, \textsf{hwsd} (HardWare
+Service
 Discovery), which uses a plugin-based approach to discover GPUs for GLX, AGL or
@@
 administrators with the tools to implement easy to use integration with cluster
-schedulers. This feature is transparent to \textsf{Equalizer} application developers.
+schedulers. This feature is transparent to \textsf{Equalizer} application
+developers.
 
@@
 it imposes a different threading model for window creation and event handling
-compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread is independently
+compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering
+thread is independently
 responsible for creating its windows, receiving the events and eventually
@@
 
-\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based on the realization
+\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based
+on the realization
 that while fully expressive, the verbatim abstraction layer of nodes, pipes,
-windows and channels in \textsf{Equalizer} requires significant learning to fully
+windows and channels in \textsf{Equalizer} requires significant learning to
+fully
 understand and exploit. In reality, a higher abstraction of \textsf{Application}
@@
 An important part of writing a parallel rendering application is the
-communication layer between the individual processes. \textsf{Equalizer} relies on the
-\textsf{Collage} network library for its internal operation. \textsf{Collage} furthermore provides
-powerful abstractions for writing \textsf{Equalizer} applications, which are introduced
+communication layer between the individual processes. \textsf{Equalizer} relies
+on the
+\textsf{Collage} network library for its internal operation. \textsf{Collage}
+furthermore provides
+powerful abstractions for writing \textsf{Equalizer} applications, which are
+introduced
 in this section.
@@
 
-\textsf{Collage} provides networking functionality of different abstraction layers,
+\textsf{Collage} provides networking functionality of different abstraction
+layers,
 gradually providing higher level functionality for the programmer. The main
@@
   wrapped in ICommands and dispatched to command handler methods. A Node is a
-  proxy for a remote LocalNode. The \textsf{Equalizer} Client object is a LocalNode.
+  proxy for a remote LocalNode. The \textsf{Equalizer} Client object is a
+LocalNode.
 \item[Object] Provides object-oriented, versioned data distribution of C++
@@
 models and class hierarchies to represent their application state. As outlined
-in \cite{EMP:09}, the parallel rendering code in an \textsf{Equalizer} application only
+in \cite{EMP:09}, the parallel rendering code in an \textsf{Equalizer}
+application only
 needs access to the data needed for rendering, as all application logic is
@@
 
-Distributed objects are not required to build \textsf{Equalizer} applications. While most
+Distributed objects are not required to build \textsf{Equalizer} applications.
+While most
 developers choose to use this abstraction for convenience, we have seen
@@
 
-\label{sec:Serializable}The \textsf{Collage} Serializable implements one convenient usage
+\label{sec:Serializable}The \textsf{Collage} Serializable implements one
+convenient usage
 pattern for object data distribution. The \textsf{co::Serializable} data
@@
 have been developed which exploit modern CPU architectures and deliver
-compression rates well above one gigabyte per second. \textsf{Collage} uses the Pression
+compression rates well above one gigabyte per second. \textsf{Collage} uses the
+Pression
 library~\cite{pression}, which provides an unified interface for a number of
@@
 transparently for the application. Applications can still use data-specific
-compression and feed the compressed data to \textsf{Collage}, typically disabling the
+compression and feed the compressed data to \textsf{Collage}, typically
+disabling the
 generic compression.
@@
 
-\textsf{Collage} implements a few generic distributed objects which are used by \textsf{Equalizer}
+\textsf{Collage} implements a few generic distributed objects which are used by
+\textsf{Equalizer}
 and applicatons. A barrier is a distributed barrier primitive used for software
-swap barriers in \textsf{Equalizer} (\sref{sec:swap}). Its implementation follows a
+swap barriers in \textsf{Equalizer} (\sref{sec:swap}). Its implementation
+follows a
 simple master-slave approach, which has shown to by sufficient for this use
@@
 distributed objects. Master versions can be registered on a central node, e.g.,
-the application node in \textsf{Equalizer}. Consumers, e.g., \textsf{Equalizer} render clients,
+the application node in \textsf{Equalizer}. Consumers, e.g., \textsf{Equalizer}
+render clients,
 can selectively map the objects they are interested in. Committing the object
@@
 
-In this section, we present some major applications built using \textsf{Equalizer}, and
+In this section, we present some major applications built using
+\textsf{Equalizer}, and
 show how they interact with the framework to solve complex parallel rendering
@@
 interactive, high quality rendering of CAD data. The RTT Scale module,
-delivering multi-GPU and distributed execution, is based on \textsf{Equalizer} and
+delivering multi-GPU and distributed execution, is based on \textsf{Equalizer}
+and
 \textsf{Collage}, and has driven many of the aforementioned features.
@@
 RTT Scale uses a master-slave execution mode, were a single running Deltagen
-instance can go into ``Scale mode'' at any time by launching an \textsf{Equalizer}
+instance can go into ``Scale mode'' at any time by launching an
+\textsf{Equalizer}
 configuration. Consequently, the whole internal representation needed for
-rendering is based on a \textsf{Collage}-based data distribution. The rendering clients
+rendering is based on a \textsf{Collage}-based data distribution. The rendering
+clients
 are separate, smaller applications which will map their scenes during
@@
 analyses data distribution performance, in particular how the various
-optimizations in \textsf{Collage} effect complex operations in realistic scenarios.
+optimizations in \textsf{Collage} effect complex operations in realistic
+scenarios.
 
@@
 volvis.org and the Visual Human Project.  This work was partially supported by
-the Swiss National Science Foundation Grant 200021-116329/1.\FIXME{SNF and Hasler as well as EU grants}
-
-We would also like to thank all supporters and contributors of \textsf{Equalizer}, most
+the Swiss National Science Foundation Grant 200021-116329/1.\FIXME{SNF and
+Hasler as well as EU grants}
+
+We would also like to thank all supporters and contributors of
+\textsf{Equalizer}, most
 notably RTT, the Blue Brain Project, the University of Siegen, EVL, Dardo, TBD.
@@
 % simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{Stefan
-Eilemann} Stefan is the technical manager of the Visualization Team in the Blue
+
+\begin{IEEEbiography}[{
+\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{
+Stefan Eilemann} Stefan is the technical manager of the Visualization Team in
+the Blue
 Brain Project. His work involves working towards large-scale visualization for
@@
 \bibliographystyle{abbrv}
-\bibliography{references}
+\bibliography{references/references}
 

### 2017-08-17 paperIEEE.tex
Fix references
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@61 db7c848b-7755-4df2-b763-8796797afa52
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@62 db7c848b-7755-4df2-b763-8796797afa52
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@63 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+\FIXME{add text}
+
 %----------------------------------------------------------------------
@@
 
-Two graphs: Time to render a 4k, 256-step MSAA image of largest ply model and
-Livre using 1..16 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps),
-tiles, chunks, pixel, subpixel) (+DFR for Livre)
+Two graphs: Time to render a 4k, 32-step MSAA image of the 1mm David model and
+Livre using 1 to 10 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps),
+tiles, chunks, pixel, subpixel) (+DFR for Livre).
+The experiments were performed on a cluster using hexacore Intel Xeon E5-2620v3 CPUs (2.4GHz), 16 GB main memory, 4 GBit/s Ethernet and QDR Infiniband.
+GCC 4.8 has been used with CMake 3.7 release mode settings to compile the software stack.	\FIXME{Not sure about Ethernet.}
 
@@
 \bibliographystyle{abbrv}
-\bibliography{references/references}
+\bibliography{references}
 

### 2017-11-30 paperIEEE.tex
RTNeuron section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@64 db7c848b-7755-4df2-b763-8796797afa52

@@
 \subsection{RTNeuron}
-\cite{HBBES:13}
+
+RTNeuron \cite{HBBES:13} is a scalable real-time rendering tool for the
+visualisation of neuronal simulations based on cable models. It is on
+OpenSceneGraph for data management and Equalizer for parallel rendering, and
+focuses not only on fast rendering times, but also on fast loading times with no
+offline preprocessing. It provides level of detail (LOD) rendering, high quality
+anti-aliasing based on jittered frusta and accumulation during still views,
+interactive modification of the visual representation of neurons in a per-neuron
+basis (full neuron vs. soma only, branch pruning depending on the branch level,
+\dots). RTNeuron implements both sort-first and sort-last rendering with order
+independent transparency.
 

### 2017-11-30 paperIEEE.tex
Omegalib section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@65 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-\FIXME{add text}
+Omegalib \cite{Omegalib} is a software framework built on top of Equalizer that
+facilitates application development for hybrid reality environments such as the
+Cave 2. Hybrid reality enviroments aim to create a seamless 2D/3D environment
+that supports both information-rich analysis (traditionally done on tiled
+display wall) as well as virtual reality simulation exploration (traditionally
+done in VR systems) at a resolution matching human visual acuity. Omegalib
+supports dynamic reconfigurability of the display environment, so that areas of
+the display can be interactively allocated to 2D or 3D workspaces as needed. It
+makes it possible to have multiple immersive applications running on a
+cluster-controlled display system, have different input sources dynamically
+routed to applications, and have rendering results optionally redirected to a
+distributed compositing manager. Omegalib supports pluggable front-ends, to
+simplify the integration of third-party libraries like OpenGL, OpenSceneGraph,
+and the Visualization Toolkit (VTK).
 

### 2017-11-30 paperIEEE.tex
Shorten RSP section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@67 db7c848b-7755-4df2-b763-8796797afa52

@@
 transport. RSP behaves similarly to TCP, it provides full reliability and
-ordering of the data, and slow receivers will eventually throttle the
-sender. This behavior is needed to guarantee delivery of data in all
-situations. Pragmatic generic multicast (PGM~\cite{pgm}) provides full ordering,
-but slow clients will disconnect from the multicast session instead of
-throttling the send rate.
+ordering of the data, and slow receivers will eventually throttle the sender
+through a sliding window algorithm. This behavior is needed to guarantee
+delivery of data in all situations. Pragmatic generic multicast (PGM~\cite{pgm})
+provides full ordering, but slow clients will disconnect from the multicast
+session instead of throttling the send rate.
 
@@
 
-Our RSP implementation uses a separate protocol thread for each RSP group,
-which handles all reads and writes on the multicast group. This thread
-implements the protocol handling and communicates with the application threads
-through thread-safe queues, as shown in \fig{fRSPProto}. These queues contain
-datagrams, which are prefixed by a type-specific header. Each connection has a
-configurable number of buffers, (1024 by default) of a configurable MTU (1470
-bytes default) which are either free for use or in use for reading or writing.
-
-\begin{figure}[ht]\center
-  \includegraphics[width=\columnwidth]{images/rspPackets}
-  \caption{\label{fRSPProto}Reliable Stream Protocol}
-\end{figure}
-
-A client opens a listening multicast RSP connection. Similar to TCP, this
-listener will create one connected RSP connection for each remote member in the
-multicast group; that is, on each node $n$ RSP connection instances exist, $1$
-for the local listener and $n-1$ for all other nodes. A multicast group
-therefore looks like a fully connected set of TCP connections between $n$ nodes
-for the user.
-
-Each member of a multicast group has a group-unique 16 bit identifier, which is
-randomly selected, announced to the group, and if no collisions are notified by
-any other member, confirmed for this connection. This mechanism allows
-addressing of connections with a low protocol overhead. The datagrams used in
-this exchange contain no payload, and they contain two bytes type (HELLO,
-HELLO\_REPLAY, HELLO\_DENY, HELLO\_CONFIRM), two bytes protocol version and two
-bytes connection identifier.
-
-The \textsf{\_appBuffers} queue, which is blocking, contains received UDP
-datagrams on incoming connections, or empty buffers for writing on outgoing
-connections. The \textsf{\_threadBuffers} queue is lock-free and used by the
-protocol thread to store empty buffers on incoming connections, and pre-packaged
-UDP datagrams on outgoing connections. The application thread pushes buffers
-into this queue, consumed by the protocol thread. A datagram is up to 64
-kilobytes (UDP limit, configurable) and contains a header (two bytes type, two
-bytes size, two bytes source identifier, two bytes sequence number) and its
-payload. The sequence number is used for positive (ack) and negative (nack)
-acknowledgments and reordering of packets.
-
-The \textsf{\_writeBuffers std::deque} is only used by the protocol
-thread. It contains already sent, but not yet acknowledged datagrams. Its
-datagrams are resent upon reception of a nack or recycled into
-\textsf{\_appBuffers} when acknowledged by all receivers. Similarly, the
-\textsf{\_recvBuffers std::deque} contains received out-of-order packets already
-received by an incoming connection. Upon each reception it is checked, and
-in-order datagrams are placed into \textsf{\_appBuffers} for consumption by the
-application thread.
+Our RSP implementation uses a separate protocol thread for each RSP group, which
+handles all reads and writes on the multicast group. This thread implements the
+protocol handling and communicates with the application threads through
+thread-safe queues. These queues contain datagrams of the application byte
+stream, prefixed by a type-specific header of at most eight bytes. Each
+connection has a configurable number of buffers, (1024 by default) of a
+configurable MTU (1470 bytes default) which are either free for use or in
+transmission.
 
@@
 actively nack'ed as soon as it is detected. Write connections continuously
-retransmit packets from nack datagrams, and advance their
-\textsf{\_writeBuffers} upon reception of acks from all readers. Furthermore,
-the writer will explicitly request an ack or nack when it runs out of empty
-buffers or finishes its write queue. Nack datagrams may contain multiple ranges
-of missed datagrams, as UDP implementations often drop multiple packets at
-once.
+retransmit packets from nack datagrams, and advance their window upon reception
+of acks from all readers. Furthermore, the writer will explicitly request an ack
+or nack when it runs out of empty buffers or finishes its write queue. Nack
+datagrams may contain multiple ranges of missed datagrams, which is motivated by
+the observation that UDP implementations often drop multiple packets at once.
 
@@
 chosen a more aggressive congestion control algorithm of additive increase and
-decrease. This has proven experimentally to be more optimal: UDP is often
-rate-limited by switches; that is, packets are discarded regularly and not
+additive decrease. This has proven experimentally to be more optimal: UDP is
+often rate-limited by switches; that is, packets are discarded regularly and not
 exceptionally. Only slowly backing of the current send rate helps to stay close

### 2017-11-30 paperIEEE.tex
Start discussion
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@68 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+We have presented a significantly improved generic parallel rendering system
+over the original publication \cite{EMP:09}. While the original publication
+motivated the system design, this publication describes a feature-rich, mature
+implementation capable of supporting a wide variety of use cases. We doubled the
+support for scalable rendering modes, many of which presented for the first
+time. We present new runtime adaptations for better load balance and
+performance, describe how common optimizations are integrated into the system,
+making Equalizer the most generically available scalable rendering system.
+
+Furthermore, we present many new features needed in parallel rendering
+applications, from advanced Virtual Reality support to advanced display system
+setup for 2D/3D integration, auto-configuration and runtime reconfiguration, and
+an advanced network data synchronization library tailored to parallel rendering
+applications. We highlight a few commercial and research applications
+underlining the generic and versatile system implemenation.
+
 RDMA needs further investigation.

### 2017-12-07 paperIEEE.tex
Review related work
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@69 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-Highlight integration of many features into a general framework vs one-off
+\FIXME{Highlight integration of many features into a general framework vs one-off
 research implementation of one algorithm. Also point out the various and very
@@
 implementation
-scenarios that can benefit from \textsf{Equalizer}.
+scenarios that can benefit from \textsf{Equalizer}.}
 
@@
 
-In~\cite{EMP:09}, we presented the initial \textsf{Equalizer} parallel
-rendering framework,
-which summarized the work in parallel rendering up to 2009. An extensive
-Programming and User Guide provides in-depth documentation on using and
-programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following we assume
-these two
-publications and their references as a starting point, and focus on the related
-work published since then.
+In 2009 we presented the initial \textsf{Equalizer} parallel rendering
+framework~\cite{EMP:09}, which introduced the architecture of a generic parallel
+rendering frameworks and summarized the work in parallel rendering. Since then,
+an extensive Programming and User Guide provides in-depth documentation on using
+and programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following related
+work we assume these two publications and their references as a starting point,
+and focus on the new work published since 2009.
 
@@
 rendering, as shown by the success of \textsf{Paraview}, \textsf{Visit} and
-\textsf{Equalizer}-based applications.
+various \textsf{Equalizer}-based applications.
 
@@
 around the concept of a shared framebuffer where all content windows are
-separate applications using pixel streaming but is no longer actively supported.
+separate applications using pixel streaming. It is no longer actively supported.
 \textsf{Sage 2} is a complete, browser-centric reimplementation where each
@@
 \textsf{Sage}, but provide a few native content applications integrated into the
-display servers. These solutions implement a scalable display environment and
+display servers. All these solutions implement a scalable display environment and
 are a target display platform for scalable 3D graphics applications.
@@
 framework side, \textsf{Omegalib}~\cite{Omegalib}, a framework used in the
-Cave2, made significant progress in integrating 2D collaborative workspaces
-like \textsf{Sage~2} with 3D immersive content.
-
-visualization for remote sensing data~\cite{LK:09}
-
-\cite{CKP:12} tiled display wall virtual environment
+Cave2, made significant progress in integrating 2D collaborative workspaces like
+\textsf{Sage~2} with 3D immersive content. Lambers et.al. developed a framework
+for visualizing remote sensing data~\cite{LK:09} on large displays and immersive
+installations based on Equalizer.
 

### 2017-12-14 paperIEEE.tex
review performance section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@70 db7c848b-7755-4df2-b763-8796797afa52

@@
 framework~\cite{EMP:09}, which introduced the architecture of a generic parallel
-rendering frameworks and summarized the work in parallel rendering. Since then,
+rendering framework and summarized the work in parallel rendering. Since then,
 an extensive Programming and User Guide provides in-depth documentation on using
@@
 optimizations to increase the rendering performance for a wide variety of
-algorithms.
+algorithms and use cases.
 
@@
 
-The initial version of \textsf{Equalizer} implemented the basic sort-first
-(2D), sort-last (DB),
-stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the following we
-present the newly added decomposition modes and motivate their use case.
-\fig{fig:compounds}
-provides an overview of the new modes. We describe how to setup compound
-descriptions in the framework in \cite{EMP:09}.
+The initial version of \textsf{Equalizer} implemented the basic sort-first (2D),
+sort-last (DB), stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the
+following we present the newly added decomposition modes and motivate their use
+case, which bring the overall feature set way beyond the typical sort-first and
+sort-last rendering modes. \fig{fig:compounds} provides an overview of the new
+modes. We describe how to setup compound descriptions in \cite{EMP:09}.
 
@@
 applications, but does require the configuration latency to be equal or greater
-than the number of source channels. Furthermore, to work in a multi-threaded,
-multi-GPU configuration, the application needs to support running the rendering
-threads asynchronously, as outlined below in \sref{sec:threading}. The output
-frame
-rate of the destination channel is smoothened using a {\em frame rate equalizer}
+than the number of source channels. Furthermore, to work in multi-threaded,
+multi-GPU configurations, the application needs to support running the rendering
+threads asynchronously (\sref{sec:threading}). The output frame rate of the
+destination channel may be smoothened using a {\em frame rate equalizer}
 (\sref{sec:framerateEq}).
@@
 decomposition which works well for fill-limited applications which are not
-geometry bound. Source channels cannot reduce geometry load through view frustum
-culling, since each source channel has almost the same frustum (only shifted by
-some pixels), but applied to a reduced 2D viewport. However, the fragment load
-on all source channels is very similar due to the interleaved distribution of
+geometry bound, for example direct volume rendering. Source channels cannot
+reduce geometry load through view frustum culling, since each source channel has
+almost the same frustum. However, the fragment load on all source channels is
+reduced linearly and well load-balanced due to the interleaved distribution of
 pixels. This functionality is transparent to \textsf{Equalizer} applications,
-and the
-default compositing implementation uses the OpenGL stencil buffer to blit pixels
-onto the destination channel.
+and the default compositing uses the stencil buffer to blit pixels onto the
+destination channel.
 
@@
 Subpixel compounds (\fig{fig:subpixel}) are similar to pixel compounds, but they
-decompose the work for a single pixel, for example when using multisampling or
-depth of field. Composition typically uses accumulation and averaging of all
-computed fragments for a pixel. This feature is not fully transparent to the
-application, since it needs to adapt (jitter or tilt) the frustum based on the
-iteration executed. Furthermore, subpixel compounds interact with idle image
-refinements, e.g., they can accelerate idle anti-aliasing of a scene when the
-camera and scene are not changed.
+decompose the work for a single pixel, for example when using FSAA or depth of
+field. Composition typically uses accumulation and averaging of all computed
+fragments for a pixel. This feature is not fully transparent to the application,
+since it needs to adapt (jitter or tilt) the frustum based on the iteration
+executed. Furthermore, subpixel compounds interact with idle image refinements,
+e.g., they can accelerate idle anti-aliasing of a scene when the camera and
+scene are not changed.
 
@@
 component for scalable parallel rendering, and therefore the eponym for the
-\textsf{Equalizer} project name. \FIXME{modified paragraph is ok; we
-have to ensure that the font differs between the Equalizer project and the
-Equalizer components consistently throughout the text}
+\textsf{Equalizer} project name.
 
@@
 in a binary tree, and then the algorithm balances the two branches of each level
-by equalizing the integral over the area on each side.
-
-We have implemented various tunable parameters allowing application developers
-to optimize the load balancing based on the characteristics of their rendering
-algorithm:
+by equalizing the integral over the area on each side. We have implemented
+various tunable parameters allowing application developers to optimize the load
+balancing based on the characteristics of their rendering algorithm:
 
 \begin{compactdesc}
-\item[Damping] reduces frame-to-frame oscillations. It is a normalized scalar
-  defining how much of the computed delta from the previous position is
-  applied. The equal load distribution within the region of interest assumed by
-  the load equalizer is in reality not equal, causing the load balancing to
-  overshoot.
+\item[Damping] reduces frame-to-frame oscillations. The equal load distribution
+within the region of interest assumed by the load equalizer is in reality not
+equal, causing the load balancing to overshoot. Damping is a normalized scalar
+defining how much of the computed delta from the previous position is applied to
+the new split.
 \item[Resistance] eliminates small deltas in the load balancing step. This might
-  help the application to cache some computations since the frustum does not
+help the application to cache visibility computations since the frustum does not
 change each frame.
@@
 position, e.g., screen door transparency, which can be eliminated by aligning
-  the boundary to the pixel repetition. Furthermore, some rendering algorithms
-  are sensitive to cache alignments, which can again be exploited by chosing the
+the boundary to the pixel repetition. Furthermore, some rendering algorithms are
+sensitive to cache alignments, which can again be exploited by chosing the
 corresponding boundary.
@@
 Cross-segment load balancing (\fig{fig:vieweq}) addresses the optimal resource
-allocation of $n$ rendering resources to $m$ output channels (with $n\geq
-m$). The view equalizer works in conjunction with load equalizer balancing the
+allocation of $n$ rendering resources to $m$ output channels (with $n\geq m$). A
+view equalizer works in conjunction with load equalizers balancing the
 individual output channels. It monitors the usage of shared source channels
-(across outputs) and activates them to balance the rendering time of all
-outputs. In \cite{EEP:11}, we provide a detailed description and evaluation of
-our algorithm.
+across outputs and activates them to balance the rendering time of all outputs.
+In \cite{EEP:11}, we provide a detailed description and evaluation of our
+algorithm.
 
@@
 The DFR equalizer (\fig{fig:dfr}) provides a functionality similar to dynamic
-video resizing \cite{MBDM:97}; that is, it maintains a constant framerate by
+video resizing \cite{MBDM:97}, that is, it maintains a constant framerate by
 adapting the rendering resolution of a fill-limited application. In
-\textsf{Equalizer},
-this works by rendering into a source channel (typically on a FBO) separate to
-the destination channel, and then scaling the rendering during the transfer
-(typically through an on-GPU texture) to the destination channel. The DFR
-equalizer monitors the rendering performance and accordingly adapts the
-resolution of the source channel and zoom factor for the source to destination
-transfer. If the performance and source channel resolutions allows, this will
-not only subsample, but also supersample the destination channel to reduce
-aliasing artefacts.
+\textsf{Equalizer}, this works by rendering into a source channel (typically on
+a FBO) separate to the destination channel, and then scaling the rendering
+during the transfer (typically through an on-GPU texture) to the destination
+channel. The DFR equalizer monitors the rendering performance and accordingly
+adapts the resolution of the source channel and zoom factor for the source to
+destination transfer. If the performance and source channel resolutions allows,
+this will not only subsample, but also supersample the destination channel to
+reduce aliasing artefacts.
 
@@
 
-The framerate equalizer smoothens the output frame rate of a destination
-channel by instructing the corresponding window to delay its buffer swap to a
-minimum time between swaps. This is regularly used for time-multiplexed
-decompostions, where source channels tend to drift and finish their rendering
-not evenly distributed over time. This equalizer is however fully independent of
-DPlex compounds, and may be used to smoothen irregular application rendering
+The framerate equalizer smoothens the output frame rate of a destination channel
+by instructing the corresponding window to delay its buffer swap to a minimum
+time between swaps. This is regularly used for time-multiplexed decompostions,
+where source channels tend to drift and finish their rendering unevenly
+distributed over time. This equalizer is however fully independent of DPlex
+compounds, and may be used to smoothen irregular application rendering
 algorithms.
@@
 mutable data has to have a copy in each render thread, which is not feasible in
-large data scenarios. To easily customize the synchronization of all threads on
-a single process, \textsf{Equalizer} implements three threading modes:  Full
-synchronization, draw synchronization and asynchronous. Note that the execution
-between nodes is always asynchronous, for up to \textsf{latency} frames.
+many complex applications and large data scenarios. To easily customize the
+synchronization of all threads on a single process, \textsf{Equalizer}
+implements three threading modes:  Full synchronization, draw synchronization
+and asynchronous. Note that the execution between node processes is always
+asynchronous, for up to \textsf{latency} frames.
 
@@
 executed after the last channel is done drawing. This allows the render threads
-to read shared data during their draw operation, but not during
-compositing. Since compositing is typically independent of the rendered data,
-this is the default mode. This mode allows to overlap compositing with data
+to read shared data during their draw operation, but not during compositing.
+Since compositing is typically independent of the rendered data, this is the
+default mode. This mode allows to overlap compositing with rendering and data
 synchronization on multi-GPU machines.
@@
 parallel rendering framework. \textsf{Equalizer} has been used in many virtual
-reality
-installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution C6 CAVE
-at the KAUST visualization laboratory and head-mounted displays
+reality installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution
+C6 CAVE at the KAUST visualization laboratory, and head-mounted displays
 (\fig{FIG_teaser}). In the following we lay out the features needed to support
@@
 
-Head tracking is the minimal feature needed to support immersive
-installations. \textsf{Equalizer} does support multiple, independent tracked
-views
-through the observer abstraction introduced in \sref{sec:observer}. Built-in
-VRPN support enables the direct, application-transparent configuration of a VRPN
-tracker device. Alternatively, an application can provide a $4\times 4$ tracking
-matrix defining the transformation from the canvas coordinate system to the
-observer. Both CAVE-like tracking with fixed projection surfaces and HMD
-tracking modes are implemented.
+Head tracking is the minimal feature needed to support immersive installations.
+\textsf{Equalizer} does support multiple, independent tracked views through the
+observer abstraction (\sref{sec:observer}). Built-in VRPN support enables the
+direct, application-transparent configuration of a VRPN tracker device.
+Alternatively, an application can provide a $4\times 4$ tracking matrix defining
+the transformation from the canvas coordinate system to the observer. Both
+CAVE-like tracking with fixed projection surfaces and HMD tracking with moving
+displays are implemented.
 
@@
 user. Initial experiments show that this is particularly effective for objects
-placed within the immersive space; that is, in front of all display segments.
+placed within the immersive space, that is, in front of the physical displays.
 
@@
 are not symmetric, and by measuring individual users a more precise frustum can
-be computed. \textsf{Equalizer} supports this through the optional
-configuration (file or
-programmatically) of individual 3D positions for the left and right eye.
+be computed. \textsf{Equalizer} supports this through the optional configuration
+(file or programmatically) of individual 3D translations for all eyes relative
+to the tracking matrix.
 
@@
 does potentially involve the start and stop of resources and processes for
-passive stereo or stereo-dependent task decompositions, as described in
-\sref{sec:reconfig}.
+passive stereo or stereo-dependent task decompositions (\sref{sec:reconfig}).
 

### 2017-12-14 paperIEEE.tex
Review Usability section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@71 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-Real-world visualization setups are often complex, and having an abstract
-representation of the display system can simplify the configuration
-process. Real-world applications often have the need to be aware of spatial
-relationship of the display setup, for example to render 2D overlays or to
-configure multiple views on a tiled display wall.
+Real-world visualization setups can be complex. An abstract representation of
+the display system simplifies the configuration process. Applications often have
+the need to be aware of spatial relationship of the display setup, for example
+to render 2D overlays or to configure multiple views on a tiled display wall.
 
@@
 each other. A segment is referencing a channel, which defines the output area of
-this segment, e.g., on a DVI connector connected to a projector.
+this segment, e.g., on a DVI connector connected to a projector. This
+abstraction covers all use cases from simple windows, tiled display walls with
+bezels, to non-planar immersive systems with edge-blending.
 
@@
 all of its segments. A segment can also define a frustum, which overrides the
-canvas frustum, e.g., for non-planar setups such as CAVEs or curved
-screens. These frusta describe a physically-correct display setup for a Virtual
-Reality installation. \label{sec:swap} A canvas may have a software or hardware
-swap barrier, which will synchronize the rendering of all contributing GPUs. The
+canvas frustum, e.g., for non-planar setups such as CAVEs or curved screens.
+These frusta describe a physically-correct display setup for a Virtual Reality
+installation. \label{sec:swap} A canvas may have a software or hardware swap
+barrier, which will synchronize the rendering of all contributing GPUs. The
 software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
 swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all
-segments, the
-swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
-the swap command. The hardware swap barrier is implemented using the
-\textsf{WGL\_NV\_swap\_group} or \textsf{GLX\_NV\_swap\_group} extensions.
-
-On each canvas, the application can display one or more views. A view is a view
-on a model, in the sense used by the MVC pattern. The view class is used by
-\textsf{Equalizer} applications to define view-specific data for rendering,
-e.g., a
-scene, viewing mode or camera. The application process manages this data, and
-the render clients receive it for rendering.
+segments, the swap buffers call followed by a \textsf{glFlush} to ensure timely
+execution of the swap command. The hardware swap barrier is implemented using a
+\textsf{NV\_swap\_group} extension.
+
+On each canvas, the application can display one or more views. A view is used in
+the sense of the MVC pattern. The view class is available to applications to
+define view-specific data for rendering, e.g., a scene, viewing mode or camera.
+The application process manages this data, and the render clients receive it for
+rendering.
 
@@
 
-A view may have a frustum description. The view's frustum overrides frusta
-specified at the canvas or segment level. This is used for non-physically
-correct rendering, e.g., to compare two models side-by-side on a tiled display
-wall. If the view does not specify a frustum, the corresponding destination
-channels will use the physically correct sub-frustum resulting from the
-view/segment intersection.
+% A view may have a frustum description. The view's frustum overrides frusta
+% specified at the canvas or segment level. This is used for non-physically
+% correct rendering, e.g., to compare two models side-by-side on a tiled display
+% wall. If the view does not specify a frustum, the corresponding destination
+% channels will use the physically correct sub-frustum resulting from the
+% view/segment intersection.
 
@@
 An observer looks at one or more views. It is described by the observer position
-in the world and its eye separation. Each observer will have its own stereo
-mode, focus distance and frame loop (framerate). This allows to have untracked
+in the world and its eye separation. Each observer has its own stereo
+mode, focus distance and eye positions. This allows to have untracked
 views and multiple tracked views, e.g., two HMDs, in the same application.
@@
 accordingly. Since starting new resources typically requires object mapping and
-associated data distribution, it is a costly operation. To improve performance
-and robustness, \textsf{Equalizer} will flush the rendering pipeline whenever a
-runtime
-reconfiguration is detected, before executing the change. This implementation
-allows the late join or early exit, transparent to application developers.
+associated data distribution, it is a costly operation.
 
@@
 administrators with the tools to implement easy to use integration with cluster
-schedulers. This feature is transparent to \textsf{Equalizer} application
-developers.
+schedulers. This feature is transparent to application developers.
 
@@
 it imposes a different threading model for window creation and event handling
-compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering
-thread is independently
-responsible for creating its windows, receiving the events and eventually
-dispatching them to the application process main thread. This design is
-motivated by the natural threading model of X11 and WGL, and allows simple
+compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread
+is independently responsible for creating its windows, receiving the events and
+eventually dispatching them to the application process main thread. This design
+is motivated by the natural threading model of X11 and WGL, and allows simple
 sequential semantics between OpenGL rendering and event handling. In contrast,
 Qt requires all windows and QOpenGLContext to be created from the Qt main
-thread. An existing Qt window or context may be moved to a different thread, and
-events are signalled from the main thread.
-
-Challenges in threading model, architecture
+thread. An existing Qt window or context may subsequently be moved to a
+different thread, and events are signalled from the main thread. For Qt windows,
+Equalizer will internally dispatch and handle the window creation from the
+render to the main thread, move the created objects back to the render thread,
+and dispatch Qt signals to the correct render threads.
 
@@
 
-\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based
-on the realization
-that while fully expressive, the verbatim abstraction layer of nodes, pipes,
-windows and channels in \textsf{Equalizer} requires significant learning to
-fully
-understand and exploit. In reality, a higher abstraction of \textsf{Application}
-and \textsf{Renderer} is sufficient for most use cases. In \textsf{Sequel}, the
-application class drives the configuration, and one renderer instance is created
-for each (pipe) render thread. They also provide the natural place to store and
-distribute data. Last, but not least, \textsf{ViewData} provides a convenient
-way to manage multiple views by storing the camera, model or any other
-view-specific information.
+\textsf{Sequel} is a simplification layer for \textsf{Equalizer}. It is based on
+the realization that while fully expressive, the verbatim abstraction layer of
+nodes, pipes, windows and channels in \textsf{Equalizer} requires significant
+learning to fully understand and exploit. In reality, a higher abstraction of
+only \textsf{Application} and \textsf{Renderer} is sufficient for many use
+cases. In \textsf{Sequel}, the application class drives the configuration, and
+one renderer instance is created for each (pipe) render thread. They also
+provide the natural place to store and distribute data. Last, but not least,
+\textsf{ViewData} provides a convenient way to manage multiple views by storing
+the camera, model or any other view-specific information.
 

### 2018-01-11 paperIEEE.tex
Some Introduction
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@72 db7c848b-7755-4df2-b763-8796797afa52

@@
 \author{Stefan~Eilemann, David~Steiner and Renato Pajarola%
-\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the
+\IEEEcompsocitemizethanks{All authors are with the
 Visualization and MultiMedia Lab, Department of Informatics, University of
-Z\"urich.
-\IEEEcompsocthanksitem S. Eilemann is also with the Blue Brain Project, Ecole
-Polytechnique Federale de Lausanne.}}
+Z\"urich.}}
 
@@
 
-\FIXME{Highlight integration of many features into a general framework vs one-off
-research implementation of one algorithm. Also point out the various and very
-diverse examples in Section 7, highlighting the variety of application and
-implementation
-scenarios that can benefit from \textsf{Equalizer}.}
+\FIXME{Highlight integration of many features into a general framework vs
+one-off research implementation of one algorithm. Also point out the various and
+very diverse examples in Section 7, highlighting the variety of application and
+implementation scenarios that can benefit from \textsf{Equalizer}.}
+
+In this systems paper, we present an overview of the integration of research and
+commercial developments into the Equalizer parallel rendering framework first
+presented in \cite{EMP:09}. While individual research using the system has been
+presented since, many important features have not previously been published and
+are presented here. In particular, the integration of parallel rendering
+research, APIs for developing complex distributed applications, and many
+individual features make Equalizer an unique, open source framework to develop
+visualization applications for any type of setup and use case.
 
@@
 Subpixel compounds (\fig{fig:subpixel}) are similar to pixel compounds, but they
-decompose the work for a single pixel, for example when using FSAA or depth of
-field. Composition typically uses accumulation and averaging of all computed
-fragments for a pixel. This feature is not fully transparent to the application,
-since it needs to adapt (jitter or tilt) the frustum based on the iteration
-executed. Furthermore, subpixel compounds interact with idle image refinements,
-e.g., they can accelerate idle anti-aliasing of a scene when the camera and
-scene are not changed.
+decompose the work for a single pixel, for example with Monte-Carlo ray tracing,
+FSAA or depth of field rendering. Composition typically uses accumulation and
+averaging of all computed fragments for a pixel. This feature is not fully
+transparent to the application, since it needs to adapt (jitter or tilt) the
+frustum based on the iteration executed. Furthermore, subpixel compounds
+interact with idle image refinements, e.g., they can accelerate idle
+anti-aliasing of a scene when the camera and scene are not changed.
 
@@
 delivering multi-GPU and distributed execution, is based on \textsf{Equalizer}
-and
-\textsf{Collage}, and has driven many of the aforementioned features.
+and \textsf{Collage}, and has driven many of the aforementioned features.
 
@@
 instance can go into ``Scale mode'' at any time by launching an
-\textsf{Equalizer}
-configuration. Consequently, the whole internal representation needed for
-rendering is based on a \textsf{Collage}-based data distribution. The rendering
-clients
-are separate, smaller applications which will map their scenes during
-startup. At runtime, any change performed in the main application is committed
-as a delta at the beginning of the next frame, following a design pattern
-similar to the \textsf{Collage} \textsf{Serializable}
+\textsf{Equalizer} configuration. Consequently, the whole internal
+representation needed for rendering is based on a \textsf{Collage}-based data
+distribution. The rendering clients are separate, smaller applications which
+will map their scenes during startup. At runtime, any change performed in the
+main application is committed as a delta at the beginning of the next frame,
+following a design pattern similar to the \textsf{Collage Serializable}
 (\sref{sec:Serializable}). Multicast (\sref{sec:RSP}) is used to keep data
@@
 application is used for virtual prototyping and design reviews in front of
-high-resolution display walls or in a CAVE and for virtual prototyping of
-human-machine interactions in a CAVE or using HMDs. For scalability, sort-first
-and tile compounds are used to achieve fast, high-quality rendering, primarily
-for interactive raytracing, both based on CPUs and GPUs. For CPU-based
-raytracing, often Linux-based rendering clients are used with a Windows-based
-application node.
+high-resolution display walls and CAVEs, as well as for virtual prototyping of
+human-machine interactions CAVEs and HMDs (\fig{FIG_teaser}(b)). For
+scalability, sort-first and tile compounds are used to achieve fast,
+high-quality rendering, primarily for interactive raytracing, both based on CPUs
+and GPUs. For CPU-based raytracing, often Linux-based rendering clients are used
+with a Windows-based application node.
 

### 2018-01-11 paperIEEE.tex
Conclusion
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@73 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-In 2009 we presented the initial \textsf{Equalizer} parallel rendering
-framework~\cite{EMP:09}, which introduced the architecture of a generic parallel
-rendering framework and summarized the work in parallel rendering. Since then,
-an extensive Programming and User Guide provides in-depth documentation on using
-and programming \textsf{Equalizer}~\cite{Eilemann:13}. In the following related
-work we assume these two publications and their references as a starting point,
-and focus on the new work published since 2009.
+In 2009 we presented \textsf{Equalizer}~\cite{EMP:09}, which introduced the
+architecture of a generic parallel rendering framework and summarized our work
+in parallel rendering. Since then, an extensive Programming and User Guide
+provides in-depth documentation on using and programming
+\textsf{Equalizer}~\cite{Eilemann:13}. In the following related work we assume
+these two publications and their references as a starting point, and focus on
+the new work published since 2009.
 
@@
 implementation capable of supporting a wide variety of use cases. We doubled the
-support for scalable rendering modes, many of which presented for the first
-time. We present new runtime adaptations for better load balance and
+support for scalable rendering modes, many of which are presented here for the
+first time. We present new runtime adaptations for better load balance and
 performance, describe how common optimizations are integrated into the system,
@@
 
-RDMA needs further investigation.
+With respect to the feature set implemented, we believe that Equalizer now
+covers almost any scenario within its scope. For future work, we would like to
+integrate new research for better scalability, new network implementations in
+particular for modern zero-copy RDMA based transports, as well as extending the
+Sequel abstraction layer for ease of use.
 

### 2018-01-11 paperIEEE.tex
More Intro
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@74 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+We do present novel algorithms for parallel rendering which did not warrant a
+separate publication: pixel and subpixel decompositions, dynamic frame
+resolution, tunable sort-first load-balancing parameters, frame rate
+equalization, thread synchronization modes for multi-threaded rendering, dynamic
+focus distance and asymmetric eye positions for Virtual Reality, parallel pixel
+streaming to tiled display walls, as well as a fully-fledged data distribution
+API with compression and reliable multicast.
+
 The rest of this paper is structured as follows: First we introduce new related
 work since the introduction of \textsf{Equalizer}. The main body of this paper
-presents
-new performance features, virtual reality algorithms, usability features to
-build complex applications, main novel features of the underlying
-\textsf{Collage}
-network library and a quick overview of the main \textsf{Equalizer}-based
-applications. A
-result section presents new experiments not previously published, followed by
-the discussion and conclusion.
+presents new performance features, virtual reality algorithms, usability
+features to build complex applications, main novel features of the underlying
+\textsf{Collage} network library and a quick overview of the main
+\textsf{Equalizer}-based applications. A result section presents new experiments
+not previously published, followed by the discussion and conclusion.
 
@@
 \begin{compactdesc}
+\item[Split Mode] configures the tile layout: horizontal or vertical stripes,
+and 2D, an binary tree split alternating the split axis on each level, resulting
+in compact 2D tiles.
 \item[Damping] reduces frame-to-frame oscillations. The equal load distribution
@@
 by instructing the corresponding window to delay its buffer swap to a minimum
-time between swaps. This is regularly used for time-multiplexed decompostions,
+time between swaps. This is regularly used for time-multiplexed decompositions,
 where source channels tend to drift and finish their rendering unevenly
@@
 We would also like to thank all supporters and contributors of
-\textsf{Equalizer}, most
-notably RTT, the Blue Brain Project, the University of Siegen, EVL, Dardo, TBD.
+\textsf{Equalizer}, most notably RTT, the Blue Brain Project, the University of
+Siegen, the Electronic Visualization Lab at the University of Illinois Chicago
+and Dardo Kleiner.
 

### 2018-01-11 paperIEEE.tex
Canvas/layout example figure and text
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@75 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-\FIXME{Highlight integration of many features into a general framework vs
-one-off research implementation of one algorithm. Also point out the various and
-very diverse examples in Section 7, highlighting the variety of application and
-implementation scenarios that can benefit from \textsf{Equalizer}.}
-
 In this systems paper, we present an overview of the integration of research and
@@
 resolution, tunable sort-first load-balancing parameters, frame rate
-equalization, thread synchronization modes for multi-threaded rendering, dynamic
-focus distance and asymmetric eye positions for Virtual Reality, parallel pixel
-streaming to tiled display walls, as well as a fully-fledged data distribution
-API with compression and reliable multicast.
+equalization, thread synchronization modes for multi-threaded rendering, a
+powerful abstraction for multi-view rendering on arbitrary display setups,
+dynamic focus distance and asymmetric eye positions for Virtual Reality,
+parallel pixel streaming to tiled display walls, as well as a fully-fledged data
+distribution API with compression and reliable multicast.
 
@@
 
-% A view may have a frustum description. The view's frustum overrides frusta
-% specified at the canvas or segment level. This is used for non-physically
-% correct rendering, e.g., to compare two models side-by-side on a tiled display
-% wall. If the view does not specify a frustum, the corresponding destination
-% channels will use the physically correct sub-frustum resulting from the
-% view/segment intersection.
-
 \label{sec:observer}
@@
 views and multiple tracked views, e.g., two HMDs, in the same application.
+
+\begin{figure}[ht]\center
+  \includegraphics[width=\columnwidth]{images/layout}
+  \caption{\label{fLayout}A 2x2 tiled display wall and control host rendering four independent views driven by an eight node visualization cluster}
+\end{figure}
+
+\fig{fLayout} shows RTT Deltagen running an example multi-segment, multi-view
+setup driven by eight rendering nodes. The main tiled display wall canvas uses
+four LCD segments showing one layout with four views, which do not align on the
+segment boundaries. This setup creates seven destination channels. The
+configuration provides multiple, run-time configurable layouts. It is driven
+from the control host on the right, which shows four views, each in their own
+canvas and segment windows with a single-view layout each. One view on the
+control host synchronizes its content (model and camera) with one view on the
+display wall through Collage objects. The control host allows full model
+modifications and all workflows supported within the standalone Deltagen
+application, and all changes are synchronized to the corresponding rendering
+nodes. Since this setup is monoscopic, no head tracking or observers are used.
 
@@
 \textsf{Collage} provides networking functionality of different abstraction
-layers,
-gradually providing higher level functionality for the programmer. The main
-primitives in \textsf{Collage} are:
+layers, gradually providing higher level functionality for the programmer. The
+main primitives in \textsf{Collage} and their relation are shown in
+\fig{fNetObject} and provide:
 
@@
 \end{figure}
-\fig{fNetObject} illustrates this architecture.
 

### 2018-01-11 paperIEEE.tex
Tuning
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@76 db7c848b-7755-4df2-b763-8796797afa52

@@
 parallel rendering framework. \textsf{Equalizer} has been used in many virtual
-reality installations, such as the Cave2 (\cite{FNTTL:13}), the high-resolution
-C6 CAVE at the KAUST visualization laboratory, and head-mounted displays
+reality installations, such as the Cave2 \cite{FNTTL:13}, the high-resolution C6
+CAVE at the KAUST visualization laboratory, and head-mounted displays
 (\fig{FIG_teaser}). In the following we lay out the features needed to support
-these installations. All features presented were motivated by application use
-cases and have been validated with their respective users.
+these installations, motivated by application use cases.
 
@@
 new dynamic focus mode, where the application defines the distance of the focal
-plane from the observer, based on the current view direction of the
-user. Initial experiments show that this is particularly effective for objects
-placed within the immersive space, that is, in front of the physical displays.
+plane from the observer, based on the current view direction of the user.
+Initial experiments show that this provides better viewing comfort, in
+particular for objects placed in front of the physical displays.
 
@@
 Traditional head tracking computes the left and right eye positions by using a
-configurable interocular distance and the tracking matrix. However, human heads
-are not symmetric, and by measuring individual users a more precise frustum can
-be computed. \textsf{Equalizer} supports this through the optional configuration
-(file or programmatically) of individual 3D translations for all eyes relative
-to the tracking matrix.
+configurable interocular distance. However, human heads are not symmetric, and
+by measuring individual users a more precise frustum can be computed.
+\textsf{Equalizer} supports this through the optional configuration (file or
+programmatically) of individual 3D translations for all eyes relative to the
+tracking matrix.
 
@@
 virtual reality. The unit is per view, allowing different scale factors within
-the same configuration. It scales both the specified projection surface as well
+the same application. It scales both the specified projection surface as well
 as the eye position (and therefore separation) to achieve the necessary effect.
@@
 Applications can switch each view between mono and stereo rendering at runtime,
-and can run both monoscopic and stereoscopic views concurrently
-(\fig{FIG_teaser} (b)). While this is trivial for active stereo rendering, it
-does potentially involve the start and stop of resources and processes for
-passive stereo or stereo-dependent task decompositions (\sref{sec:reconfig}).
+and run both monoscopic and stereoscopic views concurrently (\fig{FIG_teaser}
+(b)). This switch does potentially involve the start and stop of resources and
+processes for passive stereo or stereo-dependent task decompositions
+(\sref{sec:reconfig}).
 
@@
 compression rates well above one gigabyte per second. \textsf{Collage} uses the
-Pression
-library~\cite{pression}, which provides an unified interface for a number of
-compression libraries, such as FastLZ~\cite{jesperfast}, Snappy~\cite{snappy}
-and ZStandard~\cite{zstd}. It also contains a custom, virtually zero-cost RLE
-compressor. Pression parallelizes the compression and decompression using
-data decomposition. Note that this compression is generic, and implemented
-transparently for the application. Applications can still use data-specific
-compression and feed the compressed data to \textsf{Collage}, typically
-disabling the
-generic compression.
+Pression library~\cite{pression}, which provides an unified interface for a
+number of compression libraries, such as FastLZ~\cite{jesperfast},
+Snappy~\cite{snappy} and ZStandard~\cite{zstd}. It also contains a custom,
+virtually zero-cost RLE compressor. Pression parallelizes the compression and
+decompression using data decomposition. This compression is generic, and
+implemented transparently for the application. Applications can also use
+data-specific compression.
 

### 2018-01-15 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@82 db7c848b-7755-4df2-b763-8796797afa52

@@
   \begin{abstract}
-    Developing complex, real world graphics applications which utilize multiple
-    GPUs and computers for interactive 3D rendering tasks is a complex task, it
+    Developing complex, real world graphics applications which leverage multiple
+    GPUs and computers for interactive 3D rendering tasks is a complex task. It
     requires expertise in distributed systems and parallel rendering in addition
@@
     \textsf{Equalizer} parallel rendering framework presented in \cite{EMP:09},
-    we show how a variety of generic algorithms are integrated in the framework
+    we show how a variety of generic algorithms can be integrated in the framework
     to help application scalability and development in many different domains,
-    highlight how concrete applications benefit from the diverse aspects and use
-    cases of \textsf{Equalizer}, and show new experimental results for parallel
+    highlighting how concrete applications benefit from the diverse aspects and use
+    cases of \textsf{Equalizer}, and we present new experimental results for parallel
     rendering and data distribution.
@@
 % \IEEEraisesectionheading puts \section within a raised box.
-\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
+\IEEEraisesectionheading{\section{Introduction} \label{sec:introduction}}
 %----------------------------------------------------------------------
@@
 %------------------------------------------------------------------------------
-\section{Performance}
+\section{Performance Features}
 %------------------------------------------------------------------------------
@@
 sort-last rendering modes. \fig{fig:compounds} provides an overview of the new
-modes. We describe how to setup compound descriptions in \cite{EMP:09}.
+modes. The way how compounds are setup is still as given in \cite{EMP:09}.
 
@@
 sort-last rendering, respectively. They decompose the scene into a predefined
-set of fixed-size image tiles or database ranges. These tasks are queued and
+set of fixed-size image tiles or database ranges. These tasks, or work packages are queued and
 processed by all source channels by polling a server-central queue. Prefetching
 ensures that the task communication overlaps with rendering. As shown in~
-\cite{SPEP:16}, these modes provide better performance due to being inherently
+\cite{SPEP:16}, these modes can provide better performance due to being inherently
 load-balanced, as long as there is an insignificant overhead for the render task
@@
 %------------------------------------------------------------------------------
-\section{Virtual Reality}
+\section{Virtual Reality Features}
 %------------------------------------------------------------------------------
@@
 
+%------------------------------------------------------------------------------
 \subsection{Head Tracking}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Dynamic Focus Distance}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Asymmetric Eye Position}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Model Unit}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Runtime Stereo Switch}
@@
 %------------------------------------------------------------------------------
-\section{Usability}
+\section{Usability Features}
 %------------------------------------------------------------------------------
@@
 %----------------------------------------------------------------------
-\section{Applications\label{sApplications}}
+\section{Example Applications} \label{sApplications}
 %----------------------------------------------------------------------
@@
 
+%------------------------------------------------------------------------------
 \subsection{Livre}
@@
 
+%------------------------------------------------------------------------------
 \subsection{RTT Deltagen}
@@
 
+%------------------------------------------------------------------------------
 \subsection{RTNeuron}
@@
 
-\subsection{Raster}
+%------------------------------------------------------------------------------
+\subsection{RASTeR}
 
@@
 
+%------------------------------------------------------------------------------
 \subsection{Bino}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Omegalib}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Decomposition Modes}
@@
 
+%------------------------------------------------------------------------------
 \subsection{Object Distribution}
@@
 Michelangelo Project, Stanford 3D Scanning Repository, Cyberware Inc.,
-volvis.org and the Visual Human Project.  This work was partially supported by
-the Swiss National Science Foundation Grant 200021-116329/1.\FIXME{SNF and
-Hasler as well as EU grants}
+volvis.org and the Visual Human Project. 
+%
+This work was partially supported by
+the Swiss National Science Foundation Grants 200021-116329 and 200020-129525,
+the Swiss Commission for Technology and Innovation CTI/KTI Project 9394.2 PFES-ES,
+the EU FP7 People Programme (Marie Curie Actions) under REA Grant Agreement n$^{\circ}290227$ and
+a Hasler Stiftung grant (project number $12097$).
 
@@
 % if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{John Doe}
+\begin{IEEEbiographynophoto}{David Steiner}
 Biography text here.
@@
 
-\begin{IEEEbiographynophoto}{Jane Doe}
-Biography text here.
-\end{IEEEbiographynophoto}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/renato}}]{Prof. Dr. Renato Pajarola}
+has been a Professor in computer science at the University of Zurich since 2005, leading the Visualization and MultiMedia Lab (VMML).
+He has previously been an Assistant Professor at the University of California Irvine and a Postdoc at Georgia Tech.
+He has received his Dipl.
+Inf-Ing.
+ETH and Dr.
+sc.
+techn.
+degrees in computer science from the Swiss Federal Institute of Technology (ETH) Zurich in 1994 and 1998 respectively.
+He is a Fellow of the Eurographics Association and a Senior Member of both ACM and IEEE.
+His research interests include real-time 3D graphics, interactive data visualization and geometry processing.
+\end{IEEEbiography}
 

### 2018-01-16 paperIEEE.tex
Fixed author line; \IEEEcompsocthanksitem seems to be necessary.
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@85 db7c848b-7755-4df2-b763-8796797afa52

@@
 %% Author and Affiliation (multiple authors with multiple affiliations)
-\author{Stefan Eilemann\thanks{email: eilemann@gmail.com} \\ %
-\and Renato Pajarola~\IEEEmembership{Senior Member,~IEEE}\thanks{email:
-pajarola@acm.org}}
-
-\author{Stefan~Eilemann, David~Steiner and Renato Pajarola%
-\IEEEcompsocitemizethanks{All authors are with the
+\author{Stefan~Eilemann\thanks{email: eilemann@gmail.com},
+        David~Steiner\thanks{email: steiner@ifi.uzh.ch} \\ %
+        and~Renato~Pajarola\thanks{email:pajarola@acm.org},~\IEEEmembership{Senior Member,~IEEE}%
+\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the
 Visualization and MultiMedia Lab, Department of Informatics, University of
 Z\"urich.}}
+
+%\author{Stefan Eilemann\thanks{email: eilemann@gmail.com} \\ %
+%\and Renato Pajarola~\IEEEmembership{Senior Member,~IEEE}\thanks{email:
+%pajarola@acm.org}}
+
+%\author{Stefan~Eilemann, David~Steiner and Renato Pajarola%
+%\IEEEcompsocitemizethanks{All authors are with the
+%Visualization and MultiMedia Lab, Department of Informatics, University of
+%Z\"urich.}}
 

### 2018-01-16 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@86 db7c848b-7755-4df2-b763-8796797afa52

@@
 research, APIs for developing complex distributed applications, and many
-individual features make Equalizer an unique, open source framework to develop
+individual features make Equalizer a unique, open source framework to develop
 visualization applications for any type of setup and use case.

### 2018-01-17 paperIEEE.tex
Redone introduction.
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@89 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-In this systems paper, we present an overview of the integration of research and
-commercial developments into the Equalizer parallel rendering framework first
-presented in \cite{EMP:09}. While individual research using the system has been
-presented since, many important features have not previously been published and
-are presented here. In particular, the integration of parallel rendering
-research, APIs for developing complex distributed applications, and many
+The Equalizer parallel rendering framework as first presented in~\cite{EMP:09} has
+demonstrated its general versatility and the usefulness of its minimally-intrusive 
+programming design in a variety of applications and projects.
+In particular, the integration of large-scale parallel rendering algorithms,
+APIs for developing complex distributed applications, and many
 individual features make Equalizer a unique, open source framework to develop
-visualization applications for any type of setup and use case.
-
-We do present novel algorithms for parallel rendering which did not warrant a
-separate publication: pixel and subpixel decompositions, dynamic frame
-resolution, tunable sort-first load-balancing parameters, frame rate
+visualization applications for virtually any type of setup and use case.
+While individual applications and case studies using the framework, as well as
+new algorithms and system components extending it have been presented
+since the initial release of Equalizer, many important features and functionalities
+have not previously been published and are presented here.
+In this report, we thus present an updated comprehensive review of the
+integration of research, application use cases and commercial developments
+with respect to Equalizer as well as novel comparative experimental results
+of its scalability features.
+
+In particular, we present novel algorithms for parallel rendering which did not yet
+appear in a separate publication, including pixel and sub-pixel decompositions,
+dynamic frame resolution, tunable sort-first load-balancing parameters, frame-rate
 equalization, thread synchronization modes for multi-threaded rendering, a
 powerful abstraction for multi-view rendering on arbitrary display setups,
-dynamic focus distance and asymmetric eye positions for Virtual Reality,
+dynamic focus distance and asymmetric eye positions for VR,
 parallel pixel streaming to tiled display walls, as well as a fully-fledged data
@@
 
-The rest of this paper is structured as follows: First we introduce new related
-work since the introduction of \textsf{Equalizer}. The main body of this paper
-presents new performance features, virtual reality algorithms, usability
+The remainder of this paper is structured as follows: First we provide an update
+on related work since the introduction of \textsf{Equalizer}. The main body of this paper
+then presents new performance features, VR algorithms, usability
 features to build complex applications, main novel features of the underlying

### 2018-01-18 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@91 db7c848b-7755-4df2-b763-8796797afa52

@@
 To benchmark the data distribution we used two datasets: The david statue at 2mm
-resolution (\fig{fDavid2mm}) and 3D volumes of a spike frequencies of an
+resolution (\fig{fDavid2mm}) and 3D volumes of spike frequencies of an
 electrical simulation of three million neurons (\fig{fVolumes}).
@@
 
-RLE compression has a very low overhead, but does not compress the data, since
-it only removes ``blank space'' in the data. The Snappy compression, used as
+RLE compression has a very low overhead but
+% does not compress the data, since it only
+merely removes ``blank space'' in the data. The Snappy compression, used as
 default in \textsf{Collage}, achieves the same compression ratio as the LZ
@@
 reduction. This is due to a high entropy in the data and the dominant use of
-floating point values. Overall the profile is similar to the generic benchmark,
+floating point values. Overall, the profile is similar to the generic benchmark,
 at a smaller compression rate.
@@
 and short) values, which is easier to compress. The naive RLE implementation
-already achieves a good compression, showing that the smaller volume contains at
+already achieves a good compression rate, showing that the smaller volume contains at
 most 28\% empty space and the bigger volume at most 43\%. Snappy and ZStandard
 can reduce the spike data much further, reducing the data to a few megabytes.
-Surprisingly, the beechnut data set does not yield significant higher
+Surprisingly, the beechnut data set does not yield significantly higher
 compression with the modern Snappy and ZStandard libraries.
@@
 found in the implementation~\cite{eqObjectBench}. We used the same data sets as
-in the previous section. After eight processes, a nodes start to run two
-processes per node, which share CPU, memory and network interface bandwidth.
+in the previous section. After eight processes, a node starts to run two
+processes per node, which share CPU, memory and network interface bandwidth \FIXME{unclear}.
 
@@
 reuses uncompressed, serialized data for mappings of other nodes, and
-\textbf{compression buffered} compresses this buffer. Unbuffered operations need
+\textbf{compression buffered} compresses the buffer holding the data \FIXME{unclear}. Unbuffered operations need
 to reserialize, and potentially recompress, the master object data for each
@@
 the distribution times for this data set. Due to limited compressibility of the
-data, the results are relative similar. Compressing the data repeatedly for
+data, the results are relatively similar. Compressing the data repeatedly for
 each client leads to decreased performance, since the compression overhead
@@
 observe how the tradeoff between compression ratio and speed influences overall
-performance. Better, slower compression algorithms lead to better overall better
+performance. Better, slower compression algorithms lead to improved overall
 performance when amortized over many send operations.
@@
 The volume data sets are distributed in a single object, serializing the
-raw volume buffer. The spike volume data set has a significant compression
-ratio, which reflects in the results in \fig{fSpikesDist}. Compression for this
-data is benefitial for transmitting data over a 10~Gb/s link even for a single
-slave process only (970 MB/s vs 570 MB/s raw data transfer). Buffering has
+raw volume buffer. The Spike volume data set has a significant compression
+ratio, which is reflected by the results in \fig{fSpikesDist}. Compression for this
+data is beneficial for transmitting data over a 10~Gb/s link even for a single
+slave process only (970 MB/s vs 570 MB/s raw data transfer) \FIXME{unclear}. Buffering has
 little benefit since the serialization of volume data is trivial. Buffered
-compression makes a huge difference, since the compression cost can be amortized
+compression makes a significant difference, since the compression cost can be amortized
 over $n$ nodes, reaching raw data transmission rates of 3.7 GB/s with the
@@
 (\fig{fBeechnutDist}). Due to the larger object size, uncompressed transmission
-is slightly faster compared to the spike data set at 700 MB/s, and compressed
+is slightly faster compared to the Spike data set at 700 MB/s, and compressed
 transmission does not improve the mapping performance, likely due to increased
@@
 Finally, we compare data distribution speed using different protocols. In this
-benchmark, data synchronization time of the spike volume data set is measured,
+benchmark, data synchronization time of the Spike volume data set is measured,
 as in \fig{fSpikesDist} (middle). Buffering is enabled, and compression is
@@
   \caption{\label{fSpikesDist}Object mapping (left) and data synchronization
-    time (middle, detail view right) for the spike data set in
+    time (middle, detail view right) for the Spike data set in
     \fig{fVolumes}, left}

### 2018-01-21 paperIEEE.tex
Graphs up to date
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@93 db7c848b-7755-4df2-b763-8796797afa52
Describe tree_equalizer
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@94 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-\subsubsection{Sort-First and Sort-Last Load Equalizer}
-
-Sort-first (\fig{fig:loadeq}) and sort-last load balancing is the most obvious
-optimization for these parallel rendering modes. Our load equalizer is fully
-transparent for application developers; that is, it uses a reactive approach
+\subsubsection{Sort-First and Sort-Last Load Balancing}
+
+Sort-first (\fig{fig:loadeq}) and sort-last load balancing are the most obvious
+optimizations for these parallel rendering modes. Our load equalizers are fully
+transparent for application developers; that is, they use a reactive approach
 based on past rendering times. This assumes a reasonable frame-to-frame
-coherency. Our implementation stores a 2D or 1D grid of the load, mapping the
+coherency. Equalizer implements two different algorithms, a
+\textsf{load\_equalizer} and a \textsf{tree\_equalizer}. The result section
+provides some evidence on the strengths and weaknesses of both algorithms.
+
+The \textsf{load\_equalizer} stores a 2D or 1D grid of the load, mapping the
 load of each channel. The load is stored in normalized 2D/1D coordinates using
@@
 in a binary tree, and then the algorithm balances the two branches of each level
-by equalizing the integral over the area on each side. We have implemented
-various tunable parameters allowing application developers to optimize the load
-balancing based on the characteristics of their rendering algorithm:
+by equalizing the integral over the cost area map on each side.
+
+The \textsf{tree\_equalizer} also uses a binary tree for recursive load
+balancing. It computes the accumulated render time on all nodes of the tree, and
+uses this to allocate an equal render time to each subtree. It makes no
+assumption of the load distribution in 2D or 1D space, it only tries to correct
+the imbalance in render time.
+
+We have implemented various tunable parameters for both equalizers allowing
+application developers to optimize the load balancing based on the
+characteristics of their rendering algorithm:
 

### 2018-01-25 paperIEEE.tex
Add compound result figures
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@95 db7c848b-7755-4df2-b763-8796797afa52
Move large figures after main text
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@96 db7c848b-7755-4df2-b763-8796797afa52
Address FIXMEs
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@97 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-RLE compression has a very low overhead but
-% does not compress the data, since it only
-merely removes ``blank space'' in the data. The Snappy compression, used as
-default in \textsf{Collage}, achieves the same compression ratio as the LZ
-variants at a much higher speed. The ZStandard compressor has roughly the same
-speed as the LZ variants at the lowest compression level, but provides
-significantly better compression. At higher compression levels it can improve
-the compression ratio slightly, but at a high cost for the compression speed.
+RLE compression has a very low overhead but merely removes ``blank space'' in
+the data. The Snappy compression, used as default in \textsf{Collage}, achieves
+the same compression ratio as the LZ variants at a much higher speed. The
+ZStandard compressor has roughly the same speed as the LZ variants at the lowest
+compression level, but provides significantly better compression. At higher
+compression levels it can improve the compression ratio slightly, but at a high
+cost for the compression speed.
 
@@
 for the triangle and volume data.
-
-\begin{figure*}[ht]\center
-  \includegraphics[width=\textwidth]{images/compressorDetail}
-  \caption{\label{fCompressorDetail}Data compression for ply data (left, David
-    statue 2mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
-\end{figure*}
 
@@
 
-In this section we present how data distribution and synchronization performs in
-real-world applications. We extracted the existing data distribution code in a
-PLY (eqPly) and a volume renderer (Livre) into a benchmark application to
-measure the time to initially map all the objects on the render client nodes,
+In this section we analyse how data distribution and synchronization performs in
+real-world applications. We extracted the existing data distribution code from a
+mesh renderer (eqPly) and a volume renderer (Livre) into a benchmark application
+to measure the time to initially map all the objects on the render client nodes,
 and to perform a commit-sync of the full data set after mapping has been
@@
 services running on the cluster. The details of the benchmark algorithm can be
-found in the implementation~\cite{eqObjectBench}. We used the same data sets as
-in the previous section. After eight processes, a node starts to run two
-processes per node, which share CPU, memory and network interface bandwidth \FIXME{unclear}.
+found in the implementation~\cite{eqObjectBench}. used the same data sets as in
+the previous section. We ran the benchmark on up to eight physical nodes, that
+is, after eight processes nodes start to run two processes per node, which share
+CPU, memory and network interface bandwidth.
 
@@
 uses the Snappy compressor to distribute unbuffered data, \textbf{buffered}
-reuses uncompressed, serialized data for mappings of other nodes, and
-\textbf{compression buffered} compresses the buffer holding the data \FIXME{unclear}. Unbuffered operations need
-to reserialize, and potentially recompress, the master object data for each
-slave node. Each slave instance needs to deserialize and decompress the data.
+reuses uncompressed, serialized data for mappings from multiple nodes, and
+\textbf{compression buffered} reused the compressed buffer for multiple nodes.
+Unbuffered operations need to reserialize, and potentially recompress, the
+master object data for each slave node. Each slave instance needs to deserialize
+and decompress the data.
 
@@
 
-The volume data sets are distributed in a single object, serializing the
-raw volume buffer. The Spike volume data set has a significant compression
-ratio, which is reflected by the results in \fig{fSpikesDist}. Compression for this
+The volume data sets are distributed in a single object, serializing the raw
+volume buffer. The Spike volume data set has a significant compression ratio,
+which is reflected by the results in \fig{fSpikesDist}. Compression for this
 data is beneficial for transmitting data over a 10~Gb/s link even for a single
-slave process only (970 MB/s vs 570 MB/s raw data transfer) \FIXME{unclear}. Buffering has
-little benefit since the serialization of volume data is trivial. Buffered
-compression makes a significant difference, since the compression cost can be amortized
-over $n$ nodes, reaching raw data transmission rates of 3.7 GB/s with the
-default Snappy compressor and at best 4.4 GB/s with ZStandard at level 1.
+slave process only. Buffering has little benefit since the serialization of
+volume data is trivial. Buffered compression makes a significant difference,
+since the compression cost can be amortized over $n$ nodes, reaching raw data
+transmission rates of 3.7 GB/s with the default Snappy compressor and at best
+4.4 GB/s with ZStandard at level 1.
 
@@
 profiling analysis.
+
+%----------------------------------------------------------------------
+\section{Discussion and Conclusion}
+%----------------------------------------------------------------------
+\label{sec:conclusions}
+
+We have presented a significantly improved generic parallel rendering system
+over the original publication \cite{EMP:09}. While the original publication
+motivated the system design, this publication describes a feature-rich, mature
+implementation capable of supporting a wide variety of use cases. We doubled the
+support for scalable rendering modes, many of which are presented here for the
+first time. We present new runtime adaptations for better load balance and
+performance, describe how common optimizations are integrated into the system,
+making Equalizer the most generically available scalable rendering system.
+
+Furthermore, we present many new features needed in parallel rendering
+applications, from advanced Virtual Reality support to advanced display system
+setup for 2D/3D integration, auto-configuration and runtime reconfiguration, and
+an advanced network data synchronization library tailored to parallel rendering
+applications. We highlight a few commercial and research applications
+underlining the generic and versatile system implemenation.
+
+With respect to the feature set implemented, we believe that Equalizer now
+covers almost any scenario within its scope. For future work, we would like to
+integrate new research for better scalability, new network implementations in
+particular for modern zero-copy RDMA based transports, as well as extending the
+Sequel abstraction layer for ease of use.
+
+%----------------------------------------------------------------------
+\appendices
+% use section* for acknowledgment
+\ifCLASSOPTIONcompsoc
+  % The Computer Society usually uses the plural form
+  \section*{Acknowledgments}
+\else
+  % regular IEEE prefers the singular form
+  \section*{Acknowledgment}
+\fi
+%----------------------------------------------------------------------
+We would like to thank and acknowledge the following institutions and projects
+for providing the 3D geometry and volume test data sets: the Digital
+Michelangelo Project, Stanford 3D Scanning Repository, Cyberware Inc.,
+volvis.org and the Visual Human Project.
+%
+This work was partially supported by
+the Swiss National Science Foundation Grants 200021-116329 and 200020-129525,
+the Swiss Commission for Technology and Innovation CTI/KTI Project 9394.2 PFES-ES,
+the EU FP7 People Programme (Marie Curie Actions) under REA Grant Agreement n$^{\circ}290227$ and
+a Hasler Stiftung grant (project number $12097$).
+
+We would also like to thank all supporters and contributors of
+\textsf{Equalizer}, most notably RTT, the Blue Brain Project, the University of
+Siegen, the Electronic Visualization Lab at the University of Illinois Chicago
+and Dardo Kleiner.
+
+% trigger a \newpage just before the given reference
+% number - used to balance the columns on the last page
+% adjust value as needed - may need to be readjusted if
+% the document is modified later
+%\IEEEtriggeratref{8}
+% The "triggered" command can be changed if desired:
+%\IEEEtriggercmd{\enlargethispage{-5in}}
+
+% biography section
+%
+% If you have an EPS/PDF photo (graphicx package needed) extra braces are
+% needed around the contents of the optional argument to biography to prevent
+% the LaTeX parser from getting confused when it sees the complicated
+% \includegraphics command within an optional argument. (You could create
+% your own custom macro containing the \includegraphics command to make things
+% simpler here.)
+
+\begin{IEEEbiography}[{
+\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{
+Stefan Eilemann} Stefan is the technical manager of the Visualization Team in
+the Blue
+Brain Project. His work involves working towards large-scale visualization for
+Exascale simulations; interactive integration of simulation, analysis and
+visualization as well as flexible frameworks for data sharing and dynamic
+allocation of heterogenous resources. He is the CEO and founder of Eyescale and
+also the lead developer of the \textsf{Equalizer} parallel rendering framework.
+He has worked in SGI's Advanced Graphics Division as the technical lead of
+OpenGL Multipipe SDK - a toolkit for scalable, parallel visualization software
+on shared memory supercomputers. He received his masters degree in Computer
+Science from EPFL in 2015, and his Engineering Diploma in Computer Science in
+1998. He is currently working towards a PhD in Computer Science at the
+Visualization and MultiMedia Lab at the University of Zurich.
+\end{IEEEbiography}
+
+% if you will not have a photo at all:
+\begin{IEEEbiographynophoto}{David Steiner}
+Biography text here.
+\end{IEEEbiographynophoto}
+
+% insert where needed to balance the two columns on the last page with
+% biographies
+%\newpage
+
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/renato}}]{Prof. Dr. Renato Pajarola}
+has been a Professor in computer science at the University of Zurich since 2005, leading the Visualization and MultiMedia Lab (VMML).
+He has previously been an Assistant Professor at the University of California Irvine and a Postdoc at Georgia Tech.
+He has received his Dipl.
+Inf-Ing.
+ETH and Dr.
+sc.
+techn.
+degrees in computer science from the Swiss Federal Institute of Technology (ETH) Zurich in 1994 and 1998 respectively.
+He is a Fellow of the Eurographics Association and a Senior Member of both ACM and IEEE.
+His research interests include real-time 3D graphics, interactive data visualization and geometry processing.
+\end{IEEEbiography}
+
+\bibliographystyle{abbrv}
+\bibliography{references}
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\textwidth]{images/compounds}
+  \caption{\label{fCompounds}Compound scalability for polygonal and volume data}
+\end{figure*}
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\textwidth]{images/equalizers}
+  \caption{\label{fEqualizers}Scalability with heterogenous rendering resources}
+\end{figure*}
+
+\begin{figure*}[ht]\center
+  \includegraphics[width=\textwidth]{images/compressorDetail}
+  \caption{\label{fCompressorDetail}Data compression for ply data (left, David
+    statue 2mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
+\end{figure*}
 
@@
 
-%----------------------------------------------------------------------
-\section{Discussion and Conclusion}
-%----------------------------------------------------------------------
-\label{sec:conclusions}
-
-We have presented a significantly improved generic parallel rendering system
-over the original publication \cite{EMP:09}. While the original publication
-motivated the system design, this publication describes a feature-rich, mature
-implementation capable of supporting a wide variety of use cases. We doubled the
-support for scalable rendering modes, many of which are presented here for the
-first time. We present new runtime adaptations for better load balance and
-performance, describe how common optimizations are integrated into the system,
-making Equalizer the most generically available scalable rendering system.
-
-Furthermore, we present many new features needed in parallel rendering
-applications, from advanced Virtual Reality support to advanced display system
-setup for 2D/3D integration, auto-configuration and runtime reconfiguration, and
-an advanced network data synchronization library tailored to parallel rendering
-applications. We highlight a few commercial and research applications
-underlining the generic and versatile system implemenation.
-
-With respect to the feature set implemented, we believe that Equalizer now
-covers almost any scenario within its scope. For future work, we would like to
-integrate new research for better scalability, new network implementations in
-particular for modern zero-copy RDMA based transports, as well as extending the
-Sequel abstraction layer for ease of use.
-
-%----------------------------------------------------------------------
-\appendices
-% use section* for acknowledgment
-\ifCLASSOPTIONcompsoc
-  % The Computer Society usually uses the plural form
-  \section*{Acknowledgments}
-\else
-  % regular IEEE prefers the singular form
-  \section*{Acknowledgment}
-\fi
-%----------------------------------------------------------------------
-We would like to thank and acknowledge the following institutions and projects
-for providing the 3D geometry and volume test data sets: the Digital
-Michelangelo Project, Stanford 3D Scanning Repository, Cyberware Inc.,
-volvis.org and the Visual Human Project.
-%
-This work was partially supported by
-the Swiss National Science Foundation Grants 200021-116329 and 200020-129525,
-the Swiss Commission for Technology and Innovation CTI/KTI Project 9394.2 PFES-ES,
-the EU FP7 People Programme (Marie Curie Actions) under REA Grant Agreement n$^{\circ}290227$ and
-a Hasler Stiftung grant (project number $12097$).
-
-We would also like to thank all supporters and contributors of
-\textsf{Equalizer}, most notably RTT, the Blue Brain Project, the University of
-Siegen, the Electronic Visualization Lab at the University of Illinois Chicago
-and Dardo Kleiner.
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% biography section
-%
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-
-\begin{IEEEbiography}[{
-\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{
-Stefan Eilemann} Stefan is the technical manager of the Visualization Team in
-the Blue
-Brain Project. His work involves working towards large-scale visualization for
-Exascale simulations; interactive integration of simulation, analysis and
-visualization as well as flexible frameworks for data sharing and dynamic
-allocation of heterogenous resources. He is the CEO and founder of Eyescale and
-also the lead developer of the \textsf{Equalizer} parallel rendering framework.
-He has worked in SGI's Advanced Graphics Division as the technical lead of
-OpenGL Multipipe SDK - a toolkit for scalable, parallel visualization software
-on shared memory supercomputers. He received his masters degree in Computer
-Science from EPFL in 2015, and his Engineering Diploma in Computer Science in
-1998. He is currently working towards a PhD in Computer Science at the
-Visualization and MultiMedia Lab at the University of Zurich.
-\end{IEEEbiography}
-
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{David Steiner}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/renato}}]{Prof. Dr. Renato Pajarola}
-has been a Professor in computer science at the University of Zurich since 2005, leading the Visualization and MultiMedia Lab (VMML).
-He has previously been an Assistant Professor at the University of California Irvine and a Postdoc at Georgia Tech.
-He has received his Dipl.
-Inf-Ing.
-ETH and Dr.
-sc.
-techn.
-degrees in computer science from the Swiss Federal Institute of Technology (ETH) Zurich in 1994 and 1998 respectively.
-He is a Fellow of the Eurographics Association and a Senior Member of both ACM and IEEE.
-His research interests include real-time 3D graphics, interactive data visualization and geometry processing.
-\end{IEEEbiography}
-
-\bibliographystyle{abbrv}
-\bibliography{references}
-
 % that's all folks

### 2018-01-29 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@99 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-Two graphs: Time to render a 4k, 32-step MSAA image of the 1mm David model and
-Livre using 1 to 10 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps),
-tiles, chunks, pixel, subpixel) (+DFR for Livre).
-The experiments were performed on a cluster using hexacore Intel Xeon E5-2620v3 CPUs (2.4GHz), 16 GB main memory, 4 GBit/s Ethernet and QDR Infiniband.
-GCC 4.8 has been used with CMake 3.7 release mode settings to compile the software stack.	\FIXME{Not sure about Ethernet.}
+% Time to render a 4k, 32-step MSAA image of the 1mm David model and Livre using 1 to 10 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps), tiles, chunks, pixel, subpixel) (+DFR for Livre).
+
+We conducted performance benchmarks for the decomposition modes on a cluster using hexacore Intel Xeon E5-2620v3 CPUs (2.4GHz), 16 GB main memory, 4 GBit/s Ethernet and QDR Infiniband.
+GCC 4.8 has been used with CMake 3.7 release mode settings to compile the software stack.
+
+We tested the decomposition modes with both volume (\fig{fBenchmarkEVolve}) and polygonal data  (\fig{fBenchmarkEqPly}) and used test scenes that allowed us to finely adjust the rendering load the system has to cope with.
+In both cases the scene comprised of two rows of identical models (the 1mm David model when rendering polygonal data, the $1024^3$ Flower model when rendering volumes), with 30 models in total.
+Rendering is performed at 2560x1440 pixels and the camera is initially placed in the center of the scene and slowly moved backward over the duration of 800 frames, steadily increasing the rendering load by revealing more and more model instances.
+
+\begin{figure}[ht]\center
+  \includegraphics[width=.9\columnwidth]{images/benchmark_eVolve}
+  \caption{\label{fBenchmarkEVolve} Benchmarking decomposition modes with Flower volume model at $1024^3$ unsigned byte (1GB).}
+\end{figure}
+
+\begin{figure}[ht]\center
+  \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
+  \caption{\label{fBenchmarkEqPly} Benchmarking decomposition modes with David statue at 1mm resolution (988MB).}
+\end{figure}
 
@@
 
-To benchmark the data distribution we used two datasets: The david statue at 2mm
+To benchmark the data distribution we used two datasets: The David statue at 2mm
 resolution (\fig{fDavid2mm}) and 3D volumes of spike frequencies of an

### 2018-01-31 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@100 db7c848b-7755-4df2-b763-8796797afa52

@@
 We tested the decomposition modes with both volume (\fig{fBenchmarkEVolve}) and polygonal data  (\fig{fBenchmarkEqPly}) and used test scenes that allowed us to finely adjust the rendering load the system has to cope with.
-In both cases the scene comprised of two rows of identical models (the 1mm David model when rendering polygonal data, the $1024^3$ Flower model when rendering volumes), with 30 models in total.
-Rendering is performed at 2560x1440 pixels and the camera is initially placed in the center of the scene and slowly moved backward over the duration of 800 frames, steadily increasing the rendering load by revealing more and more model instances.
+In both cases the scene comprised of two rows of instantiated, identical models (the 1mm David model when rendering polygonal data, the $1024^3$ Flower model when rendering volumes), with 30 models in total.
+Rendering was performed
+% with the \textsf{Equalizer} example applications eqPly (for meshes) and eVolve (for volumes)
+at 2560x1440.
+The camera is initially placed in the center of the scene, between the two rows.
+It is then moved backward over the duration of 800 frames, steadily increasing the rendering load by revealing more and more model instances.
+
+We investigated the scalability of individual compounds by running the same experiment using a varying number of render nodes (2-9), as well as one dedicated application/display node.
+We subsequently summed up the duration of all rendered frames for each run (\fig{fCompounds}).
+
+In the case of volume rendering we also measured the performance of compounds under heterogeneous load, which was easily achievable by varying the number of volume samples used for each fragment (1-8), while rendering.
+This allowed for a consistent linear scaling of rendering load, which was randomly varied (\fig{fEqualizers}) either per frame or per node.
+Such a linear scaling of load per node corresponds to a scaling of resources, e.g., doubling the rendering load on a specific node corresponds to halving the available rendering resources.
+This node would then only contribute the value 0.5 in terms of {\em normalized compute resources} to the system, as illustrated by \fig{fEqualizers}.
 
@@
 
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{David Steiner}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/david_steiner_picture2}}]{David Steiner}
+received MSc degrees from the University of Applied Sciences Upper Austria (Digital Media) and the University of Zurich (Computer Science).
+He joined the Visualization and MultiMedia Lab (VMML) in 2012 and is currently pursuing his doctorate.
+His research interests include interactive large-scale data visualization, distributed parallel rendering, and load balancing.
+\end{IEEEbiography}
 
@@
 \end{IEEEbiography}
+
+% insert where needed to balance the two columns on the last page with
+% biographies
+% \newpage
 

### 2018-02-05 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@101 db7c848b-7755-4df2-b763-8796797afa52

@@
 at 2560x1440.
-The camera is initially placed in the center of the scene, between the two rows.
-It is then moved backward over the duration of 800 frames, steadily increasing the rendering load by revealing more and more model instances.
+The camera is initially placed in the center of the scene, between the two rows (\fig{fBenchmarkEVolve} and \fig{fBenchmarkEqPly}), rendering only half of the model instances.
+It is then moved backward over the duration of 800 frames, steadily increasing the rendering load by revealing more models, until all 30 instances are visible.
 
@@
 
-In the case of volume rendering we also measured the performance of compounds under heterogeneous load, which was easily achievable by varying the number of volume samples used for each fragment (1-8), while rendering.
+In the case of volume rendering we also measured the performance of compounds under heterogeneous load, which was easily achievable by varying the number of volume samples used for each fragment (1-7), while rendering.
 This allowed for a consistent linear scaling of rendering load, which was randomly varied (\fig{fEqualizers}) either per frame or per node.
-Such a linear scaling of load per node corresponds to a scaling of resources, e.g., doubling the rendering load on a specific node corresponds to halving the available rendering resources.
-This node would then only contribute the value 0.5 in terms of {\em normalized compute resources} to the system, as illustrated by \fig{fEqualizers}.
+Such a linear scaling of load per node corresponds to a scaling of resources, e.g., doubling the rendering load on a specific node corresponds to halving its available rendering resources.
+To the system this node would then contribute the value 0.5 in terms of {\em normalized compute resources}, as illustrated by \fig{fEqualizers} (left).
+
+This figure gives an impression of how individual compounds perform on heterogeneous systems.
+In this case the tree equalizer compound performs best (\fig{fEqualizers} (left)), as it allows us to a priori define how much {\em usage} it should make of individual rendering nodes, in accordance with the expected load / the available compute resources.
+\fig{fEqualizers} (right), on the other hand, illustrates how compounds perform on a system where compute resources fluctuate randomly every frame, as can arguably be the case for cloud-based applications.
+For this scenario tile equalizer seems best suited, as it performs load balancing implicitly and does not assume coherence of available resources between frames.
 

### 2018-02-05 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@102 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-We investigated the scalability of individual compounds by running the same experiment using a varying number of render nodes (2-9), as well as one dedicated application/display node.
+We investigated the scalability of individual decomposition modes by running the same experiment using a varying number of render nodes (2-9), as well as one dedicated application/display node.
 We subsequently summed up the duration of all rendered frames for each run (\fig{fCompounds}).
 
-In the case of volume rendering we also measured the performance of compounds under heterogeneous load, which was easily achievable by varying the number of volume samples used for each fragment (1-7), while rendering.
+In the case of volume rendering we also measured the performance of decomposition modes under heterogeneous load, which was easily achievable by varying the number of volume samples used for each fragment (1-7), while rendering.
 This allowed for a consistent linear scaling of rendering load, which was randomly varied (\fig{fEqualizers}) either per frame or per node.
@@
 
-This figure gives an impression of how individual compounds perform on heterogeneous systems.
-In this case the tree equalizer compound performs best (\fig{fEqualizers} (left)), as it allows us to a priori define how much {\em usage} it should make of individual rendering nodes, in accordance with the expected load / the available compute resources.
-\fig{fEqualizers} (right), on the other hand, illustrates how compounds perform on a system where compute resources fluctuate randomly every frame, as can arguably be the case for cloud-based applications.
+This figure gives an impression of how individual modes perform on heterogeneous systems.
+In this case the tree equalizer performs best (\fig{fEqualizers} (left)), as it allows us to a priori define how much {\em usage} it should make of individual rendering nodes, in accordance with the expected load / the available compute resources.
+\fig{fEqualizers} (right), on the other hand, illustrates how decomposition modes perform on a system where compute resources fluctuate randomly every frame, as can arguably be the case for cloud-based applications.
 For this scenario tile equalizer seems best suited, as it performs load balancing implicitly and does not assume coherence of available resources between frames.

### 2018-02-05 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@103 db7c848b-7755-4df2-b763-8796797afa52

@@
 \item[Split Mode] configures the tile layout: horizontal or vertical stripes,
-and 2D, an binary tree split alternating the split axis on each level, resulting
+and 2D, a binary tree split alternating the split axis on each level, resulting
 in compact 2D tiles.
@@
 This figure gives an impression of how individual modes perform on heterogeneous systems.
-In this case the tree equalizer performs best (\fig{fEqualizers} (left)), as it allows us to a priori define how much {\em usage} it should make of individual rendering nodes, in accordance with the expected load / the available compute resources.
-\fig{fEqualizers} (right), on the other hand, illustrates how decomposition modes perform on a system where compute resources fluctuate randomly every frame, as can arguably be the case for cloud-based applications.
-For this scenario tile equalizer seems best suited, as it performs load balancing implicitly and does not assume coherence of available resources between frames.
+In this case the tree equalizer performs best (\fig{fEqualizers} (left)), as it allows us to a priori define how much {\em usage} it should make of individual nodes, i.e., bias the allocation of rendering time, in accordance with the expected load / the available compute resources.
+\fig{fEqualizers} (right), on the other hand, illustrates how the tested decomposition modes perform on a system where compute resources fluctuate randomly every frame, as can arguably be the case for, e.g., cloud-based applications.
+\FIXME{Should we only mention this in discussion?} For this scenario tile equalizer seems best suited, as it performs load balancing implicitly and does not assume coherence of available resources between frames.
 

### 2018-02-08 paperIEEE.tex
Source footnote
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@104 db7c848b-7755-4df2-b763-8796797afa52
Tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@105 db7c848b-7755-4df2-b763-8796797afa52

@@
 \usepackage{amsmath}
-%\usepackage{url}
+\usepackage{hyperref}
 \usepackage{caption,subcaption}
@@
 
-We conducted performance benchmarks for the decomposition modes on a cluster using hexacore Intel Xeon E5-2620v3 CPUs (2.4GHz), 16 GB main memory, 4 GBit/s Ethernet and QDR Infiniband.
-GCC 4.8 has been used with CMake 3.7 release mode settings to compile the software stack.
-
-We tested the decomposition modes with both volume (\fig{fBenchmarkEVolve}) and polygonal data  (\fig{fBenchmarkEqPly}) and used test scenes that allowed us to finely adjust the rendering load the system has to cope with.
-In both cases the scene comprised of two rows of instantiated, identical models (the 1mm David model when rendering polygonal data, the $1024^3$ Flower model when rendering volumes), with 30 models in total.
-Rendering was performed
-% with the \textsf{Equalizer} example applications eqPly (for meshes) and eVolve (for volumes)
-at 2560x1440.
-The camera is initially placed in the center of the scene, between the two rows (\fig{fBenchmarkEVolve} and \fig{fBenchmarkEqPly}), rendering only half of the model instances.
-It is then moved backward over the duration of 800 frames, steadily increasing the rendering load by revealing more models, until all 30 instances are visible.
+We conducted performance benchmarks for the decomposition modes on a cluster
+using hexacore Intel Xeon E5-2620v3 CPUs (2.4GHz), \FIXME{GPUs}, 16 GB main
+memory, 4 GBit/s Ethernet and QDR Infiniband. GCC 4.8 has been used with CMake
+3.7 release mode settings to compile the software stack.
+
+We tested the decomposition modes with both volume data (\fig{fBenchmarkEVolve})
+and polygonal data  (\fig{fBenchmarkEqPly}) and used test scenes that allowed us
+to finely adjust the rendering load the system has to cope with. In both cases
+the scene comprised of two rows of instantiated, identical models with 30 models
+in total. Rendering was performed at a resolution of 2560x1440. The camera is
+initially placed in the center of the scene, between the two rows
+(\fig{fBenchmarkEVolve} and \fig{fBenchmarkEqPly}), rendering only half of the
+model instances. It is then moved backward over the duration of 800 frames,
+steadily increasing the rendering load by revealing more models, until all 30
+instances are visible.
 
@@
 
-In the case of volume rendering we also measured the performance of decomposition modes under heterogeneous load, which was easily achievable by varying the number of volume samples used for each fragment (1-7), while rendering.
-This allowed for a consistent linear scaling of rendering load, which was randomly varied (\fig{fEqualizers}) either per frame or per node.
-Such a linear scaling of load per node corresponds to a scaling of resources, e.g., doubling the rendering load on a specific node corresponds to halving its available rendering resources.
-To the system this node would then contribute the value 0.5 in terms of {\em normalized compute resources}, as illustrated by \fig{fEqualizers} (left).
-
-This figure gives an impression of how individual modes perform on heterogeneous systems.
-In this case the tree equalizer performs best (\fig{fEqualizers} (left)), as it allows us to a priori define how much {\em usage} it should make of individual nodes, i.e., bias the allocation of rendering time, in accordance with the expected load / the available compute resources.
-\fig{fEqualizers} (right), on the other hand, illustrates how the tested decomposition modes perform on a system where compute resources fluctuate randomly every frame, as can arguably be the case for, e.g., cloud-based applications.
-\FIXME{Should we only mention this in discussion?} For this scenario tile equalizer seems best suited, as it performs load balancing implicitly and does not assume coherence of available resources between frames.
+For volume rendering we also measured the performance of decomposition modes
+under heterogeneous load, which was easily achievable by varying the number of
+volume samples used for each fragment (1-7) while rendering. This allowed for a
+consistent linear scaling of rendering load, which was randomly varied
+(\fig{fEqualizers}) either per frame or per node. Such a linear scaling of load
+per node corresponds to a scaling of resources, e.g., doubling the rendering
+load on a specific node corresponds to halving its available rendering
+resources. To the system this node would then contribute the value 0.5 in terms
+of {\em normalized compute resources}, as illustrated by \fig{fEqualizers}
+(left).
+
+This figure gives an impression of how individual modes perform on heterogeneous
+systems. In this case the tree equalizer performs best (\fig{fEqualizers}
+(left)), as it allows us to a priori define how much {\em usage} it should make
+of individual nodes, i.e., bias the allocation of rendering time in accordance
+with the (simulated) compute resources. \fig{fEqualizers} (right), on the other
+hand, illustrates how the tested decomposition modes perform on a system where
+compute resources fluctuate randomly every frame, as can arguably be the case
+for, e.g., cloud-based applications. For this scenario tile equalizer seems best
+suited, as it performs load balancing implicitly and does not assume coherence
+of available resources between frames.
 
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eVolve}
-  \caption{\label{fBenchmarkEVolve} Benchmarking decomposition modes with Flower volume model at $1024^3$ unsigned byte (1GB).}
+  \caption{\label{fBenchmarkEVolve} Flower volume model at $1024^3$ unsigned byte (1GB).}
 \end{figure}
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
-  \caption{\label{fBenchmarkEqPly} Benchmarking decomposition modes with David statue at 1mm resolution (988MB).}
+  \caption{\label{fBenchmarkEqPly} David statue at 1mm resolution (988MB).}
 \end{figure}
@@
 To benchmark the data distribution we used two datasets: The David statue at 2mm
-resolution (\fig{fDavid2mm}) and 3D volumes of spike frequencies of an
-electrical simulation of three million neurons (\fig{fVolumes}).
-
-\begin{figure}[ht]\center
-  \includegraphics[width=.49\columnwidth]{images/david2mm}
-  \caption{\label{fDavid2mm}David statue at 2mm resolution (145MB)}
-\end{figure}
+resolution (as in \fig{fBenchmarkEqPly}, but in 2mm resolution) and 3D volumes
+of spike frequencies of an electrical simulation of three million neurons
+(\fig{fVolumes}).
 
@@
 services running on the cluster. The details of the benchmark algorithm can be
-found in the implementation~\cite{eqObjectBench}. used the same data sets as in
-the previous section. We ran the benchmark on up to eight physical nodes, that
-is, after eight processes nodes start to run two processes per node, which share
-CPU, memory and network interface bandwidth.
+found in the
+\htmladdnormallinkfoot{implementation}{https://github.com/Eyescale/Equalizer/tree/paper2018/tools/eqObjectBench}.
+used the same data sets as in the previous section. We ran the benchmark on up
+to eight physical nodes, that is, after eight processes nodes start to run two
+processes per node, which share CPU, memory and network interface bandwidth.
 
@@
 \includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{
-Stefan Eilemann} Stefan is the technical manager of the Visualization Team in
-the Blue
-Brain Project. His work involves working towards large-scale visualization for
-Exascale simulations; interactive integration of simulation, analysis and
-visualization as well as flexible frameworks for data sharing and dynamic
-allocation of heterogenous resources. He is the CEO and founder of Eyescale and
-also the lead developer of the \textsf{Equalizer} parallel rendering framework.
-He has worked in SGI's Advanced Graphics Division as the technical lead of
-OpenGL Multipipe SDK - a toolkit for scalable, parallel visualization software
-on shared memory supercomputers. He received his masters degree in Computer
-Science from EPFL in 2015, and his Engineering Diploma in Computer Science in
-1998. He is currently working towards a PhD in Computer Science at the
-Visualization and MultiMedia Lab at the University of Zurich.
-\end{IEEEbiography}
+Stefan Eilemann} Stefan was the technical manager of the Visualization Team in
+the Blue Brain Project. His work involves working towards large-scale
+visualization for Exascale simulations; interactive integration of simulation,
+analysis and visualization as well as flexible frameworks for data sharing and
+dynamic allocation of heterogenous resources. He is the CEO and founder of
+Eyescale and also the lead developer of the \textsf{Equalizer} parallel
+rendering framework. He received his masters degree in Computer Science from
+EPFL in 2015, and his Engineering Diploma in Computer Science in 1998. He is
+currently working towards a PhD in Computer Science at the Visualization and
+MultiMedia Lab at the University of Zurich. \end{IEEEbiography}
 
@@
   \caption{\label{fDavidDist}Object mapping (left) and data synchronization
-    time (middle, detail view right) for the David 2mm data set in
-    \fig{fDavid2mm}}
+    time (middle, detail view right) for the David 2mm data set}
 \end{figure*}

### 2018-02-08 paperIEEE.tex
Rework & results
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@108 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-In particular, we present novel algorithms for parallel rendering which did not yet
-appear in a separate publication, including pixel and sub-pixel decompositions,
-dynamic frame resolution, tunable sort-first load-balancing parameters, frame-rate
-equalization, thread synchronization modes for multi-threaded rendering, a
-powerful abstraction for multi-view rendering on arbitrary display setups,
-dynamic focus distance and asymmetric eye positions for VR,
-parallel pixel streaming to tiled display walls, as well as a fully-fledged data
-distribution API with compression and reliable multicast.
+We present novel algorithms for parallel rendering which did not appear in a
+separate publication, including pixel and sub-pixel decompositions, dynamic
+frame resolution, tunable sort-first load-balancing parameters, frame-rate
+equalization, thread synchronization modes for multi-GPU rendering, a powerful
+abstraction for multi-view rendering on arbitrary display setups, dynamic focus
+distance and asymmetric eye positions for VR, parallel pixel streaming to tiled
+display walls, as well as a fully-fledged data distribution API with compression
+and reliable multicast.
 
@@
 separate applications using pixel streaming. It is no longer actively supported.
-\textsf{Sage 2} is a complete, browser-centric reimplementation where each
+\textsf{Sage~2} is a complete, browser-centric reimplementation where each
 application is a web application distributed across browser instances.
@@
 for visualizing remote sensing data~\cite{LK:09} on large displays and immersive
-installations based on Equalizer.
+installations.
 
@@
 
-This section introduces new decomposition modes, runtime adaptations and other
-optimizations to increase the rendering performance for a wide variety of
-algorithms and use cases.
+% This section introduces new decomposition modes, runtime adaptations and other
+% optimizations to increase the rendering performance for a wide variety of
+% algorithms and use cases.
 
@@
 sort-last rendering modes. \fig{fig:compounds} provides an overview of the new
-modes. The way how compounds are setup is still as given in \cite{EMP:09}.
+modes. The compound concept to set up scalable rendering is presented in
+\cite{EMP:09}.
 
@@
 sort-last rendering, respectively. They decompose the scene into a predefined
-set of fixed-size image tiles or database ranges. These tasks, or work packages are queued and
-processed by all source channels by polling a server-central queue. Prefetching
-ensures that the task communication overlaps with rendering. As shown in~
-\cite{SPEP:16}, these modes can provide better performance due to being inherently
-load-balanced, as long as there is an insignificant overhead for the render task
-setup. This mode is transparent to \textsf{Equalizer} applications.
+set of fixed-size image tiles or database ranges. These tasks, or work packages,
+are queued and processed by all source channels by polling a server-central
+queue. Prefetching ensures that the task communication overlaps with rendering.
+As shown in~ \cite{SPEP:16}, these modes can provide better performance due to
+being inherently load-balanced, as long as there is an insignificant overhead
+for the render task setup. This mode is transparent to \textsf{Equalizer}
+applications.
 
@@
 transparent to the application, since it needs to adapt (jitter or tilt) the
-frustum based on the iteration executed. Furthermore, subpixel compounds
-interact with idle image refinements, e.g., they can accelerate idle
-anti-aliasing of a scene when the camera and scene are not changed.
+frustum based on the iteration executed.
+
+% Furthermore, subpixel compounds interact with idle image refinements, e.g., they
+% can accelerate idle anti-aliasing of a scene when the camera and scene are not
+% changed.
 
@@
 
-We have implemented various tunable parameters for both equalizers allowing
-application developers to optimize the load balancing based on the
-characteristics of their rendering algorithm:
+Both equalizers implement tuneable parameters allowing application developers to
+optimize the load balancing based on the characteristics of their rendering
+algorithm:
 
@@
 
-The monitor equalizer (\fig{fig:monitor}) allows to reuse the rendering on
-another channel, typically for monitoring a larger setup on a control
-workstation. Output frames on the display channels are connected to input frames
-on a single monitoring channel. The monitor equalizer changes the scaling factor
-and offset between the output and input, so that the monitor channel has the
-same, but typically downscaled view, as the originating segments.
+The monitor equalizer (\fig{fig:monitor}, \fig{fLayout}) allows to reuse the
+rendering on another channel, typically for monitoring a larger setup on a
+control workstation. Output frames on the display channels are connected to
+input frames on a single monitoring channel. The monitor equalizer changes the
+scaling factor and offset between the output and input, so that the monitor
+channel has the same, but typically downscaled view, as the originating
+segments.
 
@@
 
-The region of interest is the screen-space 2D bounding box enclosing the
-geometry rendered by a single resource. We have extended the core parallel
-rendering framework to use an application-provided ROI to optimize the load
-equalizer as well as image compositing performance. The load equalizer uses the
-ROI to refine its load grid to the regions containing data. The compositing code
-uses the ROI to minimize image readback and network transmission. In
-\cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of the algorithm, and
-show that using ROI can quadruple the rendering performance, in particular for
-the costly compositing step in sort-last rendering.
+The region of interest is the screen-space 2D bounding box enclosing the data
+rendered by a single resource. We have extended the core parallel rendering
+framework to use an application-provided ROI to optimize the
+\texsf{load\_equalizer} as well as the image compositing. The load equalizer
+uses the ROI to refine its load grid to the regions containing data. The
+compositing code uses the ROI to minimize image readback and network
+transmission. In \cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of
+the algorithm, and show that using ROI can quadruple the rendering performance,
+in particular for the costly compositing step in sort-last rendering.
 
@@
 
-Compression for the compositing step in parallel rendering is critical for
-performance. This not only applies to the well-researched network transfer step,
-but also for the transfer between GPU and CPU. \textsf{Equalizer} supports a
-variety of
-compression algorithms, from very fast RLE encoding, YUV
-subsampling on the GPU to JPEG compression. These algorithms are implemented as
-runtime-loaded
-plugins, allowing easy extension and customization to application-specific
-compression. In \cite{MEP:10}, we show this to be a critical step for
-interactive performance at scale.
+Compression for the compositing step is critical for performance. This not only
+applies to the well-researched network transfer step, but also for the transfer
+between GPU and CPU. \textsf{Equalizer} supports a variety of compression
+algorithms, from very fast RLE encoding, YUV subsampling on the GPU to JPEG
+compression. These algorithms are implemented as runtime-loaded plugins,
+allowing easy extension and customization to application-specific compression.
+In \cite{MEP:10}, we show this to be a critical step for interactive performance
+at scale.
 
@@
 mutable data has to have a copy in each render thread, which is not feasible in
-many complex applications and large data scenarios. To easily customize the
+most applications and large data scenarios. To easily customize the
 synchronization of all threads on a single process, \textsf{Equalizer}
@@
 to read shared data during their draw operation, but not during compositing.
-Since compositing is typically independent of the rendered data, this is the
-default mode. This mode allows to overlap compositing with rendering and data
+Since compositing is often independent of the rendered data, this is the default
+mode. This mode allows to overlap compositing with rendering and data
 synchronization on multi-GPU machines.
@@
 requires the application to have one instance of each mutable object in each
-render thread. It is required for using time-multiplex compounds in
-multithreaded rendering.
+render thread. It is required for scaling time-multiplex compounds on multi-GPU
+machines.
 
@@
 direct, application-transparent configuration of a VRPN tracker device.
-Alternatively, an application can provide a $4\times 4$ tracking matrix defining
-the transformation from the canvas coordinate system to the observer. Both
+Alternatively, applications can provide a $4\times 4$ tracking matrix. Both
 CAVE-like tracking with fixed projection surfaces and HMD tracking with moving
@@
 new dynamic focus mode, where the application defines the distance of the focal
-plane from the observer, based on the current view direction of the user.
-Initial experiments show that this provides better viewing comfort, in
-particular for objects placed in front of the physical displays.
+plane from the observer, based on the current \textit{lookat} distance. Initial
+experiments show that this provides better viewing comfort, in particular for
+objects placed in front of the physical displays.
 
@@
 
-Traditional head tracking computes the left and right eye positions by using a
-configurable interocular distance. However, human heads are not symmetric, and
-by measuring individual users a more precise frustum can be computed.
-\textsf{Equalizer} supports this through the optional configuration (file or
-programmatically) of individual 3D translations for all eyes relative to the
-tracking matrix.
+Traditional head tracking computes the left and right eye positions by using an
+interocular distance. However, human heads are not symmetric, and by measuring
+individual users a more precise frustum can be computed. \textsf{Equalizer}
+supports this through the optional configuration of individual 3D eye
+translations relative to the tracking matrix.
 
@@
 \cite{deflect}, we have shown interactive framerates for a 24 megapixel
-resolution over a WAN link. Events received are translated into
-\textsf{Equalizer} events and injected into the normal event flow, allowing
-seamless application integration.
+resolution over a WAN link. \textsf{Deflect} events are translated and injected
+into the \textsf{Equalizer} event flow, allowing seamless application
+integration.
 
@@
 
-We investigated the scalability of individual decomposition modes by running the same experiment using a varying number of render nodes (2-9), as well as one dedicated application/display node.
-We subsequently summed up the duration of all rendered frames for each run (\fig{fCompounds}).
+We investigated the scalability of individual decomposition modes by running the
+same experiment using a varying number of render nodes (2-9), as well as one
+dedicated application/display node. We subsequently summed up the duration of
+all rendered frames for each run (\fig{fCompounds}). We provide the results of
+the load balancer providing the better result for each application and compound.
+The simpler \textsf{tree\_equalizer} outperforms in almost all cases the
+load-grid-driven \textsf{load\_equalizer}, except for sort-first volume
+rendering where the load in the region of interest is relatively uniform. This
+counterintuitive result once more proves that simple algorithms often outperform
+theoretically better, but more complex implementations.
 
@@
 compute resources fluctuate randomly every frame, as can arguably be the case
-for, e.g., cloud-based applications. For this scenario tile equalizer seems best
-suited, as it performs load balancing implicitly and does not assume coherence
-of available resources between frames.
+for virutalized enviroments. For this scenario tile equalizer seems best suited,
+as it performs load balancing implicitly and does not assume coherence of
+available resources between frames. The simpler tree equalizer also outperforms
+the load equalizer in this experiment.
 
@@
 \includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{
-Stefan Eilemann} Stefan was the technical manager of the Visualization Team in
-the Blue Brain Project. His work involves working towards large-scale
-visualization for Exascale simulations; interactive integration of simulation,
-analysis and visualization as well as flexible frameworks for data sharing and
-dynamic allocation of heterogenous resources. He is the CEO and founder of
-Eyescale and also the lead developer of the \textsf{Equalizer} parallel
-rendering framework. He received his masters degree in Computer Science from
-EPFL in 2015, and his Engineering Diploma in Computer Science in 1998. He is
-currently working towards a PhD in Computer Science at the Visualization and
-MultiMedia Lab at the University of Zurich. \end{IEEEbiography}
+Stefan Eilemann} works towards large-scale visualization for Exascale
+simulations; interactive integration of simulation, analysis and visualization
+as well as flexible frameworks for data sharing and dynamic allocation of
+heterogenous resources. He was the technical manager of the Visualization Team
+in the Blue Brain Project, is the CEO and founder of Eyescale, and the lead
+developer of the \textsf{Equalizer} parallel rendering framework. He received
+his masters degree in Computer Science from EPFL in 2015, and his Engineering
+Diploma in Computer Science in 1998. He is currently working towards a PhD in
+Computer Science at the Visualization and MultiMedia Lab at the University of
+Zurich. \end{IEEEbiography}
 

### 2018-02-08 paperIEEE.tex
Fix tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@109 db7c848b-7755-4df2-b763-8796797afa52
Collage review
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@110 db7c848b-7755-4df2-b763-8796797afa52

@@
 framework to use an application-provided ROI to optimize the
-\texsf{load\_equalizer} as well as the image compositing. The load equalizer
+\textsf{load\_equalizer} as well as the image compositing. The load equalizer
 uses the ROI to refine its load grid to the regions containing data. The
@@
 segments, the swap buffers call followed by a \textsf{glFlush} to ensure timely
-execution of the swap command. The hardware swap barrier is implemented using a
-\textsf{NV\_swap\_group} extension.
+execution of the swap command. The hardware swap barrier is implemented using
+the \textsf{NV\_swap\_group} extension.
 
@@
 application, and all changes are synchronized to the corresponding rendering
-nodes. Since this setup is monoscopic, no head tracking or observers are used.
+nodes. For this monoscopic setup no head tracking or observers are used.
 
@@
 administrators with the tools to implement easy to use integration with cluster
-schedulers. This feature is transparent to application developers.
+schedulers.
 
@@
 
-Qt is a popular window system with many application developers. Unfortunately,
-it imposes a different threading model for window creation and event handling
+Qt is a popular window system with application developers. Unfortunately, it
+imposes a different threading model for window creation and event handling
 compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread
@@
   a set of connections by implementing output stream operators. Uses buffering
-  to aggregate data for network transmission. Performs endian swapping during
-  input if the endianness differs between the originating and local node.
+  to aggregate data for network transmission. Performs byte swapping during
+  input if the endianness differs between the remote and local node.
 \item[Node and LocalNode] The abstraction of a process in the cluster. Nodes
@@
 RSP is an implementation of a reliable multicast protocol over unreliable UDP
-transport. RSP behaves similarly to TCP, it provides full reliability and
+transport. RSP behaves similarly to TCP; it provides full reliability and
 ordering of the data, and slow receivers will eventually throttle the sender
@@
 Our RSP implementation uses a separate protocol thread for each RSP group, which
-handles all reads and writes on the multicast group. This thread implements the
-protocol handling and communicates with the application threads through
-thread-safe queues. These queues contain datagrams of the application byte
-stream, prefixed by a type-specific header of at most eight bytes. Each
-connection has a configurable number of buffers, (1024 by default) of a
-configurable MTU (1470 bytes default) which are either free for use or in
-transmission.
+handles all reads and writes on the multicast socket. It implements the protocol
+handling and communicates with the application threads through thread-safe
+queues. The queues contain datagrams of the application byte stream, prefixed by
+a header of at most eight bytes. Each connection has a configurable number of
+buffers (1024 by default) of a configurable MTU (1470 bytes default), which are
+either free or in transmission.
 
@@
 connection actively acknowledges every $n$ (17 by default) packets fully
-received. The incoming connections temporarily offset this acknowledgment by
-their connection identifier to avoid bursts of acks. Any missed datagram is
-actively nack'ed as soon as it is detected. Write connections continuously
-retransmit packets from nack datagrams, and advance their window upon reception
-of acks from all readers. Furthermore, the writer will explicitly request an ack
-or nack when it runs out of empty buffers or finishes its write queue. Nack
-datagrams may contain multiple ranges of missed datagrams, which is motivated by
-the observation that UDP implementations often drop multiple packets at once.
+received. The incoming connections offset this acknowledgment by their
+connection identifier to avoid bursts of acks. Any missed datagram is actively
+nack'ed as soon as detected. Write connections continuously retransmit packets
+for nack datagrams, and advance their window upon reception of all acks from the
+group. The writer will explicitly request an ack or nack when it runs out of
+empty buffers or finishes its write queue. Nack datagrams may contain multiple
+ranges of missed datagrams, which is motivated by the observation that UDP
+implementations often drop multiple contiguous packets.
 
 Congestion control is necessary to optimize bandwidth usage. While TCP uses the
-well-known AIMD (additive increase, multiplicative decrease) algorithm, we have
-chosen a more aggressive congestion control algorithm of additive increase and
-additive decrease. This has proven experimentally to be more optimal: UDP is
-often rate-limited by switches; that is, packets are discarded regularly and not
+well-known additive increase, multiplicative decrease algorithm, we have chosen
+a more aggressive congestion control algorithm of additive increase and additive
+decrease. This has proven experimentally to be more optimal: UDP is often
+rate-limited by switches; that is, packets are discarded regularly and not
 exceptionally. Only slowly backing of the current send rate helps to stay close
 to this limit. Furthermore, our RSP traffic is limited to the local subnet,
-making cooparation between multiple data stream less of an issue. Send rate
+making cooperation between multiple data stream less of an issue. Send rate
 limiting uses a bucket algorithm, where over time the bucket fills with send
@@
 in \cite{EMP:09}, the parallel rendering code in an \textsf{Equalizer}
-application only
-needs access to the data needed for rendering, as all application logic is
-centralized in the application main thread. We have encountered two main
-approaches to address this distribution: Using a shared filesystem for static
-data, or using data distribution for static and dynamic data.
-
-Distributed objects are not required to build \textsf{Equalizer} applications.
-While most
-developers choose to use this abstraction for convenience, we have seen
-applications using other means for data distribution, e.g., MPI.
+application only needs access to the data needed for rendering, as all
+application logic is centralized in the application main thread. We have
+encountered two main approaches to address this distribution: Using a shared
+filesystem for static data, or using data distribution for static and dynamic
+data. Distributed objects are not required to build \textsf{Equalizer}
+applications. While most developers choose to use this abstraction for
+convenience, we have seen applications using other means for data distribution,
+e.g., MPI.
 
@@
 subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
-application programmer implements serialization and deserialization of the
-distributed data. Distributed objects can be static (immutable) or
-dynamic. Objects have a universally unique identifier (UUID) to address them
-cluster-wide. A master-slave model is used to establish mapping and data
-synchronization across processes. Typically, the application main loop registers
-a master instance and communicates the UUID to the render clients, which map
-their instance to the given identifier. The following object types are
-available:
+application programmer implements serialization and deserialization. Distributed
+objects can be static (immutable) or dynamic. Objects have an universally unique
+identifier (UUID) as cluster-wide addresses. A master-slave model is used to
+establish mapping and data synchronization across processes. Typically, the
+application main loop registers a master instance and communicates the UUID to
+the render clients, which map their instance to the given identifier. The
+following object types are available:
 
@@
 
-All distributed objects have to implement \textsf{getInstanceData} and
-\textsf{applyInstanceData} to serialize and deserialize the object's distributed
-data. These methods provide an output or input stream as a parameter, which
-abstracts the data transmission and can be used like a \textsf{std::stream}.
-The data streams implement efficient buffering and compression, and
-automatically select the best connection for data transport. Custom data type
-serializers can be implemented by providing the appropriate serialization
-functions. No pointers should be directly transmitted through the data
-streams. For pointers, the corresponding object is typically a distributed
-object as well, and its identifier and potentially version is transmitted in
-place of the memory address.
-
-Dynamic objects are versioned, and have to override \textsf{getChangeType} to
-indicate how they want to have changes to be handled. Upon \textsf{commit} the
-delta data from the previous version is sent, if available using multicast, to
-all mapped slave instances. The data is queued on the remote node, and is
-applied when the application calls \textsf{sync} to synchronize the object to a
-new version. The \textsf{sync} method might block if a version has not yet been
-committed or is still in transmission. All versioned objects the following
-characteristics:
+Serialization is facilitated using output or input streams, which abstract the
+data transmission and are used like a \textsf{std::stream}. The data streams
+implement efficient buffering and compression, and automatically select the best
+connection for data transport. Custom data type serializers can be implemented
+by providing the appropriate serialization functions. No pointers should be
+directly transmitted through the data streams. For pointers, the corresponding
+object is typically a distributed object as well, and its UUID and version are
+transmitted in place of a pointer.
+
+Dynamic objects are versioned, and on \textsf{commit} the delta data from the
+previous version is sent, if available using multicast, to all mapped slave
+instances. The data is queued on the remote node, and is applied when the
+application calls \textsf{sync} to synchronize the object to a new version. The
+\textsf{sync} method might block if a version has not yet been committed or is
+still in transmission. All versioned objects the following characteristics:
 
@@
 
-Besides the instance data serialization methods used to map an object, delta and
-unbuffered objects may implement \textsf{pack} and \textsf{unpack} to serialize
-or deserialize the changes since the last version.
-
 \label{sec:Serializable}The \textsf{Collage} Serializable implements one
-convenient usage
-pattern for object data distribution. The \textsf{co::Serializable} data
-distribution is based on the concept of dirty bits, allowing inheritance with
-data distribution. Dirty bits form a 64-bit mask which marks the parts of the
-object to be distributed during the next commit. For serialization, the
-application developer implements \textsf{serialize} or \textsf{deserialize},
-which are called with the bit mask specifying which data has to be transmitted
-or received. During a commit or sync, the current dirty bits are given, whereas
-during object mapping all dirty bits are passed to the serialization methods.
+convenient usage pattern for object data distribution. The
+\textsf{co::Serializable} data distribution is based on the concept of dirty
+bits, allowing inheritance with data distribution. Dirty bits form a 64-bit mask
+which marks the parts of the object to be distributed during the next commit.
+For serialization, the application developer implements \textsf{serialize} or
+\textsf{deserialize}, which are called with the bit mask specifying which data
+has to be transmitted or received. During a commit or sync, the current dirty
+bits are given, whereas during object mapping all dirty bits are passed to the
+serialization methods.
 
@@
 the slave nodes. A token-based protocol will block the commit on the master
-instance if to many unsynchronized versions exist.
+instance if too many unsynchronized versions exist.
 
@@
 compression, chunking, caching, preloading and multicast. The results section
-evaluates these optimizations.
+evaluates some of these optimizations.
 
@@
 
-The data streaming interface also implements chunking, which aims to pipeline
-the serialization code with the network transmission. After a configurable
-number of bytes has been serialized to the internal buffer, it is transmitted
-and serialization continues. This is used both for the initial mapping data and
-for commit data.
-
-Caching retains instance data of objects in a cache, and reuses this data to
-accelarate mapping of objects. The instance cache is either filled by
+The data streaming interface implements chunking, which pipelines the
+serialization code with the network transmission. After a configurable number of
+bytes has been serialized to the internal buffer, it is transmitted and
+serialization continues. This is used both for the initial mapping data and for
+commit data.
+
+Caching retains instance data of objects in a client-side cache, and reuses this
+data to accelerate mapping of objects. The instance cache is either filled by
 ``snooping'' on multicast transmissions or by an explicit preloading when the
-master objects are registered. The preloading (``send on register'') sends
-instance data of recently registered master objects while the correspinding node
-is idle to all connected nodes. These nodes simply enter the received data to
-their cache. Preloading uses multicast when available.
+master objects are registered. The preloading sends instance data of recently
+registered master objects while the corresponding node is idle to all connected
+nodes. These nodes simply enter the received data to their cache. Preloading
+uses multicast when available.
 
@@
 
-During data synchronization, the master instance commits the object data to all
-mapped slave instances simultaneously. This is a ``push'' operation, whereas the
-mapping is a slave-triggered ``pull'' operation with additional synchronization
-points. Slave nodes queue this data and consume it during synchronization. We
-test the time to commit and sync the data using different compression engines.
+During data synchronization, the master commits the object data to all mapped
+slave instances simultaneously. This is a ``push'' operation, whereas the
+mapping is a slave-triggered ``pull'' operation. Slave nodes queue this data and
+consume it during synchronization. We test the time to commit and sync the data
+using different compression engines.
 
@@
 data is beneficial for transmitting data over a 10~Gb/s link even for a single
-slave process only. Buffering has little benefit since the serialization of
-volume data is trivial. Buffered compression makes a significant difference,
-since the compression cost can be amortized over $n$ nodes, reaching raw data
-transmission rates of 3.7 GB/s with the default Snappy compressor and at best
-4.4 GB/s with ZStandard at level 1.
+slave process. Buffering has little benefit since the serialization of volume
+data is trivial. Buffered compression makes a significant difference, since the
+compression cost can be amortized over $n$ nodes, reaching raw data transmission
+rates of 3.7 GB/s with the default Snappy compressor and at best 4.4 GB/s with
+ZStandard at level 1.
 
@@
 suspect the abstraction of a byte stream connection chosen by \textsf{Collage}
-for historical reasons is not well suited for remote DMA semantics; that is,
-one needs to design the network API around zero-copy semantics with managed
-memory for modern high-speed transports. Both Infiniband connections show
-significant measurement jitter.
-
-Our RSP multicast implementation performs as expected. \textsf{Collage} starts
-using multicast to commit new object versions when two or more clients are
-mapped, since the transmission to a single client is better done using unicast.
-RSP consistently outperforms unicast on the same physical interface and shows
-good scaling behavior (2.5 times slower on 16 vs 2 clients on ethernet, 1.8
-times slower on InfiniBand). The scaling is significantly better when only one
-process per node is used (\fig{fIFDist}, middle: 30\% slower on ethernet,
-nearly flat on InfiniBand). The increased transmission time with multiple
-clients is caused by a higher probability of packet loss, which increases
-significantly when using more than one process per node, as shown in
-\fig{fIFDist}, right. This figure plots the number of retransmissions
-divided by the number of datagrams. Infiniband outperforms ethernet slightly,
-but is largely limited by the RSP implementation throughput of preparing and
-queueing the datagrams to and from the protocol thread, which we observed in a
-profiling analysis.
+is not well suited for remote DMA semantics; that is, one needs to design the
+network API around zero-copy semantics with managed memory for modern high-speed
+transports. Both Infiniband connections show significant measurement jitter.
+
+RSP multicast performs as expected. \textsf{Collage} starts using multicast to
+commit new object versions when two or more clients are mapped, since the
+transmission to a single client is faster using unicast. RSP consistently
+outperforms unicast on the same physical interface and shows good scaling
+behavior (2.5 times slower on 16 vs 2 clients on ethernet, 1.8 times slower on
+InfiniBand). The scaling is significantly better when only one process per node
+is used (\fig{fIFDist}, middle: 30\% slower on ethernet, nearly flat on
+InfiniBand). The increased transmission time with multiple clients is caused by
+a higher probability of packet loss, which increases significantly when using
+more than one process per node. \fig{fIFDist} (right) plots the number of
+retransmissions divided by the number of datagrams. Infiniband outperforms
+ethernet slightly, but is largely limited by the RSP implementation throughput
+of preparing and queueing the datagrams to and from the protocol thread, which
+we observed in profiling.
 
@@
 applications. We highlight a few commercial and research applications
-underlining the generic and versatile system implemenation.
+underlining the generic and versatile system implementation.
 
@@
 %
-This work was partially supported by
-the Swiss National Science Foundation Grants 200021-116329 and 200020-129525,
-the Swiss Commission for Technology and Innovation CTI/KTI Project 9394.2 PFES-ES,
-the EU FP7 People Programme (Marie Curie Actions) under REA Grant Agreement n$^{\circ}290227$ and
-a Hasler Stiftung grant (project number $12097$).
+This work was partially supported by the Swiss National Science Foundation
+Grants 200021-116329 and 200020-129525, the Swiss Commission for Technology and
+Innovation CTI/KTI Project 9394.2 PFES-ES, the EU FP7 People Programme (Marie
+Curie Actions) under REA Grant Agreement n$^{\circ}290227$ and a Hasler Stiftung
+grant (project number $12097$).
 

### 2018-02-08 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@111 db7c848b-7755-4df2-b763-8796797afa52

@@
 the display system simplifies the configuration process. Applications often have
-the need to be aware of spatial relationship of the display setup, for example
+the need to be aware of spatial relationships of the display setup, for example
 to render 2D overlays or to configure multiple views on a tiled display wall.
@@
 swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all
-segments, the swap buffers call followed by a \textsf{glFlush} to ensure timely
+segments, and the swap buffers call followed by a \textsf{glFlush} to ensure timely
 execution of the swap command. The hardware swap barrier is implemented using
@@
 is independently responsible for creating its windows, receiving the events and
-eventually dispatching them to the application process main thread. This design
+eventually dispatching them to the application process' main thread. This design
 is motivated by the natural threading model of X11 and WGL, and allows simple

### 2018-02-08 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@112 db7c848b-7755-4df2-b763-8796797afa52
units & hardware
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@113 db7c848b-7755-4df2-b763-8796797afa52

@@
 layers, gradually providing higher level functionality for the programmer. The
-main primitives in \textsf{Collage} and their relation are shown in
+main primitives in \textsf{Collage} and their relations are shown in
 \fig{fNetObject} and provide:
@@
 
-% Time to render a 4k, 32-step MSAA image of the 1mm David model and Livre using 1 to 10 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps), tiles, chunks, pixel, subpixel) (+DFR for Livre).
-
-We conducted performance benchmarks for the decomposition modes on a cluster
-using hexacore Intel Xeon E5-2620v3 CPUs (2.4GHz), \FIXME{GPUs}, 16 GB main
-memory, 4 GBit/s Ethernet and QDR Infiniband. GCC 4.8 has been used with CMake
-3.7 release mode settings to compile the software stack.
+% Time to render a 4k, 32-step MSAA image of the 1\,mm David model and Livre using 1 to 10 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps), tiles, chunks, pixel, subpixel) (+DFR for Livre).
+
+We conducted performance benchmarks for the decomposition modes on a
+rendering cluster using hexacore Intel Xeon E5-2620v3 CPUs ($2.4$\,GHz), $16$\,GB DDR4 RAM per node, and nVidia GTX $970$ GPUs with $4$\,GB VRAM each.
+The nodes are connected via a $40$\,Gbit/s {\em Infiniband} network.
+GCC 4.8 has been used with CMake 3.7 release mode settings to compile the software stack.
 
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eVolve}
-  \caption{\label{fBenchmarkEVolve} Flower volume model at $1024^3$ unsigned byte (1GB).}
+  \caption{\label{fBenchmarkEVolve} Flower volume model at $1024^3$ unsigned byte (1\,GB).}
 \end{figure}
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
-  \caption{\label{fBenchmarkEqPly} David statue at 1mm resolution (988MB).}
+  \caption{\label{fBenchmarkEqPly} David statue at 1\,mm resolution (988\,MB).}
 \end{figure}
@@
 The data distribution benchmarks have been performed on a cluster using
-dual processor Intel Xeon X5690 CPUs (3.47GHz), 24 GB main memory, 10GBit/s
+dual processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory, 10\,GBit/s
 Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake 3.2 release
@@
 
-To benchmark the data distribution we used two datasets: The David statue at 2mm
-resolution (as in \fig{fBenchmarkEqPly}, but in 2mm resolution) and 3D volumes
+To benchmark the data distribution we used two datasets: The David statue at 2\,mm
+resolution (as in \fig{fBenchmarkEqPly}, but in 2\,mm resolution) and 3D volumes
 of spike frequencies of an electrical simulation of three million neurons
@@
     spike frequencies of a three million neuron electrical simulation at
-    512x437x240 unsigned byte (51MB, left) and an MicroCT scan of a beechnut at
-    1024x1024x1546 unsigned short (3GB, right) resolution}
+    512x437x240 unsigned byte (51\,MB, left) and an MicroCT scan of a beechnut at
+    1024x1024x1546 unsigned short (3\,GB, right) resolution}
 \end{figure}
@@
 
-The David statue at 2mm resolution is organized in a kD-tree for rendering. Each
+The David statue at 2\,mm resolution is organized in a kD-tree for rendering. Each
 node is a separate distributed object, which has two child node objects. A
@@
 which is reflected by the results in \fig{fSpikesDist}. Compression for this
-data is beneficial for transmitting data over a 10~Gb/s link even for a single
+data is beneficial for transmitting data over a 10\,Gb/s %why was there a tilde inbetween? Should this be ~10Gb/s instead?
+link even for a single
 slave process. Buffering has little benefit since the serialization of volume
@@
 compression cost can be amortized over $n$ nodes, reaching raw data transmission
-rates of 3.7 GB/s with the default Snappy compressor and at best 4.4 GB/s with
+rates of 3.7\,GB/s with the default Snappy compressor and at best 4.4\,GB/s with
 ZStandard at level 1.
@@
 (\fig{fBeechnutDist}). Due to the larger object size, uncompressed transmission
-is slightly faster compared to the Spike data set at 700 MB/s, and compressed
+is slightly faster compared to the Spike data set at 700\,MB/s, and compressed
 transmission does not improve the mapping performance, likely due to increased
@@
   \caption{\label{fCompressorDetail}Data compression for ply data (left, David
-    statue 2mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
+    statue 2\,mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
 \end{figure*}
@@
   \caption{\label{fDavidDist}Object mapping (left) and data synchronization
-    time (middle, detail view right) for the David 2mm data set}
+    time (middle, detail view right) for the David 2\,mm data set}
 \end{figure*}

### 2018-02-08 paperIEEE.tex
hardware again
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@114 db7c848b-7755-4df2-b763-8796797afa52

@@
 We conducted performance benchmarks for the decomposition modes on a
-rendering cluster using hexacore Intel Xeon E5-2620v3 CPUs ($2.4$\,GHz), $16$\,GB DDR4 RAM per node, and nVidia GTX $970$ GPUs with $4$\,GB VRAM each.
-The nodes are connected via a $40$\,Gbit/s {\em Infiniband} network.
+rendering cluster using hexacore Intel Xeon E5-2620v3 CPUs (2.4\,GHz), 16\,GB DDR4 RAM per node, nVidia GTX 970 GPUs with 4\,GB VRAM each, and QDR Infiniband.
 GCC 4.8 has been used with CMake 3.7 release mode settings to compile the software stack.

### 2018-02-08 paperIEEE.tex
typo, clarification
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@115 db7c848b-7755-4df2-b763-8796797afa52

@@
 compute resources fluctuate randomly every frame, as can arguably be the case
-for virutalized enviroments. For this scenario tile equalizer seems best suited,
+for virtualized enviroments. For this scenario tile equalizer seems best suited,
 as it performs load balancing implicitly and does not assume coherence of
@@
 The data distribution benchmarks have been performed on a cluster using
-dual processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory, 10\,GBit/s
+dual processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory per node, 10\,GBit/s
 Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake 3.2 release

### 2018-02-08 paperIEEE.tex
minor fixes
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@116 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-The ply file is converted into a kd-Tree for fast view frustum culling and
+The PLY file is converted into a k-d tree for fast view frustum culling and
 rendering, and the resulting data structure is serialized in binary form for
@@
 
-The ply data is little compressible, the default compressor achieves a 10\%
+The PLY data is little compressible, the default compressor achieves a 10\%
 reduction. This is due to a high entropy in the data and the dominant use of
@@
 
-The David statue at 2\,mm resolution is organized in a kD-tree for rendering. Each
+The David statue at 2\,mm resolution is organized in a k-d tree for rendering. Each
 node is a separate distributed object, which has two child node objects. A
@@
   \includegraphics[width=\textwidth]{images/compressorDetail}
-  \caption{\label{fCompressorDetail}Data compression for ply data (left, David
+  \caption{\label{fCompressorDetail}Data compression for PLY data (left, David
     statue 2\,mm) and raw volumes shown in \fig{fVolumes} (middle and right)}

### 2018-02-08 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@118 db7c848b-7755-4df2-b763-8796797afa52

--- 2018-02-08 paperIEEE.tex
@@
 analyses data distribution performance, in particular how the various
-optimizations in \textsf{Collage} effect complex operations in realistic
+optimizations in \textsf{Collage} affect complex operations in realistic
 scenarios.
@@
 
-% Time to render a 4k, 32-step MSAA image of the 1\,mm David model and Livre using 1 to 10 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps), tiles, chunks, pixel, subpixel) (+DFR for Livre).
-
-We conducted performance benchmarks for the decomposition modes on a
-rendering cluster using hexacore Intel Xeon E5-2620v3 CPUs (2.4\,GHz), 16\,GB DDR4 RAM per node, nVidia GTX 970 GPUs with 4\,GB VRAM each, and QDR Infiniband.
-GCC 4.8 has been used with CMake 3.7 release mode settings to compile the software stack.
+% Time to render a 4k, 32-step MSAA image of the 1mm David model and Livre using 1 to 10 nodes with all modes (2D, 2D LB, DB, DPlex (over AA steps), tiles, chunks, pixel, subpixel) (+DFR for Livre).
+
+We conducted new performance benchmarks for various decomposition modes on a cluster
+using hexacore Intel Xeon E5-2620v3 CPUs (2.4GHz), \FIXME{GPUs}, 16 GB main
+memory, 4 GBit/s Ethernet and QDR Infiniband. GCC 4.8 has been used with CMake
+3.7 release mode settings to compile the software stack.
 
@@
 and polygonal data  (\fig{fBenchmarkEqPly}) and used test scenes that allowed us
-to finely adjust the rendering load the system has to cope with. In both cases
+to continuously modify the rendering load the system has to cope with. In both cases
 the scene comprised of two rows of instantiated, identical models with 30 models
-in total. Rendering was performed at a resolution of 2560x1440. The camera is
+in total. Rendering was performed at an output resolution of $2560\times1440$. The camera is
 initially placed in the center of the scene, between the two rows
@@
 dedicated application/display node. We subsequently summed up the duration of
-all rendered frames for each run (\fig{fCompounds}). We provide the results of
-the load balancer providing the better result for each application and compound.
+all rendered frames for each run (\fig{fCompounds}), backing up along the alley of models.
+We provide the results of the load balancer providing the better result for each application and compound.\FIXME{what does this last sentence mean?}
 The simpler \textsf{tree\_equalizer} outperforms in almost all cases the
@@
 theoretically better, but more complex implementations.
+
+\FIXME{oscillation, pixel, clearer}
+\FIXME{I'm not sure I read Fig.10 the same way.
+Left it seems that tile\_equalizer is the best for polygonal rendering, it's constantly faster than the tree\_equalizer mentioned in the paragraph above.}
 
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eVolve}
-  \caption{\label{fBenchmarkEVolve} Flower volume model at $1024^3$ unsigned byte (1\,GB).}
+  \caption{\label{fBenchmarkEVolve} Flower alley with 30 volume models of $1024^3$ unsigned bytes (1GB) each.}
 \end{figure}
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
-  \caption{\label{fBenchmarkEqPly} David statue at 1\,mm resolution (988\,MB).}
+  \caption{\label{fBenchmarkEqPly} David alley with 30 statues of 56m triangles (988MB) each.\FIXME{check the triangle count}}
 \end{figure}
@@
 The data distribution benchmarks have been performed on a cluster using
-dual processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory, 10\,GBit/s
+dual processor Intel Xeon X5690 CPUs (3.47GHz), 24 GB main memory, 10GBit/s
 Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake 3.2 release
@@
 
-To benchmark the data distribution we used two datasets: The David statue at 2\,mm
-resolution (as in \fig{fBenchmarkEqPly}, but in 2\,mm resolution) and 3D volumes
+To benchmark the data distribution we used two datasets: The David statue at 2mm
+resolution (as in \fig{fBenchmarkEqPly}, but in 2mm resolution) and 3D volumes
 of spike frequencies of an electrical simulation of three million neurons
@@
     spike frequencies of a three million neuron electrical simulation at
-    512x437x240 unsigned byte (51\,MB, left) and an MicroCT scan of a beechnut at
-    1024x1024x1546 unsigned short (3\,GB, right) resolution}
+    512x437x240 unsigned byte (51MB, left) and an MicroCT scan of a beechnut at
+    1024x1024x1546 unsigned short (3GB, right) resolution}
 \end{figure}
@@
 
-The David statue at 2\,mm resolution is organized in a kD-tree for rendering. Each
+The David statue at 2mm resolution is organized in a kD-tree for rendering. Each
 node is a separate distributed object, which has two child node objects. A
@@
 which is reflected by the results in \fig{fSpikesDist}. Compression for this
-data is beneficial for transmitting data over a 10\,Gb/s %why was there a tilde inbetween? Should this be ~10Gb/s instead?
-link even for a single
+data is beneficial for transmitting data over a 10~Gb/s link even for a single
 slave process. Buffering has little benefit since the serialization of volume
@@
 compression cost can be amortized over $n$ nodes, reaching raw data transmission
-rates of 3.7\,GB/s with the default Snappy compressor and at best 4.4\,GB/s with
+rates of 3.7 GB/s with the default Snappy compressor and at best 4.4 GB/s with
 ZStandard at level 1.
@@
 (\fig{fBeechnutDist}). Due to the larger object size, uncompressed transmission
-is slightly faster compared to the Spike data set at 700\,MB/s, and compressed
+is slightly faster compared to the Spike data set at 700 MB/s, and compressed
 transmission does not improve the mapping performance, likely due to increased
@@
   \caption{\label{fCompressorDetail}Data compression for ply data (left, David
-    statue 2\,mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
+    statue 2mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
 \end{figure*}
@@
   \caption{\label{fDavidDist}Object mapping (left) and data synchronization
-    time (middle, detail view right) for the David 2\,mm data set}
+    time (middle, detail view right) for the David 2mm data set}
 \end{figure*}

### 2018-02-08 paperIEEE.tex
minor fixes
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@119 db7c848b-7755-4df2-b763-8796797afa52

--- 2018-02-08 paperIEEE.tex
@@
 
-We conducted performance benchmarks for the decomposition modes on a
-rendering cluster using hexacore Intel Xeon E5-2620v3 CPUs (2.4\,GHz), 16\,GB DDR4 RAM per node, nVidia GTX 970 GPUs with 4\,GB VRAM each, and QDR Infiniband.
-GCC 4.8 has been used with CMake 3.7 release mode settings to compile the software stack.
+We conducted new performance benchmarks for various decomposition modes on a cluster
+using hexacore Intel Xeon E5-2620v3 CPUs (2.4\,GHz), nVidia GTX 970 GPUs with 4\,GB VRAM each, 16\,GB main
+memory per node, 4\,GBit/s Ethernet, and QDR Infiniband. GCC 4.8 has been used with CMake
+3.7 release mode settings to compile the software stack.
 
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eVolve}
-  \caption{\label{fBenchmarkEVolve} Flower volume model at $1024^3$ unsigned byte (1\,GB).}
+  \caption{\label{fBenchmarkEVolve} Flower alley with 30 volume models of $1024^3$ unsigned bytes (1\,GB) each.}
 \end{figure}
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
-  \caption{\label{fBenchmarkEqPly} David statue at 1\,mm resolution (988\,MB).}
+  \caption{\label{fBenchmarkEqPly} David alley with 30 statues of 56m triangles (988\,MB) each.\FIXME{check the triangle count}}
 \end{figure}
@@
 The data distribution benchmarks have been performed on a cluster using
-dual processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory per node, 10\,GBit/s
+dual processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory, 10\,GBit/s
 Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake 3.2 release
@@
 which is reflected by the results in \fig{fSpikesDist}. Compression for this
-data is beneficial for transmitting data over a 10\,Gb/s %why was there a tilde inbetween? Should this be ~10Gb/s instead?
-link even for a single
+data is beneficial for transmitting data over a 10~Gb/s \FIXME{~10\,Gb/s ?} link even for a single
 slave process. Buffering has little benefit since the serialization of volume
@@
 compression cost can be amortized over $n$ nodes, reaching raw data transmission
-rates of 3.7\,GB/s with the default Snappy compressor and at best 4.4\,GB/s with
+rates of 3.7~GB/s \FIXME{~3.7\,GB/s ?} with the default Snappy compressor and at best 4.4~GB/s \FIXME{~4.4\,GB/s ?} with
 ZStandard at level 1.

### 2018-02-08 paperIEEE.tex
removed fixmes
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@121 db7c848b-7755-4df2-b763-8796797afa52

@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
-  \caption{\label{fBenchmarkEqPly} David alley with 30 statues of 56m triangles (988\,MB) each.\FIXME{check the triangle count}}
+  \caption{\label{fBenchmarkEqPly} David alley with 30 statues of 56\,M triangles (988\,MB) each.
 \end{figure}
@@
 which is reflected by the results in \fig{fSpikesDist}. Compression for this
-data is beneficial for transmitting data over a 10~Gb/s \FIXME{~10\,Gb/s ?} link even for a single
+data is beneficial for transmitting data over a 10\,Gb/s link even for a single
 slave process. Buffering has little benefit since the serialization of volume
@@
 compression cost can be amortized over $n$ nodes, reaching raw data transmission
-rates of 3.7~GB/s \FIXME{~3.7\,GB/s ?} with the default Snappy compressor and at best 4.4~GB/s \FIXME{~4.4\,GB/s ?} with
+rates of 3.7\,GB/s with the default Snappy compressor and at best 4.4\,GB/s with
 ZStandard at level 1.

### 2018-02-09 paperIEEE.tex
fixed error
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@122 db7c848b-7755-4df2-b763-8796797afa52
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@123 db7c848b-7755-4df2-b763-8796797afa52

@@
 queue. Prefetching ensures that the task communication overlaps with rendering.
-As shown in~ \cite{SPEP:16}, these modes can provide better performance due to
-being inherently load-balanced, as long as there is an insignificant overhead
-for the render task setup. This mode is transparent to \textsf{Equalizer}
-applications.
+As shown in~\cite{SPEP:16}, these modes can provide better performance due to
+being implicitly, i.e., inherently load-balanced, as long as there is an insignificant overhead for the render task setup. This mode is transparent to \textsf{Equalizer} applications.
 
@@
 
+\subsubsection{Implicit Load Balancing}
+
+The package equalizer allows to adjusts the implicit load balancing that results from rendering clients pulling work packages, i.e., tiles and chunks, from a central queue on the server.
+It is implemented as \textsf{tile\_equalizer} and \textsf{chunk\_equalizer}, respectively, and changes the affinity of work packages to individual nodes, based on an affinity model specified to the equalizer.
+In~\cite{SPEP:16}, we explore this approach in more detail.
+
 %------------------------------------------------------------------------------
@@
 all rendered frames for each run (\fig{fCompounds}), backing up along the alley of models.
-We provide the results of the load balancer providing the better result for each application and compound.\FIXME{what does this last sentence mean?}
+% We provide the results of the load balancer providing the better result for each application and compound.\FIXME{what does this last sentence mean?}
 The simpler \textsf{tree\_equalizer} outperforms in almost all cases the
@@
 rendering where the load in the region of interest is relatively uniform. This
-counterintuitive result once more proves that simple algorithms often outperform
-theoretically better, but more complex implementations.
+counterintuitive result seems to again confirm the idea that simple algorithms often outperform theoretically better, but more complex implementations.
+On the other hand, \textsf{tile\_equalizer} often outperforms \textsf{tree\_equalizer}.
+This suggests that the underlying implicit load balancing can be superior to the the explicit methods of \textsf{load\_equalizer} and \textsf{tree\_equalizer} in high load situations, where the additional overhead of tile generation and distribution is more justified.
 
 \FIXME{oscillation, pixel, clearer}
-\FIXME{I'm not sure I read Fig.10 the same way.
+% \FIXME{I'm not sure I read Fig.10 the same way.
 Left it seems that tile\_equalizer is the best for polygonal rendering, it's constantly faster than the tree\_equalizer mentioned in the paragraph above.}
@@
   \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
-  \caption{\label{fBenchmarkEqPly} David alley with 30 statues of 56\,M triangles (988\,MB) each.
+  \caption{\label{fBenchmarkEqPly} David alley with 30 statues of 56\,M triangles (988\,MB) each.}
 \end{figure}

### 2018-02-12 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@124 db7c848b-7755-4df2-b763-8796797afa52

@@
 
+\subsubsection{Dynamic Work Packages}
+
+The package equalizer allows to adjusts the implicit load balancing that results from rendering clients pulling work packages, i.e., tiles and chunks, from a central queue on the server.
+It is implemented as \textsf{tile\_equalizer} and \textsf{chunk\_equalizer}, respectively, and changes the assignment of work packages to individual nodes, based on an affinity model specified to the equalizer.
+In~\cite{SPEP:16}, we explore this approach in more detail.
+
 \subsubsection{Cross-Segment Load Balancing}
@@
 segments.
-
-\subsubsection{Implicit Load Balancing}
-
-The package equalizer allows to adjusts the implicit load balancing that results from rendering clients pulling work packages, i.e., tiles and chunks, from a central queue on the server.
-It is implemented as \textsf{tile\_equalizer} and \textsf{chunk\_equalizer}, respectively, and changes the affinity of work packages to individual nodes, based on an affinity model specified to the equalizer.
-In~\cite{SPEP:16}, we explore this approach in more detail.
 

### 2018-02-13 paperIEEE.tex
Preflight check
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@125 db7c848b-7755-4df2-b763-8796797afa52

--- 2018-02-08 paperIEEE.tex
@@
 queue. Prefetching ensures that the task communication overlaps with rendering.
-As shown in~ \cite{SPEP:16}, these modes can provide better performance due to
-being inherently load-balanced, as long as there is an insignificant overhead
-for the render task setup. This mode is transparent to \textsf{Equalizer}
-applications.
+As shown in~\cite{SPEP:16} and the results, these modes can provide better
+performance due to being implicitly, i.e., inherently load-balanced, as long as
+there is an insignificant overhead for the render task setup. This mode is
+transparent to \textsf{Equalizer} applications.
 
@@
 
+\subsubsection{Dynamic Work Packages}
+
+The package equalizers implement client-affinity models for tile and chunk
+compounds (\sref{sec:tile}). A \textsf{tile\_equalizer} and
+\textsf{chunk\_equalizer} create the packages and change the assignment of them
+to individual nodes, based on an affinity model specified in the equalizer.
+In~\cite{SPEP:16}, we explore this approach in detail.
+
 \subsubsection{Cross-Segment Load Balancing}
@@
 
-Due to the master-slave model of data distribution, multicast can be used to
+Due to the master-slave model of data distribution, multicast is used to
 optimize the transmission time of data. If the contributing nodes share a
 multicast session, and more than one slave instance is mapped, \textsf{Collage}
-automatically uses the multicast connection to send the new version
-information.
+automatically uses the multicast connection to send the new version information.
 
@@
 \textsf{Collage} implements a few generic distributed objects which are used by
-\textsf{Equalizer}
-and applicatons. A barrier is a distributed barrier primitive used for software
-swap barriers in \textsf{Equalizer} (\sref{sec:swap}). Its implementation
-follows a
-simple master-slave approach, which has shown to by sufficient for this use
-case. Queues are single producer, multiple consumer FIFOs. To hide network
-latencies, the consumers use prefetching of queue items into a local queue. They
-are used for tile and chunk compounds (\sref{sec:tile}).
+\textsf{Equalizer} and other applicatons. A barrier is a distributed barrier
+primitive used for software swap barriers in \textsf{Equalizer}
+(\sref{sec:swap}). Its implementation follows a simple master-slave approach,
+which has shown to by sufficient for this use case. Queues are single producer,
+multiple consumer FIFOs. To hide network latencies, consumers prefetch queue
+items into a local queue. They are used for tile and chunk compounds
+(\sref{sec:tile}).
 
@@
 publications~\cite{EP:07, EMP:09, MEP:10, EEP:11, EBAHMP:12, HBBES:13, deflect,
-  SPEP:16}. The first part summarizes rendering performance over all
-decomposition modes with a few representative workloads. The second part
-analyses data distribution performance, in particular how the various
-optimizations in \textsf{Collage} affect complex operations in realistic
-scenarios.
+SPEP:16}. The first part summarizes rendering performance over all decomposition
+modes with a few representative workloads. The second part analyses data
+distribution performance, in particular how the optimizations in
+\textsf{Collage} perform in realistic scenarios.
 
@@
 
-We tested the decomposition modes with both volume data (\fig{fBenchmarkEVolve})
-and polygonal data  (\fig{fBenchmarkEqPly}) and used test scenes that allowed us
-to continuously modify the rendering load the system has to cope with. In both cases
-the scene comprised of two rows of instantiated, identical models with 30 models
-in total. Rendering was performed at an output resolution of $2560\times1440$. The camera is
-initially placed in the center of the scene, between the two rows
-(\fig{fBenchmarkEVolve} and \fig{fBenchmarkEqPly}), rendering only half of the
-model instances. It is then moved backward over the duration of 800 frames,
-steadily increasing the rendering load by revealing more models, until all 30
-instances are visible.
+We tested the decomposition modes with both polygonal data and volume data
+(\fig{fBenchmarks} (middle and left)), using test scenes that allowed to adapt
+the rendering load the system has to cope with. In both cases the scene is
+comprised of two rows of instantiated, identical models with 30 models in total.
+Rendering was performed at an output resolution of $2560\times1440$. The camera
+is initially placed in the center of the scene, between the two rows, rendering
+only half of the model instances. It is then moved backward over the duration of
+800 frames, steadily increasing the rendering load by revealing more models,
+until all 30 instances are visible.
 
 We investigated the scalability of individual decomposition modes by running the
-same experiment using a varying number of render nodes (2-9), as well as one
-dedicated application/display node. We subsequently summed up the duration of
-all rendered frames for each run (\fig{fCompounds}), backing up along the alley of models.
-We provide the results of the load balancer providing the better result for each application and compound.\FIXME{what does this last sentence mean?}
+same experiment using a varying number of render nodes (2-9) and one dedicated
+application/display node. We subsequently summed up the duration of all rendered
+frames for each run (\fig{fCompounds}).
+
+For sort-first and sort-last rendering we present static and load-balanced task
+decomposition. For readability, we only present the results of the equalizer
+(load or tree) providing the better performance for each application.
+Unsurprisingly, static decompositions perform worse over load-balanced
+compounds. Sort-first polygon rendering exhibits oscillations in performance as
+nodes are added to the task, due to unfavorable assignment of a tile with a high
+work load on odd node counts. Static sort-last volume rendering has a similar
+oscillation behaviour \FIXME{due to @David?}.
+
 The simpler \textsf{tree\_equalizer} outperforms in almost all cases the
@@
 rendering where the load in the region of interest is relatively uniform. This
-counterintuitive result once more proves that simple algorithms often outperform
-theoretically better, but more complex implementations.
-
-\FIXME{oscillation, pixel, clearer}
-\FIXME{I'm not sure I read Fig.10 the same way.
-Left it seems that tile\_equalizer is the best for polygonal rendering, it's constantly faster than the tree\_equalizer mentioned in the paragraph above.}
+counterintuitive result seems to again confirm that simple algorithms often
+outperform theoretically better, but more complex implementations. On the other
+hand, the \textsf{tile\_equalizer} often outperforms \textsf{tree\_equalizer}.
+This suggests that the underlying implicit load balancing can be superior to the
+explicit methods of \textsf{load\_equalizer} and \textsf{tree\_equalizer} in
+high load situations, where the additional overhead of tile generation and
+distribution is more justified. The relative simple nature of our benchmark
+application's rendering algorithms is also favoring work packages, since it has
+a near-zero static overhead per rendering pass.
+
+Last, but not least, we also provide scalability results for pixel compounds.
+While naturally load-balanced, pixel compounds only scale fill rate and not
+geometry processing. Consequently, pixel compounds provide better performance
+for volume rendering, and a predictable scaling behaviour for both.
 
@@
 
-\begin{figure}[ht]\center
-  \includegraphics[width=.9\columnwidth]{images/benchmark_eVolve}
-  \caption{\label{fBenchmarkEVolve} Flower alley with 30 volume models of $1024^3$ unsigned bytes (1\,GB) each.}
-\end{figure}
-
-\begin{figure}[ht]\center
-  \includegraphics[width=.9\columnwidth]{images/benchmark_eqPly2}
-  \caption{\label{fBenchmarkEqPly} David alley with 30 statues of 56\,M triangles (988\,MB) each.
-\end{figure}
+\begin{figure*}[ht]\center
+    \includegraphics[height=.17\textwidth]{images/benchmark_eVolve}
+    \includegraphics[height=.17\textwidth]{images/benchmark_eqPly2}
+    \includegraphics[height=.17\textwidth]{images/spikes512}
+    \includegraphics[height=.17\textwidth]{images/beechnut}
+  \caption{\label{fBenchmarks}Benchmark data: Flower alley with 30 volume models
+  of $1024^3$ unsigned bytes (1\,GB) each (left); David alley with 30 statues of
+  56\,M triangles (988\,MB) each (middle); Volumes used for object distribution
+  benchmarks (right): spike frequencies of a three million neuron electrical
+  simulation at 512x437x240 unsigned byte (51\,MB) and an MicroCT scan of a
+  beechnut at 1024x1024x1546 unsigned short (3\,GB) resolution}
+\end{figure*}
 
@@
 
-The data distribution benchmarks have been performed on a cluster using
-dual processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory per node, 10\,GBit/s
-Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake 3.2 release
-mode settings to compile the software stack.
-
-To benchmark the data distribution we used two datasets: The David statue at 2\,mm
-resolution (as in \fig{fBenchmarkEqPly}, but in 2\,mm resolution) and 3D volumes
-of spike frequencies of an electrical simulation of three million neurons
-(\fig{fVolumes}).
+The data distribution benchmarks have been performed on a cluster using dual
+processor Intel Xeon X5690 CPUs (3.47\,GHz), 24\,GB main memory per node,
+10\,GBit/s Ethernet and QDR Infiniband. Intel ICC 2017 has been used with CMake
+3.2 release mode settings to compile the software stack. To benchmark the data
+distribution we used two datasets: The David statue at 2\,mm resolution (as in
+\fig{fBenchmarks (middle)}, but in 2\,mm resolution) and 3D volumes of spike
+frequencies of an electrical simulation of three million neurons
+(\fig{fBenchmarks} (right)).
 
@@
 rendering, and the resulting data structure is serialized in binary form for
-data transmission.
-
-\begin{figure}[ht]\center
-  \includegraphics[height=.44\columnwidth]{images/spikes512}\hfil
-  \includegraphics[height=.44\columnwidth]{images/beechnut}
-  \caption{\label{fVolumes}Volumes used for object distribution benchmarks:
-    spike frequencies of a three million neuron electrical simulation at
-    512x437x240 unsigned byte (51\,MB, left) and an MicroCT scan of a beechnut at
-    1024x1024x1546 unsigned short (3\,GB, right) resolution}
-\end{figure}
-
-The spike frequency volumes aggregate the number of spikes which happened within
-a given voxel over a given time. The absolute spike count is renormalized to an
-unsigned byte value during creation. Higher densities in the volume represent
-higher spiking activity in the voxel.
+data transmission. The spike frequency volumes aggregate the number of spikes
+which happened within a given voxel over a given time. The absolute spike count
+is renormalized to an unsigned byte value during creation. Higher densities in
+the volume represent higher spiking activity in the voxel.
 
@@
 second as well as the size of the compressed data relative to the uncompressed
-data. The different ZSTD$x$ engines use the ZStandard compression library at
-compression level $x$. The measurements were performed on a single, isolated
-node.
+data. The ZSTD$x$ engines use the ZStandard compression library at compression
+level $x$. The measurements were performed on a single, isolated node.
 
@@
 disabled to focus on the raw network performance. \fig{fIFDist} shows the
-performance using various protocols.
-
-Unsurprisingly, TCP over the faster InfiniBand link outperforms the cheaper ten
-gigabit ethernet link by more than a factor of two. Unexpectedly, the native
-RDMA connection performs worse, even though it outperforms IPoIB in a simple
-peer-to-peer connection benchmark. This needs further investigation, but we
-suspect the abstraction of a byte stream connection chosen by \textsf{Collage}
-is not well suited for remote DMA semantics; that is, one needs to design the
-network API around zero-copy semantics with managed memory for modern high-speed
-transports. Both Infiniband connections show significant measurement jitter.
+performance using various protocols. TCP over the faster InfiniBand link
+outperforms the cheaper ten gigabit ethernet link by more than a factor of two.
+Unexpectedly, the native RDMA connection performs worse, even though it
+outperforms IPoIB in a simple peer-to-peer connection benchmark. This needs
+further investigation, but we suspect the abstraction of a byte stream
+connection chosen by \textsf{Collage} is not well suited for remote DMA
+semantics; that is, one needs to design the network API around zero-copy
+semantics with managed memory for modern high-speed transports. Both Infiniband
+connections show significant measurement jitter.
 
@@
   \caption{\label{fCompressorDetail}Data compression for PLY data (left, David
-    statue 2\,mm) and raw volumes shown in \fig{fVolumes} (middle and right)}
+    statue 2\,mm) and raw volumes shown in \fig{fBenchmarks} (right)}
 \end{figure*}
@@
     time (middle, detail view right) for the Spike data set in
-    \fig{fVolumes}, left}
+    \fig{fBenchmarks} (right)}
 \end{figure*}
@@
     time (middle, detail view right) for the beechnut data set in
-    \fig{fVolumes}, right}
+    \fig{fBenchmarks} (right)}
 \end{figure*}

### 2018-02-13 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@126 db7c848b-7755-4df2-b763-8796797afa52

@@
 work load on odd node counts. Static sort-last volume rendering has a similar
-oscillation behaviour \FIXME{due to @David?}.
+oscillation behaviour, as ranges of scene geometry tend to also get unfavorably assigned under such conditions.
 
@@
 high load situations, where the additional overhead of tile generation and
-distribution is more justified. The relative simple nature of our benchmark
+distribution is more justified. The relatively simple nature of our benchmark
 application's rendering algorithms is also favoring work packages, since it has

### 2018-02-13 paperIEEE.tex
smal fix.
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@127 db7c848b-7755-4df2-b763-8796797afa52

@@
 distribution we used two datasets: The David statue at 2\,mm resolution (as in
-\fig{fBenchmarks (middle)}, but in 2\,mm resolution) and 3D volumes of spike
+\fig{fBenchmarks} (middle), but in 2\,mm resolution) and 3D volumes of spike
 frequencies of an electrical simulation of three million neurons

### 2018-02-13 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@128 db7c848b-7755-4df2-b763-8796797afa52

@@
   benchmarks (right): spike frequencies of a three million neuron electrical
-  simulation at 512x437x240 unsigned byte (51\,MB) and an MicroCT scan of a
-  beechnut at 1024x1024x1546 unsigned short (3\,GB) resolution}
+  simulation at $512\times437\times240$ unsigned byte (51\,MB) and an MicroCT scan of a
+  beechnut at $1024\times1024\times1546$ unsigned short (3\,GB) resolution}
 \end{figure*}

### 2018-02-13 paperIEEE.tex
minor fixes and fixmes.
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@129 db7c848b-7755-4df2-b763-8796797afa52

@@
 installations~\cite{EEP:11}; asynchronous compositing and NUMA
-optimizations~\cite{EBAHMP:12}; as well as work queueing~\cite{SPEP:16}.
+optimizations~\cite{EBAHMP:12}; as well as work queueing~\cite{SPEP:16}. \FIXME{Why ; instead of , ?}
 Additionally, complex large scale and out-of-core multiresolution
@@
 based on past rendering times. This assumes a reasonable frame-to-frame
-coherency. Equalizer implements two different algorithms, a
+coherency \FIXME{Shouldn't that be frame-to-frame {\em coherence}?}. Equalizer implements two different algorithms, a
 \textsf{load\_equalizer} and a \textsf{tree\_equalizer}. The result section
@@
 The package equalizers implement client-affinity models for tile and chunk
-compounds (\sref{sec:tile}). A \textsf{tile\_equalizer} and
-\textsf{chunk\_equalizer} create the packages and change the assignment of them
+compounds (\sref{sec:tile}). A \textsf{tile\_equalizer} or
+\textsf{chunk\_equalizer} creates the packages and changes the assignment of them
 to individual nodes, based on an affinity model specified in the equalizer.
@@
 adapts the resolution of the source channel and zoom factor for the source to
-destination transfer. If the performance and source channel resolutions allows,
+destination transfer. If the performance and source channel resolutions allow,
 this will not only subsample, but also supersample the destination channel to
@@
 
+\FIXME{Sentence incomplete.}
 Asynchronous compositing pipelines rendering with compositing operations, by
@@
 between GPU and CPU. \textsf{Equalizer} supports a variety of compression
-algorithms, from very fast RLE encoding, YUV subsampling on the GPU to JPEG
+algorithms, from very fast run-length encoding and YUV subsampling on the GPU to JPEG
 compression. These algorithms are implemented as runtime-loaded plugins,

### 2018-02-14 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@130 db7c848b-7755-4df2-b763-8796797afa52

@@
 based on past rendering times. This assumes a reasonable frame-to-frame
-coherency \FIXME{Shouldn't that be frame-to-frame {\em coherence}?}. Equalizer implements two different algorithms, a
+coherency. \FIXME{Shouldn't that be frame-to-frame {\em coherence}?} Equalizer implements two different algorithms, a
 \textsf{load\_equalizer} and a \textsf{tree\_equalizer}. The result section
@@
 between GPU and CPU. \textsf{Equalizer} supports a variety of compression
-algorithms, from very fast run-length encoding and YUV subsampling on the GPU to JPEG
+algorithms, from very fast run-length encoding (RLE) and YUV subsampling on the GPU to JPEG
 compression. These algorithms are implemented as runtime-loaded plugins,
@@
 sequential semantics between OpenGL rendering and event handling. In contrast,
-Qt requires all windows and QOpenGLContext to be created from the Qt main
+Qt requires all windows and each QOpenGLContext to be created from the Qt main
 thread. An existing Qt window or context may subsequently be moved to a
@@
 one renderer instance is created for each (pipe) render thread. They also
-provide the natural place to store and distribute data. Last, but not least,
+provide the natural place to store and distribute data. Finally,
 \textsf{ViewData} provides a convenient way to manage multiple views by storing
@@
 application programmer implements serialization and deserialization. Distributed
-objects can be static (immutable) or dynamic. Objects have an universally unique
-identifier (UUID) as cluster-wide addresses. A master-slave model is used to
+objects can be static (immutable) or dynamic. Objects have a universally unique
+identifier (UUID) as cluster-wide address. A master-slave model is used to
 establish mapping and data synchronization across processes. Typically, the
@@
 \item[Unbuffered] The object is versioned and unbuffered. No data is stored, and
-  no previous versions can be mapped. No old versions can be mapped and no
+  no previous versions can be mapped. \FIXME{Redundant.} No old versions can be mapped and no
   additional data is stored.
@@
 Serialization is facilitated using output or input streams, which abstract the
-data transmission and are used like a \textsf{std::stream}. The data streams
+data transmission and are used like an \textsf{std::stream}. The data streams
 implement efficient buffering and compression, and automatically select the best
@@
 \textsf{sync} method might block if a version has not yet been committed or is
-still in transmission. All versioned objects the following characteristics:
+still in transmission. All versioned objects have the following characteristics:
 
@@
 \item Slave instance versions can only be advanced; that is, \textsf{sync(
-  version )} with a version smaller than the current version will fail.
+  version)} with a version smaller than the current version will fail.
 \item Newly mapped slave instances are mapped to the oldest available
-  version by default, or to the version specified when calling
+  version by default \FIXME{Is this correct?}, or to the version specified when calling
   \textsf{mapObject}.
@@
 compression rates well above one gigabyte per second. \textsf{Collage} uses the
-Pression library~\cite{pression}, which provides an unified interface for a
+Pression library~\cite{pression}, which provides a unified interface for a
 number of compression libraries, such as FastLZ~\cite{jesperfast},
@@
 master objects are registered. The preloading sends instance data of recently
-registered master objects while the corresponding node is idle to all connected
-nodes. These nodes simply enter the received data to their cache. Preloading
+registered master objects to all connected
+nodes, while the corresponding node is idle. These nodes simply enter the received data to their cache. Preloading
 uses multicast when available.
@@
 compute resources fluctuate randomly every frame, as can arguably be the case
-for virtualized enviroments. For this scenario tile equalizer seems best suited,
+for shared rendering nodes in virtualized enviroments. For this scenario tile equalizer seems best suited,
 as it performs load balancing implicitly and does not assume coherence of
@@
 the data compression algorithm. Our microbenchmark compresses a set of binary
-files, precalculating the speed and compression ratio of the various engines.
+files \FIXME{What are these files? Fig. 7 only speaks of "generic" binary data...}, precalculating the speed and compression ratio of the various engines.
 \fig{fCompressor} shows the compression and decompression speed in gigabyte per
@@
 \htmladdnormallinkfoot{implementation}{https://github.com/Eyescale/Equalizer/tree/paper2018/tools/eqObjectBench}.
+
+\FIXME{Sentence incomplete.}
 used the same data sets as in the previous section. We ran the benchmark on up

### 2018-02-14 paperIEEE.tex
Lots of little changes and a few last  fixmes.
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@131 db7c848b-7755-4df2-b763-8796797afa52

@@
 primitive used for software swap barriers in \textsf{Equalizer}
-(\sref{sec:swap}). Its implementation follows a simple master-slave approach,
-which has shown to by sufficient for this use case. Queues are single producer,
-multiple consumer FIFOs. To hide network latencies, consumers prefetch queue
-items into a local queue. They are used for tile and chunk compounds
+(\sref{sec:swap}) \FIXME{Using the word barrier three times in one short sentence}. Its implementation follows a simple master-slave approach,
+which has shown to by sufficient for this use case. The used distributed data structures are single producer,
+multiple consumer FIFO queues. To hide network latencies, consumers prefetch stored
+items into a local queue. Queues are used for tile and chunk compounds
 (\sref{sec:tile}).
@@
 high-resolution display walls and CAVEs, as well as for virtual prototyping of
-human-machine interactions CAVEs and HMDs (\fig{FIG_teaser}(b)). For
+human-machine interactions using CAVEs and HMDs (\fig{FIG_teaser}(b)). For
 scalability, sort-first and tile compounds are used to achieve fast,
@@
 RTNeuron \cite{HBBES:13} is a scalable real-time rendering tool for the
-visualisation of neuronal simulations based on cable models. It is on
+visualisation of neuronal simulations based on cable models. It is based on
 OpenSceneGraph for data management and Equalizer for parallel rendering, and
@@
 anti-aliasing based on jittered frusta and accumulation during still views,
-interactive modification of the visual representation of neurons in a per-neuron
+interactive modification of the visual representation of neurons on a per-neuron
 basis (full neuron vs. soma only, branch pruning depending on the branch level,
@@
 distributed compositing manager. Omegalib supports pluggable front-ends, to
-simplify the integration of third-party libraries like OpenGL, OpenSceneGraph,
+simplify the integration of third-party libraries like OpenGL \FIXME{But if it's based on Equalizer then OpenGL should already be part of the backend?}, OpenSceneGraph,
 and the Visualization Toolkit (VTK).
@@
 distribution is more justified. The relatively simple nature of our benchmark
-application's rendering algorithms is also favoring work packages, since it has
+application's rendering algorithms is also favoring work packages, since they have
 a near-zero static overhead per rendering pass.
 
-Last, but not least, we also provide scalability results for pixel compounds.
+Finally, we also provide scalability results for pixel compounds.
 While naturally load-balanced, pixel compounds only scale fill rate and not

### 2018-02-15 paperIEEE.tex
Address FIXMEs
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@132 db7c848b-7755-4df2-b763-8796797afa52

@@
 been proposed: compression and region of interest during
-compositing~\cite{MEP:10}; load-balancing resources for multi-display
-installations~\cite{EEP:11}; asynchronous compositing and NUMA
-optimizations~\cite{EBAHMP:12}; as well as work queueing~\cite{SPEP:16}. \FIXME{Why ; instead of , ?}
+compositing~\cite{MEP:10}, load-balancing resources for multi-display
+installations~\cite{EEP:11}, asynchronous compositing and NUMA
+optimizations~\cite{EBAHMP:12}, as well as work queueing~\cite{SPEP:16}.
 Additionally, complex large scale and out-of-core multiresolution
@@
 based on past rendering times. This assumes a reasonable frame-to-frame
-coherency. \FIXME{Shouldn't that be frame-to-frame {\em coherence}?} Equalizer implements two different algorithms, a
+coherence. Equalizer implements two different algorithms, a
 \textsf{load\_equalizer} and a \textsf{tree\_equalizer}. The result section
@@
 
-\FIXME{Sentence incomplete.}
-Asynchronous compositing pipelines rendering with compositing operations, by
+Asynchronous compositing is pipelining the rendering with compositing operations, by
 executing the image readback, network transfer and image assembly from threads
@@
 \item[Unbuffered] The object is versioned and unbuffered. No data is stored, and
-  no previous versions can be mapped. \FIXME{Redundant.} No old versions can be mapped and no
-  additional data is stored.
+  no previous versions can be mapped.
 \end{compactdesc}
@@
 Serialization is facilitated using output or input streams, which abstract the
-data transmission and are used like an \textsf{std::stream}. The data streams
+data transmission and are used like a \textsf{std::stream}. The data streams
 implement efficient buffering and compression, and automatically select the best
@@
 \item Newly mapped slave instances are mapped to the oldest available
-  version by default \FIXME{Is this correct?}, or to the version specified when calling
+  version by default, or to the version specified when calling
   \textsf{mapObject}.
@@
 \textsf{Collage} implements a few generic distributed objects which are used by
-\textsf{Equalizer} and other applicatons. A barrier is a distributed barrier
-primitive used for software swap barriers in \textsf{Equalizer}
-(\sref{sec:swap}) \FIXME{Using the word barrier three times in one short sentence}. Its implementation follows a simple master-slave approach,
-which has shown to by sufficient for this use case. The used distributed data structures are single producer,
-multiple consumer FIFO queues. To hide network latencies, consumers prefetch stored
+\textsf{Equalizer} and other applications. A barrier is a distributed barrier
+primitive used for software swap synchronization in \textsf{Equalizer}
+(\sref{sec:swap}). Its implementation follows a simple master-slave approach,
+which has shown to by sufficient for this use case. Queues are distrbuted, single producer,
+multiple consumer FIFO queues. To hide network latencies, consumers prefetch
 items into a local queue. Queues are used for tile and chunk compounds
@@
 distributed compositing manager. Omegalib supports pluggable front-ends, to
-simplify the integration of third-party libraries like OpenGL \FIXME{But if it's based on Equalizer then OpenGL should already be part of the backend?}, OpenSceneGraph,
+simplify the integration of third-party libraries like OpenGL, OpenSceneGraph,
 and the Visualization Toolkit (VTK).
@@
 the data compression algorithm. Our microbenchmark compresses a set of binary
-files \FIXME{What are these files? Fig. 7 only speaks of "generic" binary data...}, precalculating the speed and compression ratio of the various engines.
+files, precalculating the speed and compression ratio of the various engines.
 \fig{fCompressor} shows the compression and decompression speed in gigabyte per
@@
 
-\FIXME{Sentence incomplete.}
-used the same data sets as in the previous section. We ran the benchmark on up
+We used the same data sets as in the previous section, and ran the benchmark on up
 to eight physical nodes, that is, after eight processes nodes start to run two

### 2018-02-15 paperIEEE.tex
Tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@133 db7c848b-7755-4df2-b763-8796797afa52

@@
 master objects are registered. The preloading sends instance data of recently
-registered master objects to all connected
-nodes, while the corresponding node is idle. These nodes simply enter the received data to their cache. Preloading
+registered master objects to all connected nodes, while the corresponding node
+is idle. These nodes simply enter the received data to their cache. Preloading
 uses multicast when available.
@@
 (\sref{sec:swap}). Its implementation follows a simple master-slave approach,
-which has shown to by sufficient for this use case. Queues are distrbuted, single producer,
-multiple consumer FIFO queues. To hide network latencies, consumers prefetch
-items into a local queue. Queues are used for tile and chunk compounds
-(\sref{sec:tile}).
+which has shown to by sufficient for this use case. Queues are distributed,
+single producer, multiple consumer FIFO queues. To hide network latencies,
+consumers prefetch items into a local queue. Queues are used for tile and chunk
+compounds (\sref{sec:tile}).
 
@@
 the application node in \textsf{Equalizer}. Consumers, e.g., \textsf{Equalizer}
-render clients,
-can selectively map the objects they are interested in. Committing the object
-map will commit all registered objects and sync their new version to the
-slaves. Syncing the map on the slaves will synchronize all mapped instances to
-the new version recorded in the object map. This effective design allows data
-distribution with minimal application logic. It is used by \textsf{Sequel}
-(\sref{sec:sequel}) and other \textsf{Collage} applications.
+render clients, can selectively map the objects they are interested in.
+Committing the object map will commit all registered objects and sync their new
+version to the slaves. Syncing the map on the slaves will synchronize all mapped
+instances to the new version recorded in the object map. This effective design
+allows data distribution with minimal application logic. It is used by
+\textsf{Sequel} (\sref{sec:sequel}) and other \textsf{Collage} applications.
 
@@
 In this section, we present some major applications built using
-\textsf{Equalizer}, and
-show how they interact with the framework to solve complex parallel rendering
-problems.
+\textsf{Equalizer}, and show how they interact with the framework to solve
+complex parallel rendering problems.
 
@@
 work load on odd node counts. Static sort-last volume rendering has a similar
-oscillation behaviour, as ranges of scene geometry tend to also get unfavorably assigned under such conditions.
+oscillation behaviour, as ranges of scene geometry tend to also get unfavorably
+assigned under such conditions.
 
@@
 distribution is more justified. The relatively simple nature of our benchmark
-application's rendering algorithms is also favoring work packages, since they have
-a near-zero static overhead per rendering pass.
-
-Finally, we also provide scalability results for pixel compounds.
-While naturally load-balanced, pixel compounds only scale fill rate and not
-geometry processing. Consequently, pixel compounds provide better performance
-for volume rendering, and a predictable scaling behaviour for both.
+application's rendering algorithms is also favoring work packages, since they
+have a near-zero static overhead per rendering pass.
+
+Finally, we also provide scalability results for pixel compounds. While
+naturally load-balanced, pixel compounds only scale fill rate and not geometry
+processing. Consequently, pixel compounds provide better performance for volume
+rendering, and a predictable scaling behaviour for both.
 
@@
 compute resources fluctuate randomly every frame, as can arguably be the case
-for shared rendering nodes in virtualized enviroments. For this scenario tile equalizer seems best suited,
-as it performs load balancing implicitly and does not assume coherence of
-available resources between frames. The simpler tree equalizer also outperforms
-the load equalizer in this experiment.
-
-\begin{figure*}[ht]\center
-    \includegraphics[height=.17\textwidth]{images/benchmark_eVolve}
-    \includegraphics[height=.17\textwidth]{images/benchmark_eqPly2}
-    \includegraphics[height=.17\textwidth]{images/spikes512}
-    \includegraphics[height=.17\textwidth]{images/beechnut}
+for shared rendering nodes in virtualized enviroments. For this scenario tile
+equalizer seems best suited, as it performs load balancing implicitly and does
+not assume coherence of available resources between frames. The simpler tree
+equalizer also outperforms the load equalizer in this experiment.
+
+\begin{figure*}[ht]%
+\includegraphics[height=.17\textwidth]{images/benchmark_eVolve}\hfil
+\includegraphics[height=.17\textwidth]{images/benchmark_eqPly2}\hfil
+\includegraphics[height=.17\textwidth]{images/spikes512}\hspace{.5mm}\includegraphics[height=.17\textwidth]{images/beechnut}
   \caption{\label{fBenchmarks}Benchmark data: Flower alley with 30 volume models

### 2018-02-15 paperIEEE.tex
Tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@134 db7c848b-7755-4df2-b763-8796797afa52
First submission
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@135 db7c848b-7755-4df2-b763-8796797afa52

@@
     integration for a wide range of real-world research and industry
-    applications. Based on the basic architecture and implementation of the
-    \textsf{Equalizer} parallel rendering framework presented in \cite{EMP:09},
-    we show how a variety of generic algorithms can be integrated in the framework
+    applications. Using the \textsf{Equalizer} parallel rendering framework, we
+    show how a wide set of generic algorithms can be integrated in the framework
     to help application scalability and development in many different domains,
-    highlighting how concrete applications benefit from the diverse aspects and use
-    cases of \textsf{Equalizer}, and we present new experimental results for parallel
-    rendering and data distribution.
+    highlighting how concrete applications benefit from the diverse aspects and
+    use cases of \textsf{Equalizer}. We present novel parallel rendering
+    algorithms, powerful abstractions for large visualization setups and virtual
+    reality, as well as new experimental results for parallel rendering and data
+    distribution.
   \end{abstract}
@@
 \includegraphics[height=.17\textwidth]{images/benchmark_eqPly2}\hfil
-\includegraphics[height=.17\textwidth]{images/spikes512}\hspace{.5mm}\includegraphics[height=.17\textwidth]{images/beechnut}
+\includegraphics[height=.17\textwidth]{images/spikes512}\hspace{.2mm}\includegraphics[height=.17\textwidth]{images/beechnut}
   \caption{\label{fBenchmarks}Benchmark data: Flower alley with 30 volume models

### 2018-07-30 paperIEEE.tex
Reviewer 1, Collage
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@137 db7c848b-7755-4df2-b763-8796797afa52

@@
 control host synchronizes its content (model and camera) with one view on the
-display wall through Collage objects. The control host allows full model
-modifications and all workflows supported within the standalone Deltagen
+display wall through \textsf{Collage} objects. The control host allows full
+model modifications and all workflows supported within the standalone Deltagen
 application, and all changes are synchronized to the corresponding rendering
@@
 communication layer between the individual processes. \textsf{Equalizer} relies
-on the
-\textsf{Collage} network library for its internal operation. \textsf{Collage}
-furthermore provides
-powerful abstractions for writing \textsf{Equalizer} applications, which are
-introduced
-in this section.
+on the \textsf{Collage} network library for its internal operation.
+\textsf{Collage} furthermore provides powerful abstractions for writing
+\textsf{Equalizer} applications, which are introduced in this section.
+\textsf{Collage} is a separate, standalone library heavily influenced by the
+requirements of \textsf{Equalizer}. It is used directly when implementing data
+distribution for parallel rendering applications, and independently by a few
+research projects.
 

### 2018-07-30 paperIEEE.tex
Reviewer 1, Collage tests
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@138 db7c848b-7755-4df2-b763-8796797afa52
Reviewer 1, typos
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@139 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-RTT Scale uses a master-slave execution mode, were a single running Deltagen
+RTT Scale uses a master-slave execution mode, where a single running Deltagen
 instance can go into ``Scale mode'' at any time by launching an
@@
 slave instances simultaneously. This is a ``push'' operation, whereas the
-mapping is a slave-triggered ``pull'' operation. Slave nodes queue this data and
-consume it during synchronization. We test the time to commit and sync the data
-using different compression engines.
+mapping is a slave-triggered ``pull'' operation. Furthermore, the buffers only
+have to be serialized and compressed once, and then can be sent directly to all
+mapped slave nodes. Slave nodes queue this data and consume it during
+synchronization. In contrast, object mapping needs to wait for each slave node
+to request the mapping, and then may need to reserialize and compress the
+object data. We test the time to commit and sync the data using different
+compression engines.
 

### 2018-08-01 paperIEEE.tex
Reviewer two comments
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@140 db7c848b-7755-4df2-b763-8796797afa52

@@
 
-The Equalizer parallel rendering framework as first presented in~\cite{EMP:09} has
-demonstrated its general versatility and the usefulness of its minimally-intrusive
-programming design in a variety of applications and projects.
-In particular, the integration of large-scale parallel rendering algorithms,
-APIs for developing complex distributed applications, and many
-individual features make Equalizer a unique, open source framework to develop
-visualization applications for virtually any type of setup and use case.
-While individual applications and case studies using the framework, as well as
-new algorithms and system components extending it have been presented
-since the initial release of Equalizer, many important features and functionalities
-have not previously been published and are presented here.
-In this report, we thus present an updated comprehensive review of the
-integration of research, application use cases and commercial developments
-with respect to Equalizer as well as novel comparative experimental results
-of its scalability features.
+The Equalizer parallel rendering framework, first presented in~\cite{EMP:09},
+has demonstrated its general versatility and the usefulness of its
+minimally-intrusive design in a variety of applications and projects. In
+particular, the integration of large-scale parallel rendering algorithms, APIs
+for developing complex distributed applications, and many individual features
+make Equalizer a unique, open source framework to develop visualization
+applications for virtually any type of setup and use case. While individual
+applications and case studies using the framework, as well as new algorithms
+and system components extending it have been presented since the initial
+release of Equalizer, many features and functionalities have not been published
+yet and are presented here. In this report, we thus present an updated
+comprehensive review of the integration of research, application use cases and
+commercial developments with respect to Equalizer as well as novel comparative
+experimental results of its scalability features.
 
@@
 
-The remainder of this paper is structured as follows: First we provide an update
-on related work since the introduction of \textsf{Equalizer}. The main body of this paper
-then presents new performance features, VR algorithms, usability
-features to build complex applications, main novel features of the underlying
-\textsf{Collage} network library and a quick overview of the main
-\textsf{Equalizer}-based applications. A result section presents new experiments
-not previously published, followed by the discussion and conclusion.
+The remainder of this paper is structured as follows: First we provide an
+update on related work since the introduction of \textsf{Equalizer}. The main
+body of this paper then presents new performance features, VR algorithms,
+usability features to build complex applications, an overview of the underlying
+\textsf{Collage} network library and a summary of the main
+\textsf{Equalizer}-based applications. A result section presents new
+experiments, followed by the discussion and conclusion.
 
@@
 provides in-depth documentation on using and programming
-\textsf{Equalizer}~\cite{Eilemann:13}. In the following related work we assume
-these two publications and their references as a starting point, and focus on
-the new work published since 2009.
+\textsf{Equalizer}~\cite{Eilemann:13}. We assume these two publications and
+their references as a starting point, and focus on the new work published since
+2009.
 
@@
 
+\section{Fundamentals}
+
+The core task of developing an Equalizer application is to separate the
+rendering code from the application logic. The application main thread will
+execute its main loop as before, but instead of issuing rendering calls it will
+request a frame to be rendered from Equalizer. Depending on the configuration,
+Equalizer will then decompose and synchronize the rendering tasks. The
+application rendering code is executed in parallel under a given render
+context, which is the core entity abstracting the application-specific
+rendering algorithm from the system-specific configuration. This render context
+formalizes the minimal abstraction for parallel rendering, and specifies:
+
+\begin{compactdesc}
+
+ \item[Buffer] OpenGL-style read and draw buffer as well as color mask. These
+ parameters are influenced by the current eye pass, eye separation and
+ anaglyphic stereo settings.
+
+ \item[Viewport] Two-dimensional pixel viewport restricting the rendering area.
+ The pixel viewport is influenced by the destination viewport and viewports set
+ for sort-first decompositions.
+
+ \item[Frustum] Frustum parameters as defined by \textsf{glFrustum}. Typically
+ the frustum is used to set up the OpenGL projection matrix. The frustum is
+ influenced by the destinations view definition, sort-first, pixel and subpixel
+ decomposition, tracking head matrix and the current eye pass.
+
+ \item[Head Transformation] A transformation matrix positioning the frustum.
+ For planar views this is an identity matrix. In immersive rendering, it is
+ normally used to set up the `view' part of the modelview matrix, before static
+ light sources are defined.
+
+ \item[Range] A one-dimensional range with the interval [0..1]. This parameter is
+ optional and should be used by the application to render only the appropriate
+ subset of its data for sort-last rendering.
+
+\item[View] The view object from the logical rendering rendering configuration
+(\sref{sView}). Holds view-specific data, such as camera, model or any other
+application state.
+
+\end{compactdesc}
+
+An Equalizer configuration determines how the application is executed in
+parallel. It has three parts: A resource section describing the physical
+rendering resources (node, pipe, window, channel), a section describing the
+visualization setup (\sref{sView}), and a compound section describing how these
+resources are to be used for rendering. Compounds describe the rendering task
+decomposition and result recomposition, and their algorithm has been introduced
+in \cite{EMP:09}.
+
+Our execution model is fully asynchronuous, and only introduces synchronization
+points when strictly needed. The main synchronization points are: configured
+swap barriers between a set of output channels which have to display
+simultaneously, the availability of input frames for scalable rendering, and a
+task synchronization to prevent runaway of the main loop execution. By default,
+Equalizer keeps up to one frame of latency in execution, that is, some resource
+might render the next frame while others are still finishing the current.
+Nonetheless, resources which are ready will immediately display their result.
+This asynchronous execution architecture, coupled with a frame of latency,
+allows pipelining of many operations, such as the application event processing,
+task computation and load balancing, rendering, image readback, compression,
+network transmission, and compositing. It also hides small imbalances in the
+task distribution, as they can average out over multiple frames. It is
+fundamental to provide optimal performance, in particular for larger cluster
+sizes.
+
 %------------------------------------------------------------------------------
@@
 
-% This section introduces new decomposition modes, runtime adaptations and other
-% optimizations to increase the rendering performance for a wide variety of
-% algorithms and use cases.
-
 %------------------------------------------------------------------------------
@@
 
-The initial version of \textsf{Equalizer} implemented the basic sort-first (2D),
-sort-last (DB), stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the
-following we present the newly added decomposition modes and motivate their use
+The initial version of \textsf{Equalizer} implemented the basic sort-first
+(2D), sort-last (DB), stereo (EYE) and multilevel decompositions \cite{EMP:09}.
+In the following we present new decomposition modes and motivate their use
 case, which bring the overall feature set way beyond the typical sort-first and
@@
 coherence. Equalizer implements two different algorithms, a
-\textsf{load\_equalizer} and a \textsf{tree\_equalizer}. The result section
-provides some evidence on the strengths and weaknesses of both algorithms.
+\textsf{load\_equalizer} and a \textsf{tree\_equalizer}. Both implementations
+modify the \textsf{viewport} or \textsf{range} of their direct child compounds.
+The result section provides some evidence on the strengths and weaknesses of
+both algorithms.
 
@@
 
-Cross-segment load balancing (\fig{fig:vieweq}) addresses the optimal resource
-allocation of $n$ rendering resources to $m$ output channels (with $n\geq m$). A
-view equalizer works in conjunction with load equalizers balancing the
+Cross-segment load balancing (CSLB, \fig{fig:vieweq}) optimizes resource
+allocation of $n$ rendering resources to $m$ output channels (with $n\geq m$).
+A view equalizer works in conjunction with load equalizers balancing the
 individual output channels. It monitors the usage of shared source channels
 across outputs and activates them to balance the rendering time of all outputs.
-In \cite{EEP:11}, we provide a detailed description and evaluation of our
-algorithm.
+CSLB is unique in that it supports load balancing for planar and non-planar
+visualization systems, as well as a flexible allocation of rendering resources
+to individual output channels. In \cite{EEP:11}, we provide a detailed
+description and evaluation of our algorithm and show that it significantly
+improve performance over a static allocation.
 
@@
 The region of interest is the screen-space 2D bounding box enclosing the data
-rendered by a single resource. We have extended the core parallel rendering
-framework to use an application-provided ROI to optimize the
-\textsf{load\_equalizer} as well as the image compositing. The load equalizer
-uses the ROI to refine its load grid to the regions containing data. The
-compositing code uses the ROI to minimize image readback and network
+rendered by a single resource. We use an application-provided ROI to optimize
+the \textsf{load\_equalizer} as well as image compositing in our framework. The
+load equalizer uses the ROI to refine its load grid to the regions containing
+data. The compositing code uses the ROI to minimize image readback and network
 transmission. In \cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of
@@
 mutable data has to have a copy in each render thread, which is not feasible in
-most applications and large data scenarios. To easily customize the
-synchronization of all threads on a single process, \textsf{Equalizer}
-implements three threading modes:  Full synchronization, draw synchronization
-and asynchronous. Note that the execution between node processes is always
-asynchronous, for up to \textsf{latency} frames.
+most applications and large data scenarios. We implement three threading modes:
+Full synchronization, draw synchronization and asynchronous. Note that the
+execution between node processes is always asynchronous, for up to
+\textsf{latency} frames.
 
@@
 %------------------------------------------------------------------------------
-\section{Virtual Reality Features}
-%------------------------------------------------------------------------------
-
-Virtual Reality is an important field for parallel rendering. It does however
-require special attention to support it as a first-class citizen in a generic
-parallel rendering framework. \textsf{Equalizer} has been used in many virtual
-reality installations, such as the Cave2 \cite{FNTTL:13}, the high-resolution C6
-CAVE at the KAUST visualization laboratory, and head-mounted displays
+\section{Virtual Reality}
+%------------------------------------------------------------------------------
+
+Virtual Reality is an important field for parallel rendering, which requires
+special attention to support it as a first-class citizen in a generic parallel
+rendering framework. \textsf{Equalizer} has been used in many virtual reality
+installations, such as the Cave2 \cite{FNTTL:13}, the high-resolution C6 CAVE
+at the KAUST visualization laboratory, and head-mounted displays
 (\fig{FIG_teaser}). In the following we lay out the features needed to support
-these installations, motivated by application use cases.
+these installations.
 
@@
 
-Head tracking is the minimal feature needed to support immersive installations.
-\textsf{Equalizer} does support multiple, independent tracked views through the
-observer abstraction (\sref{sec:observer}). Built-in VRPN support enables the
-direct, application-transparent configuration of a VRPN tracker device.
-Alternatively, applications can provide a $4\times 4$ tracking matrix. Both
-CAVE-like tracking with fixed projection surfaces and HMD tracking with moving
-displays are implemented.
+Head tracking is fundamental to support immersive installations. We support
+multiple, independent tracked views through the observer abstraction
+(\sref{sec:observer}). Built-in VRPN support enables direct,
+application-transparent configuration of a VRPN tracker device. Alternatively,
+applications can provide a $4\times 4$ tracking matrix. Both CAVE-like tracking
+with fixed projection surfaces and HMD tracking with moving displays are
+implemented.
 
@@
 %------------------------------------------------------------------------------
-\subsection{Physical and Logical Visualization Setup}
+\subsection{Physical and Logical Visualization Setup}\label{sView}
 
@@
 %----------------------------------------------------------------------
-\section{Example Applications} \label{sApplications}
+\section{Applications} \label{sApplications}
 %----------------------------------------------------------------------
@@
 
-RTT Deltagen (now Dassault 3D Excite) is a commercial application for
-interactive, high quality rendering of CAD data. The RTT Scale module,
-delivering multi-GPU and distributed execution, is based on \textsf{Equalizer}
-and \textsf{Collage}, and has driven many of the aforementioned features.
+RTT Deltagen (\fig{fLayout}, now Dassault 3D Excite) is a commercial
+application for interactive, high quality rendering of CAD data. The RTT Scale
+module, delivering multi-GPU and distributed execution, is based on
+\textsf{Equalizer} and \textsf{Collage}, and has driven many of the
+aforementioned features.
 
@@
 RTNeuron \cite{HBBES:13} is a scalable real-time rendering tool for the
-visualisation of neuronal simulations based on cable models. It is based on
-OpenSceneGraph for data management and Equalizer for parallel rendering, and
-focuses not only on fast rendering times, but also on fast loading times with no
-offline preprocessing. It provides level of detail (LOD) rendering, high quality
-anti-aliasing based on jittered frusta and accumulation during still views,
-interactive modification of the visual representation of neurons on a per-neuron
-basis (full neuron vs. soma only, branch pruning depending on the branch level,
-\dots). RTNeuron implements both sort-first and sort-last rendering with order
-independent transparency.
+visualisation of neuronal simulations based on cable models (\fig{FIG_teaser}
+(a)). It is based on OpenSceneGraph for data management and Equalizer for
+parallel rendering, and focuses not only on fast rendering times, but also on
+fast loading times with no offline preprocessing. It provides level of detail
+(LOD) rendering, high quality anti-aliasing based on jittered frusta and
+accumulation during still views, interactive modification of the visual
+representation of neurons on a per-neuron basis (full neuron vs. soma only,
+branch pruning depending on the branch level, \dots). RTNeuron implements both
+sort-first and sort-last rendering with order independent transparency.
 

### 2018-08-01 paperIEEE.tex
Reviewer 3, comment 1
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@141 db7c848b-7755-4df2-b763-8796797afa52

@@
 fit for distributed memory systems, since the separate memory space makes
-concurrent rendering of different frames easier to implement. While it increases
-the framerate linearly, it does not decrease the latency between user input and
-the corresponding output. Consequently, this decomposition mode is mostly useful
-for non-interactive movie generation. It is transparent to \textsf{Equalizer}
-applications, but does require the configuration latency to be equal or greater
-than the number of source channels. Furthermore, to work in multi-threaded,
-multi-GPU configurations, the application needs to support running the rendering
-threads asynchronously (\sref{sec:threading}). The output frame rate of the
-destination channel may be smoothened using a {\em frame rate equalizer}
-(\sref{sec:framerateEq}).
+concurrent rendering of different frames easier to implement. While it
+increases the framerate linearly, it does not decrease the latency between user
+input and the corresponding output. Consequently, this decomposition mode is
+mostly useful for non-interactive rendering, e.g., for movie generation. It is
+transparent to \textsf{Equalizer} applications, but does require the
+configuration latency to be equal or greater than the number of source
+channels. Furthermore, to work in multi-threaded, multi-GPU configurations, the
+application needs to support running the rendering threads asynchronously
+(\sref{sec:threading}). The output frame rate of the destination channel may be
+smoothened using a {\em frame rate equalizer} (\sref{sec:framerateEq}).
+
+DPlex is configured by restricting the source compounds to a \textit{phase}
+(offset) of a certain \textit{period} (number of skipped frames) of the
+destination channel's rendering. All output frames of the source channels may
+use the same name, as each output frame only is available at a distinct
+rendered frame for input. The period/phase decomposition may also be used for
+other means, e.g., to have one channel render at half the frame rate on an
+underpowered control workstation for a tiled display wall.
 
@@
 transparent to \textsf{Equalizer} applications.
+
+Tile and chunk compounds are configured through output and input queues, which
+behave similarly to compositing frames. A single output queue configures the
+tile or chunk size and generates the work packages on the destination compound.
+Each source has an input queue of the same name, which consumes items from the
+output queue. Each work item generates a separate draw and readback task, and
+one input assemble image on the destination channel.
 
@@
 
+Pixel compounds are configured similarly to sort-first compounds, except that
+instead of decomposing the \textit{viewport}, they decompose a \textit{pixel}
+kernel. The kernel is a 2D structure describing the shape and sampled pixel
+within the shape in the format \textit{[x y width height]}, that is, a source
+compound selects the pixel \textit{x, y} within a decomposition of size
+\textit{width, height}.
+
 \subsubsection{Subpixel}
@@
 
-% Furthermore, subpixel compounds interact with idle image refinements, e.g., they
-% can accelerate idle anti-aliasing of a scene when the camera and scene are not
-% changed.
+Subpixel compounds are configured similarly to pixel compounds, but instead of
+a \textit{pixel} kernel they decompose a \textit{subpixel} kernel in the format
+\textit{[index, size]}, describing which subpixel of \textit{size} samples a
+source produces. By default, the output frames are averaged on the destination
+channel.
 

### 2018-08-01 paperIEEE.tex
Tweaks
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@142 db7c848b-7755-4df2-b763-8796797afa52

@@
 slave instances simultaneously. This is a ``push'' operation, whereas the
-mapping is a slave-triggered ``pull'' operation. Furthermore, the buffers only
-have to be serialized and compressed once, and then can be sent directly to all
-mapped slave nodes. Slave nodes queue this data and consume it during
+mapping is a slave-triggered ``pull'' operation. During commit, the buffers
+only have to be serialized and compressed once, and then can be sent directly
+to all mapped slave nodes. Slave nodes queue this data and consume it during
 synchronization. In contrast, object mapping needs to wait for each slave node
@@
 node is a separate distributed object, which has two child node objects. A
-total of $1023$ objects is distributed and synchronized. \fig{fDavidDist} shows
+total of $1023$ objects are distributed and synchronized. \fig{fDavidDist} shows
 the distribution times for this data set. Due to limited compressibility of the

### 2018-08-07 paperIEEE.tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@143 db7c848b-7755-4df2-b763-8796797afa52

@@
 Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
-sort-last rendering, respectively. They decompose the scene into a predefined
-set of fixed-size image tiles or database ranges. These tasks, or work packages,
+sort-last rendering, respectively. They are implemented via the \textsf{tile\_equalizer} and \textsf{chunk\_equalizer} compounds, which decompose the scene into a predefined
+set of fixed-size image tiles or database ranges, respectively.
+The resulting tasks, or work packages,
 are queued and processed by all source channels by polling a server-central
@@
 assigned under such conditions.
+It seems plausible that this is also due to overheads associated with these two rendering scenarios.
+That is, volume rendering typically benefits less form sort-last decompositions, because of higher costs for compositing.
+Similarly, polygonal rendering often benefits less from sort-first decomposition, as compositing costs are typically much lower, while culling of geometry can create an additional overhead.
 
@@
 compute resources fluctuate randomly every frame, as can arguably be the case
-for shared rendering nodes in virtualized enviroments. For this scenario tile
-equalizer seems best suited, as it performs load balancing implicitly and does
+for shared rendering nodes in virtualized enviroments. For this scenario \textsf{tile\_equalizer} seems best suited, as it performs load balancing implicitly and does
 not assume coherence of available resources between frames. The simpler tree
-equalizer also outperforms the load equalizer in this experiment.
+equalizer also outperforms the \textsf{load\_equalizer} in this experiment.
 

### 2018-08-12 paperIEEE.tex
Final touches
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@144 db7c848b-7755-4df2-b763-8796797afa52

@@
 Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
-sort-last rendering, respectively. They are implemented via the \textsf{tile\_equalizer} and \textsf{chunk\_equalizer} compounds, which decompose the scene into a predefined
-set of fixed-size image tiles or database ranges, respectively.
-The resulting tasks, or work packages,
-are queued and processed by all source channels by polling a server-central
-queue. Prefetching ensures that the task communication overlaps with rendering.
-As shown in~\cite{SPEP:16} and the results, these modes can provide better
-performance due to being implicitly, i.e., inherently load-balanced, as long as
-there is an insignificant overhead for the render task setup. This mode is
-transparent to \textsf{Equalizer} applications.
+sort-last rendering, respectively. They are implemented via the
+\textsf{tile\_equalizer} and \textsf{chunk\_equalizer} compounds, which
+decompose the scene into a predefined set of fixed-size image tiles or database
+ranges, respectively. The resulting tasks, or work packages, are queued and
+processed by all source channels by polling a server-central queue. Prefetching
+ensures that the task communication overlaps with rendering. As shown
+in~\cite{SPEP:16} and the results, these modes can provide better performance
+due to being implicitly, i.e., inherently load-balanced, as long as there is an
+insignificant overhead for the render task setup. This mode is transparent to
+\textsf{Equalizer} applications.
 
@@
 compounds. Sort-first polygon rendering exhibits oscillations in performance as
-nodes are added to the task, due to unfavorable assignment of a tile with a high
-work load on odd node counts. Static sort-last volume rendering has a similar
-oscillation behaviour, as ranges of scene geometry tend to also get unfavorably
-assigned under such conditions.
-It seems plausible that this is also due to overheads associated with these two rendering scenarios.
-That is, volume rendering typically benefits less form sort-last decompositions, because of higher costs for compositing.
-Similarly, polygonal rendering often benefits less from sort-first decomposition, as compositing costs are typically much lower, while culling of geometry can create an additional overhead.
+nodes are added to the task, due to unfavorable assignment of a tile with a
+high work load on odd node counts. Static sort-last volume rendering has a
+similar oscillation behaviour, as ranges of scene geometry tend to also get
+unfavorably assigned under such conditions. For our polygonal rendering setup,
+the amount of data exceeds the GPU memory, therefore sort-last rendering
+exhibits a better scaling behaviour since it reduces the amount of geometry
+data per GPU. For volume rendering, sort-first performs better since it has
+less compositing overhead and is easy to load balance.
 

### 2018-08-12 paperIEEE.tex
Really final touches
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@145 db7c848b-7755-4df2-b763-8796797afa52

@@
 \includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{
-Stefan Eilemann} works towards large-scale visualization for Exascale
-simulations; interactive integration of simulation, analysis and visualization
-as well as flexible frameworks for data sharing and dynamic allocation of
-heterogenous resources. He was the technical manager of the Visualization Team
-in the Blue Brain Project, is the CEO and founder of Eyescale, and the lead
-developer of the \textsf{Equalizer} parallel rendering framework. He received
-his masters degree in Computer Science from EPFL in 2015, and his Engineering
-Diploma in Computer Science in 1998. He is currently working towards a PhD in
-Computer Science at the Visualization and MultiMedia Lab at the University of
-Zurich. \end{IEEEbiography}
+Stefan Eilemann} works on large data visualization and parallel rendering. He
+was the technical manager of the Visualization Team in the Blue Brain Project,
+is the CEO and founder of Eyescale, and the lead developer of the
+\textsf{Equalizer} parallel rendering framework. He received his masters degree
+in Computer Science from EPFL in 2015, and an Engineering Diploma in Computer
+Science in 1998. He is working towards a PhD in Computer Science at the
+Visualization and MultiMedia Lab at the University of Zurich.
+\end{IEEEbiography}
 

### 2018-08-13 paperIEEE.tex
Minor edits.
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@147 db7c848b-7755-4df2-b763-8796797afa52

@@
         David~Steiner\thanks{email: steiner@ifi.uzh.ch} \\ %
-        and~Renato~Pajarola\thanks{email:pajarola@acm.org},~\IEEEmembership{Senior Member,~IEEE}%
+        and~Renato~Pajarola\thanks{email: pajarola@ifi.uzh.ch},~\IEEEmembership{Senior Member,~IEEE}%
 \IEEEcompsocitemizethanks{\IEEEcompsocthanksitem All authors are with the
@@
 %\and Renato Pajarola~\IEEEmembership{Senior Member,~IEEE}\thanks{email:
-%pajarola@acm.org}}
+%pajarola@ifi.uzh.ch}}
 
@@
 description and evaluation of our algorithm and show that it significantly
-improve performance over a static allocation.
+improves performance over a static allocation.
 
@@
 unfavorably assigned under such conditions. For our polygonal rendering setup,
-the amount of data exceeds the GPU memory, therefore sort-last rendering
+the amount of data exceeds the GPU memory, therefore, sort-last rendering
 exhibits a better scaling behaviour since it reduces the amount of geometry
@@
 mapping is a slave-triggered ``pull'' operation. During commit, the buffers
-only have to be serialized and compressed once, and then can be sent directly
+only have to be serialized and compressed once, and can then be sent directly
 to all mapped slave nodes. Slave nodes queue this data and consume it during
